### 2011-08-18 paper.tex
start of paper

\documentclass{article}
\includepackage{graphicx}

\begin{document}

\title{Visual Exploration of Climate Data, Using Icon-Annotated Maps}
\author{Cook, Hofmann, Wickham, Wickham, Cheng}

# Drawing trend lines on map
#   - How to
#   - Comparison to coloring slope
#   - Reference to Grinstein icon plots
#   - Alexander Gribov's work

# Interactive graphics
# Data processing

# Perhaps the GHCN data? Not gridded
# NOAA data, almost grid, but over ocean, so maps more tricky
# NARCAP from NCAR?
# Fill in with NASA data, to start, try to put new data for final version

\section{Introduction}

Need a colorful map

\section{Generating a time series at each location}

\section{Showing seasonal effect, radially oriented icon}

\section{Effect of scale}

\section{Reference lines and boxes}

\section{Regular vs irregular locations}

\end{document}

### 2011-08-18 code/0-glyphs.r
Add code from workshop

@@ -1,0 +1,31 @@
+rescale01 <- function(x) {
+  rng <- range(x, na.rm = TRUE)
+  (x - rng[1]) / (rng[2] - rng[1])
+}
+rescale11 <- function(x) 2 * rescale01(x) - 1
+  
+glyphs <- function(data, x_major, x_minor, y_major, y_minor, polar = F, height = NULL, width = NULL) {
+  data$gid <- interaction(data[[x_major]], data[[y_major]], drop = TRUE)
+  
+  if (is.null(width)) {
+    width <- resolution(data[[x_major]]) * 0.9    
+    message("Using width ", format(width, digits = 3))
+  }
+  if (is.null(height)) {
+    height <- resolution(data[[y_major]]) * 0.9    
+    message("Using height ", format(height, digits = 3))
+  }
+  
+  if (polar) {
+    data$gx <- data[[x_major]] + rescale01(data[[y_minor]]) * 
+      cos(2 * pi * rescale01(data[[x_minor]]))
+    data$gy <- data[[y_major]] + rescale01(data[[y_minor]]) * 
+      sin(2 * pi * rescale01(data[[x_minor]]))
+    data <- data[order(data[[x_major]], data[[x_minor]]), ]
+  } else {
+    data$gx <- data[[x_major]] + rescale11(data[[x_minor]]) * width / 2
+    data$gy <- data[[y_major]] + rescale11(data[[y_minor]]) * height / 2
+  }
+  
+  data
+}

### 2011-08-18 code/0-nasa.r

@@ -1,0 +1,22 @@
+library("maps")
+library("ggplot2")
+
+if (!file.exists("nasadata.rds")) {
+  library("lubridate")
+
+  nasa <- read.csv("nasadata.csv.bz2", stringsAsFactors = FALSE)
+  nasa$date <- ymd(nasa$date)
+  nasa$day <- (nasa$date - min(nasa$date)) %/% ddays(1)
+  nasa$month <- month(nasa$date)
+  nasa$year <- year(nasa$date)
+  
+  saveRDS(nasa, "nasadata.rds")
+} else {
+  nasa <- readRDS("nasadata.rds")
+}
+
+outlines <- as.data.frame(map("world", plot = FALSE,
+  xlim = -c(113.8, 56.2), ylim = c(-21.2, 36.2))[c("x","y")])
+map <- geom_path(aes(x, y), inherit.aes = FALSE, data = outlines, 
+  alpha = 0.4, legend = FALSE)
+

### 2011-08-18 code/1-modeling-adv.r

@@ -1,0 +1,39 @@
+source("0-nasa.r")
+source("0-glyphs.r")
+
+library(ggplot2)
+library(plyr)
+library(mgcv)
+
+space <- dlply(nasa, "time", function(df) { 
+   gam(surftemp ~ s(lat, long), data =df)
+})
+
+loc_grid <- unique(nasa[c("lat", "long")])
+
+space_preds <- ldply(space, function(mod) {
+  loc_grid$pred <- predict(mod, newdata = loc_grid)
+  loc_grid
+})
+
+space_preds <- glyphs(space_preds, "long", "time", "lat", "pred")
+qplot(gx, gy, data = space_preds, geom = "line", group = gid) + map
+
+qplot(long, lat, data = space_preds, fill = pred, geom = "tile") + facet_wrap(~ time)
+
+
+big_mod <- gam(surftemp ~ s(long, lat, time) + factor(month), data = nasa)
+loc_grid <- expand.grid(
+  lat = unique(nasa$lat), 
+  long = unique(nasa$long), 
+  month = 1, 
+  time = seq(0, 2161, length = 50))
+loc_grid$pred <- predict(big_mod, newdata = loc_grid)
+
+loc_grid <- glyphs(loc_grid, "long", "time", "lat", "pred")
+qplot(gx, gy, data = loc_grid, geom = "line", group = gid) + map
+
+
+nasa$resid <- resid(big_mod)
+nasa <- glyphs(nasa, "long", "time", "lat", "resid")
+qplot(gx, gy, data = nasa, geom = "line", group = gid) + map

### 2011-08-18 code/1-modeling.r

@@ -1,0 +1,97 @@
+library("ggplot2")
+library("plyr")
+library("mgcv")
+
+source("0-glyphs.r")
+source("0-nasa.r")
+
+# Coefficients ---------------------------------------------------------------
+
+temp_models <- dlply(nasa, c("lat", "long"), function(df) {
+  # lm(ozone ~ factor(month) - 1, data = df)
+  lm(surftemp ~ factor(month), data = df)
+})
+
+
+coefs <- ldply(temp_models, function(x) {
+  data.frame(month = 2:12, coef = coef(x)[-1])
+})
+
+# Show seasonal patterns
+coefs <- glyphs(coefs, "long", "month", "lat", "coef") 
+qplot(gx, gy, data = coefs, geom = "line", group = gid) + map 
+qplot(gx, gy, data = coefs, geom = "line", group = gid) + 
+  geom_path(aes(y = lat), colour = "grey50", alpha = 0.75)
+
+coefs <- glyphs(coefs, "long", "month", "lat", "coef", polar = T) 
+qplot(gx, gy, data = coefs, geom = "path", group = gid) + map 
+ggplot(coefs, aes(gx, gy, group = gid)) +
+  map + 
+  geom_point(aes(long, lat), colour = "red", shape = ".") + 
+  geom_path()
+
+# Predictions ----------------------------------------------------------------
+
+temp_models <- dlply(nasa, c("lat", "long"), function(df) {
+  lm(surftemp ~ year + factor(month), data = df)
+})
+
+month_grid <- expand.grid(year = 2000, month = 1:12)
+month_preds <- ldply(temp_models, function(mod) {
+  month_grid$pred <- predict(mod, newdata = month_grid)
+  month_grid
+})
+
+year_grid <- expand.grid(year = unique(nasa$year), month = 1)
+year_preds <- ldply(temp_models, function(mod) {
+  year_grid$pred <- predict(mod, newdata = year_grid)
+  year_grid
+})
+
+month_preds <- glyphs(month_preds, "long", "month", "lat", "pred") 
+year_preds <- glyphs(year_preds, "long", "year", "lat", "pred") 
+
+qplot(gx, gy, data = month_preds, geom = "path", group = gid) + map 
+qplot(gx, gy, data = year_preds, geom = "path", group = gid) + map 
+
+temp_smooth <- dlply(nasa, c("lat", "long"), function(df) {
+  gam(surftemp ~ s(day) + factor(month), data = df)
[…]

### 2011-08-18 code/1-modeling.r
Fix scaling typo

@@ -68,7 +68,7 @@ qplot(gx, gy, data = day_preds, geom = "path", group = gid) + map
 # Rescale predictions to individual scales
 day_preds2 <- ddply(day_preds, c("lat", "long"), mutate, 
   pred_s = rescale01(pred))
-day_preds2 <- glyphs(day_preds, "long", "day", "lat", "pred_s") 
+day_preds2 <- glyphs(day_preds2, "long", "day", "lat", "pred_s") 
 qplot(gx, gy, data = day_preds2, geom = "path", group = gid) + map 
 
 

### 2011-08-18 code/1-modeling.r
Scaling alternative

@@ -67,10 +67,13 @@ qplot(gx, gy, data = day_preds, geom = "path", group = gid) + map
 
 # Rescale predictions to individual scales
 day_preds2 <- ddply(day_preds, c("lat", "long"), mutate, 
-  pred_s = rescale01(pred))
+  pred_s = rescale01(pred),
+  pred_m = pred / max(pred))
 day_preds2 <- glyphs(day_preds2, "long", "day", "lat", "pred_s") 
 qplot(gx, gy, data = day_preds2, geom = "path", group = gid) + map 
 
+day_preds2 <- glyphs(day_preds2, "long", "day", "lat", "pred_m") 
+qplot(gx, gy, data = day_preds2, geom = "path", group = gid) + map 
 
 # Linear trend is not a good fit!
 

### 2011-08-18 paper.tex
references

@@
 #   - Reference to Grinstein icon plots
-#   - Alexander Gribov's work
+# Pickett RM and Grinstein GG (1988). Iconographics 
+# displays for visualizing multidimensional data. 
+# In Proc. IEEE Conference on Systems, Man, and Cybernetics, pages 514-19.
+# Note the later work on metaphoric data display, garbage work!
+#   - Alexander Gribov's work http://rosuda.org/software/Gauguin/gauguin.html
 

### 2011-08-24 paper.tex
tweak
Working on the plane

--- 2011-08-18 paper.tex
@@
-\documentclass{article}
-\includepackage{graphicx}
+\documentclass[oneside]{article}
+\usepackage{fullpage}
+\usepackage[pdftex]{graphicx}
+\DeclareGraphicsExtensions{.png,.pdf}
+\graphicspath{{images/}}
+\usepackage{hyperref}
+\usepackage{verbatim}
+\usepackage[format=plain,font=small]{caption}
+\usepackage[small]{titlesec}
+\usepackage[round,sectionbib]{natbib}
+\bibliographystyle{plainnat}
+\renewcommand\rmdefault{bch}
+\linespread{1.07}
 
@@
 
-\title{Visual Exploration of Climate Data, Using Icon-Annotated Maps}
+\title{Glyph-maps for visualising climate data and models}
 \author{Cook, Hofmann, Wickham, Wickham, Cheng}
 
-# Drawing trend lines on map
-#   - How to
-#   - Comparison to coloring slope
-#   - Reference to Grinstein icon plots
-#   - Alexander Gribov's work
-
-# Interactive graphics
-# Data processing
-
-# Perhaps the GHCN data? Not gridded
-# NOAA data, almost grid, but over ocean, so maps more tricky
-# NARCAP from NCAR?
-# Fill in with NASA data, to start, try to put new data for final version
+% Drawing trend lines on map
+%   - How to
+%   - Comparison to coloring slope
+%   - Reference to Grinstein icon plots
+% Pickett RM and Grinstein GG (1988). Iconographics 
+% displays for visualizing multidimensional data. 
+% In Proc. IEEE Conference on Systems, Man, and Cybernetics, pages 514-19.
+% Note the later work on metaphoric data display, garbage work!
+%   - Alexander Gribov's work http://rosuda.org/software/Gauguin/gauguin.html
+%   - Spatial star plots by Andrews http://www.udallas.edu:8080/~andrews/software/software.html
+%   - Look at what Bertin does
+%
+% Interactive graphics
+% Data processing
+%
+% Perhaps the GHCN data? Not gridded
+% NOAA data, almost grid, but over ocean, so maps more tricky
+% NARCAP from NCAR?
+% Fill in with NASA data, to start, try to put new data for final version
+\maketitle
 
@@
 
-Need a colorful map
+Need a colourful map, facetted by year \& month
+
+Simple, but effective, technique that combines aspects of glyphs, facetted plots and map-charts. Particularly well suited for the type of temporal-spatial data, commonly seen in climate.
+
+Glyph-maps have attributes of both glyphs (stars, faces, profiles, ...), inviting exploration at the global level, and small-multiple time series, allowing local exploration of individual series.
+Allow to see both geographic context and individual locations simultaneously.
+
+An interesting attribute of these glyph plots is that the position of graphical elements is determined by major and minor axes.
+For maps, the major axes are latitude and longitude, and the minor axes are typically time on the x-axis and some measurement or model prediction on the y-axis.
+
+Avoids one problem that plagues glyph displays: the ordering of the variables.
+Although some techniques exist to ameliorate this problem \citep{kleiner:1981,hurley:2010}, they are not necessary with time data because it is intrinsically ordered.
+
+These displays tend to be particularly effective when printed in large format, as the much higher resolution of print (600 dpi) compared to computer  screen (72-120 dpi) allows for much finer reading.
+These are high-density data displays in the best tradition of Tufte.
+Wall size maps are not usually practical during analysis, but are very engaging.
+
+Will first explore for regularly gridded data, a simple starting point and a common output of climate models, and then show a few simple extensions to deal with non-rectangular grids and irregular locations, as exemplified by raw climate data which is typically collect at irregular locations.
 
@@
 
-\section{Showing seasonal effect, radially oriented icon}
+A big advantage of glyph-maps is that they can be generated with existing graphics software, after a simple pre-processing step.
+This makes them immediately accessible to a wide audience.
+The graphics in this paper were created in R \citep{R} using ggplot2 \citep{me:ggplot2,wickham:2007d} and \citep{me:plyr}, but are readily implemented in any statistical programming environment.
+Code and data is available in the online supplementary materials.
+
+Because the x and y locations are linear combinations of two variables, these types of displays can also be built interactively by software that supports tours, such as DataViewer, XGobi or GGobi.
+This technique is used to good effect in \citet{buja:1996a}.
+
+\section{Reference lines}
+
+Space-time-glyphs often require reference lines: it is difficult to compare the relative location of glyphs in different regions.
+As noted by \citet{cleveland:1993a}, a \emph{visual reference grid} is important for accurate perception of values, particularly of important properties of lines like slope and minima.
+This is illustrated well by the previous plot of monthly coefficients.
+When we add a reference to each location, Figure~\ref{fig:ref-basic}, it becomes much more clear that not only is there a difference in shape between the Northern and Southern hemispheres, but there is also a location shift: values in the North are much higher than values in the South.
+It is difficult to see this without reference lines.
+
+Figure~\ref{fig:ref-basic} shows two variants of a reference, a line drawn at the overall mid-range, and a box drawn around each grid cell.
+We prefer using white for such references because it is minimally perceptible: you can see it easily if you concentrate on it, but it does not otherwise distract from the perception of the graphic.
+Note the careful choice of layering: the map is drawn under mid-range lines, but over boxes.
+
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=0.5\linewidth]{ref-line}%
+  \includegraphics[width=0.5\linewidth]{ref-box}
+  \caption{caption}
+  \label{fig:ref-basic}
+\end{figure}
+
+Other types of reference information can also be usefully added. 
+Other ways of display additional data:
+
+\begin{itemize}
+
+  \item Map variable to colour or transparency of line.
+
+  \item Colour of reference lines/boxes.
+
+  \item Highlight special points in an additional layer.
+
+  \item Fill colour of reference boxes.
+
+\end{itemize}
+
+A particularly useful application of this technique is when display model predictions to also display some measure of model fit, so that appropriate care can be taken to avoid over-interpreting predictions from models that did not fit well.
+
+Figure~\ref{fig:ref-adv} shows two applications of these ideas, adding the month with the highest average temperature in either the foreground or background.
+Careful colour choice is necessary so that this information is perceptible, but not distracting.
+
+\begin{figure}[htbp]
+  \centering
+    \includegraphics[width=0.5\linewidth]{ref-max-1}%
+    \includegraphics[width=0.5\linewidth]{ref-max-2}
+  \caption{caption}
+  \label{fig:ref-adv}
+\end{figure}
+
+\section{Radial icons/star glyphs}
+
+These time series glyphs can be readily altered to support circular variables, by projecting each line into polar coordinates and joining the ends.
+This type of display is best suited for variables that are truly circular, such as seasonal patterns or angles (such as for wind direction).
+
+Cite star glyphs.
+
+The alteration to the data transformation is quite simple.
+Without loss of generality, we assume that $y_{minor}$ becomes $r$, the radius, and $x_{minor}$ becomes $theta$ the angle, both rescaled to lie between 0 and 1.
+Then the glyph $x$ coordinate is given by $x_{major} + \frac{w \cdot r}{2} \cos(2 \pi \theta)$, and the $y$ coordinate by $y_{major} + \frac{h \cdot r}{2} \sin(2 \pi \theta)$.
+
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=2in]{month-cartesian}
+  \includegraphics[width=2in]{month-polar}
+
+  \caption{Monthly averages in (left) Cartesian coordinates and (right) polar coordinates.}
+
+  \label{fig:cycle}
+\end{figure}
+
+\section{Scatterplots}
+
+So far we have limited ourselves to always mapping some aspect of time on the x-axis.
+But the method is much more general.
+Figure~\ref{fig:cloud} shows how by mapping another variable to the x-axis, in this case low-cloud level on x and high-cloud level on y, we can explore how this 2d distribution varies over space.
+Again, summarising the data with models is very useful: the second plot displays a loess curve fit to the data instead of the individual points.
+
+\begin{figure}[htbp]
+  \centering
+    \includegraphics[width=0.5\linewidth]{clouds}%
+    \includegraphics[width=0.5\linewidth]{clouds-smooth}
+  \caption{caption}
+  \label{fig:cloud}
+\end{figure}
 
@@
 
-\section{Reference lines and boxes}
+Partitioning is a extremely useful technique: product plots, ANOVA, ...
+For glyph plots, it's often useful to partition the glyphs into shape and scale components, decomposing the values into a product of maximum and scaled values.
+This is revealing for smooth models of surface temperature.
+
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=0.5\linewidth]{month-rescale-max}%
+  \includegraphics[width=0.5\linewidth]{month-rescale01}
+  \caption{caption}
+  \label{fig:label}
+\end{figure}
+
+Other types of scaling: scale range to 0-1 (subtract min and divide by range), standardise mean and standard deviation (or robust variants).
+
+Equivalent to using individual scales in facetted graphics.
 
@@
 
+\subsection{Non-rectangular grids}
+
+Rectangular, but not in this projection.  Other types of regular grids.
+
+\subsection{Irregular locations}
+
+\section{Conclusions}
+
+\bibliography{references}
+
 \end{document}

### 2011-08-24 code/0-glyphs.r

@@ -17,9 +17,9 @@ glyphs <- function(data, x_major, x_minor, y_major, y_minor, polar = F, height =
   }
   
   if (polar) {
-    data$gx <- data[[x_major]] + rescale01(data[[y_minor]]) * 
+    data$gx <- data[[x_major]] + width / 2 * rescale01(data[[y_minor]]) * 
       cos(2 * pi * rescale01(data[[x_minor]]))
-    data$gy <- data[[y_major]] + rescale01(data[[y_minor]]) * 
+    data$gy <- data[[y_major]] + height / 2 * rescale01(data[[y_minor]]) * 
       sin(2 * pi * rescale01(data[[x_minor]]))
     data <- data[order(data[[x_major]], data[[x_minor]]), ]
   } else {

### 2011-08-24 code/0-nasa.r

@@ -17,6 +17,17 @@ if (!file.exists("nasadata.rds")) {
 
 outlines <- as.data.frame(map("world", plot = FALSE,
   xlim = -c(113.8, 56.2), ylim = c(-21.2, 36.2))[c("x","y")])
-map <- geom_path(aes(x, y), inherit.aes = FALSE, data = outlines, 
-  alpha = 0.4, legend = FALSE)
+map <- list(
+  geom_path(aes(x, y), inherit.aes = FALSE, data = outlines, alpha = 0.4,
+    legend = FALSE),
+  scale_x_continuous(breaks = NA),
+  scale_y_continuous(breaks = NA), 
+  xlab(NULL),
+  ylab(NULL),
+  coord_cartesian(xlim = c(-114.93, -55.07), ylim = c(-21.53, 36.64), wise = T))
+
+
+tl 
+map_mini <- c(map, list(
+  coord_cartesian(xlim = c(-114.93, -85), ylim = c(7.55, 36.64), wise = T)))
 

### 2011-08-24 code/1-modeling.r

@@ -4,36 +4,42 @@ library("mgcv")
 
 source("0-glyphs.r")
 source("0-nasa.r")
+source("0-cache.r")
 
-# Coefficients ---------------------------------------------------------------
+# Scatteplots ----------------------------------------------------------------
 
-temp_models <- dlply(nasa, c("lat", "long"), function(df) {
-  # lm(ozone ~ factor(month) - 1, data = df)
-  lm(surftemp ~ factor(month), data = df)
+temp <- glyphs(nasa, "long", "temperature", "lat", "surftemp") 
+qplot(gx, gy, data = temp, shape = I(".")) + map 
+
+cloud <- glyphs(nasa, "long", "cloudlow", "lat", "cloudhigh") 
+qplot(gx, gy, data = cloud, shape = I(".")) + 
+  geom_tile(aes(long, lat), colour = "white", fill = NA) + map 
+ggsave("../images/clouds.png", width = 6, height = 6)
+
+grid_along <- function(x, n = 20) {
+  rng <- range(x, na.rm = TRUE)
+  seq(rng[1], rng[2], length = n)
+}
+
+cloud_mod <- ddply(nasa, c("lat", "long"), function(df) {
+  grid_low <- data.frame(cloudlow = grid_along(df$cloudlow))
+  mod <- loess(cloudhigh ~ cloudlow, data = df, na = na.exclude)
+  mutate(grid_low, cloudhigh = predict(mod, newdata = grid_low))
 })
+cloud_mod <- glyphs(cloud_mod, "long", "cloudlow", "lat", "cloudhigh") 
+qplot(gx, gy, data = cloud_mod, geom = "line", group = gid) + 
+  geom_tile(aes(long, lat), colour = "white", fill = NA) + map 
+ggsave("../images/clouds-smooth.png", width = 6, height = 6)
 
 
-coefs <- ldply(temp_models, function(x) {
-  data.frame(month = 2:12, coef = coef(x)[-1])
-})
+# Model fitting --------------------------------------------------------------
 
-# Show seasonal patterns
-coefs <- glyphs(coefs, "long", "month", "lat", "coef") 
-qplot(gx, gy, data = coefs, geom = "line", group = gid) + map 
-qplot(gx, gy, data = coefs, geom = "line", group = gid) + 
-  geom_path(aes(y = lat), colour = "grey50", alpha = 0.75)
-
-coefs <- glyphs(coefs, "long", "month", "lat", "coef", polar = T) 
[…]

### 2011-08-25 paper.tex
shaded map instead of outline
filling in paper

--- 2011-08-18 paper.tex
@@
 \documentclass{article}
-\includepackage{graphicx}
+\usepackage{graphicx}
 
@@
 
-# Drawing trend lines on map
-#   - How to
-#   - Comparison to coloring slope
-#   - Reference to Grinstein icon plots
-#   - Alexander Gribov's work
+% Drawing trend lines on map
+%   - How to
+%   - Comparison to coloring slope
+%   - Reference to Grinstein icon plots
+% Pickett RM and Grinstein GG (1988). Iconographics 
+% displays for visualizing multidimensional data. 
+% In Proc. IEEE Conference on Systems, Man, and Cybernetics, pages 514-19.
+% Note the later work on metaphoric data display, garbage work!
+%   - Alexander Gribov's work http://rosuda.org/software/Gauguin/gauguin.html
 
-# Interactive graphics
-# Data processing
+% Interactive graphics
+% Data processing
 
-# Perhaps the GHCN data? Not gridded
-# NOAA data, almost grid, but over ocean, so maps more tricky
-# NARCAP from NCAR?
-# Fill in with NASA data, to start, try to put new data for final version
+% Perhaps the GHCN data? Not gridded
+% NOAA data, almost grid, but over ocean, so maps more tricky
+% NARCAP from NCAR?
+% Fill in with NASA data, to start, try to put new data for final version
+
+\maketitle
 
@@
 
-Need a colorful map
+Need a colorful map, facetted by year \& month
+
+\begin{figure*}[htp]
+\includegraphics[width=6in]{images/nasa-colored-map.png}
+\caption{Facetted map of de-seasonalized temperature.}
+\end{figure*}
 
 \section{Generating a time series at each location}
+
+\begin{figure*}[htp]
+\centerline{\includegraphics[width=4in]{images/nasa-deseas-trend.pdf}}
+\caption{Time series icon at each location, showing de-seasonalized trend.}
+\end{figure*}
 

### 2011-08-25 code/0-glyphs.r

@@ -11,8 +11,17 @@ glyphs <- function(data, x_major, x_minor, y_major, y_minor, polar = F, height =
     width <- resolution(data[[x_major]]) * 0.9    
     message("Using width ", format(width, digits = 3))
   }
+  else {
+    width <- resolution(data[[x_major]]) * width    
+    message("Using width ", format(width, digits = 3))
+  }
+    
   if (is.null(height)) {
     height <- resolution(data[[y_major]]) * 0.9    
+    message("Using height ", format(height, digits = 3))
+  }
+  else {
+    height <- resolution(data[[y_major]]) * height    
     message("Using height ", format(height, digits = 3))
   }
   

### 2011-08-25 paper.tex

--- 2011-08-24 paper.tex
@@
 %   - Alexander Gribov's work http://rosuda.org/software/Gauguin/gauguin.html
+
 %   - Spatial star plots by Andrews http://www.udallas.edu:8080/~andrews/software/software.html
@@
 
+\begin{figure*}[htp]
+\includegraphics[width=6in]{images/nasa-colored-map.png}
+\caption{Facetted map of de-seasonalized temperature.}
+\end{figure*}
+
 \section{Generating a time series at each location}
+
+\begin{figure*}[htp]
+\centerline{\includegraphics[width=4in]{images/nasa-deseas-trend.pdf}}
+\caption{Time series icon at each location, showing de-seasonalized trend.}
+\end{figure*}
+
+\section{Showing seasonal effect, radially oriented icon}
 
@@
 
+\begin{figure}[htbp]
+  \centering
+    \includegraphics[width=0.5\linewidth]{clouds-heike}%
+  \caption{Just as an alternative: shaded map instead of the outlines. }
+  \label{fig:cloud-alt}
+\end{figure}
+
 \section{Effect of scale}

### 2011-08-25 paper.tex
fixing spelling

@@
 
-\title{Glyph-maps for visualising climate data and models}
+\title{Glyph-maps for visualizing climate data and models}
 \author{Cook, Hofmann, Wickham, Wickham, Cheng}
@@
 
-Need a colourful map, facetted by year \& month
+Need a colorful map, facetted by year \& month
 
@@
   \centering
-  \includegraphics[width=0.5\linewidth]{ref-line}%
-  \includegraphics[width=0.5\linewidth]{ref-box}
+  %\includegraphics[width=0.5\linewidth]{ref-line}%
+  %\includegraphics[width=0.5\linewidth]{ref-box}
   \caption{caption}
@@
   \centering
-    \includegraphics[width=0.5\linewidth]{ref-max-1}%
-    \includegraphics[width=0.5\linewidth]{ref-max-2}
+    %\includegraphics[width=0.5\linewidth]{ref-max-1}%
+    %\includegraphics[width=0.5\linewidth]{ref-max-2}
   \caption{caption}
@@
   \centering
-  \includegraphics[width=2in]{month-cartesian}
-  \includegraphics[width=2in]{month-polar}
+  %\includegraphics[width=2in]{month-cartesian}
+  %\includegraphics[width=2in]{month-polar}
 
@@
 Figure~\ref{fig:cloud} shows how by mapping another variable to the x-axis, in this case low-cloud level on x and high-cloud level on y, we can explore how this 2d distribution varies over space.
-Again, summarising the data with models is very useful: the second plot displays a loess curve fit to the data instead of the individual points.
+Again, summarizing the data with models is very useful: the second plot displays a loess curve fit to the data instead of the individual points.
 
@@
   \centering
-    \includegraphics[width=0.5\linewidth]{clouds}%
-    \includegraphics[width=0.5\linewidth]{clouds-smooth}
+   % \includegraphics[width=0.5\linewidth]{clouds}%
+    %\includegraphics[width=0.5\linewidth]{clouds-smooth}
   \caption{caption}
@@
   \centering
-    \includegraphics[width=0.5\linewidth]{clouds-heike}%
+    %\includegraphics[width=0.5\linewidth]{clouds-heike}%
   \caption{Just as an alternative: shaded map instead of the outlines. }
@@
   \centering
-  \includegraphics[width=0.5\linewidth]{month-rescale-max}%
-  \includegraphics[width=0.5\linewidth]{month-rescale01}
+  %\includegraphics[width=0.5\linewidth]{month-rescale-max}%
+  %\includegraphics[width=0.5\linewidth]{month-rescale01}
   \caption{caption}
@@
 
-Other types of scaling: scale range to 0-1 (subtract min and divide by range), standardise mean and standard deviation (or robust variants).
+Other types of scaling: scale range to 0-1 (subtract min and divide by range), standardize mean and standard deviation (or robust variants).
 

### 2011-08-25 nasa.r

@@ -1,0 +1,89 @@
+setwd("/Users/dicook/papers/ClimateChange")
+nasa <- readRDS(file.choose()) # nasadata.rds
+
+library("maps")
+library("ggplot2")
+
+outlines <- as.data.frame(map("world", plot = FALSE,
+  xlim = -c(113.8, 56.2), ylim = c(-21.2, 36.2))[c("x","y")])
+map <- geom_path(aes(x, y), inherit.aes = FALSE, data = outlines, 
+  alpha = 0.2, legend = FALSE)
+
+# Raw data
+qplot(long, lat, data=nasa, geom="tile", fill=surftemp, facets=year~month) + scale_fill_gradient2(midpoint=296.9, mid="white") + map
+
+# Fit models
+# Seasonal only
+temp_seas_models <- dlply(nasa, c("lat", "long"), function(df) {
+  # lm(ozone ~ factor(month) - 1, data = df)
+  lm(surftemp ~ factor(month), data = df)
+})
+
+# Including trend
+temp_trend_models <- dlply(nasa, c("lat", "long"), function(df) {
+  lm(surftemp ~ year + factor(month), data = df)
+})
+
+locs <- dlply(nasa, c("lat", "long"))
+
+resids <- mdply(cbind(d = locs, m = temp_models), function(d, m) {
+  d$temp_resid <- d$surftemp - predict(m, newdata = d)
+  d
+})
+
+preds <- mdply(cbind(d = locs, m = temp_models), function(d, m) {
+  d$temp_pred <- predict(m, newdata = d)
+  d
+})
+
+# De-seasonalized
+theme_nothing <- function (base_size = 12) {
+  structure(list(
+	panel.grid.minor=theme_blank(), 
+	panel.grid.major=theme_blank(), 
+	panel.background=theme_blank(), 
+	panel.border = theme_blank(), 
+	panel.margin = unit(0, "lines"), 
+	axis.title.x=theme_blank(), 
+	axis.title.y=theme_blank(), 
+	axis.line=theme_blank(), 
+	axis.ticks=theme_blank(), 
+	axis.text.y = theme_blank(), 
+	axis.text.x = theme_blank(),
+	legend.position = "none"
+  ), class = "options")
+}
+
+qplot(long, lat, data=resids, geom="tile", fill=temp_resid, facets=year~month) + scale_fill_gradient2(midpoint=0, mid="white", high="red", low="blue") + map + theme_nothing()
+ggsave("nasa-colored-map.png", height=10, width=18)
+       
[…]

### 2011-08-25 code/1-heike-scatter.R
shaded map in the background

@@ -1,0 +1,32 @@
+library(ggplot2)
+library(maps)
+
+getbox <- function (map, xlim, ylim) {
+	# identify all regions involved
+	small <- subset(map, (long > xlim[1]) & (long < xlim[2]) & (lat > ylim[1]) & (lat < ylim[2]))
+	regions <- unique(small$region)
+	small <- subset(map, region %in% regions)	
+	
+	# now shrink all nodes back to the bounding box
+	small$long <- pmax(small$long, xlim[1])
+	small$long <- pmin(small$long, xlim[2])
+	small$lat <- pmax(small$lat, ylim[1])
+	small$lat <- pmin(small$lat, ylim[2])
+
+	return(small)
+}
+
+world <- map_data("world")
+nasa <- getbox(world, xlim = c(-114.93, -55.07), ylim = c(-21.53, 36.64))
+
+
+map <- list(
+  geom_polygon(aes(long, lat, group=group), inherit.aes = FALSE, data = nasa, fill="grey80", 
+    legend = FALSE),
+  scale_x_continuous(breaks = NA),
+  scale_y_continuous(breaks = NA), 
+  xlab(NULL),
+  ylab(NULL),
+  coord_cartesian(xlim = c(-114.93, -55.07), ylim = c(-21.53, 36.64), wise = T))
+
+ggplot() + map + geom_point(aes(gx, gy), data = temp, shape = ".")

### 2011-08-25 paper.tex
working on intro

@@
 \title{Glyph-maps for visualizing climate data and models}
-\author{Cook, Hofmann, Wickham, Wickham, Cheng}
+\author{Cook, Hofmann, Wickham, Wickham}
 
@@
 
+Climate data is usually composed of measurements on variables such as temperature, precipitation, and winds that have been collected from a spatial and temporal context.
+The classic display for spatial data is to use color or grey scale overlaid on geographic location, typically including a map.
+When measurements are made a different time periods also a common way to display the data is in small multiples, such as in Figure \ref{fig:facetted-map}.
+In this plot, a separate map is drawn for each month and year, over a 6 year period of remotely sensed temperature data above Central America.
+Color is used to display the de-seasonalized temperature measurements, with red being high values and blue being low values.
+The data  comes from \cite{NASA-Data-Expo}.
+
+\begin{figure*}[htp]
+\includegraphics[width=6in]{nasa-colored-map.png}
+\caption{Facetted map of de-seasonalized temperature.}
+\label{fig:facetted-map}
+\end{figure*}
+
 Need a colorful map, facetted by year \& month
+
+\subsection{Glyph plots}
 
@@
 Will first explore for regularly gridded data, a simple starting point and a common output of climate models, and then show a few simple extensions to deal with non-rectangular grids and irregular locations, as exemplified by raw climate data which is typically collect at irregular locations.
-
-\begin{figure*}[htp]
-\includegraphics[width=6in]{images/nasa-colored-map.png}
-\caption{Facetted map of de-seasonalized temperature.}
-\end{figure*}
-
 \section{Generating a time series at each location}
@@
 \begin{figure*}[htp]
-\centerline{\includegraphics[width=4in]{images/nasa-deseas-trend.pdf}}
+\centerline{\includegraphics[width=4in]{nasa-deseas-trend.pdf}}
 \caption{Time series icon at each location, showing de-seasonalized trend.}
@@
   \centering
-  %\includegraphics[width=0.5\linewidth]{ref-line}%
-  %\includegraphics[width=0.5\linewidth]{ref-box}
+  \includegraphics[width=0.5\linewidth]{ref-line}%
+  \includegraphics[width=0.5\linewidth]{ref-box}
   \caption{caption}

### 2011-08-25 nasa.r
tweaking my code

@@ -49,12 +49,12 @@ theme_nothing <- function (base_size = 12) {
 	axis.line=theme_blank(), 
 	axis.ticks=theme_blank(), 
 	axis.text.y = theme_blank(), 
-	axis.text.x = theme_blank(),
-	legend.position = "none"
+	axis.text.x = theme_blank()#,
+	#legend.position = "none"
   ), class = "options")
 }
 
-qplot(long, lat, data=resids, geom="tile", fill=temp_resid, facets=year~month) + scale_fill_gradient2(midpoint=0, mid="white", high="red", low="blue") + map + theme_nothing()
+qplot(long, lat, data=resids, geom="tile", fill=temp_resid, facets=year~month) + scale_fill_gradient2(midpoint=0, mid="white", high="red", low="blue") + map + theme_nothing() + coord_map()
 ggsave("nasa-colored-map.png", height=10, width=18)
        
 resids <- glyphs(resids, "long", "day", "lat", "temp_resid") 

### 2011-08-25 paper.tex
working on intro

@@
 
-Climate data is usually composed of measurements on variables such as temperature, precipitation, and winds that have been collected from a spatial and temporal context.
+Climate data is composed of measurements on variables such as temperature, precipitation, and winds that have been collected from a spatial and temporal context.
 The classic display for spatial data is to use color or grey scale overlaid on geographic location, typically including a map.
 When measurements are made a different time periods also a common way to display the data is in small multiples, such as in Figure \ref{fig:facetted-map}.
-In this plot, a separate map is drawn for each month and year, over a 6 year period of remotely sensed temperature data above Central America.
+In this plot, a separate map is drawn for each month (columns) and year (rows), over a 6 year period of remotely sensed temperature data above Central America.
 Color is used to display the de-seasonalized temperature measurements, with red being high values and blue being low values.
 The data  comes from \cite{NASA-Data-Expo}.
+The most noticeable feature in this display is the strong red in the equatorial region beginning in mid-1997 and tapering out during 1998.
+This is the pattern of an El Ni\~no event, a major temperature anomaly.
+With some work other smaller more localized patterns can be perceived, for example, cooler over the land mass in the early year, but warmer in the latter years.
 
 \begin{figure*}[htp]
-\includegraphics[width=6in]{nasa-colored-map.png}
+\centerline{\includegraphics[width=6in]{nasa-colored-map.png}}
 \caption{Facetted map of de-seasonalized temperature.}
@@
 
-Need a colorful map, facetted by year \& month
+In this paper we explore the use of a different type of display, that uses icons on the map.
+Icon plots date back to the star glyphs of  \citep{kleiner}, made infamous by \citep{Chernoff}.
+They are used, infrequently to examine multivariate data.
 
-\subsection{Glyph plots}
+%\subsection{Glyph plots}
 
@@
   \centering
-    %\includegraphics[width=0.5\linewidth]{ref-max-1}%
-    %\includegraphics[width=0.5\linewidth]{ref-max-2}
+    \includegraphics[width=0.5\linewidth]{ref-max-1}%
+    \includegraphics[width=0.5\linewidth]{ref-max-2}
   \caption{caption}
@@
   \centering
-  %\includegraphics[width=2in]{month-cartesian}
-  %\includegraphics[width=2in]{month-polar}
+  \includegraphics[width=2in]{month-cartesian}
+  \includegraphics[width=2in]{month-polar}
 
@@
   \centering
-   % \includegraphics[width=0.5\linewidth]{clouds}%
-    %\includegraphics[width=0.5\linewidth]{clouds-smooth}
+   \includegraphics[width=0.5\linewidth]{clouds}%
+   \includegraphics[width=0.5\linewidth]{clouds-smooth}
   \caption{caption}
@@
   \centering
-    %\includegraphics[width=0.5\linewidth]{clouds-heike}%
+    \includegraphics[width=0.5\linewidth]{clouds-heike}%
   \caption{Just as an alternative: shaded map instead of the outlines. }
@@
   \centering
-  %\includegraphics[width=0.5\linewidth]{month-rescale-max}%
-  %\includegraphics[width=0.5\linewidth]{month-rescale01}
+  \includegraphics[width=0.5\linewidth]{month-rescale-max}%
+  \includegraphics[width=0.5\linewidth]{month-rescale01}
   \caption{caption}

### 2011-08-25 paper.tex
tweaking intro

@@
 
-In this paper we explore the use of a different type of display, that uses icons on the map.
+In this paper we explore the use of a different type of display, that uses icons on the map, for climate data.
 Icon plots date back to the star glyphs of  \citep{kleiner}, made infamous by \citep{Chernoff}.
-They are used, infrequently to examine multivariate data.
+They are used, infrequently, to examine multivariate data.
+An icon is produced by mapping each variable to some feature of the icon.
+One icon is generated for every case in the data set.
+For large data sets it is infeasible to digest the quantity of icons, and for multivariate data there is no natural order so it can be like having a mountain of little stamps to rearrange and organize, in order to understand the structures present in the data.
+For spatiotemporal data ordering icons is naturally done by spatial location.
+There is also usually some spatial trend, or continuity, which will make digesting patterns among the icons simpler.
 

### 2011-08-26 paper.tex
tweaking intro

@@
 With some work other smaller more localized patterns can be perceived, for example, cooler over the land mass in the early year, but warmer in the latter years.
+What the reader is required to do is play the game "Spot the difference" from one small image to another.
+It's a very hard task for readers to do.
+Large structure such as the El Ni\~no event are clear but readers will not be able to effectively glue the images together sufficiently to read off the trend over the years or notice small local deviations between plots.
 

### 2011-08-26 paper.tex
Reference lines

@@
 
-Space-time-glyphs often require reference lines: it is difficult to compare the relative location of glyphs in different regions.
-As noted by \citet{cleveland:1993a}, a \emph{visual reference grid} is important for accurate perception of values, particularly of important properties of lines like slope and minima.
-This is illustrated well by the previous plot of monthly coefficients.
-When we add a reference to each location, Figure~\ref{fig:ref-basic}, it becomes much more clear that not only is there a difference in shape between the Northern and Southern hemispheres, but there is also a location shift: values in the North are much higher than values in the South.
-It is difficult to see this without reference lines.
+Space-time-glyphs usually require a \emph{visual reference grid} \citet{cleveland:1993a} to make it easier to compare locations.
+This is important for accurate perception of important line properties like slope and minima.
+This need is well illustrated by Figure~\ref{}.
+When we add a reference to each location, as in Figure~\ref{fig:ref-basic}, we see not only a difference in shape between Northern and Southern hemispheres, but also average value.
+Without reference lines, it is difficult to see that average seasonal coefficients are much higher in the North than in the South.
 
-Figure~\ref{fig:ref-basic} shows two variants of a reference, a line drawn at the overall mid-range, and a box drawn around each grid cell.
-We prefer using white for such references because it is minimally perceptible: you can see it easily if you concentrate on it, but it does not otherwise distract from the perception of the graphic.
-Note the careful choice of layering: the map is drawn under mid-range lines, but over boxes.
+Figure~\ref{fig:ref-basic} shows two reference grid variants: a line drawn at the overall mid-range, and a box drawn around each grid cell.
+We prefer using white for these references because it is minimally perceptible: you can see the grid easily when looking for it, but it does not otherwise distract from the perception of the graphic \citep{carr:1994}.
 
@@
   \includegraphics[width=0.5\linewidth]{ref-box}
-  \caption{caption}
+
+  \caption{Glyph-maps of seasonal temperature patterns.
+Adding reference lines, (left) mid-range lines and (right) grid-cell boxes, makes it easier to see differences in glyph position, not just shape.}
   \label{fig:ref-basic}
@@
 
-Other types of reference information can also be usefully added. 
-Other ways of display additional data:
-
-\begin{itemize}
-
-  \item Map variable to colour or transparency of line.
-
-  \item Colour of reference lines/boxes.
-
-  \item Highlight special points in an additional layer.
-
-  \item Fill colour of reference boxes.
-
-\end{itemize}
-
-A particularly useful application of this technique is when display model predictions to also display some measure of model fit, so that appropriate care can be taken to avoid over-interpreting predictions from models that did not fit well.
-
-Figure~\ref{fig:ref-adv} shows two applications of these ideas, adding the month with the highest average temperature in either the foreground or background.
+It's also often useful to display other types of reference information.
+You are only limited by your imagination (and the capabilities of your graphics software), but two simple ideas are to (1) vary the colour or transparency of glyphs or references, or (2) use an additional layer to highlight special points.
+Figure~\ref{fig:ref-adv} illustrates these ideas to display the month with the highest temperature at each locations.
 Careful colour choice is necessary so that this information is perceptible, but not distracting.
@@
   \includegraphics[width=0.5\linewidth]{ref-max-2}
-  \caption{caption}
+  \caption{Glyph-maps of seasonal temperature patterns also showing the month with the highest temperature using (left) an additional layer of points and (right) changing the background colour of the reference grid.}
   \label{fig:ref-adv}
 \end{figure}
+
+A particularly useful summary when to add when displaying model predictions, is some measure of model fit. This makes it easier to avoid over-interpreting predictions from poorly-fitting models.
 

### 2011-08-26 paper.tex
Tidy up scatterplot section

@@
 
-\section{Scatterplots}
+\section{Generalization: scatterplots}
 
 So far we have limited ourselves to always mapping some aspect of time on the x-axis.
-But the method is much more general.
-Figure~\ref{fig:cloud} shows how by mapping another variable to the x-axis, in this case low-cloud level on x and high-cloud level on y, we can explore how this 2d distribution varies over space.
-Again, summarizing the data with models is very useful: the second plot displays a loess curve fit to the data instead of the individual points.
+But glyph-maps are not limited to only this type of data and can  be easily generalized to create many other types of display.
+Figure~\ref{fig:cloud} shows a generalisation to scatterplots: by mapping another variable to the x-axis, in this case low-cloud level on x and high-cloud level on y, we can explore how a bivariate relationship varies spatially.
+It is again useful to use a model to extract the essence of the relationship, summarizing the data with models is very useful: the second plot displays a loess curve fit to the data instead of the individual points.
 
@@
   \includegraphics[width=0.5\linewidth]{clouds-smooth}
-  \caption{caption}
+
+  \caption{Glyph-map showing (left) scatterplot of high-cloud vs.
+low-cloud and (right) smoothed (loess) curve fit to each location.
+The relationship between these two variables varies considerably over the spatial domain (particularly between land and ocean), but neighbouring locations tend to similar.
+}
   \label{fig:cloud}
-\end{figure}
-
-\begin{figure}[htbp]
-  \centering
-    \includegraphics[width=0.5\linewidth]{clouds-heike}%
-  \caption{Just as an alternative: shaded map instead of the outlines. }
-  \label{fig:cloud-alt}
 \end{figure}

### 2011-08-26 code/1-modeling.r

@@ -8,15 +8,11 @@ source("0-cache.r")
 
 # Scatteplots ----------------------------------------------------------------
 
-temp <- glyphs(nasa, "long", "temperature", "lat", "surftemp") 
-ggplot(temp, aes(gx, gy, group = gid)) + 
-  map +
-  geom_point(shape = ".")
-
 cloud <- glyphs(nasa, "long", "cloudlow", "lat", "cloudhigh") 
 ggplot(cloud, aes(gx, gy, group = gid)) + 
   map +
-  geom_point(shape = ".")
+  geom_point(shape = ".") +
+  geom_tile(aes(long, lat), colour = "white", fill = NA)
 ggsave("../images/clouds.png", width = 6, height = 6)
 
 grid_along <- function(x, n = 20) {

### 2011-08-26 nasa.r
tweak intro

@@ -87,3 +87,29 @@ ggsave("nasa-deseas-trend.pdf", height=8, width=8)
 year_resids <- glyphs(year_preds, "long", "year", "lat", "pred", height=3) 
 qplot(gx, gy, data = year_preds, geom = "path", group = gid, xlab="", ylab="") + map + coord_map()
 ggsave("nasa-deseas-trend.pdf", height=8, width=8)
+
+# Explain icons
+nasa.temp<-data.frame(time=nasa$time, temp=nasa$surftemp)
+nasa.temp<-cast(nasa.temp, formula=temp~time)
+nasa.oneloc <- subset(nasa, x == 22 & y == 21, select=c("id","time","surftemp"))
+nasa.twoloc <- subset(nasa, x == 7 & y == 12, select=c("id","time","surftemp"))
+nasa.threeloc <- subset(nasa, x == 12 & y == 3, select=c("id","time","surftemp"))
+nasa.sub<-rbind(nasa.oneloc, nasa.twoloc, nasa.threeloc)
+#nasa.oneloc<-t(as.matrix(nasa.oneloc$surftemp))
+#colnames(nasa.oneloc)<-paste("t", 1:72, sep="")
+nasa.oneloc$gx<-cos(2*pi*(nasa.oneloc$time/72)) * ((nasa.oneloc$surftemp-293)/(302-293))
+nasa.oneloc$gy<-sin(2*pi*(nasa.oneloc$time/72)) * ((nasa.oneloc$surftemp-293)/(302-293))
+qplot(time, surftemp, data=nasa.oneloc, geom="path", xlab="Time", ylab="Temperature (Kelvin)")
+ggsave("nasa-oneloc-ts.pdf", height=2, width=8)
+qplot(gx, gy, data=nasa.oneloc, geom="path") + theme_nothing()
+
+nasa.sub$lon<-rep(c(2,4,6), rep(72,3))
+nasa.sub$lat<-rep(1,216)
+
+nasa.sub<-glyphs(nasa.sub, "lon", "time", "lat", "surftemp", width=0.5, height=3, polar=TRUE)
+
+qplot(time, surftemp, data=nasa.sub, geom="path", xlab="Time", ylab="Temperature (Kelvin)", group=id, colour=gid)
+ggsave("nasa-glyph-ts.pdf", height=2, width=8)
+
[…]

### 2011-08-26 paper.tex
tweak intro
glyph plot explanation

@@
 In this paper we explore the use of a different type of display, that uses icons on the map, for climate data.
-Icon plots date back to the star glyphs of  \citep{kleiner}, made infamous by \citep{Chernoff}.
+Icon plots date back to the star glyphs of  \citep{kleiner}, and were made infamous by \citep{Chernoff}.
 They are used, infrequently, to examine multivariate data.
@@
 There is also usually some spatial trend, or continuity, which will make digesting patterns among the icons simpler.
+
+\begin{figure*}[htp]
+\centerline{\includegraphics[width=6in]{nasa-glyph-ts.pdf}}
+\centerline{\includegraphics[width=6in]{nasa-glyph.pdf}}
+\caption{Glyphs of temperature for three different locations.}
+\label{fig:glyphs}
+\end{figure*}
 

### 2011-08-26 paper.tex
tweak intro

@@
 There is also usually some spatial trend, or continuity, which will make digesting patterns among the icons simpler.
+
+Figure \ref{fig:glyphs} illustrates the idea of icon plots, with emphasis on temporal data.
+The time measurements are treated as variables in a multivariate data set.
+Here we have six years of monthly data, giving 72 time values which are considered to be 72 variables.
+Three different locations are selected, for demontration purposes.
+The top plot shows the time series of these three locations.
+Location ``1.1'' (red) has medium values of temperature, with strong seasonality, the six years corresponding to the six eaks.
+Location ``2.1'' (green) has generally high temperature values, with weaker seasonality in most years and a lack of low temperatures in the middle of the time period when lows would be expected.
+Location ``3.1'' (blue) has strong seasonality but generally low temperatures.
+The bottom plot shows these same cases as star glyphs.
+A separate glyph is created for each location.
+The 72 time variables form axes organized radially.
+Each temperature value is plotted relative to the overall range of temperature values.
+The seasonal patterns are clearly visible in this type of display, forming six ``petals'' in the icon.
+Location ``2.1'' has a different pattern than the other two, with only four petals, and location ``3.1'' is a much smaller ico indicating overall lower values.
 

### 2011-08-26 nasa.r
glyph plots code

@@ -108,6 +108,12 @@ nasa.sub$lat<-rep(1,216)
 
 nasa.sub<-glyphs(nasa.sub, "lon", "time", "lat", "surftemp", width=0.5, height=3, polar=TRUE)
 
+nasa.sub$gid<-as.character(nasa.sub$gid)
+nasa.sub$gid[nasa.sub$gid=="2.1"]<-"1.1"
+nasa.sub$gid[nasa.sub$gid=="4.1"]<-"2.1"
+nasa.sub$gid[nasa.sub$gid=="6.1"]<-"3.1"
+nasa.sub$gid<-factor(nasa.sub$gid)
+
 qplot(time, surftemp, data=nasa.sub, geom="path", xlab="Time", ylab="Temperature (Kelvin)", group=id, colour=gid)
 ggsave("nasa-glyph-ts.pdf", height=2, width=8)
 

### 2011-08-26 paper.tex
Tweak rescaling graphics

@@
 
-\section{Effect of scale}
+\section{Partitioning patterns into shape and size}
 
@@
   \centering
-  \includegraphics[width=0.5\linewidth]{month-rescale-max}%
-  \includegraphics[width=0.5\linewidth]{month-rescale01}
-  \caption{caption}
+  \includegraphics[width=0.33\linewidth]{month-rescale-none}%
+  \includegraphics[width=0.33\linewidth]{month-rescale-max}%
+  \includegraphics[width=0.33\linewidth]{month-rescale01}
+  \caption{Smoothed de-seasonalized daily temperature glyph-map. (Left) unscaled, (center) scaled to have maximum 1, (right) scaled to have maximum 1 and minimum 0.}
   \label{fig:label}
@@
 
-Other types of scaling: scale range to 0-1 (subtract min and divide by range), standardize mean and standard deviation (or robust variants).
+Of course other types of scaling are possible. Another classical scaling method is to standardize mean and standard deviation to 0 and 1 (or robust variants).
 

### 2011-08-26 code/0-nasa.r

@@ -47,6 +47,32 @@ map <- list(
   scale_y_continuous(breaks = NA, expand = c(0.02, 0)), 
   xlab(NULL),
   ylab(NULL))
+  
+theme_fullframe <- function (base_size = 12){
+  structure(list(
+    axis.line = theme_blank(), 
+    axis.text.x = theme_blank(), 
+    axis.text.y = theme_blank(),
+    axis.ticks = theme_blank(), 
+    axis.title.x = theme_blank(), 
+    axis.title.y = theme_blank(), 
+    axis.ticks.length = unit(0, "lines"), 
+    axis.ticks.margin = unit(0, "lines"), 
+    legend.position = "none", 
+    panel.border = theme_blank(), 
+    panel.grid.major = theme_blank(), 
+    panel.grid.minor = theme_blank(), 
+    panel.margin = unit(0, "lines"), 
+    plot.background = theme_blank(), 
+    plot.margin = unit(rep(2, 4), "mm")
+  ), class = "options")
+}
+
+ref_boxes <- list(
+  geom_tile(aes(long, lat), colour = "white", fill = NA, inherit.aes = FALSE),
+  scale_x_continuous(breaks = NA, expand = c(0, 0)),
+  scale_y_continuous(breaks = NA, expand = c(0, 0))
+)
 
 
 map_mini <- c(map, list(

### 2011-08-26 code/1-modeling.r

@@ -131,6 +131,16 @@ ggplot(month_preds, aes(gx, gy, group = gid)) +
 ggsave("../images/month-polar.pdf", width = 4, height = 4)
 
 # Rescale predictions to individual scales -----------------------------------
+
+day_preds <- glyphs(day_preds, "long", "day", "lat", "pred") 
+ggplot(day_preds, aes(gx, gy, group = gid)) + 
+  map + 
+  geom_path() +
+  ref_boxes +
+  theme_fullframe()
+ggsave("../images/month-rescale-none.pdf", width = 4, height = 4)
+
+
 day_preds2 <- ddply(day_preds, c("lat", "long"), mutate, 
   pred_s = rescale01(pred),
   pred_m = pred / max(pred))
@@ -139,14 +149,16 @@ day_preds2 <- glyphs(day_preds2, "long", "day", "lat", "pred_s")
 ggplot(day_preds2, aes(gx, gy, group = gid)) + 
   map + 
   geom_path() +
-  geom_tile(aes(long, lat), colour = "white", fill = NA)
+  ref_boxes +
+  theme_fullframe()
 ggsave("../images/month-rescale01.pdf", width = 4, height = 4)
 
 day_preds2 <- glyphs(day_preds2, "long", "day", "lat", "pred_m") 
 ggplot(day_preds2, aes(gx, gy, group = gid)) + 
   map + 
   geom_path() +
-  geom_tile(aes(long, lat), colour = "white", fill = NA)
+  ref_boxes +
+  theme_fullframe()
 ggsave("../images/month-rescale-max.pdf", width = 4, height = 4)
 
 # Linear trend is not a good fit!

### 2011-08-26 paper.tex
first glyph-map added

@@
 In this paper we explore the use of a different type of display, that uses icons on the map, for climate data.
-Icon plots date back to the star glyphs of  \citep{kleiner}, and were made infamous by \citep{Chernoff}.
-They are used, infrequently, to examine multivariate data.
+Icon plots date back to the star glyphs of  \citep{kleiner:1981}, and were made infamous by \citep{Chernoff}.
 An icon is produced by mapping each variable to some feature of the icon.
 One icon is generated for every case in the data set.
-For large data sets it is infeasible to digest the quantity of icons, and for multivariate data there is no natural order so it can be like having a mountain of little stamps to rearrange and organize, in order to understand the structures present in the data.
-For spatiotemporal data ordering icons is naturally done by spatial location.
-There is also usually some spatial trend, or continuity, which will make digesting patterns among the icons simpler.
-
 Figure \ref{fig:glyphs} illustrates the idea of icon plots, with emphasis on temporal data.
@@
 
+Icon plots are used, infrequently, to examine multivariate data.
+For large data sets it is infeasible to digest the quantity of icons, and for multivariate data there is no natural order so it can be like having a mountain of little stamps to rearrange and organize, in order to understand the structures present in the data.
+The initial paper on star glyphs \citep{kleiner:1981} and a recent paper \citep{hurley:2010} propose some approaches for organizing the icons.
+With spatiotemporal data ordering of icons is naturally done by spatial location.
+There is also usually some spatial trend, or continuity, which will make digesting patterns among the icons simpler.
+
+Figure \ref{fig:glyph-map} shows the glyphs drawn for each spatial location of the raw temperature data.
+
+\begin{figure*}[htp]
+\centerline{\includegraphics[width=5in]{nasa-glyph-all.pdf}}
+\caption{Glyph-map of raw data.}
+\label{fig:glyph-map}
+\end{figure*}
+
 %\subsection{Glyph plots}
 
-Simple, but effective, technique that combines aspects of glyphs, facetted plots and map-charts. Particularly well suited for the type of temporal-spatial data, commonly seen in climate.
+%Simple, but effective, technique that combines aspects of glyphs, facetted plots and map-charts. Particularly well suited for the type of temporal-spatial data, commonly seen in climate.
 
-Glyph-maps have attributes of both glyphs (stars, faces, profiles, ...), inviting exploration at the global level, and small-multiple time series, allowing local exploration of individual series.
+%Glyph-maps have attributes of both glyphs (stars, faces, profiles, ...), inviting exploration at the global level, and small-multiple time series, allowing local exploration of individual series.
 Allow to see both geographic context and individual locations simultaneously.
 
-An interesting attribute of these glyph plots is that the position of graphical elements is determined by major and minor axes.
+%An interesting attribute of these glyph plots is that the position of graphical elements is determined by major and minor axes.
 For maps, the major axes are latitude and longitude, and the minor axes are typically time on the x-axis and some measurement or model prediction on the y-axis.
 
-Avoids one problem that plagues glyph displays: the ordering of the variables.
+%Avoids one problem that plagues glyph displays: the ordering of the variables.
 Although some techniques exist to ameliorate this problem \citep{kleiner:1981,hurley:2010}, they are not necessary with time data because it is intrinsically ordered.
 
-These displays tend to be particularly effective when printed in large format, as the much higher resolution of print (600 dpi) compared to computer  screen (72-120 dpi) allows for much finer reading.
+%These displays tend to be particularly effective when printed in large format, as the much higher resolution of print (600 dpi) compared to computer  screen (72-120 dpi) allows for much finer reading.
 These are high-density data displays in the best tradition of Tufte.
@@
 
-Will first explore for regularly gridded data, a simple starting point and a common output of climate models, and then show a few simple extensions to deal with non-rectangular grids and irregular locations, as exemplified by raw climate data which is typically collect at irregular locations.
+%Will first explore for regularly gridded data, a simple starting point and a common output of climate models, and then show a few simple extensions to deal with non-rectangular grids and irregular locations, as exemplified by raw climate data which is typically collect at irregular locations.
 \section{Generating a time series at each location}

### 2011-08-26 nasa.r

@@ -119,3 +119,7 @@ ggsave("nasa-glyph-ts.pdf", height=2, width=8)
 
 qplot(gx, gy, data=nasa.sub, geom="polygon", colour=gid, fill=I(NA)) + theme_nothing()
 ggsave("nasa-glyph.pdf", height=2, width=6)
+
+nasa.gly<-glyphs(nasa, "long", "time", "lat", "surftemp", height=3, polar=TRUE)
+ggplot() +  map + geom_polygon(data=nasa.gly, mapping=aes(x=gx, y=gy, group=id), colour=I("black"), fill=I(NA)) + theme_nothing() + coord_map()
+ggsave("nasa-glyph-all.pdf", height=6, width=6)

### 2011-08-26 nasa.r
theme_fullframe

@@ -121,5 +121,5 @@ qplot(gx, gy, data=nasa.sub, geom="polygon", colour=gid, fill=I(NA)) + theme_not
 ggsave("nasa-glyph.pdf", height=2, width=6)
 
 nasa.gly<-glyphs(nasa, "long", "time", "lat", "surftemp", height=3, polar=TRUE)
-ggplot() +  map + geom_polygon(data=nasa.gly, mapping=aes(x=gx, y=gy, group=id), colour=I("black"), fill=I(NA)) + theme_nothing() + coord_map()
+ggplot() +  map + geom_polygon(data=nasa.gly, mapping=aes(x=gx, y=gy, group=id), colour=I("black"), fill=I(NA)) + theme_fullframe() + coord_map()
 ggsave("nasa-glyph-all.pdf", height=6, width=6)

### 2011-08-26 nasa.r
first glyph-maps

@@ -123,3 +123,7 @@ ggsave("nasa-glyph.pdf", height=2, width=6)
 nasa.gly<-glyphs(nasa, "long", "time", "lat", "surftemp", height=3, polar=TRUE)
 ggplot() +  map + geom_polygon(data=nasa.gly, mapping=aes(x=gx, y=gy, group=id), colour=I("black"), fill=I(NA)) + theme_fullframe() + coord_map()
 ggsave("nasa-glyph-all.pdf", height=6, width=6)
+
+nasa.gly<-glyphs(nasa, "long", "time", "lat", "surftemp", height=1.5)
+ggplot() +  map + geom_path(data=nasa.gly, mapping=aes(x=gx, y=gy, group=id), colour=I("black")) + theme_fullframe() + coord_map()
+ggsave("nasa-glyph-ts-all.pdf", height=6, width=6)

### 2011-08-26 paper.tex
theme_fullframe
first glyph-maps
healey referenced for change blindness

@@
 In this plot, a separate map is drawn for each month (columns) and year (rows), over a 6 year period of remotely sensed temperature data above Central America.
-Color is used to display the de-seasonalized temperature measurements, with red being high values and blue being low values.
+Color is used to display the deseasonalized temperature measurements, with red being high values and blue being low values.
 The data  comes from \cite{NASA-Data-Expo}.
@@
 With some work other smaller more localized patterns can be perceived, for example, cooler over the land mass in the early year, but warmer in the latter years.
-What the reader is required to do is play the game "Spot the difference" from one small image to another.
-It's a very hard task for readers to do.
+What the reader is required to do is play the game ``Spot the difference" from one small image to another,.
+From a cognitive perspective, this is a very hard task to do.
+It falls into the framework of {\it change blindess} (see e.g.
+\citet{heahley:perception}  and often leads to failure without readers realizing, in particular for small areas of change or subtle changes.
 Large structure such as the El Ni\~no event are clear but readers will not be able to effectively glue the images together sufficiently to read off the trend over the years or notice small local deviations between plots.
@@
 \centerline{\includegraphics[width=6in]{nasa-colored-map.png}}
-\caption{Facetted map of de-seasonalized temperature.}
+\caption{Facetted map of deseasonalized temperature.}
 \label{fig:facetted-map}
@@
 Here we have six years of monthly data, giving 72 time values which are considered to be 72 variables.
-Three different locations are selected, for demontration purposes.
+Three different locations are selected, for demonstration purposes.
 The top plot shows the time series of these three locations.
-Location ``1.1'' (red) has medium values of temperature, with strong seasonality, the six years corresponding to the six eaks.
+Location ``1.1'' (red) has medium values of temperature, with strong seasonality, the six years corresponding to the six peaks.
 Location ``2.1'' (green) has generally high temperature values, with weaker seasonality in most years and a lack of low temperatures in the middle of the time period when lows would be expected.
@@
 
+% Add time series as well as radial
 \begin{figure*}[htp]
-\centerline{\includegraphics[width=5in]{nasa-glyph-all.pdf}}
-\caption{Glyph-map of raw data.}
+\centerline{\includegraphics[width=3.5in]{nasa-glyph-all.pdf}\includegraphics[width=3.5in]{nasa-glyph-ts-all.pdf}}
+\caption{Glyph-map of raw data, using star glyphs (left), and time series as glyphs (right).}
 \label{fig:glyph-map}

### 2011-08-26 paper.tex
Fleshing out scales section

@@
 
+Another impact of varying averages is that it tends to diminish the perceptibility of shape: if some series have uniformly high values and some have uniformly low, then it is difficult to see if the shape is the same independent of the average.
 Partitioning is a extremely useful technique: product plots, ANOVA, ...
 For glyph plots, it's often useful to partition the glyphs into shape and scale components, decomposing the values into a product of maximum and scaled values.
-This is revealing for smooth models of surface temperature.
+In this section we explore how decomposing each series into scale and shape components can be useful.
+
+This is revealing for smooth models of surface temperature, shown in Figure~\ref{fig:scaling}.
+Each location is modelled using a generalised additive model \citep{wood:2006} with a seasonal term and a long-term smooth term.
+The three plots in the figure show the long-term smooth term.
+Because some locations are uniformly warmer than others, the impact of changes across time are diminished.
+In the centre plot, each location is scaled by dividing out the maximum, and on the right, each location is scaled to range $[0, 1]$.
+These emphasise individual changes but lose the ability to make global comparisons, and need caution when interpreting - a small absolute change gets blown up into a large relative change.
 
@@
   \caption{Smoothed de-seasonalized daily temperature glyph-map. (Left) unscaled, (center) scaled to have maximum 1, (right) scaled to have maximum 1 and minimum 0.}
-  \label{fig:label}
+  \label{fig:scaling}
+\end{figure}
+
+A possible compromise is to use line colour to display the standardising variable, as shown in Figure~\ref{fig:scaling-col}.
+This makes it possible to see the individual patterns while still retaining some ability to see which shapes represent small absolute changes.
+
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=0.33\linewidth]{month-rescale01-col}
+  \caption{Adding colour to scaled plots. Here each location has been scaled to range $[0, 1]$ and colour mapped to the range of the original predictions.}
+  \label{fig:scaling-col}
 \end{figure}

### 2011-08-26 code/1-modeling.r

@@ -138,12 +138,15 @@ ggplot(day_preds, aes(gx, gy, group = gid)) +
   geom_path() +
   ref_boxes +
   theme_fullframe()
-ggsave("../images/month-rescale-none.pdf", width = 4, height = 4)
+ggsave("../images/month-rescale-none.png", width = 4, height = 4)
 
 
 day_preds2 <- ddply(day_preds, c("lat", "long"), mutate, 
   pred_s = rescale01(pred),
-  pred_m = pred / max(pred))
+  pred_m = pred / max(pred),
+  max = max(pred),
+  range = diff(range(pred))
+)
 day_preds2 <- glyphs(day_preds2, "long", "day", "lat", "pred_s") 
 
 ggplot(day_preds2, aes(gx, gy, group = gid)) + 
@@ -151,7 +154,7 @@ ggplot(day_preds2, aes(gx, gy, group = gid)) +
   geom_path() +
   ref_boxes +
   theme_fullframe()
-ggsave("../images/month-rescale01.pdf", width = 4, height = 4)
+ggsave("../images/month-rescale01.png", width = 4, height = 4)
 
 day_preds2 <- glyphs(day_preds2, "long", "day", "lat", "pred_m") 
 ggplot(day_preds2, aes(gx, gy, group = gid)) + 
@@ -159,9 +162,18 @@ ggplot(day_preds2, aes(gx, gy, group = gid)) +
   geom_path() +
   ref_boxes +
   theme_fullframe()
-ggsave("../images/month-rescale-max.pdf", width = 4, height = 4)
+ggsave("../images/month-rescale-max.png", width = 4, height = 4)
 
-# Linear trend is not a good fit!
+
+day_preds2 <- glyphs(day_preds2, "long", "day", "lat", "pred_s") 
+ggplot(day_preds2, aes(gx, gy, group = gid)) + 
+  map + 
+  geom_path(aes(colour = range)) +
+  ref_boxes +
+  theme_fullframe() + 
+  scale_colour_gradient(high = "black", low = "grey60")
+ggsave("../images/month-rescale01-col.png", width = 4, height = 4)
+
 
 # Residuals ------------------------------------------------------------------
 

### 2011-08-26 paper.tex
explained first glyph-maps

@@
 With some work other smaller more localized patterns can be perceived, for example, cooler over the land mass in the early year, but warmer in the latter years.
-What the reader is required to do is play the game ``Spot the difference" from one small image to another,.
+What the reader is required to do is play the game ``Spot the difference'' from one small image to another.
 From a cognitive perspective, this is a very hard task to do.
@@
 Here we have six years of monthly data, giving 72 time values which are considered to be 72 variables.
-Three different locations are selected, for demonstration purposes.
+Three different locations are selected, for demontration purposes.
 The top plot shows the time series of these three locations.
-Location ``1.1'' (red) has medium values of temperature, with strong seasonality, the six years corresponding to the six peaks.
+Location ``1.1'' (red) has medium values of temperature, with strong seasonality, the six years corresponding to the six eaks.
 Location ``2.1'' (green) has generally high temperature values, with weaker seasonality in most years and a lack of low temperatures in the middle of the time period when lows would be expected.
@@
 The seasonal patterns are clearly visible in this type of display, forming six ``petals'' in the icon.
-Location ``2.1'' has a different pattern than the other two, with only four petals, and location ``3.1'' is a much smaller ico indicating overall lower values.
+Location ``2.1'' has a different pattern than the other two, with only four petals, and location ``3.1'' is a much smaller icon indicating overall lower values.
 
@@
 
-Figure \ref{fig:glyph-map} shows the glyphs drawn for each spatial location of the raw temperature data.
+Figure \ref{fig:glyph-map} (left) shows the star glyphs drawn for each spatial location of the raw temperature data.
+The most seasonality occurs in north American.
+Some seasonal effects can also be seen in the Caribbean and in south American locations.
+Around the equator the temperatures are relatively warm for all seasons.
+Instead of radially displaying the temperature a glyph can be a regular time series, time displayed horizontally and temperature vertically (right plot).
+The structure perceived is different with this change in glyph type: locations at the equator have flat series, on land more varied temperatures.
 
@@
 \end{figure*}
+
+Glyph-maps have been used in
+
+This paper describes how to make glyph-maps.
 
@@
 %Will first explore for regularly gridded data, a simple starting point and a common output of climate models, and then show a few simple extensions to deal with non-rectangular grids and irregular locations, as exemplified by raw climate data which is typically collect at irregular locations.
-\section{Generating a time series at each location}
+\section{Exploring trend}
 

### 2011-08-26 paper.tex
Fleshing out computation. Move to new location.  Add more references

@@
 
-%An interesting attribute of these glyph plots is that the position of graphical elements is determined by major and minor axes.
-For maps, the major axes are latitude and longitude, and the minor axes are typically time on the x-axis and some measurement or model prediction on the y-axis.
-
 %Avoids one problem that plagues glyph displays: the ordering of the variables.
@@
 %Will first explore for regularly gridded data, a simple starting point and a common output of climate models, and then show a few simple extensions to deal with non-rectangular grids and irregular locations, as exemplified by raw climate data which is typically collect at irregular locations.
-\section{Exploring trend}
 
-\begin{figure*}[htp]
-\centerline{\includegraphics[width=4in]{nasa-deseas-trend.pdf}}
-\caption{Time series icon at each location, showing de-seasonalized trend.}
-\end{figure*}
-
-\section{Showing seasonal effect, radially oriented icon}
-
-A big advantage of glyph-maps is that they can be generated with existing graphics software, after a simple pre-processing step.
-This makes them immediately accessible to a wide audience.
-The graphics in this paper were created in R \citep{R} using ggplot2 \citep{me:ggplot2,wickham:2007d} and \citep{me:plyr}, but are readily implemented in any statistical programming environment.
 Code and data is available in the online supplementary materials.
 
-Because the x and y locations are linear combinations of two variables, these types of displays can also be built interactively by software that supports tours, such as DataViewer, XGobi or GGobi.
+\section{Computation}
+
+A big advantage of glyph-maps is that they can be generated with existing graphics software after performing a simple pre-processing step described in this section.
+This makes them immediately accessible to a wide audience.
+The graphics in this paper were created in R \citep{R} using ggplot2 \citep{me:ggplot2,wickham:2007d} and \citep{me:plyr}, but glyph-maps are readily implemented in any statistical programming environment.
+
+The pre-processing step is simple, once you notice that the position of graphical elements is determined by a major and minor axis in each direction (x and y).
+For maps, the major axes are latitude and longitude, and the minor axes are typically time on the x-axis and some measurement or model prediction on the y-axis.
+The final locations are a simple linear combination.
+After rescaling the minor axis to have range $[0, 1]$, the $x$ coordinate is given by $x_{major} + w \cdot x_{minor}$ and $y$ by $y_{major} + h \cdot y_{minor}$, where $w$ and $h$ are width and height respectively
+
+Because the x and y locations are linear combinations of two variables, these types of displays can also be built interactively by software that supports tours \citep{cook:2006}, such as DataViewer \citep{buja:1986}, XGobi \citep{swayne:1991} or GGobi \citep{swayne:2003}.
 This technique is used to good effect in \citet{buja:1996a}.
+
+Computing the coordinates for star glyphs is equally easy.
+Without loss of generality, we assume that $y_{minor}$ becomes $r$, the radius, and $x_{minor}$ becomes $\theta$, the angle, both rescaled to lie between 0 and 1.
+Then the glyph $x$ coordinate is given by $x_{major} + \frac{w \cdot r}{2} \cos(2 \pi \theta)$, and the $y$ coordinate by $y_{major} + \frac{h \cdot r}{2} \sin(2 \pi \theta)$.
+
+Figure~\ref{fig:cycle} shows monthly temperature effects as time (Cartesian coordinates) and star (polar coordinates) glyphs.
+In this case, the star glyphs are not as effective as the linear glyphs: perhaps because of the absence of cyclical pattern.
+
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=2in]{month-cartesian}
+  \includegraphics[width=2in]{month-polar}
+
+  \caption{Monthly averages in (left) Cartesian coordinates and (right) polar coordinates.}
+
+  \label{fig:cycle}
+\end{figure}
 
@@
 A particularly useful summary when to add when displaying model predictions, is some measure of model fit. This makes it easier to avoid over-interpreting predictions from poorly-fitting models.
-
-\section{Radial icons/star glyphs}
-
-These time series glyphs can be readily altered to support circular variables, by projecting each line into polar coordinates and joining the ends.
-This type of display is best suited for variables that are truly circular, such as seasonal patterns or angles (such as for wind direction).
-
-Cite star glyphs.
-
-The alteration to the data transformation is quite simple.
-Without loss of generality, we assume that $y_{minor}$ becomes $r$, the radius, and $x_{minor}$ becomes $theta$ the angle, both rescaled to lie between 0 and 1.
-Then the glyph $x$ coordinate is given by $x_{major} + \frac{w \cdot r}{2} \cos(2 \pi \theta)$, and the $y$ coordinate by $y_{major} + \frac{h \cdot r}{2} \sin(2 \pi \theta)$.
-
-\begin{figure}[htbp]
-  \centering
-  \includegraphics[width=2in]{month-cartesian}
-  \includegraphics[width=2in]{month-polar}
-
-  \caption{Monthly averages in (left) Cartesian coordinates and (right) polar coordinates.}
-
-  \label{fig:cycle}
-\end{figure}
 

### 2011-08-26 paper.tex
Fix references

@@
 Color is used to display the deseasonalized temperature measurements, with red being high values and blue being low values.
-The data comes from \cite{NASA-Data-Expo}.
+The data comes from \citet{murrell:2010}.
 The most noticeable feature in this display is the strong red in the equatorial region beginning in mid-1997 and tapering out during 1998.
@@
 It falls into the framework of {\it change blindess} (see e.g.
-\citet{heahley:perception} and often leads to failure without readers realizing, in particular for small areas of change or subtle changes.
+\citet{healey:2011} and often leads to failure without readers realizing, in particular for small areas of change or subtle changes.
 Large structure such as the El Ni\~no event are clear but readers will not be able to effectively glue the images together sufficiently to read off the trend over the years or notice small local deviations between plots.
@@
 In this paper we explore the use of a different type of display, that uses icons on the map, for climate data.
-Icon plots date back to the star glyphs of  \citep{kleiner:1981}, and were made infamous by \citep{Chernoff}.
+Icon plots date back to the star glyphs of  \citep{kleiner:1981}, and were made infamous by \citep{chernoff:1973}.
 An icon is produced by mapping each variable to some feature of the icon.

### 2011-08-26 paper.tex
conflict?
Add Anderson reference
including Galton's weather map

@@
 From a cognitive perspective, this is a very hard task to do.
-It falls into the framework of {\it change blindess} (see e.g.
+It falls into the framework of {\it change blindness} (see e.g.
 \citet{healey:2011} and often leads to failure without readers realizing, in particular for small areas of change or subtle changes.
-Large structure such as the El Ni\~no event are clear but readers will not be able to effectively glue the images together sufficiently to read off the trend over the years or notice small local deviations between plots.
+Large structures such as the El Ni\~no event are clear but readers will not be able to effectively glue the images together sufficiently to read off the trend over the years or notice small local deviations between plots.
 
@@
 In this paper we explore the use of a different type of display, that uses icons on the map, for climate data.
-Icon plots date back to the star glyphs of  \citep{kleiner:1981}, and were made infamous by \citep{chernoff:1973}.
+For weather maps, icon plots go back as far as \citet{galton:weathermap}.
+Modern icon plots date back to the semi-graphic display of \citet{anderson:1960} and star glyphs of  \citep{kleiner:1981}, and were made infamous by \citep{chernoff:1973}.
 An icon is produced by mapping each variable to some feature of the icon.

### 2011-08-26 paper.tex
included reference to gradual change

@@
 Large structures such as the El Ni\~no event are clear but readers will not be able to effectively glue the images together sufficiently to read off the trend over the years or notice small local deviations between plots.
+Rendering these small multiples as movie helps with picking up strong trends, but small trends still fail to draw readers' attention and escape unnoticed \citep{simons:gradual}.
 

### 2011-08-26 code/0-glyphs.r
polar coordinate in glyphs start at 12 o'clock and go clockwise

@@ -27,9 +27,9 @@ glyphs <- function(data, x_major, x_minor, y_major, y_minor, polar = F, height =
   
   if (polar) {
     data$gx <- data[[x_major]] + width / 2 * rescale01(data[[y_minor]]) * 
-      cos(2 * pi * rescale01(data[[x_minor]]))
-    data$gy <- data[[y_major]] + height / 2 * rescale01(data[[y_minor]]) * 
       sin(2 * pi * rescale01(data[[x_minor]]))
+    data$gy <- data[[y_major]] + height / 2 * rescale01(data[[y_minor]]) * 
+      cos(2 * pi * rescale01(data[[x_minor]]))
     data <- data[order(data[[x_major]], data[[x_minor]]), ]
   } else {
     data$gx <- data[[x_major]] + rescale11(data[[x_minor]]) * width / 2

### 2011-08-26 paper.tex
math for polar coordinates adjusted to code

@@
 \usepackage[round,sectionbib]{natbib}
+\usepackage{amstext}
 \bibliographystyle{plainnat}
@@
 \linespread{1.07}
+
+\newcommand\amin{\text{minor}}
+\newcommand\amaj{\text{major}}
 
@@
 
-The pre-processing step is simple, once you notice that the position of graphical elements is determined by a major and minor axis in each direction (x and y).
-For maps, the major axes are latitude and longitude, and the minor axes are typically time on the x-axis and some measurement or model prediction on the y-axis.
+The pre-processing step is simple, once you notice that the position of graphical elements is determined by a major and minor axis in each direction ($x$ and $y$).
+For maps, the major axes are latitude and longitude, and the minor axes are typically time on the $x$-axis and some measurement or model prediction on the $y$-axis.
 The final locations are a simple linear combination.
-After rescaling the minor axis to have range $[0, 1]$, the $x$ coordinate is given by $x_{major} + w \cdot x_{minor}$ and $y$ by $y_{major} + h \cdot y_{minor}$, where $w$ and $h$ are width and height respectively
+After rescaling the minor axis to have range $[0, 1]$, the $x$ coordinate is given by $x_{\amaj} + w \cdot x_{\amin}$ and $y$ by $y_{\amaj} + h \cdot y_{\amin}$, where $w$ and $h$ are width and height respectively
 
-Because the x and y locations are linear combinations of two variables, these types of displays can also be built interactively by software that supports tours \citep{cook:2006}, such as DataViewer \citep{buja:1986}, XGobi \citep{swayne:1991} or GGobi \citep{swayne:2003}.
+Because the $x$ and $y$ locations are linear combinations of two variables, these types of displays can also be built interactively by software that supports tours \citep{cook:2006}, such as DataViewer \citep{buja:1986}, XGobi \citep{swayne:1991} or GGobi \citep{swayne:2003}.
 This technique is used to good effect in \citet{buja:1996a}.
@@
 Computing the coordinates for star glyphs is equally easy.
-Without loss of generality, we assume that $y_{minor}$ becomes $r$, the radius, and $x_{minor}$ becomes $\theta$, the angle, both rescaled to lie between 0 and 1.
-Then the glyph $x$ coordinate is given by $x_{major} + \frac{w \cdot r}{2} \cos(2 \pi \theta)$, and the $y$ coordinate by $y_{major} + \frac{h \cdot r}{2} \sin(2 \pi \theta)$.
+Without loss of generality, we assume that $y_{\amin}$ becomes $r$, the radius, and $x_{\amin}$ becomes $\theta$, the angle, both rescaled to lie between 0 and 1.
+Then the glyph $x$ coordinate is given by $x_{\amaj} + \frac{w \cdot r}{2} \sin(2 \pi \theta)$, and the $y$ coordinate by $y_{\amaj} + \frac{h \cdot r}{2} \cos(2 \pi \theta)$.
 

### 2011-08-29 paper.tex
references
finished tweaking intro

@@
 
+%% Reproduce all of the icon displays using the fixed glyph code
+
 In this paper we explore the use of a different type of display, that uses icons on the map, for climate data.
-For weather maps, icon plots go back as far as \citet{galton:weathermap}.
-Modern icon plots date back to the semi-graphic display of \citet{anderson:1960} and star glyphs of  \citep{kleiner:1981}, and were made infamous by \citep{chernoff:1973}.
+Icon plot associated with weather maps date back as far as \citet{galton:weathermap}, described in \citet{friendlydenis:2001}.
+Icon plots date back to the star plot introduced by \citet{mayr:1877}, and resurged in the modern era with the semi-graphic displays of \citet{anderson:1960}, and were made infamous by \citep{chernoff:1973}.
 An icon is produced by mapping each variable to some feature of the icon.
 One icon is generated for every case in the data set.
-Figure \ref{fig:glyphs} illustrates the idea of icon plots, with emphasis on temporal data.
+Figure \ref{fig:glyphs} illustrates the construction of star plot, with emphasis on temporal data.
 The time measurements are treated as variables in a multivariate data set.
-Here we have six years of monthly data, giving 72 time values which are considered to be 72 variables.
+Variables are Here we have six years of monthly data, giving 72 time values which are considered to be 72 variables.
+The star plot is formed by considering the 72 axes radiating from a common midpoint, and the data values for the row are plotted on each axis relative to the locations of the minimum and maximum of the variable.
+For the temporal data, all variables are considered to be on a common scale so the global minimum and maximum of temperature are used to scale each data value.
 Three different locations are selected, for demonstration purposes.
@@
 For large data sets it is infeasible to digest the quantity of icons, and for multivariate data there is no natural order so it can be like having a mountain of little stamps to rearrange and organize, in order to understand the structures present in the data.
-The initial paper on star glyphs \citep{kleiner:1981} and a recent paper \citep{hurley:2010} propose some approaches for organizing the icons.
-With spatiotemporal data ordering of icons is naturally done by spatial location.
-There is also usually some spatial trend, or continuity, which will make digesting patterns among the icons simpler.
+An early paper on multivariate graphics \citep{kleiner:1981} and a recent paper \citep{hurley:2010} propose some approaches for organizing the icons.
+Ordering of icons is not an issue with spatiotemporal data, because it is naturally handled by spatial location.
+The spatial trend, or continuity, usually gices the plot an appearance of texturing on a landscape, make it easier to digest patterns among the icons.
 
@@
 
-Glyph-maps have been used in
-
-This paper describes how to make glyph-maps.
+The next section (\ref{sec:construction}) describes how to construct glyph-maps.
+Section \ref{sec:reference} discusses the use of reference lines and boxes, and Section \ref{sec:scale} visits issues of scaling values for different information purposes.
+Generalizations to different types of icons are described in Section \ref{sec:scatter}.
+Glyph-maps work nicely with gridded data, but may be less effective with irregular spatial locations.
+Possible adjustments are discussed in Section \ref{sec:irregular}.
+Code and data is available in the online supplementary materials.
 
@@
 
-Code and data is available in the online supplementary materials.
-
-\section{Computation}
+\section{Computation}~\label{sec:construction}
 
@@
 
-\section{Reference lines}
+\section{Reference lines}~\label{sec:reference}
 
@@
 
-\section{Generalization: scatterplots}
-
-So far we have limited ourselves to always mapping some aspect of time on the x-axis.
-But glyph-maps are not limited to only this type of data and can  be easily generalized to create many other types of display.
-Figure~\ref{fig:cloud} shows a generalisation to scatterplots: by mapping another variable to the x-axis, in this case low-cloud level on x and high-cloud level on y, we can explore how a bivariate relationship varies spatially.
-It is again useful to use a model to extract the essence of the relationship, summarizing the data with models is very useful: the second plot displays a loess curve fit to the data instead of the individual points.
-
-\begin{figure}[htbp]
-  \centering
-  \includegraphics[width=0.5\linewidth]{clouds}%
-  \includegraphics[width=0.5\linewidth]{clouds-smooth}
-
-  \caption{Glyph-map showing (left) scatterplot of high-cloud vs.
-low-cloud and (right) smoothed (loess) curve fit to each location.
-The relationship between these two variables varies considerably over the spatial domain (particularly between land and ocean), but neighbouring locations tend to similar.
-}
-  \label{fig:cloud}
-\end{figure}
-
-\section{Partitioning patterns into shape and size}
+\section{Partitioning patterns into shape and size}~\label{sec:scale}
 
@@
 
-\section{Regular vs irregular locations}
+\section{Generalization: scatterplots}~\label{sec:scatter}
+
+So far we have limited ourselves to always mapping some aspect of time on the x-axis.
+But glyph-maps are not limited to only this type of data and can  be easily generalized to create many other types of display.
+Figure~\ref{fig:cloud} shows a generalisation to scatterplots: by mapping another variable to the x-axis, in this case low-cloud level on x and high-cloud level on y, we can explore how a bivariate relationship varies spatially.
+It is again useful to use a model to extract the essence of the relationship, summarizing the data with models is very useful: the second plot displays a loess curve fit to the data instead of the individual points.
+
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=0.5\linewidth]{clouds}%
+  \includegraphics[width=0.5\linewidth]{clouds-smooth}
+
+  \caption{Glyph-map showing (left) scatterplot of high-cloud vs.
+low-cloud and (right) smoothed (loess) curve fit to each location.
+The relationship between these two variables varies considerably over the spatial domain (particularly between land and ocean), but neighbouring locations tend to similar.
+}
+  \label{fig:cloud}
+\end{figure}
+
+\section{Regular vs irregular locations}~\label{sec:irregular}
 

### 2011-08-29 paper.tex
tweaking reference lines

@@
 
-Space-time-glyphs usually require a \emph{visual reference grid} \citet{cleveland:1993a} to make it easier to compare locations.
+Graphical displays benefit from a \emph{visual reference grid} \citep{cleveland:1993a} to compare values mapped to position along a line.
+The same is true for glyph-maps, but it is not feasible to make a full set of axies for each icon.
+A complete axis would dominate the picture, and draw the eye away from the pattern of the data.
+The types of comparisons that are needed across icons are:
+
+\begin{itemize} \itemsep 0in
+\item Slope, or trend
+\item Intercept, or average value
+\item Variance, or size
+\end{itemize}
+
+Space-time-glyphs usually require a  to make it easier to compare locations.
 This is important for accurate perception of important line properties like slope and minima.
-This need is well illustrated by Figure~\ref{}.
+This need is evident in Figure~\ref{fig:glyph-map}.
+There is a hint that the average value is not the same from right time series to the next in the map (right), or if the center of the icons is the same in the polar time series (left).
+We have no frame of reference on which to judge this.
+The regular grid provides a very rudimentary frame of reference, but it is not sufficient.
+
+Figure~\ref{fig:ref-basic} shows two reference grid variants: a line drawn at the overall mid-range, and a box drawn around each grid cell.
+We prefer using white for these references because it is minimally perceptible: you can see the grid easily when looking for it, but it does not otherwise distract from the perception of the graphic \citep{carr:1994}.
 When we add a reference to each location, as in Figure~\ref{fig:ref-basic}, we see not only a difference in shape between Northern and Southern hemispheres, but also average value.
@@
 
-Figure~\ref{fig:ref-basic} shows two reference grid variants: a line drawn at the overall mid-range, and a box drawn around each grid cell.
-We prefer using white for these references because it is minimally perceptible: you can see the grid easily when looking for it, but it does not otherwise distract from the perception of the graphic \citep{carr:1994}.
+% Think we might need to show the original figures of data with reference lines
 
@@
 
+\section{What to plot: raw data, trend lines, residuals,  seasonal}
+
 \section{Generalization: scatterplots}~\label{sec:scatter}
@@
 
+\section{Long time series}
+
 \section{Conclusions}

### 2011-08-30 paper.tex
explicit formulae for glyph coordinates

@@
 The final locations are a simple linear combination.
-After rescaling the minor axis to have range $[0, 1]$, the $x$ coordinate is given by $x_{\amaj} + w \cdot x_{\amin}$ and $y$ by $y_{\amaj} + h \cdot y_{\amin}$, where $w$ and $h$ are width and height respectively
+After rescaling the minor axis to have range $[0, 1]$, the coordinates are given by
+\begin{eqnarray}\label{coords.eqn}
+x_\text{ coordinate}&=& x_{\amaj} + w \cdot x_{\amin}\\
+y_\text{ coordinate}&=& y_{\amaj} + h \cdot y_{\amin},
+\end{eqnarray)
+where $w$ and $h$ are width and height respectively.
 
-Because the $x$ and $y$ locations are linear combinations of two variables, these types of displays can also be built interactively by software that supports tours \citep{cook:2006}, such as DataViewer \citep{buja:1986}, XGobi \citep{swayne:1991} or GGobi \citep{swayne:2003}.
+Since the $x$ and $y$ locations are linear combinations of two variables, these types of displays can also be built interactively by software that supports tours \citep{cook:2006}, such as DataViewer \citep{buja:1986}, XGobi \citep{swayne:1991} or GGobi \citep{swayne:2003}.
 This technique is used to good effect in \citet{buja:1996a}.
@@
 Without loss of generality, we assume that $y_{\amin}$ becomes $r$, the radius, and $x_{\amin}$ becomes $\theta$, the angle, both rescaled to lie between 0 and 1.
-Then the glyph $x$ coordinate is given by $x_{\amaj} + \frac{w \cdot r}{2} \sin(2 \pi \theta)$, and the $y$ coordinate by $y_{\amaj} + \frac{h \cdot r}{2} \cos(2 \pi \theta)$.
+Then the glyph coordinates are given by
+\begin{eqnarray}\label{coords.polar.eqn}
+x_\text{ coordinate}&=& x_{\amaj} + \frac{w \cdot r}{2} \sin(2 \pi \theta)\\
+y_\text{ coordinate}&=& y_{\amaj} + \frac{h \cdot r}{2} \cos(2 \pi \theta).
+\end{eqnarray)
 

### 2011-08-30 paper.tex
added more references, tweaked computation

@@
 The classic display for spatial data is to use color or grey scale overlaid on geographic location, typically including a map.
-When measurements are made a different time periods also a common way to display the data is in small multiples, such as in Figure \ref{fig:facetted-map}.
+When measurements are made at different time periods it is common to display the data as small multiples, such as in Figure \ref{fig:facetted-map}.
 In this plot, a separate map is drawn for each month (columns) and year (rows), over a 6 year period of remotely sensed temperature data above Central America.
-Color is used to display the deseasonalized temperature measurements, with red being high values and blue being low values.
+Color is used to display the de-seasonalized temperature measurements, with red being high values and blue being low values.
 The data comes from \citet{murrell:2010}.
@@
 \centerline{\includegraphics[width=6in]{nasa-colored-map.png}}
-\caption{Facetted map of deseasonalized temperature.}
+\caption{Facetted map of de-seasonalized temperature.}
 \label{fig:facetted-map}
@@
 \centerline{\includegraphics[width=6in]{nasa-glyph.pdf}}
-\caption{Glyphs of temperature for three different locations.}
+\caption{Icons of raw temperature for three different locations.}
 \label{fig:glyphs}
@@
 
-Figure \ref{fig:glyph-map} (left) shows the star glyphs drawn for each spatial location of the raw temperature data.
+Figure \ref{fig:glyph-map} (left) shows the star glyphs drawn for each spatial location of the raw temperature data, we call it a \emph{glyph-map}.
 The most seasonality occurs in north American.
@@
 The structure perceived is different with this change in glyph type: locations at the equator have flat series, on land more varied temperatures.
+\citet{pickett:1988} introduced the idea of using icons on maps to display multivariate spatial data.
+This area of work has morphed into the field of metaphorical data displays, which creates abstract landscapes of spatial data, a digression from glyph-maps.
+\citet{gribov:2006} describes the use ogf glyph-maps for multivariate data also, with emphasis on the graphics software Gauguin.
 
@@
 
+In this paper we describe the algorithm for creating glyphs of temporal data on maps, generally in R, and the various modifcations that can enable different features in the data to be detected.
 The next section (\ref{sec:construction}) describes how to construct glyph-maps.
@@
 After rescaling the minor axis to have range $[0, 1]$, the coordinates are given by
-\begin{eqnarray}\label{coords.eqn}
+
+\begin{eqnarray*}\label{coords.eqn}
 x_\text{ coordinate}&=& x_{\amaj} + w \cdot x_{\amin}\\
-y_\text{ coordinate}&=& y_{\amaj} + h \cdot y_{\amin},
-\end{eqnarray)
+y_\text{ coordinate}&=& y_{\amaj} + h \cdot y_{\amin}
+\end{eqnarray*}
 where $w$ and $h$ are width and height respectively.
+The width and height will typically take into account the resolution of the major and minor axes, that is the minimum difference between consecutive locations.
 
@@
 
-Computing the coordinates for star glyphs is equally easy.
-Without loss of generality, we assume that $y_{\amin}$ becomes $r$, the radius, and $x_{\amin}$ becomes $\theta$, the angle, both rescaled to lie between 0 and 1.
+Computing the coordinates for star glyphs is achieved by using polar coordinates of $y_{\amin}$, and $x_{\amin}$,  $r$, the radius scaled to be between 0 and 1, and $\theta$, the angle.
 Then the glyph coordinates are given by
-\begin{eqnarray}\label{coords.polar.eqn}
+
+\begin{eqnarray*}\label{coords.polar.eqn}
 x_\text{ coordinate}&=& x_{\amaj} + \frac{w \cdot r}{2} \sin(2 \pi \theta)\\
-y_\text{ coordinate}&=& y_{\amaj} + \frac{h \cdot r}{2} \cos(2 \pi \theta).
-\end{eqnarray)
+y_\text{ coordinate}&=& y_{\amaj} + \frac{h \cdot r}{2} \cos(2 \pi \theta)
+\end{eqnarray*}
 
@@
 \end{figure}
+
+% Show the glyph-map for de-seasonalized data, and describe what we
+% see differently from the facetted map
 
@@
 
+\section{Perception of structure}~\label{sec:scale}
+
+Compare the facetted map Fig 1 with the glyph-map of de-seasonalized data
+
 \section{Partitioning patterns into shape and size}~\label{sec:scale}
+
+*** This section might benefit from some equations also.
 

### 2011-08-31 paper.tex
computation section

@@
 
-A big advantage of glyph-maps is that they can be generated with existing graphics software after performing a simple pre-processing step described in this section.
-This makes them immediately accessible to a wide audience.
-The graphics in this paper were created in R \citep{R} using ggplot2 \citep{me:ggplot2,wickham:2007d} and \citep{me:plyr}, but glyph-maps are readily implemented in any statistical programming environment.
-
+A big advantage of glyph-maps is that they can be generated with existing graphics software after performing a simple pre-processing step, making them easily accessible to a wide audience.
 The pre-processing step is simple, once you notice that the position of graphical elements is determined by a major and minor axis in each direction ($x$ and $y$).
-For maps, the major axes are latitude and longitude, and the minor axes are typically time on the $x$-axis and some measurement or model prediction on the $y$-axis.
-The final locations are a simple linear combination.
-After rescaling the minor axis to have range $[0, 1]$, the coordinates are given by
+For maps, the major axes are latitude and longitude, and the minor axes are typically time and some measurement or model prediction.
+In addition, the minor axes are rescaled to have range $[0, 1]$.
+The final data values are a simple linear combination, as given by
 
@@
 
-Since the $x$ and $y$ locations are linear combinations of two variables, these types of displays can also be built interactively by software that supports tours \citep{cook:2006}, such as DataViewer \citep{buja:1986}, XGobi \citep{swayne:1991} or GGobi \citep{swayne:2003}.
-This technique is used to good effect in \citet{buja:1996a}.
-
-Computing the coordinates for star glyphs is achieved by using polar coordinates of $y_{\amin}$, and $x_{\amin}$,  $r$, the radius scaled to be between 0 and 1, and $\theta$, the angle.
-Then the glyph coordinates are given by
+Computing the coordinates for star glyphs is achieved by using polar coordinates,  $r=y_{\amin}$, and $\theta=x_{\amin}$, both scaled to be between 0 and 1. Then the glyph coordinates are given by
 
@@
 
-Figure~\ref{fig:cycle} shows monthly temperature effects as time (Cartesian coordinates) and star (polar coordinates) glyphs.
+Figure~\ref{fig:cycle} shows average monthly temperature as time (Cartesian coordinates) and star (polar coordinates) glyphs, for a subset of the spatial region.
 In this case, the star glyphs are not as effective as the linear glyphs: perhaps because of the absence of cyclical pattern.
@@
   \centering
-  \includegraphics[width=2in]{month-cartesian}
-  \includegraphics[width=2in]{month-polar}
+  \includegraphics[width=3in]{month-cartesian}
+  \includegraphics[width=3in]{month-polar}
 
@@
 
+Since the $x$ and $y$ locations are linear combinations of two variables, these types of displays can also be built interactively by software that supports tours \citep{cook:2006}, such as DataViewer \citep{buja:1986}, XGobi \citep{swayne:1991} or GGobi \citep{swayne:2003}.
+This technique is used to good effect in \citet{buja:1996a}.
+
 % Show the glyph-map for de-seasonalized data, and describe what we
 % see differently from the facetted map
+
+The graphics in this paper were created in R \citep{R} using {\tt ggplot2} \citep{me:ggplot2} and {\tt plyr} \citep{me:plyr}.
 

### 2011-08-31 code/0-glyphs.r
changed 10% margin to just 5% on each side, introduced an exponent for rescaling the radius

@@ -4,11 +4,11 @@ rescale01 <- function(x) {
 }
 rescale11 <- function(x) 2 * rescale01(x) - 1
   
-glyphs <- function(data, x_major, x_minor, y_major, y_minor, polar = F, height = NULL, width = NULL) {
+glyphs <- function(data, x_major, x_minor, y_major, y_minor, polar = F, height = NULL, width = NULL, x=2) {
   data$gid <- interaction(data[[x_major]], data[[y_major]], drop = TRUE)
   
   if (is.null(width)) {
-    width <- resolution(data[[x_major]]) * 0.9    
+    width <- resolution(data[[x_major]]) * 0.95    
     message("Using width ", format(width, digits = 3))
   }
   else {
@@ -17,7 +17,7 @@ glyphs <- function(data, x_major, x_minor, y_major, y_minor, polar = F, height =
   }
     
   if (is.null(height)) {
-    height <- resolution(data[[y_major]]) * 0.9    
+    height <- resolution(data[[y_major]]) * 0.95    
     message("Using height ", format(height, digits = 3))
   }
   else {
@@ -26,8 +26,8 @@ glyphs <- function(data, x_major, x_minor, y_major, y_minor, polar = F, height =
   }
   
   if (polar) {
-    theta <- 2 * pi * rescale01(data[[x_minor]]
-    r <- rescale01(data[[y_minor]])
+    theta <- 2 * pi * rescale01(data[[x_minor]])
+    r <- rescale01(data[[y_minor]])^x
     
     data$gx <- data[[x_major]] + width  / 2 * r * sin(theta) 
     data$gy <- data[[y_major]] + height / 2 * r * cos(theta)

### 2011-08-31 paper.tex
massaged math

@@
 
-\begin{eqnarray*}\label{coords.eqn}
+\begin{equation}
+\begin{array}{lll}
 x_\text{ coordinate}&=& x_{\amaj} + w \cdot x_{\amin}\\
-y_\text{ coordinate}&=& y_{\amaj} + h \cdot y_{\amin}
-\end{eqnarray*}
+y_\text{ coordinate}&=& y_{\amaj} + h \cdot y_{\amin}, 
+\end{array}\label{coords.eqn}
+\end{equation}
 where $w$ and $h$ are width and height respectively.
@@
 Computing the coordinates for star glyphs is achieved by using polar coordinates,  $r=y_{\amin}$, and $\theta=x_{\amin}$, both scaled to be between 0 and 1. Then the glyph coordinates are given by 
-
-\begin{eqnarray*}\label{coords.polar.eqn}
-x_\text{ coordinate}&=& x_{\amaj} + \frac{w \cdot r}{2} \sin(2 \pi \theta)\\
-y_\text{ coordinate}&=& y_{\amaj} + \frac{h \cdot r}{2} \cos(2 \pi \theta)
-\end{eqnarray*}
+\begin{equation}
+\begin{array}{lll}
+x_\text{ coordinate}&=& x_{\amaj} + \frac{w \cdot r}{2} \sin(2 \pi \theta) \\
+y_\text{ coordinate}&=& y_{\amaj} + \frac{h \cdot r}{2} \cos(2 \pi \theta).
+\end{array}\label{coords.polar.eqn}
+\end{equation}
+Note that this is is a non-standard conversion to polar coordinates, but this conversion provides us with a time line that starts at 12 o'clock and wraps around the origin in a clockwise fashion.
 

### 2011-09-01 code/2-examples.R
small examples showing polar vs cartesian examples

@@ -4,11 +4,10 @@ mu <- 330
 t <- seq(0, pi, length=12)
 x1 <- mu+amp*sin(t)
 x2 <- mu+30-amp*sin(t)
-x3 <- rep(mu+15, length(t))
+x3 <- seq(330,360, length=12)
 
 locs <- data.frame(cbind(rep(t,3), c(x1,x2,x3), rep(1:3, each=12), rep(1,3*12)))
 names(locs) <- c("time","temp","x_max", "y_max")
-qplot(time,temp, data=locs, colour=x_max)
 
 locs2 <- data.frame(cbind(rep(t+pi,3), rep(c(mean(x1),mean(x2),mean(x3)), each=12), rep(1:3, each=12), rep(1,3*12)))
 names(locs2) <- c("time","temp","x_max", "y_max")
@@ -19,16 +18,28 @@ locs <- glyphs(locs, "x_max", "time", "y_max", "temp")
 locs2 <- glyphs(locs2, "x_max", "time", "y_max", "temp", ylim=range(locs$temp)) 
 ggplot(locs, aes(gx, gy, group = gid)) + 
   geom_path() +
-  geom_path(aes(gx, gy, group = gid), data=locs2, colour="red") +
+  geom_path(aes(gx, gy, group = gid), data=locs2, colour="grey80") +
   geom_tile(aes(x_max, y_max), colour = "white", fill = NA)
-
+ggsave("../images/euclid-to-polar-1.pdf")
 
 
 # now in polar coordinates:
 locs <- glyphs(locs, "x_max", "time", "y_max", "temp", polar = T) 
 locs2 <- glyphs(locs2, "x_max", "time", "y_max", "temp", ylim=range(locs$temp), polar = T) 
 ggplot(locs, aes(gx, gy, group = gid)) + 
+  geom_path(aes(gx, gy, group = gid), data=locs2, colour="grey80") +
   geom_path() +
-  geom_path(aes(gx, gy, group = gid), data=locs2, colour="red") +
   geom_tile(aes(x_max, y_max), colour = "white", fill = NA)
+ggsave("../images/euclid-to-polar-2.pdf")
 
+
+
+
+# overall average temp as reference circle
+locs2 <- data.frame(cbind(rep(t+pi,3), rep(c(mean(locs$temp)), each=3*12), rep(1:3, each=12), rep(1,3*12)))
+names(locs2) <- c("time","temp","x_max", "y_max")
[…]

### 2011-09-01 paper.tex
polar coordinates: more detailed explanation & example

@@
 
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=0.5\linewidth]{euclid-to-polar-1}%
+  \includegraphics[width=0.5\linewidth]{euclid-to-polar-2}
+
+  \caption{Examples for some small time series with strongly expressed trends in Euclidean coordinates (left) and polar coordinates (right).}
+  \label{fig:ref-basic-polar}
+\end{figure}
+
+Figure \ref{fig:ref-basic-polar} gives some small examples of time series with strongly expressed trends, shown both in cartesian coordinates and polar coordinates.
+What might be surprising is that time series that seem to have opposing trends in cartesian coordinates don't necessarily show this symmetry  in polar coordinates.
+What becomes apparent, though is that the area of glyphs in polar coordinates mainly reflect the average value of a time series, while its shape shows deviations from average values.
+Adding a circular reference line helps to pick out these deviations.
+
 Since the $x$ and $y$ locations are linear combinations of two variables, these types of displays can also be built interactively by software that supports tours \citep{cook:2006}, such as DataViewer \citep{buja:1986}, XGobi \citep{swayne:1991} or GGobi \citep{swayne:2003}.

### 2011-09-05 paper.tex
working on scaling section

@@
 Graphical displays benefit from a \emph{visual reference grid} \citep{cleveland:1993a} to compare values mapped to position along a line.
-The same is true for glyph-maps, but it is not feasible to make a full set of axies for each icon.
+The same is true for glyph-maps, but it is not feasible to make a full set of axes for each icon.
 A complete axis would dominate the picture, and draw the eye away from the pattern of the data.
@@
 
-\section{Partitioning patterns into shape and size}~\label{sec:scale}
+Glyph-maps using trend lines enable the reading of trends across
+spatial locations directly from the plot. In comparison, to read trend
+from facetted colored maps one needs to spot the difference between
+the different maps. A more direct comparison would be to make a single
+map with the slope of the trend coded as color. Cleveland's
+\citep{cleveland:1993a} perceptual tests suggests that the reader can
+perceive the positional difference of the lines better than the color
+differences. Accurate perception of color differences is difficult.
+
+\section{Effect of scale}~\label{sec:scale}
 
 *** This section might benefit from some equations also.
+
+*** Global/local min/max, remove intercept/average, data vs model scale
+
+Scaling of the data can produce dramatically different emphases on
+structure. For spatiotemporal data the most natural scale is to use
+the global minumum and maximum. The resulting icons will have size
+relative to the magnitude of the full range of the data. Locations
+with overall large or small temperature will be the focus. Figure
+\ref{fig:glyph-map} shows data scaled to -1 to 1 based on the global
+minimum and maximum, using:
+
+\begin{equation}
+y_{\amaj} = \frac{y_{\amaj}-y_{min}}{y_{max}-y_{min}} \times 2 - 1
+\end{equation}
+
+\noindent where $y_{min} = \mbox{minimum}\{y_{st}; s=1, \dots S, t=1, ..., T \}, y_{max} = \mbox{maximum}\{y_{st}; s=1, \dots S, t=1, ..., T\}$, $S=$ is the number of spatial locations and $T=$ is the number of time points.
+*** Reference other plots that use global scale too.
+
+Alternatively, scaling locally at each spatial location enables
+exploring the seasonal scale, or shape differences, from one location
+to another. Fig \ref{***} uses local scaling using:
+
+\begin{equation}
+y_{\amaj} = \frac{y_{\amaj}-y_{s,min}}{y_{s,max}-y_{s,min}} \times 2 - 1
+\end{equation}
+
+\noindent where $y_{s, min} = \mbox{minimum}\{y_{st}; t=1, ..., T \}, y_{s,max} = \mbox{maximum}\{y_{st}; t=1, ..., T\}$.
 
@@
 
+*** Connection to small multiples
+
+*** Scale of icons relative to data or model range
+
 \section{Regular vs irregular locations}~\label{sec:irregular}

### 2011-09-06 paper.tex
filling in sections

@@
 
-\section{Reference lines}~\label{sec:reference}
+\section{Reference frames}~\label{sec:reference}
 
@@
 \begin{equation}
-y_{\amaj} = \frac{y_{\amaj}-y_{min}}{y_{max}-y_{min}} \times 2 - 1
+y_{st} = \frac{y_{st}-y_{min}}{y_{max}-y_{min}} \times 2 - 1
+\label{scale1}
 \end{equation}
 
-\noindent where $y_{min} = \mbox{minimum}\{y_{st}; s=1, \dots S, t=1, ..., T \}, y_{max} = \mbox{maximum}\{y_{st}; s=1, \dots S, t=1, ..., T\}$, $S=$ is the number of spatial locations and $T=$ is the number of time points.
+\noindent where $y$ is the variable used on the minor axis in equations \ref{coords.eqn}, \ref{coords.polar.eqn}, $y_{min} = \mbox{minimum}\{y_{st}; s=1, \dots S, t=1, ..., T \}, y_{max} = \mbox{maximum}\{y_{st}; s=1, \dots S, t=1, ..., T\}$, $S=$ is the number of spatial locations and $T=$ is the number of time points.
 *** Reference other plots that use global scale too.
@@
 \begin{equation}
-y_{\amaj} = \frac{y_{\amaj}-y_{s,min}}{y_{s,max}-y_{s,min}} \times 2 - 1
+y_{st} = \frac{y_{st}-y_{s,min}}{y_{s,max}-y_{s,min}} \times 2 - 1
+\label{scale2}
 \end{equation}
@@
 \noindent where $y_{s, min} = \mbox{minimum}\{y_{st}; t=1, ..., T \}, y_{s,max} = \mbox{maximum}\{y_{st}; t=1, ..., T\}$.
+With the local scaling, the shape of the icon, for example, differences in the seasonal patterns over spatial locations, are emphasized.
+Figure \ref{***} shows data that has been locally scaled. We can see *** The local scale is equivalent to using individual scales in the facetted graphics of \texttt{ggplot2}.
 
-Another impact of varying averages is that it tends to diminish the perceptibility of shape: if some series have uniformly high values and some have uniformly low, then it is difficult to see if the shape is the same independent of the average.
-Partitioning is a extremely useful technique: product plots, ANOVA, ...
-For glyph plots, it's often useful to partition the glyphs into shape and scale components, decomposing the values into a product of maximum and scaled values.
-In this section we explore how decomposing each series into scale and shape components can be useful.
+It can also be advantageous to relocate the data uniformly, by extracting the average at each location.
+Particularly for polar glyphs the average value can lead to some unintuitive patterns in the glyphs.
+In this situation we would use $y_{st}-\bar{y}_s$ instead of $y_{st}$ in equations \ref{scale1}, \ref{scale2}, with minima and maxima calculated on the mean-shifted data.
 
-This is revealing for smooth models of surface temperature, shown in Figure~\ref{fig:scaling}.
-Each location is modelled using a generalised additive model \citep{wood:2006} with a seasonal term and a long-term smooth term.
-The three plots in the figure show the long-term smooth term.
-Because some locations are uniformly warmer than others, the impact of changes across time are diminished.
-In the centre plot, each location is scaled by dividing out the maximum, and on the right, each location is scaled to range $[0, 1]$.
-These emphasise individual changes but lose the ability to make global comparisons, and need caution when interpreting - a small absolute change gets blown up into a large relative change.
+Another way to think about these scale options is partitioning the data structure into shape, magnitude and shift components.
+The different ways to scale emphasize these different components.
+The shift component can be extracted from the data in different ways, using the intercept from a linear model, for example, instead of the average.
+
+**** This is different - I thnk it just confuses the topic of scaling.
+Because the GAM is not used to rescale the data, you still have to do that using the local or global scling equations above.
+Moving the discussion to ``What to plot''.
 
@@
 
-Of course other types of scaling are possible. Another classical scaling method is to standardize mean and standard deviation to 0 and 1 (or robust variants).
+Other types of scaling may be interesting to explore, such as a classical scaling using the mean and standard deviation.
 
@@
 
-\section{Generalization: scatterplots}~\label{sec:scatter}
+Structure in temporal data decomposed into different parts, global trend, seasonality and error.
+Extracting each of these for visualization improves the focus on different aspects of the time series.
+A generalised additive model \citep{wood:2006} with a seasonal term and a long-term smooth term.
+The three plots in the figure show the long-term smooth term.
+Because some locations are uniformly warmer than others, the impact of changes across time are diminished.
+
+ is a extremely useful technique: product plots, ANOVA, ...
+For glyph plots, it's often useful to partition the glyphs into shape and scale components, decomposing the values into a product of maximum and scaled values.
+In this section we explore how decomposing each series into scale and shape components can be useful.
+
+This is revealing for smooth models of surface temperature, shown in Figure~\ref{fig:scaling}.
+Each location is modelled using a generalised additive model \citep{wood:2006} with a seasonal term and a long-term smooth term.
+The three plots in the figure show the long-term smooth term.
+Because some locations are uniformly warmer than others, the impact of changes across time are diminished.
+In the centre plot, each location is scaled by dividing out the maximum, and on the right, each location is scaled to range $[0, 1]$.
+These emphasise individual changes but lose the ability to make global comparisons, and need caution when interpreting - a small absolute change gets blown up into a large relative change.
+
+Issues of scale relative to modeling.
+Switching from the data to model summaries bringsnew scaling issues.
+Often the scale of the raw data is ignored when the model is plotted, one usually shows just the line showing the trend, in the range of the trend.
+This has the effect of exaggerating the signficance of the model.
+Scaling the trend lines rlative to the range of the data may be useful sometimes.
+
+\section{Generalization}~\label{sec:scatter}
 
@@
 
-*** Connection to small multiples
+In the Gauguin software one alternative was an icon based on a histogram.
 
-*** Scale of icons relative to data or model range
+*** Connection to small multiples, facetted plots
 
 \section{Regular vs irregular locations}~\label{sec:irregular}
+
+Regularly gridded spatial data helps with the perception of structure, because it helps to provide the reference frames upon which to make comparisons among the icons.
+When spatial locations are not on a regular grid, some difficulties arise.
+In this situation it is not clear what the scale of icons should be, there is no regular spacing.
+Some locations may be close to each other which would generate icons that overlap.
 

### 2011-09-06 code/1-modeling.r
not sure what i changed

@@ -123,7 +123,7 @@ ggplot(month_preds, aes(gx, gy, group = gid)) +
   geom_tile(aes(long, lat), colour = "white", fill = NA)
 ggsave("../images/month-cartesian.pdf", width = 4, height = 4)
 
-month_preds <- glyphs(month_preds, "long", "month", "lat", "pred", polar = T) 
+month_preds <- glyphs(month_preds, "long", "month", "lat", "pred", polar = T, height=2, width=1) 
 ggplot(month_preds, aes(gx, gy, group = gid)) + 
   map_mini + 
   geom_path() +

### 2011-09-15 gridded.temp.r
data example

@@ -1,0 +1,46 @@
+fo<-open.ncdf("air.mon.anom.nc")
+fo$nvars
+lon <- get.var.ncdf(fo, "lon")
+lat <- get.var.ncdf(fo, "lat")
+time <- get.var.ncdf(fo, "time")
+temp <- get.var.ncdf(fo, "air")
+
+temp.usa<-temp[10:24, 8:14,] # lon -130, 65; lat 25, 50
+temp.usa.melt<-melt(temp.usa)
+colnames(temp.usa.melt)<-c("gridx", "gridy", "time", "temp")
+temp.usa.melt$lon<-lon[temp.usa.melt$gridx+9]
+temp.usa.melt$lat<-lat[temp.usa.melt$gridy+7]
+
+temp.usa.melt.gly<-glyphs(temp.usa.melt, "lon", "time", "lat", "temp")
+qplot(gx, gy, data=temp.usa.melt.gly, group=gid, geom="line")
+
+library(maps)
+
+outlines <- as.data.frame(map("state",
+                              plot=FALSE)[c("x","y")])
+map <- c(
+  geom_path(aes(x, y), inherit.aes = FALSE, 
+    data = outlines, colour = alpha("grey60", 0.4))
+)
+
+qplot(gx, gy, data=temp.usa.melt.gly, group=gid, geom="line") + map
+
+fo<-open.ncdf("air.2x2.250.mon.anom.land.nc")
+fo$nvars
+lon <- get.var.ncdf(fo, "lon")
+lat <- get.var.ncdf(fo, "lat")
+time <- get.var.ncdf(fo, "time")
+temp <- get.var.ncdf(fo, "air")
+
+length(lon)
+dim(temp)
+
+temp.usa<-temp[123:155, 20:33,] # lon -130, 65; lat 25, 50
+temp.usa.melt<-melt(temp.usa)
+colnames(temp.usa.melt)<-c("gridx", "gridy", "time", "temp")
+temp.usa.melt$lon<--lon[temp.usa.melt$gridx+122]+180
+temp.usa.melt$lat<-lat[temp.usa.melt$gridy+19]
+
+temp.usa.melt.gly<-glyphs(temp.usa.melt, "lon", "time", "lat", "temp")
+qplot(gx, gy, data=temp.usa.melt.gly, group=gid, geom="line") + map
+

### 2011-09-19 paper.tex
data example
putting better climate data examples in

@@
 \author{Cook, Hofmann, Wickham, Wickham}
+
+% Found data: http://www.esrl.noaa.gov/psd/data/gridded/data.HADCRUT2.html
+%     air.mon.anom.nc, var adj, this is 5x5 degree grid, a bit sparse
+% Or: http://www.esrl.noaa.gov/psd/data/gridded/data.gisstemp.html
+% air.2x2.mon.anom.land.nc 250k smooth data, is on a 2x2 grid
 
@@
 
+*** These figures need to be placed yet, but are here as a start on better climate data examples:
+
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=1\linewidth]{gistemp-raw}%
+
+  \includegraphics[width=1\linewidth]{gistemp-polar-raw}
+
+  \caption{Glyph-maps of gridded raw temperature anomaly data}
+  \label{fig:ref-gistemp}
+\end{figure}
+
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=1\linewidth]{gistemp-pred}%
+
+  \includegraphics[width=1\linewidth]{gistemp-polar-pred}
+
+  \caption{Glyph-maps of gridded predicted temperature anomaly data}
+  \label{fig:ref-gistemp}
+\end{figure}
+
 \section{Reference frames}~\label{sec:reference}
@@
 
-**** This is different - I thnk it just confuses the topic of scaling.
+**** This is different - I think it just confuses the topic of scaling.
 Because the GAM is not used to rescale the data, you still have to do that using the local or global scling equations above.
@@
 
+\section*{Acknowledgements}
+
+GISTEMP data provided by the NOAA/OAR/ESRL PSD, Boulder, Colorado,
+USA, from their Web site at \url{http://www.esrl.noaa.gov/psd/}.
+
 \bibliography{references}

### 2011-09-19 code/0-glyphs.r

@@ -7,7 +7,12 @@ rescale01 <- function(x, xlim=NULL) {
    (x - rng[1]) / (rng[2] - rng[1])
 }
 rescale11 <- function(x, xlim=NULL) 2 * rescale01(x, xlim) - 1
-  
+
+# Wish list
+# remove intercept/average, default
+# scale individually, globally
+# Absolute ylims, for use when plotting model predictions on the data
+#    scale - got it with the ylim argument
 glyphs <- function(data, x_major, x_minor, y_major, y_minor, polar = F, height = NULL, width = NULL, ylim=NULL) {
   data$gid <- interaction(data[[x_major]], data[[y_major]], drop = TRUE)
   
@@ -21,7 +26,7 @@ glyphs <- function(data, x_major, x_minor, y_major, y_minor, polar = F, height =
   }
     
   if (is.null(height)) {
-    height <- resolution(data[[y_major]]) * 0.95    
+    height <- resolution(data[[y_major]]) * 0.95     
     message("Using height ", format(height, digits = 3))
   }
   else {
@@ -39,8 +44,12 @@ glyphs <- function(data, x_major, x_minor, y_major, y_minor, polar = F, height =
     data <- data[order(data[[x_major]], data[[x_minor]]), ] 
   } else {
     data$gx <- data[[x_major]] + rescale11(data[[x_minor]]) * width / 2
-    data$gy <- data[[y_major]] + rescale11(data[[y_minor]], ylim) * height / 2
+    data$gy <- data[[y_major]] + rescale11(data[[y_minor]], ylim) * height / 2 
   }
   
   data
 }
+
+# Plot function, wish list
+# Add ref lines, ref boxes, both, ref circle
+# 

### 2011-09-19 code/0-nasa.r

@@ -69,7 +69,7 @@ theme_fullframe <- function (base_size = 12){
 }
 
 ref_boxes <- list(
-  geom_tile(aes(long, lat), colour = "white", fill = NA, inherit.aes = FALSE),
+  geom_tile(aes(lon, lat), colour = "white", fill = NA, inherit.aes = FALSE),
   scale_x_continuous(breaks = NA, expand = c(0, 0)),
   scale_y_continuous(breaks = NA, expand = c(0, 0))
 )

### 2011-09-19 code/1-modeling.r

@@ -6,7 +6,7 @@ source("0-glyphs.r")
 source("0-nasa.r")
 source("0-cache.r")
 
-# Scatteplots ----------------------------------------------------------------
+# Scatterplots ----------------------------------------------------------------
 
 cloud <- glyphs(nasa, "long", "cloudlow", "lat", "cloudhigh") 
 ggplot(cloud, aes(gx, gy, group = gid)) + 

### 2011-09-19 gridded.temp.r

@@ -46,8 +46,63 @@ temp.usa.melt$lon<-lon[temp.usa.melt$gridx+113]
 temp.usa.melt$lat<-lat[temp.usa.melt$gridy+19]
 
 temp.usa.melt.gly<-glyphs(temp.usa.melt, "lon", "time", "lat", "temp")
-qplot(gx, gy, data=temp.usa.melt.gly, group=gid, geom="line") + map + coord_map()
+qplot(gx, gy, data=temp.usa.melt.gly, group=gid, geom="line") + map + coord_map() # Testing
+ggplot(temp.usa.melt.gly, aes(gx, gy, group = gid)) + 
+  map + 
+  geom_path() +
+  ref_boxes +
+  theme_fullframe()
+ggsave("../images/gistemp-raw.png", width = 8, height = 4)
 
+temp.usa.melt.gly<-glyphs(temp.usa.melt, "lon", "time", "lat", "temp", polar=T)
+ggplot(temp.usa.melt.gly, aes(gx, gy, group = gid)) + 
+  map + 
+  geom_path() +
+  ref_boxes +
+  theme_fullframe()
+ggsave("../images/gistemp-polar-raw.png", width = 8, height = 4)
+
+# Predictions
+library(lubridate)
+temp.usa.melt$date <- ymd(18800101) + months(temp.usa.melt$time-1)
+temp.usa.melt$year <- year(temp.usa.melt$date)
+temp.usa.melt$month <- month(temp.usa.melt$date)
+# Just use 1950-2010, cleaner data
+temp.usa.melt.sub<-subset(temp.usa.melt, year>1950 & year<2011)
+temp.usa.melt.sub<-subset(temp.usa.melt.sub, !is.na(temp))
+#temp_models %<-cache% dlply(temp.usa.melt.sub, c("lat", "lon"), function(df) {
+#  lm(temp ~ year + factor(month), data = df)
+#}) This uses old data!!!
+temp_models <- dlply(temp.usa.melt.sub, c("lat", "lon"), function(df) {
+  lm(temp ~ year + factor(month), data = df)
+})
+month_grid <- expand.grid(year = 1950:2010, month = 1:12)
+month_preds <- ldply(temp_models, function(mod) {
+  month_grid$pred <- predict(mod, newdata = month_grid)
+  month_grid
+})
+year_grid <- expand.grid(year = unique(temp.usa.melt.sub$year), month = 1)
[…]

### 2011-09-20 paper.tex
reworking examples

@@
 One icon is generated for every case in the data set.
-Figure \ref{fig:glyphs} illustrates the construction of star plot, with emphasis on temporal data.
-The time measurements are treated as variables in a multivariate data set.
-Variables are Here we have six years of monthly data, giving 72 time values which are considered to be 72 variables.
-The star plot is formed by considering the 72 axes radiating from a common midpoint, and the data values for the row are plotted on each axis relative to the locations of the minimum and maximum of the variable.
+%Figure \ref{fig:glyphs} illustrates the construction of star plot, with emphasis on temporal data.
+For temporal data, time measurements can be treated as variables in a multivariate data set, fitting this into the icon plot framework.
+Figure \ref{templates} displays 12 iconic time series shapes, and how they would look if the icon was a time series, or displayed as a star glyph.
+The data underlying each glyph has 36 time variables.
+The star glyphs is formed by considering the 36 axes radiating from a common midpoint, and the data values for the row are plotted on each axis relative to the locations of the minimum and maximum of the variable.
 For the temporal data, all variables are considered to be on a common scale so the global minimum and maximum of temperature are used to scale each data value.
-Three different locations are selected, for demonstration purposes.
+%Three different locations are selected, for demonstration purposes.
 The top plot shows the time series of these three locations.
@@
 
-\begin{figure*}[htp]
-\centerline{\includegraphics[width=6in]{nasa-glyph-ts.pdf}}
-\centerline{\includegraphics[width=6in]{nasa-glyph.pdf}}
-\caption{Icons of raw temperature for three different locations.}
-\label{fig:glyphs}
-\end{figure*}
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=0.5\linewidth]{euclid-to-polar-1}%
+  \includegraphics[width=0.5\linewidth]{euclid-to-polar-2}
+
+  \caption{Icon plots for 12 iconic time series shapes (linear increasing, decreasing, shifted, single peak, single dip, combined linear and nonlinear, seasonal trends with different scales, and a combined linear and seasonal trend) in Euclidean coordinates, time series icons (left) and polar coordinates, star plots (right).}
+  \label{fig:templates}
+\end{figure}
+
+%\begin{figure*}[htp]
+%\centerline{\includegraphics[width=6in]{nasa-glyph-ts.pdf}}
+%\centerline{\includegraphics[width=6in]{nasa-glyph.pdf}}
+%\caption{Icons of raw temperature for three different locations.}
+%\label{fig:glyphs}
+%\end{figure*}
 
@@
 Ordering of icons is not an issue with spatiotemporal data, because it is naturally handled by spatial location.
-The spatial trend, or continuity, usually gices the plot an appearance of texturing on a landscape, make it easier to digest patterns among the icons.
+The spatial trend, or continuity, usually gives the plot an appearance of texturing on a landscape, make it easier to digest patterns among the icons.
 
-Figure \ref{fig:glyph-map} (left) shows the star glyphs drawn for each spatial location of the raw temperature data, we call it a \emph{glyph-map}.
-The most seasonality occurs in north American.
+% Reference Tufte's small multiples
+Figure \ref{fig:gistemp-pred} (left) shows the time series icons and star glyphs drawn for each spatial location of the predicted temperature data across the USA, we call it a \emph{glyph-map}.
+The data contains 2$^o$x2$^o$ gridded temperature anomalies 1950-2010, from the GISTEMP surface temperature data provided by the Earth System Research Laboratory, Physical Sciences Division, National Oceanic and Atmospheric Administration.
+Most locations observe an increasing temperature, to varying degrees, over this time period.
+A few locations show a decline.
+These locations tend to be different from their neighbors, which might suggest isolated data problems.
+%The most seasonality occurs in north American.
 Some seasonal effects can also be seen in the Caribbean and in south American locations.
@@
 
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=1\linewidth]{gistemp-pred}%
+
+  %\includegraphics[width=1\linewidth]{gistemp-polar-pred}
+
+  \caption{Glyph-maps of gridded predicted temperature anomaly data}
+  \label{fig:gistemp-pred}
+\end{figure}
+
 % Add time series as well as radial
-\begin{figure*}[htp]
-\centerline{\includegraphics[width=3.5in]{nasa-glyph-all.pdf}\includegraphics[width=3.5in]{nasa-glyph-ts-all.pdf}}
-\caption{Glyph-map of raw data, using star glyphs (left), and time series as glyphs (right).}
-\label{fig:glyph-map}
-\end{figure*}
+%\begin{figure*}[htp]
+%\centerline{\includegraphics[width=3.5in]{nasa-glyph-all.pdf}\includegraphics[width=3.5in]{nasa-glyph-ts-all.pdf}}
+%\caption{Glyph-map of raw data, using star glyphs (left), and time series as glyphs (right).}
+%\label{fig:glyph-map}
+%\end{figure*}
 
@@
 
-\begin{figure}[htbp]
-  \centering
-  \includegraphics[width=0.5\linewidth]{euclid-to-polar-1}%
-  \includegraphics[width=0.5\linewidth]{euclid-to-polar-2}
-
-  \caption{Examples for some small time series with strongly expressed trends in Euclidean coordinates (left) and polar coordinates (right).}
-  \label{fig:ref-basic-polar}
-\end{figure}
-
 Figure \ref{fig:ref-basic-polar} gives some small examples of time series with strongly expressed trends, shown both in cartesian coordinates and polar coordinates.
@@
   \caption{Glyph-maps of gridded raw temperature anomaly data}
-  \label{fig:ref-gistemp}
-\end{figure}
-
-\begin{figure}[htbp]
-  \centering
-  \includegraphics[width=1\linewidth]{gistemp-pred}%
-
-  \includegraphics[width=1\linewidth]{gistemp-polar-pred}
-
-  \caption{Glyph-maps of gridded predicted temperature anomaly data}
   \label{fig:ref-gistemp}
@@
 GISTEMP data provided by the NOAA/OAR/ESRL PSD, Boulder, Colorado,
-USA, from their Web site at \url{http://www.esrl.noaa.gov/psd/}.
+USA, from their web site at \url{http://www.esrl.noaa.gov/psd/}.
 

### 2011-09-20 code/glyph-utils.R

@@ -1,0 +1,46 @@
+library("maps")
+library("ggplot2")
+
+getbox <- function (map, xlim, ylim) {
+	# identify all regions involved
+	small <- subset(map, (long > xlim[1]) & (long < xlim[2]) & (lat > ylim[1]) & (lat < ylim[2]))
+	regions <- unique(small$region)
+	small <- subset(map, region %in% regions)	
+	
+	# now shrink all nodes back to the bounding box
+	small$long <- pmax(small$long, xlim[1])
+	small$long <- pmin(small$long, xlim[2])
+	small$lat <- pmax(small$lat, ylim[1])
+	small$lat <- pmin(small$lat, ylim[2])
+
+	return(small)
+}
+
+theme_fullframe <- function (base_size = 12){
+  structure(list(
+    axis.line = theme_blank(), 
+    axis.text.x = theme_blank(), 
+    axis.text.y = theme_blank(),
+    axis.ticks = theme_blank(), 
+    axis.title.x = theme_blank(), 
+    axis.title.y = theme_blank(), 
+    axis.ticks.length = unit(0, "lines"), 
+    axis.ticks.margin = unit(0, "lines"), 
+    legend.position = "none", 
+    panel.border = theme_blank(), 
+    panel.grid.major = theme_blank(), 
+    panel.grid.minor = theme_blank(), 
+    panel.margin = unit(0, "lines"), 
+    plot.background = theme_blank(), 
+    plot.margin = unit(rep(2, 4), "mm")
+  ), class = "options")
+}
+
+ref_boxes <- list(
+  geom_tile(aes(lon, lat), colour = "white", fill = NA, inherit.aes = FALSE),
+  scale_x_continuous(breaks = NA, expand = c(0, 0)),
+  scale_y_continuous(breaks = NA, expand = c(0, 0))
+)
+
+# Need to have a function for the circles references for polar coords
+# Need to have a function for the reference lines

### 2011-09-20 code/glyphs.R

@@ -1,0 +1,53 @@
+# Scale: 0-1
+rescale01 <- function(x, xlim=NULL) {
+  if (is.null(xlim)) {
+	  rng <- range(x, na.rm = TRUE)
+   } else {
+   	  rng <- xlim
+   }
+   (x - rng[1]) / (rng[2] - rng[1])
+}
+# Scale: -1 to 1
+rescale11 <- function(x, xlim=NULL) 2 * rescale01(x, xlim) - 1
+
+# Wish list
+# remove intercept/average, default
+# scale individually, globally
+# Absolute ylims, for use when plotting model predictions on the data
+#    scale - got it with the ylim argument
+glyphs <- function(data, x_major, x_minor, y_major, y_minor, polar = F, height = NULL, width = NULL, ylim=NULL) {
+  data$gid <- interaction(data[[x_major]], data[[y_major]], drop = TRUE)
+  
+  if (is.null(width)) {
+    width <- resolution(data[[x_major]]) * 0.95    
+    message("Using width ", format(width, digits = 3))
+  }
+  else {
+    width <- resolution(data[[x_major]]) * width    
+    message("Using width ", format(width, digits = 3))
+  }
+    
+  if (is.null(height)) {
+    height <- resolution(data[[y_major]]) * 0.95     
+    message("Using height ", format(height, digits = 3))
+  }
+  else {
+    height <- resolution(data[[y_major]]) * height    
+    message("Using height ", format(height, digits = 3))
+  }
+  
+  if (polar) {
+  	xlim <- range(data[[x_minor]]) + c(-0.5, 0.5)*resolution(data[[x_minor]])
+    theta <- 2 * pi * rescale01(data[[x_minor]], xlim)
+    r <- rescale01(data[[y_minor]], ylim)
+    
+    data$gx <- data[[x_major]] + width  / 2 * r * sin(theta) 
+    data$gy <- data[[y_major]] + height / 2 * r * cos(theta)
+    data <- data[order(data[[x_major]], data[[x_minor]]), ] 
+  } else {
+    data$gx <- data[[x_major]] + rescale11(data[[x_minor]]) * width / 2
+    data$gy <- data[[y_major]] + rescale11(data[[y_minor]], ylim) * height / 2 
+  }
+  
+  data
+}

### 2011-09-20 code/maps.R

@@ -1,0 +1,23 @@
+library("maps")
+library("ggplot2")
+
+world <- map_data("world")
+# Enter the coords of the data for the limits
+world <- getbox(world, xlim = c(-114.93, -55.07), ylim = c(-21.53, 36.64))
+# Remove slivvers
+world <- ddply(world, "group", function(df) {
+  if (diff(range(df$long)) < 1e-6) return(NULL)
+  if (diff(range(df$lat)) < 1e-6) return(NULL)
+  df
+})
+
+map <- list(
+  geom_polygon(aes(long, lat, group = group), inherit.aes = FALSE, 
+    data = world, legend = FALSE, fill = "grey80", colour = "grey90"),
+  scale_x_continuous(breaks = NA, expand = c(0.02, 0)),
+  scale_y_continuous(breaks = NA, expand = c(0.02, 0)), 
+  xlab(NULL),
+  ylab(NULL))
+
+map_mini <- c(map, list(
+  coord_cartesian(xlim = c(-114.93, -85), ylim = c(7.55, 36.64), wise = T)))

### 2011-09-20 gridded.temp.r

@@ -16,13 +16,23 @@ temp.usa.melt.gly<-glyphs(temp.usa.melt, "lon", "time", "lat", "temp")
 qplot(gx, gy, data=temp.usa.melt.gly, group=gid, geom="line")
 
 library(maps)
+world <- map_data("world")
+world <- getbox(world, xlim = c(-138, -55), ylim = c(22, 52))
+world <- ddply(world, "group", function(df) {
+  if (diff(range(df$long)) < 1e-6) return(NULL)
+  if (diff(range(df$lat)) < 1e-6) return(NULL)
+  df
+})
 
-outlines <- as.data.frame(map("state",
-                              plot=FALSE)[c("x","y")])
-map <- c(
-  geom_path(aes(x, y), inherit.aes = FALSE, 
-    data = outlines, colour = alpha("grey60", 0.4))
-)
+colnames(world)[1]<-"lon"
+
+map <- list(
+  geom_polygon(aes(lon, lat, group = group), inherit.aes = FALSE, 
+    data = world, legend = FALSE, fill = "grey80", colour = "grey90"),
+  scale_x_continuous(breaks = NA, expand = c(0.02, 0)),
+  scale_y_continuous(breaks = NA, expand = c(0.02, 0)), 
+  xlab(NULL),
+  ylab(NULL))
 
 qplot(gx, gy, data=temp.usa.melt.gly, group=gid, geom="line") + map
 
@@ -38,6 +48,7 @@ lon<-(lon+180)%%360 - 180
 length(lon)
 dim(temp)
 
+library(ggplot2)
 # Ok, this now works for transforming the lon, and selecting the USA
 temp.usa<-temp[114:150, 20:33,] # lon -130, -65; lat 25, 50
 temp.usa.melt<-melt(temp.usa)
@@ -47,11 +58,12 @@ temp.usa.melt$lat<-lat[temp.usa.melt$gridy+19]
 
 temp.usa.melt.gly<-glyphs(temp.usa.melt, "lon", "time", "lat", "temp")
 qplot(gx, gy, data=temp.usa.melt.gly, group=gid, geom="line") + map + coord_map() # Testing
-ggplot(temp.usa.melt.gly, aes(gx, gy, group = gid)) + 
-  map + 
-  geom_path() +
-  ref_boxes +
-  theme_fullframe()
+p <-ggplot(temp.usa.melt.gly, aes(gx, gy, group = gid)) + 
+  map
[…]

### 2011-09-21 paper.tex
re-worked up to computation

@@
 
+In this paper we describe the algorithm for creating glyphs of temporal data on maps, generally in R, and the various modifcations that can enable different features in the data to be detected.
+The next section (\ref{sec:glyph-map}) describes the use of glyphs on maps, and Section \ref{sec:construction} explains how to construct glyph-maps.
+Section \ref{sec:reference} discusses the use of reference lines and boxes, and Section \ref{sec:scale} visits issues of scaling values for different information purposes.
+Generalizations to different types of icons are described in Section \ref{sec:scatter}.
+Glyph-maps work nicely with gridded data, but many spaiottemporaral data sets have irregular spatial locations.
+Possible adjustments are discussed in Section \ref{sec:irregular}.
+Code and data is available in the online supplementary materials.
+
+\section{Glyph-maps}~\label{sec:glyph-map}
+
 % Reference Tufte's small multiples
-Figure \ref{fig:gistemp-pred} (left) shows the time series icons and star glyphs drawn for each spatial location of the predicted temperature data across the USA, we call it a \emph{glyph-map}.
-The data contains 2$^o$x2$^o$ gridded temperature anomalies 1950-2010, from the GISTEMP surface temperature data provided by the Earth System Research Laboratory, Physical Sciences Division, National Oceanic and Atmospheric Administration.
+\emph{Glyph-maps} are maps where a glyph is displayed at each spatial location to represent multiple data values.
+To explain the glyph-map, we use data of 2$^o$x2$^o$ gridded temperature anomalies, from the GISTEMP surface temperature data provided by the Earth System Research Laboratory, Physical Sciences Division, National Oceanic and Atmospheric Administration.
+Ground station data was de-seasonalized, differenced from ***, and spatially averaged to obtain gridded measurements.
+Figure \ref{fig:gistemp-pred} shows the time series icons for each spatial location of predicted temperature for 1950-2010 across the USA.
 Most locations observe an increasing temperature, to varying degrees, over this time period.
@@
 These locations tend to be different from their neighbors, which might suggest isolated data problems.
+There is a difference in the temperature increase at different regions: steeper inclines can be seen in the north and western mountain region and across the Great Lakes, but inclines are flatter in the southeastern USA.
 %The most seasonality occurs in north American.
@@
 
-  \caption{Glyph-maps of gridded predicted temperature anomaly data}
+  \caption{Glyph-map of predicted temperature anomaly data, from
+    1950-2010. Increasing trends can be seen over most of the USA,
+    with varying degrees of increase. Some isolated locations exhibit
+    steep declines, and inclines, perpahs indicative of some isolated
+    data problems.}
   \label{fig:gistemp-pred}
 \end{figure}
+
+*** Remove the strange data over ocean, and make the reference lines white?
 
@@
 
-In this paper we describe the algorithm for creating glyphs of temporal data on maps, generally in R, and the various modifcations that can enable different features in the data to be detected.
-The next section (\ref{sec:construction}) describes how to construct glyph-maps.
-Section \ref{sec:reference} discusses the use of reference lines and boxes, and Section \ref{sec:scale} visits issues of scaling values for different information purposes.
-Generalizations to different types of icons are described in Section \ref{sec:scatter}.
-Glyph-maps work nicely with gridded data, but may be less effective with irregular spatial locations.
-Possible adjustments are discussed in Section \ref{sec:irregular}.
-Code and data is available in the online supplementary materials.
+Figure \ref{fig:gistemp-pred} shows the star glyphs for each spatial location of raw temperature for 1880-2011 across the USA.
+The major feature to notice is the pattern of missing values, icons that do not complete the full circle.
+The high noon position of the circle is the first time point, Jan 1880, and it is this early time period that has missing values are a lot of locations.
+It was for this reason that the temporal model was fit just to 1950-2010 data.
+Another important feature is variability, indicated by the thickness of the circle.
+Some locations, particularly in the north western mountain region, have big variability in temperature, both low minima, and high maxima.
+At these locaations there are large fluctuations from the long term mean.
+
+\begin{figure}[htbp]
+  \centering
+  %\includegraphics[width=1\linewidth]{gistemp-pred}%
+
+  \includegraphics[width=1\linewidth]{gistemp-polar-raw}
+
+  \caption{Glyph-map of raw temperature anomaly data, from 1880-2011. Gaps in the full circle indicate missing values, and thickness of the circle indicates more variability in the measurements.}
+  \label{fig:gistemp-raw}
+\end{figure}
+
+*** Need to add reference circles.
 
@@
 
-\begin{figure}[htbp]
-  \centering
-  \includegraphics[width=1\linewidth]{gistemp-raw}%
+%\begin{figure}[htbp]
+%  \centering
+%  \includegraphics[width=1\linewidth]{gistemp-raw}%
 
-  \includegraphics[width=1\linewidth]{gistemp-polar-raw}
+%  \includegraphics[width=1\linewidth]{gistemp-polar-raw}
 
-  \caption{Glyph-maps of gridded raw temperature anomaly data}
-  \label{fig:ref-gistemp}
-\end{figure}
+%  \caption{Glyph-maps of gridded raw temperature anomaly data}
+%  \label{fig:ref-gistemp}
+%\end{figure}
 

### 2011-09-21 gridded.temp.r

@@ -67,11 +67,12 @@ p <- p + theme_fullframe() + coord_map() # but not as slow as this part
 ggsave("../images/gistemp-raw.png", width = 8, height = 4)
 
 temp.usa.melt.gly<-glyphs(temp.usa.melt, "lon", "time", "lat", "temp", polar=T)
-ggplot(temp.usa.melt.gly, aes(gx, gy, group = gid)) + 
-  map + 
-  geom_path() +
-  ref_boxes +
-  theme_fullframe() + coord_map()
+p <- ggplot(temp.usa.melt.gly, aes(gx, gy, group = gid)) + 
+  map  
+p <- p + geom_path() 
+p <- p +  ref_boxes 
+#  theme_fullframe() + coord_map()
+p
 ggsave("../images/gistemp-polar-raw.png", width = 8, height = 4)
 
 # Predictions

### 2011-09-21 paper.tex
worked through first half of paper, and added irregular grid example

@@
 
-A big advantage of glyph-maps is that they can be generated with existing graphics software after performing a simple pre-processing step, making them easily accessible to a wide audience.
-The pre-processing step is simple, once you notice that the position of graphical elements is determined by a major and minor axis in each direction ($x$ and $y$).
-For maps, the major axes are latitude and longitude, and the minor axes are typically time and some measurement or model prediction.
-In addition, the minor axes are rescaled to have range $[0, 1]$.
-The final data values are a simple linear combination, as given by
+%A big advantage of glyph-maps is that they can be generated with existing graphics software after performing a simple pre-processing step, making them easily accessible to a wide audience. 
+Creating a glyph-map, involves recognizing that they combine two structural components of the data: spatial location and data values.
+The spatial location is the major positioning component, and the data values are minor adjustments to these positions.
+For spatiotemporal data, the major axes are latitude ($y_{\amaj}$) and longitude ($x_{\amaj}$), and the minor axes are time ($x_{\amin}$) and some measurement ($y_{\amin}$), for example, temperature, or predicted temperature.
+The minor axes are rescaled to have range $[0, 1]$, and further adjusted to fit into the resolution of the spatial coordinates.
+The final values to be plotted are a simple linear combination, as given by
 
@@
 where $w$ and $h$ are width and height respectively.
-The width and height will typically take into account the resolution of the major and minor axes, that is the minimum difference between consecutive locations.
+The width and height will typically take into account the resolution of the major axes, that is, the minimum difference between consecutive locations.
 
@@
 \end{equation}
-Note that this is is a non-standard conversion to polar coordinates, but this conversion provides us with a time line that starts at 12 o'clock and wraps around the origin in a clockwise fashion.
+Note that, this is is a non-standard conversion to polar coordinates, but this conversion provides a time line that starts at 12 o'clock and wraps around the origin in a clockwise fashion.
 
@@
 In this case, the star glyphs are not as effective as the linear glyphs: perhaps because of the absence of cyclical pattern.
+
+*** Perhaps replace these figures with something from the GISTEMP data.
 
@@
 
-Figure \ref{fig:ref-basic-polar} gives some small examples of time series with strongly expressed trends, shown both in cartesian coordinates and polar coordinates.
+Figure \ref{fig:templates} gives some small examples of time series with prototypical trends, shown both in cartesian coordinates and polar coordinates.
+Differences between linear and nonlinear trend are more apparent in the time series icons, and are effectively lost in polar coordinates.
+The star plots are most effective in exposing cyclical patterns due to seasonality.
+Seasonality appears like floral cartoons in the polar coordinates, with peaks forming petals.
 What might be surprising is that time series that seem to have opposing trends in cartesian coordinates don't necessarily show this symmetry in polar coordinates.
-What becomes apparent, though is that the area of glyphs in polar coordinates mainly reflect the average value of a time series, while its shape shows deviations from average values.
-Adding a circular reference line helps to pick out these deviations.
+The area of glyphs in polar coordinates mainly reflect the average value of a time series, while its shape shows deviations from average values.
 
-Since the $x$ and $y$ locations are linear combinations of two variables, these types of displays can also be built interactively by software that supports tours \citep{cook:2006}, such as DataViewer \citep{buja:1986}, XGobi \citep{swayne:1991} or GGobi \citep{swayne:2003}.
-This technique is used to good effect in \citet{buja:1996a}.
+Since the coordinates are linear combinations of two variables, it is possible to build these these time series icon displays interactively, illustrated by software packages that have incorporated tours \citep{cook:2006}, such as DataViewer \citep{buja:1986}, XGobi \citep{swayne:1991} or GGobi \citep{swayne:2003}.
+This technique is used to good effect in \citet{buja:1996a}, and was used to examine the climate data in \citet{hobbs:2010}.
 
@@
 % see differently from the facetted map
-
-The graphics in this paper were created in R \citep{R} using {\tt ggplot2} \citep{me:ggplot2} and {\tt plyr} \citep{me:plyr}.
-
-*** These figures need to be placed yet, but are here as a start on better climate data examples:
 
@@
 
-Graphical displays benefit from a \emph{visual reference grid} \citep{cleveland:1993a} to compare values mapped to position along a line.
-The same is true for glyph-maps, but it is not feasible to make a full set of axes for each icon.
-A complete axis would dominate the picture, and draw the eye away from the pattern of the data.
-The types of comparisons that are needed across icons are:
+A \emph{visual reference grid} \citep{cleveland:1993a} is fundamental to a good data visualization, in order to compare values mapped to position along a line.
+For glyph-maps, it is not feasible to make a full set of axes for each icon, because these would dominate the picture.
+Visual reference grids need to be post-attentive, de-emphasized, and serve as posterior look up after the patterns in the data have been read.
+For glyph-maps, we need to be able to make the following types of comparisons across icons:
 
@@
 
-Space-time-glyphs usually require a  to make it easier to compare locations.
-This is important for accurate perception of important line properties like slope and minima.
-This need is evident in Figure~\ref{fig:glyph-map}.
-There is a hint that the average value is not the same from right time series to the next in the map (right), or if the center of the icons is the same in the polar time series (left).
-We have no frame of reference on which to judge this.
-The regular grid provides a very rudimentary frame of reference, but it is not sufficient.
+The structured spatial arrangement of icons in a glyph-map helps to compare the patterns of shape, like slope, intercept, or size, across icons.
+But, additional clues would make it comparisons easier.
+For example glyph-map in Figure~\ref{fig:gistemp-pred}, contains two additional guides, a reference grid, and reference line.
+Both are drawn in white, because it is minimally perceptible (post-attentive), relative to the dark black ink of the data.
 
-Figure~\ref{fig:ref-basic} shows two reference grid variants: a line drawn at the overall mid-range, and a box drawn around each grid cell.
+The reference grid is a box, framing each icon, representing the spatial grid.
+It helps to read differences in slope and intercept.
+Slope is read by comparing the left and right endpoints of the predictions (line) against the relative location on the frame.
+Intercept is compared by reading the ``average'' position of the line in the box, near the top, or near the bottom.
+The reference line, drawn as a horizontal line at the average predicted temperature value, provides for better reading of the intercept.
+The average value for the temperature anomaly data is not the same across the map: at higher latitudes the temperatures are uniformly higher, but at lower latitudes they are uniformly lower.
+(**** Can this be right?
+Would it mean that at these locations the temperatures 1950-2010 are much higher than the reference year temperatures?)
+
+%Figure~\ref{fig:ref-basic} shows two reference grid variants: a line drawn at the overall mid-range, and a box drawn around each grid cell.
 We prefer using white for these references because it is minimally perceptible: you can see the grid easily when looking for it, but it does not otherwise distract from the perception of the graphic \citep{carr:1994}.
@@
 % Think we might need to show the original figures of data with reference lines
+
+*** Re-do these figures with zooms in on the GISTEMP data
 
@@
 \end{figure}
+
+For star glyphs, the reference frames are also useful, but the reference line would be replaced by a reference circle, drawn at the average data value.
+Figures \ref{fig:templates} and \ref{fig:gistemp-raw} show examples.
+
+*** Re-visit this next paragraph - its a little too airy-fairy, and the plot with color needs to have the color dampened more. Its way too dominant.
 
@@
 
+*** Adding a circular reference line helps to pick out these deviations.
+
 \section{Perception of structure}~\label{sec:scale}
@@
 
-*** This section might benefit from some equations also.
-
 *** Global/local min/max, remove intercept/average, data vs model scale
+*** This section needs to have plots re-done, once we have decided on the key pieces
 
@@
 
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=0.8\linewidth]{ghcn-mountains}%
+  \caption{Glyph-maps of ground station temperature values, 2001-2010, for mountain states.}
+  \label{fig:irregular}
+\end{figure}
+
 \section{Long time series}
@@
 
+The graphics in this paper were created in R \citep{R} using {\tt ggplot2} \citep{me:ggplot2} and {\tt plyr} \citep{me:plyr}.
+
 \bibliography{references}

### 2011-09-21 code/ghcn.r

@@ -33,10 +33,37 @@ qplot(lon, lat, data=colorado)
 colorado.gly <- glyphs(colorado, "lon", "time", "lat", "temp", width=100, height=100) 
 qplot(gx, gy, data=colorado.gly, group=gid, geom="line")
 
+# Mountain states: Colorado, Wyoming, Idaho, Montana, Nevada, Utah
+mountains <- subset(temp.01.10.melt, state == "CO" | state =="WY" | state =="ID" | state == "MT" | state == "NV"| state == "UT")
+ddply(mountains, "stn", summarise, nstns = length(temp))
+mountains <- subset(mountains, stn != "261071")
+mountains <- subset(mountains, stn != "426686")
+mountains <- subset(mountains, stn != "421731")
+mountains <- subset(mountains, stn != "428973")
+mountains$state<-factor(mountains$state)
+qplot(lon, lat, data=mountains, colour=state)
+models <- dlply(mountains, c("lat", "lon"), function(df) {
+  lm(temp ~ year + month, data=df)
+})
+year_grid <- expand.grid(year = unique(mountains$year), month = "jan")
+year_preds <- ldply(models, function(mod) {
+  year_grid$pred <- predict(mod, newdata = year_grid)
+  year_grid
+})
+mountains.gly<-glyphs(year_preds, "lon", "year", "lat", "pred", width=1000, height=4000)
+ggplot(mountains.gly, aes(gx, gy, group = gid)) + 
+  #map +
+  geom_line(aes(y = lat), colour = "white", size = 1) +
+  geom_path() +
+  ref_boxes +
+  theme_fullframe() + coord_map()
+ggsave("../images/ghcn-mountains.png")
 
-models <- dlply(temp.01.10.melt, .(stn), function(df) {
-  df$mpi <- mpi(df$time)
-  lm(temp ~ time + sin(mpi) + cos(mpi), data=df)
+temp.01.10.melt <- subset(temp.01.10.melt, !is.na(temp))
+models <- dlply(temp.01.10.melt, c("lat", "lon"), function(df) {
+  #df$mpi <- mpi(df$time)
+  #lm(temp ~ time + sin(mpi) + cos(mpi), data=df)
+  lm(temp ~ year + month, data=df)
 })
 
 preds <- ldply(models, function(m) {

### 2011-09-21 code/glyph-utils.R

@@ -44,3 +44,4 @@ ref_boxes <- list(
 
 # Need to have a function for the circles references for polar coords
 # Need to have a function for the reference lines
+# Need to have reference boxes for the irregular grid, too

### 2011-09-21 code/glyphs.R

@@ -15,6 +15,7 @@ rescale11 <- function(x, xlim=NULL) 2 * rescale01(x, xlim) - 1
 # scale individually, globally
 # Absolute ylims, for use when plotting model predictions on the data
 #    scale - got it with the ylim argument
+# Resolution needs to take into account irregular grid
 glyphs <- function(data, x_major, x_minor, y_major, y_minor, polar = F, height = NULL, width = NULL, ylim=NULL) {
   data$gid <- interaction(data[[x_major]], data[[y_major]], drop = TRUE)
   

### 2011-09-21 code/templates.R

@@ -19,7 +19,7 @@ x12 <- mu + seq(0, 30, length=36) + amp*sin(t*6)/2 # seasonal + smaller amplitud
 locs <- data.frame(cbind(rep(t,12), c(x1,x2,x3,x4,x5,x6,x7,x8,x9,x10,x11,x12), c(rep(1:6, each=36), rep(1:6, each=36)), rep(2:1,each=36*6)))
 names(locs) <- c("time","temp","x_max", "y_max")
 
-locs2 <- data.frame(cbind(rep(t+pi,12), rep(c(mean(x1),mean(x2),mean(x3),mean(x4),mean(x5),mean(x6),mean(x7),mean(x8),mean(x9),mean(x10),mean(x11),mean(x12)), each=36), c(rep(1:6, each=36), rep(1:6, each=36)), rep(2:1,each=36*6)))
+locs2 <- data.frame(cbind(rep(t+pi,12), rep(mean(locs$temp), 12*36), c(rep(1:6, each=36), rep(1:6, each=36)), rep(2:1,each=36*6)))
 names(locs2) <- c("time","temp","x_max", "y_max")
 
 library(ggplot2)
@@ -29,7 +29,7 @@ qplot(time,temp, data=subset(locs, x_max==1), geom="line")
 locs <- glyphs(locs, "x_max", "time", "y_max", "temp") 
 locs2 <- glyphs(locs2, "x_max", "time", "y_max", "temp", ylim=range(locs$temp)) 
 ggplot(locs2, aes(gx, gy, group = gid)) + 
-  geom_path(colour="grey80") +
+  geom_path(colour="white", size=1.5) +
   geom_path(aes(gx, gy, group = gid), data=locs, colour="black") +
   geom_tile(aes(x_max, y_max), colour = "white", fill = NA) +
   theme_fullframe() + coord_equal()
@@ -40,7 +40,7 @@ ggsave("../images/euclid-to-polar-1.pdf")
 locs <- glyphs(locs, "x_max", "time", "y_max", "temp", polar = T) 
 locs2 <- glyphs(locs2, "x_max", "time", "y_max", "temp", ylim=range(locs$temp), polar = T) 
 ggplot(locs, aes(gx, gy, group = gid)) + 
-  geom_path(aes(gx, gy, group = gid), data=locs2, colour="grey80") +
+  geom_path(aes(gx, gy, group = gid), data=locs2, colour="white", size=1.5) +
   geom_path() +
[…]

### 2011-09-21 gridded.temp.r

@@ -60,11 +60,11 @@ temp.usa.melt.gly<-glyphs(temp.usa.melt, "lon", "time", "lat", "temp")
 qplot(gx, gy, data=temp.usa.melt.gly, group=gid, geom="line") + map + coord_map() # Testing
 p <-ggplot(temp.usa.melt.gly, aes(gx, gy, group = gid)) + 
   map
-p <- p + geom_line(aes(y = lat), colour = "grey90", size = 1.5) 
+p <- p + geom_line(aes(y = lat), colour = "white", size = 1.2) 
 p <- p + geom_path() 
 p <- p + ref_boxes # This part is really slow
 p <- p + theme_fullframe() + coord_map() # but not as slow as this part
-ggsave("../images/gistemp-raw.png", width = 8, height = 4)
+ghcn
 
 temp.usa.melt.gly<-glyphs(temp.usa.melt, "lon", "time", "lat", "temp", polar=T)
 p <- ggplot(temp.usa.melt.gly, aes(gx, gy, group = gid)) + 
@@ -101,7 +101,7 @@ year_preds <- ldply(temp_models, function(mod) {
 })
 temp.usa.melt.sub.gly<-glyphs(year_preds, "lon", "year", "lat", "pred")
 ggplot(temp.usa.melt.sub.gly, aes(gx, gy, group = gid)) + 
-  map + geom_line(aes(y = lat), colour = "grey80", size = 1.5) +
+  map + geom_line(aes(y = lat), colour = "white", size = 1.5) +
   geom_path() +
   ref_boxes +
   theme_fullframe() + coord_map()

### 2011-09-22 paper.tex
perception section

@@
 \centerline{\includegraphics[width=6in]{nasa-colored-map.png}}
-\caption{Facetted map of de-seasonalized temperature.}
+\caption{Facetted map of de-seasonalized temperature.
+The dominant feature is the El Ni\~no warming in the qouthern equatorial region in the last half of 1997 and first half of 1998.
+Smaller features are only noticeable on closer inspection, or if pointed out: such as the relative warming on the mountain regions in south and north America in the later years.}
 \label{fig:facetted-map}
@@
 
-*** Re-visit this next paragraph - its a little too airy-fairy, and the plot with color needs to have the color dampened more. Its way too dominant.
+%*** Re-visit this next paragraph - its a little too airy-fairy, and the plot with color needs to have the color dampened more.
+Its way too dominant.
+This paragraph is too speculative for the paper, so removing it from first draft.
 
-It's also often useful to display other types of reference information.
+%It's also often useful to display other types of reference information.
 You are only limited by your imagination (and the capabilities of your graphics software), but two simple ideas are to (1) vary the colour or transparency of glyphs or references, or (2) use an additional layer to highlight special points.
@@
 
-\begin{figure}[htbp]
-  \centering
-  \includegraphics[width=0.5\linewidth]{ref-max-1}%
-  \includegraphics[width=0.5\linewidth]{ref-max-2}
-  \caption{Glyph-maps of seasonal temperature patterns also showing the month with the highest temperature using (left) an additional layer of points and (right) changing the background colour of the reference grid.}
-  \label{fig:ref-adv}
-\end{figure}
+%\begin{figure}[htbp]
+%  \centering
+%  \includegraphics[width=0.5\linewidth]{ref-max-1}%
+%  \includegraphics[width=0.5\linewidth]{ref-max-2}
+%  \caption{Glyph-maps of seasonal temperature patterns also showing the month with the highest temperature using (left) an additional layer of points and (right) changing the background colour of the reference grid.}
+%  \label{fig:ref-adv}
+%\end{figure}
 
-A particularly useful summary when to add when displaying model predictions, is some measure of model fit. This makes it easier to avoid over-interpreting predictions from poorly-fitting models.
+%A particularly useful summary when to add when displaying model predictions, is some measure of model fit. This makes it easier to avoid over-interpreting predictions from poorly-fitting models.
 
-*** Adding a circular reference line helps to pick out these deviations.
+%*** Adding a circular reference line helps to pick out these deviations.
 
@@
 
-Compare the facetted map Fig 1 with the glyph-map of de-seasonalized data
+Compare the facetted map Figure \ref{facetted-map} with the glyph-map
+of de-seasonalized data in Figure \ref{fig:nasa-deseas-glyph}.
 
@@
 differences. Accurate perception of color differences is difficult.
+
+\begin{figure*}[htp]
+\centerline{\includegraphics[width=6in]{nasa-deseas-glyph.png}}
+\caption{Glyph-map of de-seasonalized temperature, to compare with Figure \ref{facetted-map}. }
+\label{fig:nasa-deseas-glyph}
+\end{figure*}
 

### 2011-09-22 code/facetted-colored-map.r

@@ -1,0 +1,54 @@
+# This file contains the code for first plot in the paper
+# The facetted, colored map
+nasa <- readRDS(file.choose()) # nasadata.rds
+
+library("maps")
+library("ggplot2")
+
+outlines <- as.data.frame(map("world", plot = FALSE,
+  xlim = -c(113.8, 56.2), ylim = c(-21.2, 36.2))[c("x","y")])
+map <- geom_path(aes(x, y), inherit.aes = FALSE, data = outlines, 
+  alpha = 0.2, legend = FALSE)
+
+# Fit models
+# Seasonal only
+temp_seas_models <- dlply(nasa, c("lat", "long"), function(df) {
+  # lm(ozone ~ factor(month) - 1, data = df)
+  lm(surftemp ~ factor(month), data = df)
+})
+
+# Including trend
+temp_trend_models <- dlply(nasa, c("lat", "long"), function(df) {
+  lm(surftemp ~ year + factor(month), data = df)
+})
+
+locs <- dlply(nasa, c("lat", "long"))
+
+resids <- mdply(cbind(d = locs, m = temp_seas_models), function(d, m) {
+  d$temp_resid <- d$surftemp - predict(m, newdata = d)
+  d
+})
+
+preds <- mdply(cbind(d = locs, m = temp_seas_models), function(d, m) {
+  d$temp_pred <- predict(m, newdata = d)
+  d
+})
+
+theme_nothing <- function (base_size = 12) {
+  structure(list(
+	panel.grid.minor=theme_blank(), 
+	panel.grid.major=theme_blank(), 
+	panel.background=theme_blank(), 
+	panel.border = theme_blank(), 
+	panel.margin = unit(0, "lines"), 
+	axis.title.x=theme_blank(), 
+	axis.title.y=theme_blank(), 
+	axis.line=theme_blank(), 
+	axis.ticks=theme_blank(), 
+	axis.text.y = theme_blank(), 
+	axis.text.x = theme_blank()#,
+	#legend.position = "none"
+  ), class = "options")
+}
+qplot(long, lat, data=resids, geom="tile", fill=temp_resid, facets=year~month) + scale_fill_gradient2(midpoint=0, mid="white", high="red", low="blue") + map + theme_nothing() + coord_map()
+ggsave("../images/nasa-colored-map.png", height=10, width=18)

### 2011-09-22 code/glyph-examples.r

@@ -1,0 +1,28 @@
+# This file contains the glyph-map for nasa data Might think about
+# putting all of the glyph-map examples, using gistemp and ushcn data
+# here too
+world <- map_data("world")
+world <- getbox(world, xlim = c(-114.93, -55.07), ylim = c(-21.53, 36.64))
+# Remove slivvers
+
+world <- ddply(world, "group", function(df) {
+  if (diff(range(df$long)) < 1e-6) return(NULL)
+  if (diff(range(df$lat)) < 1e-6) return(NULL)
+  df
+})
+
+map <- list(
+  geom_polygon(aes(long, lat, group = group), inherit.aes = FALSE, 
+    data = world, legend = FALSE, fill = "grey80", colour = "grey90"),
+  scale_x_continuous(breaks = NA, expand = c(0.02, 0)),
+  scale_y_continuous(breaks = NA, expand = c(0.02, 0)), 
+  xlab(NULL),
+  ylab(NULL))
+
+# De-seasonalized nasa data: matches the facetted, coloured map, need
+# to run that code before this
+resids.gly <- glyphs(resids, "long", "day", "lat", "temp_resid") 
+colnames(resids.gly)[6] <- "lon"
+ggplot(resids.gly, aes(gx, gy, group = gid)) + map + geom_line(aes(y = lat), colour = "white", size = 1.5) + geom_path() + ref_boxes + theme_fullframe() + coord_map()
+#qplot(gx, gy, data = resids, geom = "line", group = gid) + map
+ggsave("../images/nasa-deseas-glyph.png")

### 2011-09-23 paper.tex
perception section

@@
 
-Compare the facetted map Figure \ref{facetted-map} with the glyph-map
-of de-seasonalized data in Figure \ref{fig:nasa-deseas-glyph}.
-
 Glyph-maps using trend lines enable the reading of trends across
@@
 from facetted colored maps one needs to spot the difference between
-the different maps. A more direct comparison would be to make a single
-map with the slope of the trend coded as color. Cleveland's
-\citep{cleveland:1993a} perceptual tests suggests that the reader can
-perceive the positional difference of the lines better than the color
-differences. Accurate perception of color differences is difficult.
+the different maps.  Compare the facetted map in Figure
+\ref{fig:facetted-map} with the glyph-map of de-seasonalized data in
+Figure \ref{fig:nasa-deseas-glyph}. The glyph-map provides a texturing
+over the map. The El Ni\~no event is still primary structure, visible
+as the region that has a bump in the middle of the time series, the
+hot 1997-1998 period. We can also see that the temperatures are
+generally more varied over land. There are a few locations with
+dramatically increasing temperature in the later years
+
+Another common approach is to compute the slope at each location code
+this as color on the map. Cleveland's \citep{cleveland:1993a}
+perceptual tests suggests that the reader can perceive the positional
+difference of the lines better than the color differences. Accurate
+perception of color differences is difficult.
 
 \begin{figure*}[htp]
-\centerline{\includegraphics[width=6in]{nasa-deseas-glyph.png}}
-\caption{Glyph-map of de-seasonalized temperature, to compare with Figure \ref{facetted-map}. }
+\centerline{\includegraphics[width=5in]{nasa-deseas-glyph.png}}
+\caption{Glyph-map of de-seasonalized temperature, to compare with Figure \ref{fig:facetted-map}. }
 \label{fig:nasa-deseas-glyph}

### 2011-09-23 code/glyph-examples.r

@@ -21,7 +21,7 @@ map <- list(
 
 # De-seasonalized nasa data: matches the facetted, coloured map, need
 # to run that code before this
-resids.gly <- glyphs(resids, "long", "day", "lat", "temp_resid") 
+resids.gly <- glyphs(resids, "long", "day", "lat", "temp_resid", height=2.5) 
 colnames(resids.gly)[6] <- "lon"
 ggplot(resids.gly, aes(gx, gy, group = gid)) + map + geom_line(aes(y = lat), colour = "white", size = 1.5) + geom_path() + ref_boxes + theme_fullframe() + coord_map()
 #qplot(gx, gy, data = resids, geom = "line", group = gid) + map

### 2011-09-23 paper.tex
finished perception section

@@
 
-A \emph{visual reference grid} \citep{cleveland:1993a} is fundamental to a good data visualization, in order to compare values mapped to position along a line.
-For glyph-maps, it is not feasible to make a full set of axes for each icon, because these would dominate the picture.
-Visual reference grids need to be post-attentive, de-emphasized, and serve as posterior look up after the patterns in the data have been read.
-For glyph-maps, we need to be able to make the following types of comparisons across icons:
+A \emph{visual reference grid} \citep{cleveland:1993a} is fundamental
+to a good data visualization, in order to compare values mapped to
+position along a line. For glyph-maps, it is not feasible to make a
+full set of axes for each icon, because these would dominate the
+picture. Visual reference grids need to be post-attentive
+\citep{healey}, de-emphasized, and serve as posterior look up after
+the patterns in the data have been read. For glyph-maps, we need to be
+able to make the following types of comparisons across icons:
 
@@
 Figure \ref{fig:nasa-deseas-glyph}. The glyph-map provides a texturing
-over the map. The El Ni\~no event is still primary structure, visible
-as the region that has a bump in the middle of the time series, the
-hot 1997-1998 period. We can also see that the temperatures are
-generally more varied over land. There are a few locations with
-dramatically increasing temperature in the later years
+over the map. The El Ni\~no event is still the primary visble
+structure, the region that has a bump in the middle of the time
+series, correspondingto the hot 1997-1998 period. We can also see
+other temporal features: that the temperatures are generally more
+varied over land, there are a few locations with dramatically
+increasing temperature in the later years typically in locations that
+are the high elevation areas, along the Andes, and Sierra Madre
+Occiddental in Mexico.
 
-Another common approach is to compute the slope at each location code
-this as color on the map. Cleveland's \citep{cleveland:1993a}
-perceptual tests suggests that the reader can perceive the positional
-difference of the lines better than the color differences. Accurate
-perception of color differences is difficult.
+The glyph-map can be considered to be small multiples as named by
+\citet{tufte:2001}, in this case, small multiples of time series. By
+laying the time series out on the spatial grid, the main task expected
+of the viewer is comparing the temporal patterns from one location to
+the next. In contrast, in the small multiples used in the facetted
+colored map, the viewer is expected to compare the patterns in the
+maps from one year to the next, for corresponding months. (See
+\citet{carr:1999} for a discussion on perceptual grouping.) That is,
+compare and contrast maps with factted colred map, or compare and
+contrast time series using the glyph-map. One draw-back of the
+glyph-maps, particularly if linear trend is displayed as an icon, is
+they may suffer from the Z\"ollner Illusion \citep{Zollner} which is
+visible in Figure \ref{fig:gistemp-pred}. On the other hand, comparing
+images require that we play a spot the difference game, which is
+perceptually difficult \citet{busey}.
+
+An alternative, to both of these approaches, to focus attention on
+specific parts of the data calculate the relevant statistic for each
+location and display these values as color on a map. For example, to
+study the warming or cooling trends, we might calculate the slope of a
+linear model at each location and color tiles across the map by the
+value of the slope. From this we might see the increasing temperatures
+at higher altitudes in the NASA temperature data. However, we would
+fail to see the finer points of the time series, that the increase is
+not as simple as a linear increase. In addition, Cleveland's
+\citep{cleveland:1993a} perceptual tests suggest that the reader can
+perceive the positional difference of the lines better than the color
+differences. Accurate perception of color differences is difficult.
 

### 2011-09-23 paper.tex
finished scale section

@@
 
-Glyph-maps using trend lines enable the reading of trends across
-spatial locations directly from the plot. In comparison, to read trend
-from facetted colored maps one needs to spot the difference between
-the different maps.  Compare the facetted map in Figure
-\ref{fig:facetted-map} with the glyph-map of de-seasonalized data in
-Figure \ref{fig:nasa-deseas-glyph}. The glyph-map provides a texturing
-over the map. The El Ni\~no event is still the primary visble
-structure, the region that has a bump in the middle of the time
-series, correspondingto the hot 1997-1998 period. We can also see
-other temporal features: that the temperatures are generally more
-varied over land, there are a few locations with dramatically
-increasing temperature in the later years typically in locations that
-are the high elevation areas, along the Andes, and Sierra Madre
-Occiddental in Mexico.
+Glyph-maps using trend lines enable the reading of trends across spatial locations directly from the plot.
+In comparison, to read trend from facetted colored maps one needs to spot the difference between the different maps.
+Compare the facetted map in Figure \ref{fig:facetted-map} with the glyph-map of de-seasonalized data in Figure \ref{fig:nasa-deseas-glyph}.
+The glyph-map provides a texturing over the map.
+The El Ni\~no event is still the primary visble structure, the region that has a bump in the middle of the time series, correspondingto the hot 1997-1998 period.
+We can also see other temporal features: that the temperatures are generally more varied over land, there are a few locations with dramatically increasing temperature in the later years typically in locations that are the high elevation areas, along the Andes, and Sierra Madre Occiddental in Mexico.
 
-The glyph-map can be considered to be small multiples as named by
-\citet{tufte:2001}, in this case, small multiples of time series. By
-laying the time series out on the spatial grid, the main task expected
-of the viewer is comparing the temporal patterns from one location to
-the next. In contrast, in the small multiples used in the facetted
-colored map, the viewer is expected to compare the patterns in the
-maps from one year to the next, for corresponding months. (See
-\citet{carr:1999} for a discussion on perceptual grouping.) That is,
-compare and contrast maps with factted colred map, or compare and
-contrast time series using the glyph-map. One draw-back of the
-glyph-maps, particularly if linear trend is displayed as an icon, is
-they may suffer from the Z\"ollner Illusion \citep{Zollner} which is
-visible in Figure \ref{fig:gistemp-pred}. On the other hand, comparing
-images require that we play a spot the difference game, which is
-perceptually difficult \citet{busey}.
+The glyph-map can be considered to be small multiples as named by \citet{tufte:2001}, in this case, small multiples of time series.
+By laying the time series out on the spatial grid, the main task expected of the viewer is comparing the temporal patterns from one location to the next.
+In contrast, in the small multiples used in the facetted colored map, the viewer is expected to compare the patterns in the maps from one year to the next, for corresponding months.
+(See \citet{carr:1999} for a discussion on perceptual grouping.) That is, compare and contrast maps with factted colred map, or compare and contrast time series using the glyph-map.
+One draw-back of the glyph-maps, particularly if linear trend is displayed as an icon, is they may suffer from the Z\"ollner Illusion \citep{Zollner} which is visible in Figure \ref{fig:gistemp-pred}.
+On the other hand, comparing images require that we play a spot the difference game, which is perceptually difficult \citet{busey}.
 
-An alternative, to both of these approaches, to focus attention on
-specific parts of the data calculate the relevant statistic for each
-location and display these values as color on a map. For example, to
-study the warming or cooling trends, we might calculate the slope of a
-linear model at each location and color tiles across the map by the
-value of the slope. From this we might see the increasing temperatures
-at higher altitudes in the NASA temperature data. However, we would
-fail to see the finer points of the time series, that the increase is
-not as simple as a linear increase. In addition, Cleveland's
-\citep{cleveland:1993a} perceptual tests suggest that the reader can
-perceive the positional difference of the lines better than the color
-differences. Accurate perception of color differences is difficult.
+An alternative, to both of these approaches, to focus attention on specific parts of the data calculate the relevant statistic for each location and display these values as color on a map.
+For example, to study the warming or cooling trends, we might calculate the slope of a linear model at each location and color tiles across the map by the value of the slope.
+From this we might see the increasing temperatures at higher altitudes in the NASA temperature data.
+However, we would fail to see the finer points of the time series, that the increase is not as simple as a linear increase.
+In addition, Cleveland's \citep{cleveland:1993a} perceptual tests suggest that the reader can perceive the positional difference of the lines better than the color differences.
+Accurate perception of color differences is difficult.
 
@@
 
-*** Global/local min/max, remove intercept/average, data vs model scale
-*** This section needs to have plots re-done, once we have decided on the key pieces
+%*** Global/local min/max, remove intercept/average, data vs model scale
+%*** This section needs to have plots re-done, once we have decided on the key pieces
 
-Scaling of the data can produce dramatically different emphases on
-structure. For spatiotemporal data the most natural scale is to use
-the global minumum and maximum. The resulting icons will have size
-relative to the magnitude of the full range of the data. Locations
-with overall large or small temperature will be the focus. Figure
-\ref{fig:glyph-map} shows data scaled to -1 to 1 based on the global
-minimum and maximum, using:
+Scaling of the data, into the values used for each glyph, emphasizes different aspects of the data.
+For spatiotemporal data the most natural scale is to use the global minumum and maximum, because the temporal variables are measuring the same information, and we want to make comparisons across the spatial region.
+The icons will have size relative to the magnitude of the full range of the data.
+Locations with overall large temperature will catch the focus.
+Figure \ref{fig:nasa-deseas-glyph} shows data scaled to 0 to 1 based on the global minimum and maximum, using:
 
 \begin{equation}
-y_{st} = \frac{y_{st}-y_{min}}{y_{max}-y_{min}} \times 2 - 1
+y_{st} = \frac{y_{st}-y_{min}}{y_{max}-y_{min}} %\times 2 - 1
 \label{scale1}
@@
 \noindent where $y$ is the variable used on the minor axis in equations \ref{coords.eqn}, \ref{coords.polar.eqn}, $y_{min} = \mbox{minimum}\{y_{st}; s=1, \dots S, t=1, ..., T \}, y_{max} = \mbox{maximum}\{y_{st}; s=1, \dots S, t=1, ..., T\}$, $S=$ is the number of spatial locations and $T=$ is the number of time points.
-*** Reference other plots that use global scale too.
+In addition, $y_{st}$ may be further scaled by doubling and subtracting 1 to give values between -0.5 and 0.5 for using in equations \ref{coords.eqn}, \ref{coords.polar.eqn}.
 
-Alternatively, scaling locally at each spatial location enables
-exploring the seasonal scale, or shape differences, from one location
-to another. Fig \ref{***} uses local scaling using:
+Alternatively, scaling locally at each spatial location enables exploring shape differences, particularly seasonal patterns, from one location to another:
 
 \begin{equation}
-y_{st} = \frac{y_{st}-y_{s,min}}{y_{s,max}-y_{s,min}} \times 2 - 1
+y_{st} = \frac{y_{st}-y_{s,min}}{y_{s,max}-y_{s,min}} %\times 2 - 1
 \label{scale2}
@@
 With the local scaling, the shape of the icon, for example, differences in the seasonal patterns over spatial locations, are emphasized.
-Figure \ref{***} shows data that has been locally scaled. We can see *** The local scale is equivalent to using individual scales in the facetted graphics of \texttt{ggplot2}.
+Fig \ref{fig:scaling} shows glyph-maps generated usng both global scaling (left) and local scaling (right) for the NASA temperature data.
+On the global scale, the El Ni\~no blip in temperature is primarily visible, along with the generally flat series over most water locations, and somewhat increasing or variable temerpatures over land.
+The local scale emhpasizes the individual shapes: The El Ni\~no pattern is visible, but extend to a larger spatial region including the Caribbean, dramatic linear increases across the Andes, decreases in the Gulf of Mexico.
+(The local scale is equivalent to using individual scales in the facetted graphics of \texttt{ggplot2}).
+One caution, with local scale, the viewer needs to realize that the scale is not relative, big patterns in some locations might be just tiny effects.
+To indicate this on the locally scaled data plots, using a very pale coloring to code the original scale in the tiles, might help puts patterns in perspective.
 
-It can also be advantageous to relocate the data uniformly, by extracting the average at each location.
-Particularly for polar glyphs the average value can lead to some unintuitive patterns in the glyphs.
+Other types of shifting and scaling can be useful.
+The mean and standard deviation might be used to standardize values, or the median and interquartile range, might be useful with some types of data.
+Shifting the values at each location to have mean zero, places more focus on the trend.
+This is particularly important for the star glyphs, where differences in the average value lead to unintuitive patterns in the resulting glyphs.
 In this situation we would use $y_{st}-\bar{y}_s$ instead of $y_{st}$ in equations \ref{scale1}, \ref{scale2}, with minima and maxima calculated on the mean-shifted data.
 
-Another way to think about these scale options is partitioning the data structure into shape, magnitude and shift components.
+%Another way to think about these scale options is partitioning the data structure into shape, magnitude and shift components.
 The different ways to scale emphasize these different components.
@@
 
-**** This is different - I think it just confuses the topic of scaling.
+%**** This is different - I think it just confuses the topic of scaling.
 Because the GAM is not used to rescale the data, you still have to do that using the local or global scling equations above.
@@
   \centering
-  \includegraphics[width=0.33\linewidth]{month-rescale-none}%
-  \includegraphics[width=0.33\linewidth]{month-rescale-max}%
-  \includegraphics[width=0.33\linewidth]{month-rescale01}
-  \caption{Smoothed de-seasonalized daily temperature glyph-map. (Left) unscaled, (center) scaled to have maximum 1, (right) scaled to have maximum 1 and minimum 0.}
+  %\includegraphics[width=0.33\linewidth]{month-rescale-none}%
+  \includegraphics[width=0.5\linewidth]{month-rescale-max}%
+  \includegraphics[width=0.5\linewidth]{month-rescale01}
+  \caption{Smoothed de-seasonalized daily temperature glyph-map. (Left) %unscaled, (center) 
+globally scaled, (right) locally scaled.}
   \label{fig:scaling}
@@
 
-A possible compromise is to use line colour to display the standardising variable, as shown in Figure~\ref{fig:scaling-col}.
+%A possible compromise is to use line colour to display the standardising variable, as shown in Figure~\ref{fig:scaling-col}.
 This makes it possible to see the individual patterns while still retaining some ability to see which shapes represent small absolute changes.
 
-\begin{figure}[htbp]
-  \centering
-  \includegraphics[width=0.33\linewidth]{month-rescale01-col}
-  \caption{Adding colour to scaled plots. Here each location has been scaled to range $[0, 1]$ and colour mapped to the range of the original predictions.}
-  \label{fig:scaling-col}
-\end{figure}
+%\begin{figure}[htbp]
+%  \centering
+%  \includegraphics[width=0.33\linewidth]{month-rescale01-col}
+%  \caption{Adding colour to scaled plots. Here each location has been scaled to range $[0, 1]$ and colour mapped to the range of the original predictions.}
+%  \label{fig:scaling-col}
+%\end{figure}
 
-Other types of scaling may be interesting to explore, such as a classical scaling using the mean and standard deviation.
+%Other types of scaling may be interesting to explore, such as a classical scaling using the mean and standard deviation.
 
-Equivalent to using individual scales in facetted graphics.
+%Equivalent to using individual scales in facetted graphics.
+
+One more important point: if a model summary is displayed in the glyph-map, such as the predicted values, it may be important to force the scale to that of the original data.
+Displaying the predicted values in the absence of the raw data scale can skew the interpretation of the magnitude of the effect.
 
@@
 
-*** Connection to small multiples, facetted plots
-
 \section{Regular vs irregular locations}~\label{sec:irregular}

### 2011-09-23 paper.tex
adding some irregular data text

@@
 \subsection{Irregular locations}
+Figure \ref{fig:irregular} shows linear trends for monthly average surface temperature at stations in the USHCN (Version 2) ground station network.
+It illustrates three challenges in working with irregular locations: there is no natural choice for icon size, icons can potentially overlap, and comparisons rely more heavily on the reference guides.
+The icon sizes were chosen manually for these two plots, although the icons are displayed at about the same size, they correspond to XX degrees and YY degrees square respectively in the geographical coordinates.
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=0.4\linewidth]{ghcn-mountains}%
+  \includegraphics[width=0.6\linewidth]{usa-lin-overlap}%
+  \caption{Glyph-maps of ground station temperature values, 2001-2010, for the mountain states and the entire United States.} % these aren't actually plotting the same thing.
+  \label{fig:irregular}
+\end{figure}
+The two panels illustrate one approach to the overlapping problem: zooming.
+There is an interplay between icon size (in geographical units), plotting area (in geographical units) and plot size (in display units).
+Icons need to be big enough to be readable on the display, but small enough to avoid overlapping.
+Readability can be maintained and overlapping reduced by decreasing the icon size and simultaneously increasing the display size or decreasing the plotting area.
+However, there are generally limitations in display size (or resolution) and the utility of plotting very small areas.
+
+An alternative approach is to combine overlapping locations and plot a single summary icon.
+In Figure \ref{fig:irregular-collapsed} nearby locations are collapsed to a single icon.
+In the left panel each station still has an individual glpyh but locations close to each other share a plotting area.
+Not all glyphs are suited to this kind of space sharing, another option is to perform a summary of the locations that are collapsed and display that instead of the raw data.
+The right panel shows an example of this where the average trend for the collapsed locations is displayed.
+It is helpful to differentiate glyphs that involve summaries of more than one location, here a heavier line indicates at least two locations have been averaged.
 
@@
   \centering
-  \includegraphics[width=0.8\linewidth]{ghcn-mountains}%
-  \caption{Glyph-maps of ground station temperature values, 2001-2010, for mountain states.}
-  \label{fig:irregular}
+  \includegraphics[width=0.5\linewidth]{usa-season-collapsed}%
+  \includegraphics[width=0.5\linewidth]{usa-lin-collapse}%
+  \caption{Glyph-maps of ground station temperature values, with nearby locations collapsed.1950-2010.}
+  \label{fig:irregular-collapsed}
 \end{figure}
+
+In both examples of collapsing locations, the icon locations still represent real geographical locations of data, but contain data from up to an icon width away.
+An alternative way to combine stations is to round locations to the icon size.
+This produces a regular grid with the advantage of easier perception of structure.
+However, it hides the irregular nature of the locations and icons may end up centered over nonsense places (i.e.
+land measurements in the ocean).
+
+Reference guides become especially important in the irregular location case.
+Comparisons are harder in this case, as the ability to easily compare icons along a common baseline, be it along a row or column of glyphs, is absent.
+Without reference it is hard to tell if short series are missing data at the start or end.
 

### 2011-09-23 code/ghcn.r

@@ -50,7 +50,7 @@ year_preds <- ldply(models, function(mod) {
   year_grid$pred <- predict(mod, newdata = year_grid)
   year_grid
 })
-mountains.gly<-glyphs(year_preds, "lon", "year", "lat", "pred", width=1, height=1)
+mountains.gly<-glyphs(year_preds, "lon", "year", "lat", "pred", width=1, height=1, scale_y = mean0)
 
 ggplot(mountains.gly, aes(gx, gy, group = gid)) + 
   #map +

### 2011-09-23 code/usa.r

@@ -7,6 +7,27 @@ source("glyph.r")
 source("glyph-utils.r")
 source("overlapping.r")
 
+states <- map_data("state")
+world <- map_data("world")
+states$group <- max(world$group) + states$group
+both <- rbind(world, states)
+both <- getbox(both, xlim = c(-126, -55.07), ylim = c(24, 50))
+
+both <- ddply(both, "group", function(df) {
+  if (diff(range(df$long)) < 1e-6) return(NULL)
+  if (diff(range(df$lat)) < 1e-6) return(NULL)
+  df
+})
+
+us.map <- list(
+  geom_polygon(aes(long, lat, group = group), inherit.aes = FALSE, 
+    data = both, legend = FALSE, fill = "grey80", colour = "grey90"),
+  scale_x_continuous(breaks = NA, expand = c(0.02, 0)),
+  scale_y_continuous(breaks = NA, expand = c(0.02, 0)), 
+  xlab(NULL),
+  ylab(NULL))
+
+
 # All
 temp.all<-read.table("../data/9641C_201012_raw.avg")
 colnames(temp.all)<-c("year", "jan", "feb", "mar", "apr", "may", "jun", "jul", "aug", "sep", "oct", "nov", "dec", "ann")
@@ -74,10 +95,11 @@ lin.pred.gly <- glyphs(lin.pred.post1950, "lon", "year", "lat", "pred",
   width=w, height=h, y_scale = mean0)
 
 ggplot(lin.pred.gly, aes(gx, gy, group = gid)) + 
+  us.map + 
   add_ref_lines(lin.pred.gly) +
   geom_path() +
-  theme_fullframe()
-ggsave("../images/usa-lin-overlap.pdf", width = 8, height = 6)
+  theme_fullframe() 
+ggsave("../images/usa-lin-overlap.png", width = 5.5, height = 4)
 
 #== one solution combine overlapping glyphs
 stn.dists <- get.stn.dists(stn.all, "lon", "lat")
@@ -118,7 +140,7 @@ ggplot(lin.pred.sum, aes(gx, gy, group = gid)) +
   geom_path(aes(size = n > 1)) +
   theme_fullframe() +
   scale_size_manual(values = c("TRUE" = 1, "FALSE" = 0.25))
-ggsave("../images/usa-lin-collapse.pdf", width = 8, height = 6)
+ggsave("../images/usa-lin-collapse.png", width = 4, height = 3)
 
 
 

### 2011-09-23 paper.tex
finished first draft

@@
 Generalizations to different types of icons are described in Section \ref{sec:scatter}.
-Glyph-maps work nicely with gridded data, but many spaiottemporaral data sets have irregular spatial locations.
+Glyph-maps work nicely with gridded data, but many spatiotemporaral data sets have irregular spatial locations.
 Possible adjustments are discussed in Section \ref{sec:irregular}.
@@
 The structure perceived is different with this change in glyph type: locations at the equator have flat series, on land more varied temperatures.
-\citet{pickett:1988} introduced the idea of using icons on maps to display multivariate spatial data.
-This area of work has morphed into the field of metaphorical data displays, which creates abstract landscapes of spatial data, a digression from glyph-maps.
-\citet{gribov:2006} describes the use ogf glyph-maps for multivariate data also, with emphasis on the graphics software Gauguin.
 
@@
 *** Need to add reference circles.
+
+Icons have been displayed on maps since the earliest displays weather forecasts, in the 1800s.
+In recent times, \citet{pickett:1988} introduced the idea of using icons on maps to display multivariate spatial data.
+This area of work has morphed into the field of metaphorical data displays, which creates abstract landscapes of spatial data, a digression from glyph-maps.
+\citet{gribov:2006} describes the use of glyph-maps for multivariate data also, with emphasis on the graphics software Gauguin.
 
@@
 \begin{figure*}[htp]
-\centerline{\includegraphics[width=5in]{nasa-deseas-glyph.png}}
+\centerline{\includegraphics[width=4in]{nasa-deseas-glyph.png}}
 \caption{Glyph-map of de-seasonalized temperature, to compare with Figure \ref{fig:facetted-map}. }
@@
 
-Structure in temporal data decomposed into different parts, global trend, seasonality and error.
+Structure in temporal data is composed of different parts: global trend, seasonality and error.
 Extracting each of these for visualization improves the focus on different aspects of the time series.
-A generalised additive model \citep{wood:2006} with a seasonal term and a long-term smooth term.
+A generalized additive model \citep{wood:2006} with a seasonal term and a long-term smooth term can be useful.
+This is revealing for smooth models of surface temperature, shown in Figure~\ref{fig:scaling}.
+Each location is modelled using a generalized additive model \citep{wood:2006} with a seasonal term and a long-term smooth term.
 The three plots in the figure show the long-term smooth term.
-Because some locations are uniformly warmer than others, the impact of changes across time are diminished.
+%Because some locations are uniformly warmer than others, the impact of changes across time are diminished.
+In the center plot, each location is scaled by dividing out the maximum, and on the right, each location is scaled to range $[0, 1]$.
+These emphasize individual changes but lose the ability to make global comparisons, and need caution when interpreting - a small absolute change gets blown up into a large relative change.
+ %The three plots in the figure show the long-term smooth term. Because some locations are uniformly warmer than others, the impact of changes across time are diminished.
 
- is a extremely useful technique: product plots, ANOVA, ...
+% is a extremely useful technique: product plots, ANOVA, ...
 For glyph plots, it's often useful to partition the glyphs into shape and scale components, decomposing the values into a product of maximum and scaled values.
@@
 
-This is revealing for smooth models of surface temperature, shown in Figure~\ref{fig:scaling}.
-Each location is modelled using a generalised additive model \citep{wood:2006} with a seasonal term and a long-term smooth term.
-The three plots in the figure show the long-term smooth term.
-Because some locations are uniformly warmer than others, the impact of changes across time are diminished.
-In the centre plot, each location is scaled by dividing out the maximum, and on the right, each location is scaled to range $[0, 1]$.
-These emphasise individual changes but lose the ability to make global comparisons, and need caution when interpreting - a small absolute change gets blown up into a large relative change.
+*** Need three plots here: trend, seasonality and residuals, shall we do this with NASA ozone data or stick to temperature?
 
@@
 
-So far we have limited ourselves to always mapping some aspect of time on the x-axis.
-But glyph-maps are not limited to only this type of data and can  be easily generalized to create many other types of display.
-Figure~\ref{fig:cloud} shows a generalisation to scatterplots: by mapping another variable to the x-axis, in this case low-cloud level on x and high-cloud level on y, we can explore how a bivariate relationship varies spatially.
-It is again useful to use a model to extract the essence of the relationship, summarizing the data with models is very useful: the second plot displays a loess curve fit to the data instead of the individual points.
+Glyph-maps are not limited to time series plots.
+They are readily generalized to use any other type of display as an icon.
+Figure~\ref{fig:cloud} shows scatterplots as icons: low-cloud level is displayed horizontally and high-cloud level vertically.
+The bivariate relationship between the two types of clouds is explored spatially, and it is clear that it is different across the region.
+The second plot displays a loess curve fit to the data instead of the individual points, sharpening the view of association: most locations have a negative or zero association.
+In the Gauguin software \citep{gribov:2006} one alternative is an icon based on a histogram, small univariate distribution displays for each glyph.
+
+*** Need to use local scaling here.
 
@@
 
-In the Gauguin software one alternative was an icon based on a histogram.
-
 \section{Regular vs irregular locations}~\label{sec:irregular}
@@
 
-Rectangular, but not in this projection.  Other types of regular grids.
+Glyph-maps work best in mercator projection, where the icons are organized into rectangular microplots.
+Spatial coordinates are often not given in this coordinate system, and so may require a transformation into mercator projection coordinates.
 
@@
 
+A lot of climate data records contain long temporal history.
+For temperature records often extend from the 1800s to present day.
+This places some difficulties, one being computational because of the volume of records, and the second being visual display of a lot of data.
+Plotting the full raw time series in miniature for the glyph-maps often reveals little more than an irrelavant observation that ``temperature is quite varied!'' Careful use of de-seasonalizing, and models that can reflect nuances in global trend are needed to improve the visualization.
+And efficient computation to calculate and store predicted values will be required.
+
 \section{Conclusions}

### 2011-09-23 paper.tex
tweak

@@
 Shifting the values at each location to have mean zero, places more focus on the trend.
-This is particularly important for the star glyphs, where differences in the average value lead to unintuitive patterns in the resulting glyphs.
+This is particularly important for the star glyphs, where differences in the average value lead to unintuitive patterns in the resulting glyphs (eg Figure \ref{fig:cycle} star glyphs obscure the seasonal pattern).
 In this situation we would use $y_{st}-\bar{y}_s$ instead of $y_{st}$ in equations \ref{scale1}, \ref{scale2}, with minima and maxima calculated on the mean-shifted data.

### 2011-09-23 paper.tex
fixes to irregular data plots

@@
 It illustrates three challenges in working with irregular locations: there is no natural choice for icon size, icons can potentially overlap, and comparisons rely more heavily on the reference guides.
-The icon sizes were chosen manually for these two plots, although the icons are displayed at about the same size, they correspond to XX degrees and YY degrees square respectively in the geographical coordinates.
+The icon sizes were chosen manually for these two plots, although the icons are displayed at about the same size, they correspond to 1 degrees and 0.5 degrees square respectively in the geographical coordinates.
 \begin{figure}[htbp]
   \centering
-  \includegraphics[width=0.4\linewidth]{ghcn-mountains}%
-  \includegraphics[width=0.6\linewidth]{usa-lin-overlap}%
-  \caption{Glyph-maps of ground station temperature values, 2001-2010, for the mountain states and the entire United States.} % these aren't actually plotting the same thing.
+  \includegraphics[width=0.55\linewidth]{usa-lin-overlap}%
+  \includegraphics[width=0.35\linewidth]{ghcn-mountains}%
+  \caption{Glyph-maps of the linear trend in ground station temperature values, 1950-2010, of all continental US USHCN stations (left) and stations in the mountain states (right).} 
   \label{fig:irregular}
@@
 Icons need to be big enough to be readable on the display, but small enough to avoid overlapping.
-Readability can be maintained and overlapping reduced by decreasing the icon size and simultaneously increasing the display size or decreasing the plotting area.
-However, there are generally limitations in display size (or resolution) and the utility of plotting very small areas.
+Readability can be maintained and overlapping reduced by decreasing the icon size and simultaneously increasing the display size or decreasing the plotting area (as done in the right panel in Figure \ref{fig:irregular}).
+However, there are generally limitations in the maximum display size (or resolution) and the utility of plotting very small areas.
 

### 2011-09-23 code/usa.r

@@ -11,7 +11,7 @@ states <- map_data("state")
 world <- map_data("world")
 states$group <- max(world$group) + states$group
 both <- rbind(world, states)
-both <- getbox(both, xlim = c(-126, -55.07), ylim = c(24, 50))
+both <- getbox(both, xlim = c(-126, -65), ylim = c(24, 50))
 
 both <- ddply(both, "group", function(df) {
   if (diff(range(df$long)) < 1e-6) return(NULL)
@@ -21,7 +21,7 @@ both <- ddply(both, "group", function(df) {
 
 us.map <- list(
   geom_polygon(aes(long, lat, group = group), inherit.aes = FALSE, 
-    data = both, legend = FALSE, fill = "grey80", colour = "grey90"),
+    data = subset(both, region != "Great Lakes"), legend = FALSE, fill = "grey80", colour = "grey90"),
   scale_x_continuous(breaks = NA, expand = c(0.02, 0)),
   scale_y_continuous(breaks = NA, expand = c(0.02, 0)), 
   xlab(NULL),
@@ -101,6 +101,24 @@ ggplot(lin.pred.gly, aes(gx, gy, group = gid)) +
   theme_fullframe() 
 ggsave("../images/usa-lin-overlap.png", width = 5.5, height = 4)
 
+# Mountain states: Colorado, Wyoming, Idaho, Montana, Nevada, Utah
+mountain.states <- c("CO", "WY", "ID", "MT", "NV", "UT")
+mountains.gly<-glyphs(subset(lin.pred.post1950, state %in% mountain.states),
+  "lon", "year", "lat", "pred", width=.5, height=.5, y_scale = mean0)
+
+ggplot(mountains.gly, aes(gx, gy, group = gid)) + 
+  geom_polygon(aes(long, lat, group = group), inherit.aes = FALSE, 
+    data = both, legend = FALSE, fill = "grey80", colour = "grey90") +
+  scale_x_continuous(breaks = NA, expand = c(0.02, 0)) +
+  scale_y_continuous(breaks = NA, expand = c(0.02, 0)) +
+  xlab(NULL) +
+  ylab(NULL) +
+  add_ref_lines(mountains.gly) +
+  geom_path() +
+  theme_fullframe() +
+  coord_cartesian(xlim = c(-121, -101), ylim = c(34, 50)) 
[…]

### 2011-09-23 paper.tex
cleaned up the generalization seciton some more, different examples

--- 2011-09-23 paper.tex
@@
 They are readily generalized to use any other type of display as an icon.
-Figure~\ref{fig:cloud} shows scatterplots as icons: low-cloud level is displayed horizontally and high-cloud level vertically.
-The bivariate relationship between the two types of clouds is explored spatially, and it is clear that it is different across the region.
-The second plot displays a loess curve fit to the data instead of the individual points, sharpening the view of association: most locations have a negative or zero association.
+Figure~\ref{fig:cloud} shows scatterplots as icons: temperature values are displayed horizontally and high-cloud values are displayed vertically, both locally scaled.
+The bivariate relationship between the two variables is explored spatially, and it is clear that it is different across the region.
+The equatorial Pacific and Caribbean have positve association between temperature and high cloud, while the south American continent has negative association.
+The second plot displays a loess curve fit to the data instead of the individual points, sharpening the view of association.
 In the Gauguin software \citep{gribov:2006} one alternative is an icon based on a histogram, small univariate distribution displays for each glyph.
-
-*** Need to use local scaling here.
 
@@
   \centering
-  \includegraphics[width=0.5\linewidth]{clouds}%
-  \includegraphics[width=0.5\linewidth]{clouds-smooth}
+  \includegraphics[width=0.5\linewidth]{nasa-scat-glyph}%
+  \includegraphics[width=0.5\linewidth]{nasa-loess-glyph}
 
-  \caption{Glyph-map showing (left) scatterplot of high-cloud vs.
-low-cloud and (right) smoothed (loess) curve fit to each location.
-The relationship between these two variables varies considerably over the spatial domain (particularly between land and ocean), but neighbouring locations tend to similar.
+  \caption{Glyph-map showing (left) scatterplot of temperature vs high-cloud and (right) smoothed (loess) curve fit to each location.
+The relationship between these two variables varies considerably over the spatial domain.
 }

### 2011-09-23 code/glyph-examples.r

@@ -23,8 +23,32 @@ map <- list(
 # to run that code before this
 resids.gly <- glyphs(resids, "long", "day", "lat", "temp_resid", height=2.5) 
 ggplot(resids.gly, aes(gx, gy, group = gid)) + map +
-  add_ref_lines(resids.gly) +
-  add_ref_boxes(resids.gly) +
+  ref_lines(resids.gly) +
+  ref_boxes(resids.gly) +
   geom_path() + theme_fullframe() + coord_map()
 #qplot(gx, gy, data = resids, geom = "line", group = gid) + map
 ggsave("../images/nasa-deseas-glyph.png")
+
+# Scatterplot
+nasa.scale<-ddply(nasa, c("long","lat"), summarise,
+                  cloudlow.s=scale(cloudlow), 
+                  cloudhigh.s=scale(cloudhigh),
+                  temp.s=scale(surftemp))
+#nasa.scat.gly <- glyphs(nasa.scale, "long", "cloudlow.s", "lat", "cloudhigh.s") 
+#ggplot(nasa.scat.gly, aes(gx, gy, group = gid)) + map +
+#  add_ref_boxes(nasa.scat.gly) +
+#  geom_point(size=I(0.7), alpha=I(0.5)) + theme_fullframe() + coord_map()
+#ggsave("../images/nasa-scat-glyph.png")
+nasa.scat.gly <- glyphs(nasa.scale, "long", "temp.s", "lat", "cloudhigh.s") 
+ggplot(nasa.scat.gly, aes(gx, gy, group = gid)) + map +
+  add_ref_boxes(nasa.scat.gly) +
+  geom_point(size=I(0.7), alpha=I(0.5)) + theme_fullframe() + coord_map()
+ggsave("../images/nasa-scat-glyph.png")
+nasa.loess<-ddply(nasa.scale, c("long","lat"), summarise,
+                  temp.s=temp.s,
+                  pcloud=predict(loess(cloudhigh.s~temp.s)))
+nasa.loess.gly <- glyphs(nasa.loess, "long", "temp.s", "lat", "pcloud") 
+ggplot(nasa.loess.gly, aes(gx, gy, group = gid)) + map +
+  add_ref_boxes(nasa.loess.gly) +
+  geom_line() + theme_fullframe() + coord_map()
+ggsave("../images/nasa-loess-glyph.png")

### 2011-09-23 paper.tex
add gridded examples

@@
 
-Glyph-maps work best in mercator projection, where the icons are organized into rectangular microplots.
-Spatial coordinates are often not given in this coordinate system, and so may require a transformation into mercator projection coordinates.
+Glyph-maps work best when displayed in the coordinate system in which the grid is rectangular and regular.
+For example, the GISTEMP grid is 2$^o$ square and is displayed treating longitude and latitude as cartesian coordinates in Figure \ref{fig:gistemp-pred}.
+For regularly gridded data, this results in icons that are rectangles of equal size.
+It is often preferable to display maps as a projection of the geographical coordinates which preserves a property of interest, area for example.
+When a projection is desired a choice needs to be made between applying the projection to the completed glyph-map, or to just the spatial locations.
+When the completed glyph-map is projected the icons are no longer rectangles of the same size but they do retain their space filling property.
+When applying a projection to the spatial locations and then constructing a glyph-map, the grid becomes irregular.
+
+*** Example?
 
@@
 It illustrates three challenges in working with irregular locations: there is no natural choice for icon size, icons can potentially overlap, and comparisons rely more heavily on the reference guides.
-The icon sizes were chosen manually for these two plots, although the icons are displayed at about the same size, they correspond to 1 degrees and 0.5 degrees square respectively in the geographical coordinates.
+The icon sizes were chosen manually for these two plots, although the icons are displayed at about the same size, they correspond to 1$^o$ and 0.5$^o$ square respectively in the geographical coordinates.
 \begin{figure}[htbp]
@@
 In both examples of collapsing locations, the icon locations still represent real geographical locations of data, but contain data from up to an icon width away.
-An alternative way to combine stations is to round locations to the icon size.
+An alternative way to combine stations is to round locations to the icon size, Figure \ref{fig:irregular-grid}.
 This produces a regular grid with the advantage of easier perception of structure.
@@
 land measurements in the ocean).
+
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=0.5\linewidth]{usa-season-grid}%
+  \includegraphics[width=0.5\linewidth]{usa-lin-grid}%
+  \caption{Glyph-maps of ground station temperature values, with locations rounded to 1$^o$. 1950-2010.}
+  \label{fig:irregular-grid}
+\end{figure}
 

### 2011-09-23 code/usa.r

@@ -166,13 +166,21 @@ ggsave("../images/usa-lin-collapse.png", width = 6, height = 4)
 
 #== gridding
 lin.pred.grid <- grid.all(lin.pred.post1950, "lon", "lat", w, h)
-glyph.lin.grid <- glyphs(lin.pred.grid, "grid.x", "year", "grid.y",   
-  "pred", width = 0.95 * w, height = 0.95 * h, y_scale = mean0)
 
-ggplot(glyph.lin.grid) +
-  add_ref_lines(glyph.lin.grid) +
-  geom_line(aes(gx, gy, group = stn)) +
-  theme_fullframe() 
+lin.pred.grid.sum  <- ddply(lin.pred.grid, c("year", "grid.x", "grid.y"),
+  summarise, pred = mean(pred, na.rm = TRUE), n = length(pred), 
+  grid.x = grid.x[1], grid.y = grid.y[1])
+
+lin.pred.grid.gly <- glyphs(lin.pred.grid.sum, "grid.x", "year", "grid.y",
+  "pred",  width = w,  height = h, y_scale = mean0)
+
+ggplot(lin.pred.grid.gly,aes(gx, gy, group = gid)) +
+  us.map + 
+  add_ref_lines(lin.pred.grid.gly) +
+  geom_path(aes(size = n > 1)) +
+  theme_fullframe() +
+  scale_size_manual(values = c("TRUE" = 1, "FALSE" = 0.25))
+ggsave("../images/usa-lin-grid.png", width = 6, height = 4)
 
 #=== SEASONAL ===#
 
@@ -218,11 +226,15 @@ ggsave("../images/usa-season-collapsed.png", width = 6, height = 4)
 
 #== gridding
 seas.pred.grid <- grid.all(seas.pred.post1950, "lon", "lat", w, h)
+seas.pred.grid <- ddply(seas.pred.grid, c("grid.x", "grid.y"), mutate,
+  avg = mean(unique(avg)))
+
 glyph.seasonal.grid <- glyphs(seas.pred.grid, "grid.x", "time", "grid.y",   
   "pred", width = 0.95 * w, height = 0.95 * h, y_scale = mean0)
 
 ggplot(glyph.seasonal.grid) +
   us.map + 
-  add_ref_boxes(glyph.seasonal.grid) +
+  add_ref_boxes(glyph.seasonal.grid, "avg", alpha = 0.2, colour = NA) +
   geom_line(aes(gx, gy, group = stn)) +
[…]

### 2011-09-23 paper.tex
caption edits

@@
   \includegraphics[width=0.35\linewidth]{ghcn-mountains}%
-  \caption{Glyph-maps of the linear trend in ground station temperature values, 1950-2010, of all continental US USHCN stations (left) and stations in the mountain states (right).} 
+  \caption{Glyph-maps of the linear trend in ground station temperature values, 1950--2010, for all continental US USHCN stations (left) and stations in the mountain states (right).} 
   \label{fig:irregular}
@@
   \includegraphics[width=0.5\linewidth]{usa-lin-collapse}%
-  \caption{Glyph-maps of ground station temperature values, with nearby locations collapsed.1950-2010.}
+  \caption{Glyph-maps of seasonal patterns (left) and linear trends (right) of USHCN stations, 1950--2010.
+Nearby stations whose icons would overlap have been collapsed.
+Each location has a glpyh in the seasonal plot and colour of the tiles corresponds to the average temperature of the locations (red=warm, blue=cool).
+Heavy lines in the linear trend plot correspond to average trends over more than one location.}
   \label{fig:irregular-collapsed}
@@
   \includegraphics[width=0.5\linewidth]{usa-lin-grid}%
-  \caption{Glyph-maps of ground station temperature values, with locations rounded to 1$^o$. 1950-2010.}
+  \caption{Glyph-maps of seasonal patterns (left) and linear trends (right) of USHCN stations, 1950--2010.
+Station locations have been rounded to the nearest degree Each location has a glpyh in the seasonal plot and colour of the tiles corresponds to the average temperature of the locations (red=warm, blue=cool).
+Heavy lines in the linear trend plot correspond to average trends over more than one location.}
   \label{fig:irregular-grid}
@@
 
+USHCN data provided by NOAA NCDC from their website \url{http://cdiac.ornl.gov/ftp/ushcn_v2_monthly/}.
+
 The graphics in this paper were created in R \citep{R} using {\tt ggplot2} \citep{me:ggplot2} and {\tt plyr} \citep{me:plyr}.

### 2011-09-24 paper.tex
fixed fig reference
spell check, including NZ to US conventions

@@
 \caption{Facetted map of de-seasonalized temperature.
-The dominant feature is the El Ni\~no warming in the qouthern equatorial region in the last half of 1997 and first half of 1998.
+The dominant feature is the El Ni\~no warming in the southern equatorial region in the last half of 1997 and first half of 1998.
 Smaller features are only noticeable on closer inspection, or if pointed out: such as the relative warming on the mountain regions in south and north America in the later years.}
@@
 For temporal data, time measurements can be treated as variables in a multivariate data set, fitting this into the icon plot framework.
-Figure \ref{templates} displays 12 iconic time series shapes, and how they would look if the icon was a time series, or displayed as a star glyph.
+Figure \ref{fig:templates} displays 12 iconic time series shapes, and how they would look if the icon was a time series, or displayed as a star glyph.
 The data underlying each glyph has 36 time variables.
@@
 
-In this paper we describe the algorithm for creating glyphs of temporal data on maps, generally in R, and the various modifcations that can enable different features in the data to be detected.
+In this paper we describe the algorithm for creating glyphs of temporal data on maps, generally in R, and the various modifications that can enable different features in the data to be detected.
 The next section (\ref{sec:glyph-map}) describes the use of glyphs on maps, and Section \ref{sec:construction} explains how to construct glyph-maps.
@@
 Generalizations to different types of icons are described in Section \ref{sec:scatter}.
-Glyph-maps work nicely with gridded data, but many spatiotemporaral data sets have irregular spatial locations.
+Glyph-maps work nicely with gridded data, but many spatiotemporal data sets have irregular spatial locations.
 Possible adjustments are discussed in Section \ref{sec:irregular}.
@@
     with varying degrees of increase. Some isolated locations exhibit
-    steep declines, and inclines, perpahs indicative of some isolated
+    steep declines, and inclines, perhaps indicative of some isolated
     data problems.}
@@
 Some locations, particularly in the north western mountain region, have big variability in temperature, both low minima, and high maxima.
-At these locaations there are large fluctuations from the long term mean.
+At these locations there are large fluctuations from the long term mean.
 
@@
 The glyph-map provides a texturing over the map.
-The El Ni\~no event is still the primary visble structure, the region that has a bump in the middle of the time series, correspondingto the hot 1997-1998 period.
+The El Ni\~no event is still the primary visible structure, the region that has a bump in the middle of the time series, corresponding to the hot 1997-1998 period.
 We can also see other temporal features: that the temperatures are generally more varied over land, there are a few locations with dramatically increasing temperature in the later years typically in locations that are the high elevation areas, along the Andes, and Sierra Madre Occiddental in Mexico.
@@
 In contrast, in the small multiples used in the facetted colored map, the viewer is expected to compare the patterns in the maps from one year to the next, for corresponding months.
-(See \citet{carr:1999} for a discussion on perceptual grouping.) That is, compare and contrast maps with factted colred map, or compare and contrast time series using the glyph-map.
+(See \citet{carr:1999} for a discussion on perceptual grouping.) That is, compare and contrast maps with facetted colored map, or compare and contrast time series using the glyph-map.
 One draw-back of the glyph-maps, particularly if linear trend is displayed as an icon, is they may suffer from the Z\"ollner Illusion \citep{Zollner} which is visible in Figure \ref{fig:gistemp-pred}.
@@
 Scaling of the data, into the values used for each glyph, emphasizes different aspects of the data.
-For spatiotemporal data the most natural scale is to use the global minumum and maximum, because the temporal variables are measuring the same information, and we want to make comparisons across the spatial region.
+For spatiotemporal data the most natural scale is to use the global minimum and maximum, because the temporal variables are measuring the same information, and we want to make comparisons across the spatial region.
 The icons will have size relative to the magnitude of the full range of the data.
@@
 With the local scaling, the shape of the icon, for example, differences in the seasonal patterns over spatial locations, are emphasized.
-Fig \ref{fig:scaling} shows glyph-maps generated usng both global scaling (left) and local scaling (right) for the NASA temperature data.
-On the global scale, the El Ni\~no blip in temperature is primarily visible, along with the generally flat series over most water locations, and somewhat increasing or variable temerpatures over land.
-The local scale emhpasizes the individual shapes: The El Ni\~no pattern is visible, but extend to a larger spatial region including the Caribbean, dramatic linear increases across the Andes, decreases in the Gulf of Mexico.
+Fig \ref{fig:scaling} shows glyph-maps generated using both global scaling (left) and local scaling (right) for the NASA temperature data.
+On the global scale, the El Ni\~no blip in temperature is primarily visible, along with the generally flat series over most water locations, and somewhat increasing or variable temperatures over land.
+The local scale emphasizes the individual shapes: The El Ni\~no pattern is visible, but extend to a larger spatial region including the Caribbean, dramatic linear increases across the Andes, decreases in the Gulf of Mexico.
 (The local scale is equivalent to using individual scales in the facetted graphics of \texttt{ggplot2}).
@@
 This is revealing for smooth models of surface temperature, shown in Figure~\ref{fig:scaling}.
-Each location is modelled using a generalized additive model \citep{wood:2006} with a seasonal term and a long-term smooth term.
+Each location is modeled using a generalized additive model \citep{wood:2006} with a seasonal term and a long-term smooth term.
 The three plots in the figure show the long-term smooth term.
@@
 Issues of scale relative to modeling.
-Switching from the data to model summaries bringsnew scaling issues.
+Switching from the data to model summaries brings new scaling issues.
 Often the scale of the raw data is ignored when the model is plotted, one usually shows just the line showing the trend, in the range of the trend.
-This has the effect of exaggerating the signficance of the model.
-Scaling the trend lines rlative to the range of the data may be useful sometimes.
+This has the effect of exaggerating the significance of the model.
+Scaling the trend lines relative to the range of the data may be useful sometimes.
 
@@
 The bivariate relationship between the two variables is explored spatially, and it is clear that it is different across the region.
-The equatorial Pacific and Caribbean have positve association between temperature and high cloud, while the south American continent has negative association.
+The equatorial Pacific and Caribbean have positive association between temperature and high cloud, while the south American continent has negative association.
 The second plot displays a loess curve fit to the data instead of the individual points, sharpening the view of association.
@@
 In Figure \ref{fig:irregular-collapsed} nearby locations are collapsed to a single icon.
-In the left panel each station still has an individual glpyh but locations close to each other share a plotting area.
+In the left panel each station still has an individual glyph but locations close to each other share a plotting area.
 Not all glyphs are suited to this kind of space sharing, another option is to perform a summary of the locations that are collapsed and display that instead of the raw data.
@@
 Nearby stations whose icons would overlap have been collapsed.
-Each location has a glpyh in the seasonal plot and colour of the tiles corresponds to the average temperature of the locations (red=warm, blue=cool).
+Each location has a glyph in the seasonal plot and color of the tiles corresponds to the average temperature of the locations (red=warm, blue=cool).
 Heavy lines in the linear trend plot correspond to average trends over more than one location.}
@@
   \caption{Glyph-maps of seasonal patterns (left) and linear trends (right) of USHCN stations, 1950--2010.
-Station locations have been rounded to the nearest degree Each location has a glpyh in the seasonal plot and colour of the tiles corresponds to the average temperature of the locations (red=warm, blue=cool).
+Station locations have been rounded to the nearest degree Each location has a glyph in the seasonal plot and color of the tiles corresponds to the average temperature of the locations (red=warm, blue=cool).
 Heavy lines in the linear trend plot correspond to average trends over more than one location.}
@@
 This places some difficulties, one being computational because of the volume of records, and the second being visual display of a lot of data.
-Plotting the full raw time series in miniature for the glyph-maps often reveals little more than an irrelavant observation that ``temperature is quite varied!'' Careful use of de-seasonalizing, and models that can reflect nuances in global trend are needed to improve the visualization.
+Plotting the full raw time series in miniature for the glyph-maps often reveals little more than an irrelevant observation that ``temperature is quite varied!'' Careful use of de-seasonalizing, and models that can reflect nuances in global trend are needed to improve the visualization.
 And efficient computation to calculate and store predicted values will be required.

### 2011-09-27 paper.tex
Changes to first two sections

@@
 Climate data is composed of measurements on variables such as temperature, precipitation, and winds that have been collected from a spatial and temporal context.
-The classic display for spatial data is to use color or grey scale overlaid on geographic location, typically including a map.
-When measurements are made at different time periods it is common to display the data as small multiples, such as in Figure \ref{fig:facetted-map}.
-In this plot, a separate map is drawn for each month (columns) and year (rows), over a 6 year period of remotely sensed temperature data above Central America.
+The classic display of this type of is to use color or grey scale overlaid on geographic location, typically including a map.
+When measurements are made at different time periods it is common to display the data as small multiples, such as in Figure~\ref{fig:facetted-map}.
+In this plot, a separate map is drawn for each month (columns) and year (rows), over a 6 year period of remotely sensed temperature data above Central America \citet{murrell:2010}.
 Color is used to display the de-seasonalized temperature measurements, with red being high values and blue being low values.
-The data comes from \citet{murrell:2010}.
 The most noticeable feature in this display is the strong red in the equatorial region beginning in mid-1997 and tapering out during 1998.
 This is the pattern of an El Ni\~no event, a major temperature anomaly.
-With some work other smaller more localized patterns can be perceived, for example, cooler over the land mass in the early year, but warmer in the latter years.
-What the reader is required to do is play the game ``Spot the difference'' from one small image to another.
-From a cognitive perspective, this is a very hard task to do.
-It falls into the framework of {\it change blindness} (see e.g.
-\citet{healey:2011} and often leads to failure without readers realizing, in particular for small areas of change or subtle changes.
-Large structures such as the El Ni\~no event are clear but readers will not be able to effectively glue the images together sufficiently to read off the trend over the years or notice small local deviations between plots.
-Rendering these small multiples as movie helps with picking up strong trends, but small trends still fail to draw readers' attention and escape unnoticed \citep{simons:gradual}.
+With some work other more localized patterns can be perceived, for example, cooler over the land mass in the early year, but warmer in the latter years.
 
-\begin{figure*}[htp]
-\centerline{\includegraphics[width=6in]{nasa-colored-map.png}}
+Reading this plot is cognitively challenging: the reader must play ``spot the difference'' from one small image to another.
+Large structures such as the El Ni\~no event are clear but it is very difficult to glue the images together to read off the trend over the years or to notice local deviations.
+Rendering these small multiples as a movie can help with picking up strong trends, but small trends still fail to draw readers' attention and escape unnoticed \citep{simons:gradual}.
+
+\begin{figure*}[htbp]
+  \centering
+  \includegraphics[width=5.5in]{nasa-colored-map.png}
   \caption{Facetted map of de-seasonalized temperature.
@@
 
-In this paper we explore the use of a different type of display, that uses icons on the map, for climate data.
-Icon plot associated with weather maps date back as far as \citet{galton:weathermap}, described in \citet{friendlydenis:2001}.
-Icon plots date back to the star plot introduced by \citet{mayr:1877}, and resurged in the modern era with the semi-graphic displays of \citet{anderson:1960}, and were made infamous by \citep{chernoff:1973}.
-An icon is produced by mapping each variable to some feature of the icon.
-One icon is generated for every case in the data set.
-%Figure \ref{fig:glyphs} illustrates the construction of star plot, with emphasis on temporal data.
-For temporal data, time measurements can be treated as variables in a multivariate data set, fitting this into the icon plot framework.
-Figure \ref{fig:templates} displays 12 iconic time series shapes, and how they would look if the icon was a time series, or displayed as a star glyph.
+In this paper we develop a different type of display that resolves these problems: the \emph{glyph-map}.
+The glyph-map uses a small glyph, or icon, to represent multiple values at each location.
+Glyph maps are an adaptation of glyphs, tools developed for display multivariate data, including the star plots of \citet{mayr:1877}, the semi-graphic displays of \citet{anderson:1960}, and the infamous Chernoff faces \citep{chernoff:1973}.
+An glyph is produced by mapping each variable to some graphical feature.
+Figure~\ref{fig:nasa-deseas-glyph} shows the glyph-map equivalent of Figure~\ref{fig:facetted-map}: each location is represented by a tiny time series glyph.
+The glyph-map can also be thought of as a small multiple display \citet{tufte:2001} of time series.
+
+\begin{figure*}[htbp]
+  \centering
+  \includegraphics[width=3.5in]{nasa-deseas-glyph.png}
+  \caption{Glyph-map of de-seasonalized temperature, to compare with Figure \ref{fig:facetted-map}. }
+  \label{fig:nasa-deseas-glyph}
+\end{figure*}
+
+One problem with glyph displays of multivariate data is that there is no natural ordering of each glyph, or of the mapping between variables and glyph properties.
+This leads to a combinatorial explosion of possibilities which may have dramatically different perceptual properties.
+Despite some proposed remedies \citep{kleiner:1981,hurley:2010}, this problem has prevented the widespread adoption of glyph displays.
+Using glyphs for space-time data resolves this problem: we place glyphs according to the measurement location of the data, and we order variables in time.
+Additionally, climate data usually has high-correlations between nearby locations and times.
+This imposes a certain degree of smoothness that gives the plot an appearance of texturing on a landscape, making it easier to digest patterns.
+Others \citet{pickett:1988} have used glyphs on maps to display multivariate spatial data.
+This area of work has morphed into the field of metaphorical data displays, which creates abstract landscapes of spatial data, a digression from glyph-maps.
+\citet{gribov:2006} describes the use of glyph-maps for multivariate data also, with emphasis on the graphics software Gauguin.
+
+We focus our attention on two glyphs: lines and stars.
+Figure~\ref{fig:templates} displays 12 iconic time series shapes with line and star glyphs.
 The data underlying each glyph has 36 time variables.
 The star glyphs is formed by considering the 36 axes radiating from a common midpoint, and the data values for the row are plotted on each axis relative to the locations of the minimum and maximum of the variable.
-For the temporal data, all variables are considered to be on a common scale so the global minimum and maximum of temperature are used to scale each data value.
-%Three different locations are selected, for demonstration purposes.
-The top plot shows the time series of these three locations.
-Location ``1.1'' (red) has medium values of temperature, with strong seasonality, the six years corresponding to the six peaks.
-Location ``2.1'' (green) has generally high temperature values, with weaker seasonality in most years and a lack of low temperatures in the middle of the time period when lows would be expected.
-Location ``3.1'' (blue) has strong seasonality but generally low temperatures.
-The bottom plot shows these same cases as star glyphs.
-A separate glyph is created for each location.
-The 72 time variables form axes organized radially.
-Each temperature value is plotted relative to the overall range of temperature values.
-The seasonal patterns are clearly visible in this type of display, forming six ``petals'' in the icon.
-Location ``2.1'' has a different pattern than the other two, with only four petals, and location ``3.1'' is a much smaller icon indicating overall lower values.
 
@@
 
-%\begin{figure*}[htp]
-%\centerline{\includegraphics[width=6in]{nasa-glyph-ts.pdf}}
-%\centerline{\includegraphics[width=6in]{nasa-glyph.pdf}}
-%\caption{Icons of raw temperature for three different locations.}
-%\label{fig:glyphs}
-%\end{figure*}
+%Simple, but effective, technique that combines aspects of glyphs, facetted plots and map-charts. Particularly well suited for the type of temporal-spatial data, commonly seen in climate.
 
-Icon plots are used, infrequently, to examine multivariate data.
-For large data sets it is infeasible to digest the quantity of icons, and for multivariate data there is no natural order so it can be like having a mountain of little stamps to rearrange and organize, in order to understand the structures present in the data.
-An early paper on multivariate graphics \citep{kleiner:1981} and a recent paper \citep{hurley:2010} propose some approaches for organizing the icons.
-Ordering of icons is not an issue with spatiotemporal data, because it is naturally handled by spatial location.
-The spatial trend, or continuity, usually gives the plot an appearance of texturing on a landscape, make it easier to digest patterns among the icons.
+%Glyph-maps have attributes of both glyphs (stars, faces, profiles, ...), inviting exploration at the global level, and small-multiple time series, allowing local exploration of individual series.
+Allow to see both geographic context and individual locations simultaneously.
 
-In this paper we describe the algorithm for creating glyphs of temporal data on maps, generally in R, and the various modifications that can enable different features in the data to be detected.
-The next section (\ref{sec:glyph-map}) describes the use of glyphs on maps, and Section \ref{sec:construction} explains how to construct glyph-maps.
-Section \ref{sec:reference} discusses the use of reference lines and boxes, and Section \ref{sec:scale} visits issues of scaling values for different information purposes.
-Generalizations to different types of icons are described in Section \ref{sec:scatter}.
+%These displays tend to be particularly effective when printed in large format, as the much higher resolution of print (600 dpi) compared to computer  screen (72-120 dpi) allows for much finer reading.
+These are high-density data displays in the best tradition of Tufte.
+Wall size maps are not usually practical during analysis, but are very engaging.
+
+In this paper we describe the algorithm for creating glyphs of temporal data on maps, and the various modifications that can enable different features in the data to be detected.
+Section~\ref{sec:glyph-map} describes the use of glyphs on maps, and Section~\ref{sec:construction} explains how to construct glyph-maps.
+Section~\ref{sec:reference} discusses the use of reference lines and boxes, and Section~\ref{sec:scale} visits issues of scaling values for different information purposes.
+Generalizations to different types of icons are described in Section~\ref{sec:scatter}.
 Glyph-maps work nicely with gridded data, but many spatiotemporal data sets have irregular spatial locations.
 Possible adjustments are discussed in Section \ref{sec:irregular}.
-Code and data is available in the online supplementary materials.
 
-\section{Glyph-maps}~\label{sec:glyph-map}
+To demonstrate the glyph-map we will use three datasets:
 
-% Reference Tufte's small multiples
-\emph{Glyph-maps} are maps where a glyph is displayed at each spatial location to represent multiple data values.
-To explain the glyph-map, we use data of 2$^o$x2$^o$ gridded temperature anomalies, from the GISTEMP surface temperature data provided by the Earth System Research Laboratory, Physical Sciences Division, National Oceanic and Atmospheric Administration.
-Ground station data was de-seasonalized, differenced from ***, and spatially averaged to obtain gridded measurements.
-Figure \ref{fig:gistemp-pred} shows the time series icons for each spatial location of predicted temperature for 1950-2010 across the USA.
-Most locations observe an increasing temperature, to varying degrees, over this time period.
-A few locations show a decline.
-These locations tend to be different from their neighbors, which might suggest isolated data problems.
-There is a difference in the temperature increase at different regions: steeper inclines can be seen in the north and western mountain region and across the Great Lakes, but inclines are flatter in the southeastern USA.
-%The most seasonality occurs in north American.
-Some seasonal effects can also be seen in the Caribbean and in south American locations.
-Around the equator the temperatures are relatively warm for all seasons.
-Instead of radially displaying the temperature a glyph can be a regular time series, time displayed horizontally and temperature vertically (right plot).
-The structure perceived is different with this change in glyph type: locations at the equator have flat series, on land more varied temperatures.
+\begin{itemize}
 
-\begin{figure}[htbp]
-  \centering
-  \includegraphics[width=1\linewidth]{gistemp-pred}%
+  \item The ASA 2009 data expo data which consists of monthly observations of
+  several atmospheric variables from the International Satellite Cloud
+  Climatology Project. The dataset includes observations over 72 months
+  (1995--2000) on a 24 x 24 grid (576 locations) stretching from
+  $113.75^{\circ}$W to $56.25^{\circ}$W longitude and $21.25^{\circ}$S to
+  $36.25{^\circ}$N latitude.
 
-  %\includegraphics[width=1\linewidth]{gistemp-polar-pred}
+  \item GISTEMP surface temperature data provided by the Earth System Research
+  Laboratory, Physical Sciences Division, National Oceanic and Atmospheric
+  Administration. This data is provided on $2^{\circ}$ x $2^{\circ}$ grid over
+  the continental United States (XXX locations), measured monthly (???) from
+  XXX to YYY (ZZZ time points in total). Ground station data was
+  de-seasonalized, differenced from ***, and spatially averaged to obtain
+  gridded measurements
 
-  \caption{Glyph-map of predicted temperature anomaly data, from
-    1950-2010. Increasing trends can be seen over most of the USA,
-    with varying degrees of increase. Some isolated locations exhibit
-    steep declines, and inclines, perhaps indicative of some isolated
-    data problems.}
-  \label{fig:gistemp-pred}
-\end{figure}
+  \item USHCN (Version 2) ground station network.
 
-*** Remove the strange data over ocean, and make the reference lines white?
+\end{itemize}
 
-% Add time series as well as radial
-%\begin{figure*}[htp]
-%\centerline{\includegraphics[width=3.5in]{nasa-glyph-all.pdf}\includegraphics[width=3.5in]{nasa-glyph-ts-all.pdf}}
-%\caption{Glyph-map of raw data, using star glyphs (left), and time series as glyphs (right).}
-%\label{fig:glyph-map}
-%\end{figure*}
+All datasets and R code used to produce the plots are available in the online supplementary materials.
 
-Figure \ref{fig:gistemp-pred} shows the star glyphs for each spatial location of raw temperature for 1880-2011 across the USA.
-The major feature to notice is the pattern of missing values, icons that do not complete the full circle.
+\section{Glyph plots for large data}
+\label{sec:construction}
+
+A lot of climate data records contain long temporal history.
+For temperature records often extend from the 1800s to present day.
+This places some difficulties, one being computational because of the volume of records, and the second being visual display of a lot of data.
+Plotting the full raw time series in miniature for the glyph-maps often reveals little more than an irrelevant observation that ``temperature is quite varied!'' Careful use of de-seasonalizing, and models that can reflect nuances in global trend are needed to improve the visualization.
+And efficient computation to calculate and store predicted values will be required.
+
+Figure~\ref{fig:gistemp-pred} shows the a glyph plot for a dataset that represents a more typical size of climate data, the GISTEMP data.
+Star glyphs at each location represent temperature from 1880--2011 across the USA.
+The major feature to notice is the pattern of missing values, glyphs that do not complete the full circle.
 The high noon position of the circle is the first time point, Jan 1880, and it is this early time period that has missing values are a lot of locations.
-It was for this reason that the temporal model was fit just to 1950-2010 data.
 Another important feature is variability, indicated by the thickness of the circle.
@@
 At these locations there are large fluctuations from the long term mean.
+
+*** Need to add reference circles.
 
@@
 
-*** Need to add reference circles.
+Large datasets often required the display of summaries other than the raw data.
+Structure in temporal data is composed of different parts: global trend, seasonality and error.
+Extracting each of these for visualization improves the focus on different aspects of the time series.
+Figure~\ref{fig:gistemp-pred} shows the time series icons for each spatial location of predicted temperature for 1950--2010 across the USA (limited time period used due to missing data).
+Most locations observe an increasing temperature, to varying degrees, over this time period.
+A few locations show a decline.
+These locations tend to be different from their neighbors, which might suggest isolated data problems.
+There is a difference in the temperature increase at different regions: steeper inclines can be seen in the north and western mountain region and across the Great Lakes, but inclines are flatter in the southeastern USA.
 
-Icons have been displayed on maps since the earliest displays weather forecasts, in the 1800s.
-In recent times, \citet{pickett:1988} introduced the idea of using icons on maps to display multivariate spatial data.
-This area of work has morphed into the field of metaphorical data displays, which creates abstract landscapes of spatial data, a digression from glyph-maps.
-\citet{gribov:2006} describes the use of glyph-maps for multivariate data also, with emphasis on the graphics software Gauguin.
+*** Remove the strange data over ocean, and make the reference lines white?
 
-%\subsection{Glyph plots}
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=1\linewidth]{gistemp-pred}%
+  %\includegraphics[width=1\linewidth]{gistemp-polar-pred}
 
-%Simple, but effective, technique that combines aspects of glyphs, facetted plots and map-charts. Particularly well suited for the type of temporal-spatial data, commonly seen in climate.
+  \caption{Glyph-map of predicted temperature anomaly data, from
+    1950-2010. Increasing trends can be seen over most of the USA,
+    with varying degrees of increase. Some isolated locations exhibit
+    steep declines, and inclines, perhaps indicative of some isolated
+    data problems.}
+  \label{fig:gistemp-pred}
+\end{figure}
 
-%Glyph-maps have attributes of both glyphs (stars, faces, profiles, ...), inviting exploration at the global level, and small-multiple time series, allowing local exploration of individual series.
-Allow to see both geographic context and individual locations simultaneously.
-
-%Avoids one problem that plagues glyph displays: the ordering of the variables.
-Although some techniques exist to ameliorate this problem \citep{kleiner:1981,hurley:2010}, they are not necessary with time data because it is intrinsically ordered.
-
-%These displays tend to be particularly effective when printed in large format, as the much higher resolution of print (600 dpi) compared to computer  screen (72-120 dpi) allows for much finer reading.
-These are high-density data displays in the best tradition of Tufte.
-Wall size maps are not usually practical during analysis, but are very engaging.
-
-%Will first explore for regularly gridded data, a simple starting point and a common output of climate models, and then show a few simple extensions to deal with non-rectangular grids and irregular locations, as exemplified by raw climate data which is typically collect at irregular locations.
+%The most seasonality occurs in north American.
+Some seasonal effects can also be seen in the Caribbean and in south American locations.
+Around the equator the temperatures are relatively warm for all seasons.
+Instead of radially displaying the temperature a glyph can be a regular time series, time displayed horizontally and temperature vertically (right plot).
+The structure perceived is different with this change in glyph type: locations at the equator have flat series, on land more varied temperatures.
 
@@
 
-%A big advantage of glyph-maps is that they can be generated with existing graphics software after performing a simple pre-processing step, making them easily accessible to a wide audience. 
+Glyph-maps can be generated with existing graphics software after performing a simple pre-processing step, making them easily accessible to a wide audience.
 Creating a glyph-map, involves recognizing that they combine two structural components of the data: spatial location and data values.
@@
   y_\text{ coordinate}&=& y_{\amaj} + h \cdot y_{\amin}, 
-\end{array}\label{coords.eqn}
+  \end{array}
+  \label{coords.eqn}
 \end{equation}
-where $w$ and $h$ are width and height respectively.
+
+\noindent where $w$ and $h$ are width and height respectively.
 The width and height will typically take into account the resolution of the major axes, that is, the minimum difference between consecutive locations.
 
-Computing the coordinates for star glyphs is achieved by using polar coordinates,  $r=y_{\amin}$, and $\theta=x_{\amin}$, both scaled to be between 0 and 1. Then the glyph coordinates are given by 
+Computing the coordinates for star glyphs is achieved with a simple polar transformation:
+
 \begin{equation}
   \begin{array}{lll}
-x_\text{ coordinate}&=& x_{\amaj} + \frac{w \cdot r}{2} \sin(2 \pi \theta) \\
-y_\text{ coordinate}&=& y_{\amaj} + \frac{h \cdot r}{2} \cos(2 \pi \theta).
-\end{array}\label{coords.polar.eqn}
+  x_\text{ coordinate}&=& x_{\amaj} + y_{\amin} \cdot \frac{w}{2} \sin(2 \pi x_{\amin}) \\
+  y_\text{ coordinate}&=& y_{\amaj} + y_{\amin} \cdot \frac{h}{2} \cos(2 \pi x_{\amin}).
+  \end{array}
+  \label{coords.polar.eqn}
 \end{equation}
-Note that, this is is a non-standard conversion to polar coordinates, but this conversion provides a time line that starts at 12 o'clock and wraps around the origin in a clockwise fashion.
+
+\noindent Note that this is is a non-standard conversion to polar coordinates, but it provides a timeline that starts at 12 o'clock and travels clockwise.
 
@@
 This technique is used to good effect in \citet{buja:1996a}, and was used to examine the climate data in \citet{hobbs:2010}.
-
-% Show the glyph-map for de-seasonalized data, and describe what we
-% see differently from the facetted map
-
-%\begin{figure}[htbp]
-%  \centering
-%  \includegraphics[width=1\linewidth]{gistemp-raw}%
-
-%  \includegraphics[width=1\linewidth]{gistemp-polar-raw}
-
-%  \caption{Glyph-maps of gridded raw temperature anomaly data}
-%  \label{fig:ref-gistemp}
-%\end{figure}
 
@@
 
+% From a cognitive perspective, this is a very hard task to do.
+It falls into the framework of {\it change blindness} (see e.g.
+\citet{healey:2011} and often leads to failure without readers realizing, in particular for small areas of change or subtle changes.
+
 Glyph-maps using trend lines enable the reading of trends across spatial locations directly from the plot.
@@
 
-The glyph-map can be considered to be small multiples as named by \citet{tufte:2001}, in this case, small multiples of time series.
 By laying the time series out on the spatial grid, the main task expected of the viewer is comparing the temporal patterns from one location to the next.
@@
 Accurate perception of color differences is difficult.
-
-\begin{figure*}[htp]
-\centerline{\includegraphics[width=4in]{nasa-deseas-glyph.png}}
-\caption{Glyph-map of de-seasonalized temperature, to compare with Figure \ref{fig:facetted-map}. }
-\label{fig:nasa-deseas-glyph}
-\end{figure*}
 
@@
 
-Structure in temporal data is composed of different parts: global trend, seasonality and error.
-Extracting each of these for visualization improves the focus on different aspects of the time series.
 A generalized additive model \citep{wood:2006} with a seasonal term and a long-term smooth term can be useful.
@@
 
-\section{Long time series}
-
-A lot of climate data records contain long temporal history.
-For temperature records often extend from the 1800s to present day.
-This places some difficulties, one being computational because of the volume of records, and the second being visual display of a lot of data.
-Plotting the full raw time series in miniature for the glyph-maps often reveals little more than an irrelevant observation that ``temperature is quite varied!'' Careful use of de-seasonalizing, and models that can reflect nuances in global trend are needed to improve the visualization.
-And efficient computation to calculate and store predicted values will be required.
-
 \section{Conclusions}

### 2011-09-27 paper.tex
Move generalisations to future work

@@
 
-\section{Generalization}~\label{sec:scatter}
-
-Glyph-maps are not limited to time series plots.
-They are readily generalized to use any other type of display as an icon.
-Figure~\ref{fig:cloud} shows scatterplots as icons: temperature values are displayed horizontally and high-cloud values are displayed vertically, both locally scaled.
-The bivariate relationship between the two variables is explored spatially, and it is clear that it is different across the region.
-The equatorial Pacific and Caribbean have positive association between temperature and high cloud, while the south American continent has negative association.
-The second plot displays a loess curve fit to the data instead of the individual points, sharpening the view of association.
-In the Gauguin software \citep{gribov:2006} one alternative is an icon based on a histogram, small univariate distribution displays for each glyph.
-
-\begin{figure}[htbp]
-  \centering
-  \includegraphics[width=0.5\linewidth]{nasa-scat-glyph}%
-  \includegraphics[width=0.5\linewidth]{nasa-loess-glyph}
-
-  \caption{Glyph-map showing (left) scatterplot of temperature vs high-cloud and (right) smoothed (loess) curve fit to each location.
-The relationship between these two variables varies considerably over the spatial domain.
-}
-  \label{fig:cloud}
-\end{figure}
-
 \section{Regular vs irregular locations}~\label{sec:irregular}
@@
 
+There is much more work to be done on this type of display.
+Glyph-maps are not limited to time series plots.
+They are readily generalized to use any other type of display as an icon.
+Figure~\ref{fig:cloud} shows scatterplots as icons: temperature values are displayed horizontally and high-cloud values are displayed vertically, both locally scaled.
+The bivariate relationship between the two variables is explored spatially, and it is clear that it is different across the region.
+The equatorial Pacific and Caribbean have positive association between temperature and high cloud, while the south American continent has negative association.
+The second plot displays a loess curve fit to the data instead of the individual points, sharpening the view of association.
+Another option is available in Gauguin \citep{gribov:2006}, a icon based on a histogram, giving small univariate distribution displays for each location.
+
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=0.5\linewidth]{nasa-scat-glyph}%
+  \includegraphics[width=0.5\linewidth]{nasa-loess-glyph}
+
+  \caption{Glyph-map showing (left) scatterplot of temperature vs high-cloud and (right) smoothed (loess) curve fit to each location.
+The relationship between these two variables varies considerably over the spatial domain.
+}
+  \label{fig:cloud}
+\end{figure}
+
 \section*{Acknowledgements}

### 2011-09-27 paper.tex
Fix section labels
cleaned up references

--- 2011-09-24 paper.tex
@@
 
-\title{Glyph-maps for visualizing climate data and models}
-\author{Cook, Hofmann, Wickham, Wickham}
+\title{Glyph-maps for Visualizing Climate Data and Models}
+\author{Dianne Cook$^1$, Heike Hofmann$^1$, Hadley Wickham$^2$, Charlotte Wickham$^3$\\
+$^1$Department of Statistics, Iowa State University\\
+$^2$ Department of Statistics, Rice University\\
+$^3$ Department of Statistics, Oregon State University}
+\date{}
 
@@
 \maketitle
+
+\section*{Abstract}
 
@@
 To explain the glyph-map, we use data of 2$^o$x2$^o$ gridded temperature anomalies, from the GISTEMP surface temperature data provided by the Earth System Research Laboratory, Physical Sciences Division, National Oceanic and Atmospheric Administration.
-Ground station data was de-seasonalized, differenced from ***, and spatially averaged to obtain gridded measurements.
+Ground station data was de-seasonalized, differenced from the 1951-1980 temperatue averages, and spatially averaged to obtain gridded measurements.
 Figure \ref{fig:gistemp-pred} shows the time series icons for each spatial location of predicted temperature for 1950-2010 across the USA.

### 2011-09-27 paper.tex

--- 2011-09-27 paper.tex
@@
 
+% Reference Tufte's small multiples
+%\emph{Glyph-maps} are maps where a glyph is displayed at each spatial location to represent multiple data values.
+To explain the glyph-map, we use data of 2$^o$x2$^o$ gridded temperature anomalies, from the GISTEMP surface temperature data provided by the Earth System Research Laboratory, Physical Sciences Division, National Oceanic and Atmospheric Administration.
+Ground station data was de-seasonalized, differenced from the 1951-1980 temperatue averages, and spatially averaged to obtain gridded measurements.
+Figure \ref{fig:gistemp-pred} shows the time series icons for each spatial location of predicted temperature for 1950-2010 across the USA.
+Most locations observe an increasing temperature, to varying degrees, over this time period.
+A few locations show a decline.
+These locations tend to be different from their neighbors, which might suggest isolated data problems.
+There is a difference in the temperature increase at different regions: steeper inclines can be seen in the north and western mountain region and across the Great Lakes, but inclines are flatter in the southeastern USA.
+%The most seasonality occurs in north American.
+Some seasonal effects can also be seen in the Caribbean and in south American locations.
+Around the equator the temperatures are relatively warm for all seasons.
+Instead of radially displaying the temperature a glyph can be a regular time series, time displayed horizontally and temperature vertically (right plot).
+The structure perceived is different with this change in glyph type: locations at the equator have flat series, on land more varied temperatures.
 \begin{itemize}
@@
 
-  \item GISTEMP surface temperature data provided by the Earth System Research
-  Laboratory, Physical Sciences Division, National Oceanic and Atmospheric
-  Administration. This data is provided on $2^{\circ}$ x $2^{\circ}$ grid over
-  the continental United States (XXX locations), measured monthly (???) from
-  XXX to YYY (ZZZ time points in total). Ground station data was
-  de-seasonalized, differenced from ***, and spatially averaged to obtain
-  gridded measurements
+\item GISTEMP surface temperature data provided by the Earth System
+  Research Laboratory, Physical Sciences Division, National Oceanic
+  and Atmospheric Administration. This data is provided on $2^{\circ}$
+  x $2^{\circ}$ grid over the entire globe, continental United States
+  (XXX locations), measured monthly.  Ground station data was
+  de-seasonalized, differenced from from the 1951-1980 temperature
+  averages, and spatially averaged to obtain gridded measurements. We
+  extracted the locations corresponding to the continental USA.
 
@@
 \section{Glyph plots for large data}
-\label{sec:construction}
+\label{sec:large-data}
 
@@
 
-\section{Computation}~\label{sec:construction}
+\section{Construction}~\label{sec:construction}
 
@@
 
-\section{Perception of structure}~\label{sec:scale}
+\section{Perception of structure}~\label{sec:perception}
 

### 2011-09-27 paper.tex
Draft abstract and conclusions

@@
 
-\section*{Abstract}
+\begin{abstract}
+
+Climate data is challenging to visualise because it is multivariate and has both spatial (2d) and temporal (1d) components.
+In this paper we propose an new display for climate data: glyph-maps.
+Glyph-maps are a specialisation of multivariate glyph plots; each spatial location is display with one glyph that represents the measurements recorded over time at that location.
+Glyph-maps allow the discovery of both local and global structure, making discovery much easier than the classical display using colours.
+
+We include a description of how to create glyph-maps using existing statistical software, our initial explorations of their perceptual features, and a discussion of reference frames.
+Our techniques are inspired by rectangular gridded data, but we provide generalisation to deal with non-rectangular grids and irregular locations.
+
+\end{abstract}
 
@@
 \end{figure}
-
-%Simple, but effective, technique that combines aspects of glyphs, facetted plots and map-charts. Particularly well suited for the type of temporal-spatial data, commonly seen in climate.
 
@@
 
-There is much more work to be done on this type of display.
+We have only scratched the surface of the potential of this type of display for climate data.
 Glyph-maps are not limited to time series plots.
@@
 
+There is also much work to be done on the perceptual underpinnings of these displays.
+We have noticed that periodic trends are easier to perceive in star-glyphs, and long-term trends in line-glyphs.
+Is this always the case?
+Are their certain types of periodicity that are particularly easy to spot?
+These questions need to be answered with rigorous perceptual studies to help guide the use of these plots.
+
 \section*{Acknowledgements}

### 2011-09-27 paper.tex
More intro tweaking

@@
 
-Climate data is composed of measurements on variables such as temperature, precipitation, and winds that have been collected from a spatial and temporal context.
-The classic display of this type of is to use color or grey scale overlaid on geographic location, typically including a map.
-When measurements are made at different time periods it is common to display the data as small multiples, such as in Figure~\ref{fig:facetted-map}.
+Climate data is composed of measurements, such as temperature, precipitation, and winds, with a spatial and temporal context.
+The classic display for data of this type of is a tiled plot using color to display the value of the variable of interest as each location.
+When measurements are made at multiple time points it is common to display the data with small multiples, such as in Figure~\ref{fig:nasa-facet}.
 In this plot, a separate map is drawn for each month (columns) and year (rows), over a 6 year period of remotely sensed temperature data above Central America \citet{murrell:2010}.
-Color is used to display the de-seasonalized temperature measurements, with red being high values and blue being low values.
-The most noticeable feature in this display is the strong red in the equatorial region beginning in mid-1997 and tapering out during 1998.
-This is the pattern of an El Ni\~no event, a major temperature anomaly.
-With some work other more localized patterns can be perceived, for example, cooler over the land mass in the early year, but warmer in the latter years.
+Color is used to display de-seasonalized temperature, with red mapped to high values and blue to low.
+The most noticeable feature is the strong red patch in the equatorial Pacific beginning mid-1997 and tapering out during 1998.
+This is the El Ni\~no event, a major temperature anomaly.
+More work is required to see more localized patterns, like the cooler land temperatures seen in early years.
 
-Reading this plot is cognitively challenging: the reader must play ``spot the difference'' from one small image to another.
-Large structures such as the El Ni\~no event are clear but it is very difficult to glue the images together to read off the trend over the years or to notice local deviations.
-Rendering these small multiples as a movie can help with picking up strong trends, but small trends still fail to draw readers' attention and escape unnoticed \citep{simons:gradual}.
+Reading this type of plot is cognitively challenging: the reader must play ``spot the difference'' from one small image to another.
+Large structures such as El Ni\~no are clear but it is very difficult to mentally difference the images and read off the long-term trend or to notice local deviations.
+Using a movie to render these small multiples can help, but small trends still fail to draw readers' attention and escape unnoticed \citep{simons:gradual}.
 
@@
 Smaller features are only noticeable on closer inspection, or if pointed out: such as the relative warming on the mountain regions in south and north America in the later years.}
-  \label{fig:facetted-map}
+  \label{fig:nasa-facet}
 \end{figure*}
@@
 
-In this paper we develop a different type of display that resolves these problems: the \emph{glyph-map}.
+In this paper we develop a new type of display to resolves these problems: the \emph{glyph-map}.
 The glyph-map uses a small glyph, or icon, to represent multiple values at each location.
-Glyph maps are an adaptation of glyphs, tools developed for display multivariate data, including the star plots of \citet{mayr:1877}, the semi-graphic displays of \citet{anderson:1960}, and the infamous Chernoff faces \citep{chernoff:1973}.
-An glyph is produced by mapping each variable to some graphical feature.
-Figure~\ref{fig:nasa-deseas-glyph} shows the glyph-map equivalent of Figure~\ref{fig:facetted-map}: each location is represented by a tiny time series glyph.
-The glyph-map can also be thought of as a small multiple display \citet{tufte:2001} of time series.
+Glyph maps are an adaptation of glyphs, tools developed for display multivariate data.
+Multivariate glyphs include the star plots of \citet{mayr:1877}, the semi-graphic displays of \citet{anderson:1960}, and the infamous Chernoff faces \citep{chernoff:1973}.
+An glyph is produced by mapping each variable to some graphical feature, such as the length of a line.
+Figure~\ref{fig:nasa-glyph} shows the glyph-map equivalent of Figure~\ref{fig:nasa-facet}: each location is represented by a small time series glyph.
+As suggested by this plot, the glyph-map can also be thought of as a small multiple display \citet{tufte:2001} of time series with a geographic context.
+
+% The most seasonality occurs in north American.
+Some seasonal effects can also be seen in the Caribbean and in south American locations.
+Around the equator the temperatures are relatively warm for all seasons.
+Instead of radially displaying the temperature a glyph can be a regular time series, time displayed horizontally and temperature vertically (right plot).
+The structure perceived is different with this change in glyph type: locations at the equator have flat series, on land more varied temperatures.
 
@@
   \centering
-  \includegraphics[width=3.5in]{nasa-deseas-glyph.png}
-  \caption{Glyph-map of de-seasonalized temperature, to compare with Figure \ref{fig:facetted-map}. }
-  \label{fig:nasa-deseas-glyph}
+  \includegraphics[width=3.5in]{nasa-glyph.png}
+  \caption{Glyph-map of de-seasonalized temperature, to compare with Figure \ref{fig:nasa-facet}. }
+  \label{fig:nasa-glyph}
 \end{figure*}
 
-One problem with glyph displays of multivariate data is that there is no natural ordering of each glyph, or of the mapping between variables and glyph properties.
-This leads to a combinatorial explosion of possibilities which may have dramatically different perceptual properties.
-Despite some proposed remedies \citep{kleiner:1981,hurley:2010}, this problem has prevented the widespread adoption of glyph displays.
-Using glyphs for space-time data resolves this problem: we place glyphs according to the measurement location of the data, and we order variables in time.
-Additionally, climate data usually has high-correlations between nearby locations and times.
-This imposes a certain degree of smoothness that gives the plot an appearance of texturing on a landscape, making it easier to digest patterns.
+One problem with multivariate glyph is that lack of natural ordering of each glyph, or of the mapping between variables and glyph properties.
+This leads to a combinatorial explosion of possibilities, each of which may have dramatically different perceptual properties.
+This problem has prevented the widespread adoption of glyph displays, despite some proposed remedies \citep{kleiner:1981,hurley:2010}.
+Using glyphs with space-time data makes the problem go away: we place glyphs according to the location of the measurement, and we order variables by the time they were collected.
+Additionally, climate data usually has high correlations between nearby locations and times.
+This imposes a degree of smoothness that gives the plot an appearance of a textured  landscape, making it easier to digest patterns.
 Others \citet{pickett:1988} have used glyphs on maps to display multivariate spatial data.
@@
 
-We focus our attention on two glyphs: lines and stars.
-Figure~\ref{fig:templates} displays 12 iconic time series shapes with line and star glyphs.
-The data underlying each glyph has 36 time variables.
-The star glyphs is formed by considering the 36 axes radiating from a common midpoint, and the data values for the row are plotted on each axis relative to the locations of the minimum and maximum of the variable.
+We focus our attention on two types of glyph: lines and stars.
+Figure~\ref{fig:templates} displays 12 iconic time series shapes with line- and star-glyphs.
+The data underlying each glyph is measured at 36 time points.
+The line-glyphs are a simple time series plot.
+The star-glyphs are formed by considering the 36 axes radiating from a common midpoint, and the data values for the row are plotted on each axis relative to the locations of the minimum and maximum of the variable.
+This is a polar transformation of the line-glyph.
 
@@
 
-In this paper we describe the algorithm for creating glyphs of temporal data on maps, and the various modifications that can enable different features in the data to be detected.
-Section~\ref{sec:glyph-map} describes the use of glyphs on maps, and Section~\ref{sec:construction} explains how to construct glyph-maps.
-Section~\ref{sec:reference} discusses the use of reference lines and boxes, and Section~\ref{sec:scale} visits issues of scaling values for different information purposes.
-Generalizations to different types of icons are described in Section~\ref{sec:scatter}.
-Glyph-maps work nicely with gridded data, but many spatiotemporal data sets have irregular spatial locations.
-Possible adjustments are discussed in Section \ref{sec:irregular}.
+In this paper we describe the algorithm for creating glyphs-maps, and various modifications that can enable different features in the data to be detected.
+Section~\ref{sec:large-data} describes how glyph-maps scale to large climate data sets, and Section~\ref{sec:construction} explains how to construct them.
+Section~\ref{sec:reference} discusses the use of reference lines and boxes, Section~\ref{sec:perception}, their perceptual properties, and Section~\ref{sec:scale} visits scaling issues.
+We originally developed glyph-maps for data on a rectangular grid, but many spatiotemporal data sets have irregular spatial locations.
+Section~\ref{sec:irregular} discusses how glyph-maps can be adjusted for this type of data.
 
@@
 
-% Reference Tufte's small multiples
-%\emph{Glyph-maps} are maps where a glyph is displayed at each spatial location to represent multiple data values.
-To explain the glyph-map, we use data of 2$^o$x2$^o$ gridded temperature anomalies, from the GISTEMP surface temperature data provided by the Earth System Research Laboratory, Physical Sciences Division, National Oceanic and Atmospheric Administration.
-Ground station data was de-seasonalized, differenced from the 1951-1980 temperatue averages, and spatially averaged to obtain gridded measurements.
-Figure \ref{fig:gistemp-pred} shows the time series icons for each spatial location of predicted temperature for 1950-2010 across the USA.
-Most locations observe an increasing temperature, to varying degrees, over this time period.
-A few locations show a decline.
-These locations tend to be different from their neighbors, which might suggest isolated data problems.
-There is a difference in the temperature increase at different regions: steeper inclines can be seen in the north and western mountain region and across the Great Lakes, but inclines are flatter in the southeastern USA.
-%The most seasonality occurs in north American.
-Some seasonal effects can also be seen in the Caribbean and in south American locations.
-Around the equator the temperatures are relatively warm for all seasons.
-Instead of radially displaying the temperature a glyph can be a regular time series, time displayed horizontally and temperature vertically (right plot).
-The structure perceived is different with this change in glyph type: locations at the equator have flat series, on land more varied temperatures.
 \begin{itemize}
@@
 
-\item GISTEMP surface temperature data provided by the Earth System
-  Research Laboratory, Physical Sciences Division, National Oceanic
-  and Atmospheric Administration. This data is provided on $2^{\circ}$
-  x $2^{\circ}$ grid over the entire globe, continental United States
-  (XXX locations), measured monthly.  Ground station data was
-  de-seasonalized, differenced from from the 1951-1980 temperature
-  averages, and spatially averaged to obtain gridded measurements. We
-  extracted the locations corresponding to the continental USA.
+  \item GISTEMP surface temperature data provided by the Earth System Research
+  Laboratory, Physical Sciences Division, National Oceanic and Atmospheric
+  Administration. This data is provided on $2^{\circ}$ x $2^{\circ}$ grid over
+  the entire globe, continental United States (XXX locations), measured
+  monthly. Ground station data was de-seasonalized, differenced from from the
+  1951-1980 temperature averages, and spatially averaged to obtain gridded
+  measurements. We extracted the locations corresponding to the continental
+  USA.
 
@@
 In comparison, to read trend from facetted colored maps one needs to spot the difference between the different maps.
-Compare the facetted map in Figure \ref{fig:facetted-map} with the glyph-map of de-seasonalized data in Figure \ref{fig:nasa-deseas-glyph}.
+Compare the facetted map in Figure \ref{fig:nasa-facet} with the glyph-map of de-seasonalized data in Figure \ref{fig:nasa-glyph}.
 The glyph-map provides a texturing over the map.
@@
 Locations with overall large temperature will catch the focus.
-Figure \ref{fig:nasa-deseas-glyph} shows data scaled to 0 to 1 based on the global minimum and maximum, using:
+Figure \ref{fig:nasa-glyph} shows data scaled to 0 to 1 based on the global minimum and maximum, using:
 

### 2011-09-27 paper.tex
More large data tweaking

@@
 
-\section{Glyph plots for large data}
+\section{Glyph-maps for large data}
 \label{sec:large-data}
 
-A lot of climate data records contain long temporal history.
-For temperature records often extend from the 1800s to present day.
-This places some difficulties, one being computational because of the volume of records, and the second being visual display of a lot of data.
-Plotting the full raw time series in miniature for the glyph-maps often reveals little more than an irrelevant observation that ``temperature is quite varied!'' Careful use of de-seasonalizing, and models that can reflect nuances in global trend are needed to improve the visualization.
-And efficient computation to calculate and store predicted values will be required.
+Climate data is usually large: many locations in space, and many points in time.
+This makes climate data challenging to work with computationally and to visualise.
+Plotting the raw data often reveals little more than a well-known observation that ``temperature is quite varied!'' Careful use of models than Careful use of models to decompose data into long-term trend, seasonal effects and error is vital for to make the problem both computationally and visually tractable.
 
-Figure~\ref{fig:gistemp-pred} shows the a glyph plot for a dataset that represents a more typical size of climate data, the GISTEMP data.
-Star glyphs at each location represent temperature from 1880--2011 across the USA.
-The major feature to notice is the pattern of missing values, glyphs that do not complete the full circle.
-The high noon position of the circle is the first time point, Jan 1880, and it is this early time period that has missing values are a lot of locations.
-Another important feature is variability, indicated by the thickness of the circle.
-Some locations, particularly in the north western mountain region, have big variability in temperature, both low minima, and high maxima.
-At these locations there are large fluctuations from the long term mean.
+Figure~\ref{fig:gistemp-pred} shows the a glyph-map for the large GISTEMP dataset.
+Glyphs at each location represent temperature from 1880--2011, starting with Jan 1880 at 12 o'clock and proceeding clockwise to 2011 at 11:59.
+The most important feature to notice is the pattern of missing values.
+Most glyphs do not complete the full circle, typically missing points between 12 and 1, suggesting missing values in the earliest part of the data.
+Another important feature is variability, shown by the thickness of the circle.
+Some locations, particularly in the north western mountain region, have large variations in temperature.
 
@@
 
-Large datasets often required the display of summaries other than the raw data.
-Structure in temporal data is composed of different parts: global trend, seasonality and error.
-Extracting each of these for visualization improves the focus on different aspects of the time series.
-Figure~\ref{fig:gistemp-pred} shows the time series icons for each spatial location of predicted temperature for 1950--2010 across the USA (limited time period used due to missing data).
-Most locations observe an increasing temperature, to varying degrees, over this time period.
-A few locations show a decline.
+The raw data shows us missing values and gives some sense of variability, but gives no information about long-term trend.
+We can look trend by fitting a linear model to each location and displaying the predicted values.
+Figure~\ref{fig:gistemp-pred} shows this for 1950--2010 (limited time period used due to missing data).
+Most locations have increasing temperature over this time period.
+Only a few locations show a decline.
 These locations tend to be different from their neighbors, which might suggest isolated data problems.
-There is a difference in the temperature increase at different regions: steeper inclines can be seen in the north and western mountain region and across the Great Lakes, but inclines are flatter in the southeastern USA.
+The rate of increase differs across regions: steeper inclines in the north and western mountain regions and across the Great Lakes, and flatter inclines in the southeastern USA.
 

### 2011-09-27 paper.tex
Fix careless search-and-replace mistake
Tweaking construction section

@@
   \centering
-  \includegraphics[width=3.5in]{nasa-glyph.png}
+  \includegraphics[width=3.5in]{nasa-deseas-glyph}
   \caption{Glyph-map of de-seasonalized temperature, to compare with Figure \ref{fig:nasa-facet}. }
@@
 Glyph-maps can be generated with existing graphics software after performing a simple pre-processing step, making them easily accessible to a wide audience.
-Creating a glyph-map, involves recognizing that they combine two structural components of the data: spatial location and data values.
-The spatial location is the major positioning component, and the data values are minor adjustments to these positions.
+Creating a glyph-map requires a recognition they combine two structural components of the data: spatial location and data values.
+The spatial location is the major positioning component, and the data values are minor adjustments to those positions.
 For spatiotemporal data, the major axes are latitude ($y_{\amaj}$) and longitude ($x_{\amaj}$), and the minor axes are time ($x_{\amin}$) and some measurement ($y_{\amin}$), for example, temperature, or predicted temperature.
-The minor axes are rescaled to have range $[0, 1]$, and further adjusted to fit into the resolution of the spatial coordinates.
-The final values to be plotted are a simple linear combination, as given by
+Assuming the minor axes are rescaled to $[-1, 1]$, the final positions are a simple linear combination:
 
@@
 \noindent where $w$ and $h$ are width and height respectively.
-The width and height will typically take into account the resolution of the major axes, that is, the minimum difference between consecutive locations.
+For gridded data, the width and height be the resolution of the major axes, that is, the minimum difference between consecutive locations.
+Since the coordinates are linear combinations of two variables, it is possible to build these these line-glyphs interactively, illustrated by software packages that have incorporated tours \citep{cook:2006}, such as DataViewer \citep{buja:1986}, XGobi \citep{swayne:1991} or GGobi \citep{swayne:2003}.
+This technique is used to good effect in \citet{buja:1996a}, and was used to examine the climate data in \citet{hobbs:2010}.
 
-Computing the coordinates for star glyphs is achieved with a simple polar transformation:
+Computing the coordinates for star glyphs is achieved with a simple polar transformation, with minor axes scaled to $[0, 1]$:
 
@@
 
-\noindent Note that this is is a non-standard conversion to polar coordinates, but it provides a timeline that starts at 12 o'clock and travels clockwise.
+\noindent This is a non-standard conversion to polar coordinates, but it creates a timeline that starts at 12 o'clock and proceeds clockwise.
 
-Figure~\ref{fig:cycle} shows average monthly temperature as time (Cartesian coordinates) and star (polar coordinates) glyphs, for a subset of the spatial region.
+% HW: I removed this because I don't think it adds anything at this point
+%
+% Figure~\ref{fig:cycle} shows average monthly temperature as time (Cartesian coordinates) and star (polar coordinates) glyphs, for a subset of the spatial region.
 In this case, the star glyphs are not as effective as the linear glyphs: perhaps because of the absence of cyclical pattern.
+% 
+% *** Perhaps replace these figures with something from the GISTEMP data.
+% 
+% \begin{figure}[htbp]
+%   \centering
+%   \includegraphics[width=3in]{month-cartesian}
+%   \includegraphics[width=3in]{month-polar}
+%     
+%   \caption{Monthly averages in (left) Cartesian coordinates and (right) polar coordinates.}
+%   
+%   \label{fig:cycle}
+% \end{figure}
 
-*** Perhaps replace these figures with something from the GISTEMP data.
-
-\begin{figure}[htbp]
-  \centering
-  \includegraphics[width=3in]{month-cartesian}
-  \includegraphics[width=3in]{month-polar}
-
-  \caption{Monthly averages in (left) Cartesian coordinates and (right) polar coordinates.}
-
-  \label{fig:cycle}
-\end{figure}
-
-Figure \ref{fig:templates} gives some small examples of time series with prototypical trends, shown both in cartesian coordinates and polar coordinates.
-Differences between linear and nonlinear trend are more apparent in the time series icons, and are effectively lost in polar coordinates.
-The star plots are most effective in exposing cyclical patterns due to seasonality.
-Seasonality appears like floral cartoons in the polar coordinates, with peaks forming petals.
-What might be surprising is that time series that seem to have opposing trends in cartesian coordinates don't necessarily show this symmetry in polar coordinates.
-The area of glyphs in polar coordinates mainly reflect the average value of a time series, while its shape shows deviations from average values.
-
-Since the coordinates are linear combinations of two variables, it is possible to build these these time series icon displays interactively, illustrated by software packages that have incorporated tours \citep{cook:2006}, such as DataViewer \citep{buja:1986}, XGobi \citep{swayne:1991} or GGobi \citep{swayne:2003}.
-This technique is used to good effect in \citet{buja:1996a}, and was used to examine the climate data in \citet{hobbs:2010}.
+Figure~\ref{fig:templates} gives some small examples of time series with prototypical trends, shown both in cartesian coordinates and polar coordinates.
+Differences between linear and nonlinear trend are more apparent in the line-glyphs, and are effectively lost in the star-glyphs, which are most effective in exposing cyclical patterns due to seasonality.
+Star-glyphs show seasonality as floral cartoons, with peaks forming petals.
+It is surprising that time series with opposing trends in line-glyphs don't show this symmetry when displayed with star-glyphs.
+The area of star-glyphs coordinates mainly reflect the average value of a time series, while its shape shows deviations from the average.
 

### 2011-09-27 paper.tex
Tweaking references section

@@
 \title{Glyph-maps for Visualizing Climate Data and Models}
-\author{Dianne Cook$^1$, Heike Hofmann$^1$, Hadley Wickham$^2$, Charlotte Wickham$^3$\\
+\author{Dianne Cook$^1$, Heike Hofmann$^1$, Charlotte Wickham$^2$, Hadley Wickham$^3$, \\
 $^1$Department of Statistics, Iowa State University\\
-$^2$ Department of Statistics, Rice University\\
-$^3$ Department of Statistics, Oregon State University}
+$^2$ Department of Statistics, Oregon State University\\
+$^3$ Department of Statistics, Rice University}
 \date{}
@@
 
-A \emph{visual reference grid} \citep{cleveland:1993a} is fundamental
-to a good data visualization, in order to compare values mapped to
-position along a line. For glyph-maps, it is not feasible to make a
-full set of axes for each icon, because these would dominate the
-picture. Visual reference grids need to be post-attentive
-\citep{healey}, de-emphasized, and serve as posterior look up after
-the patterns in the data have been read. For glyph-maps, we need to be
-able to make the following types of comparisons across icons:
+The structured spatial arrangement of icons in a glyph-map helps to compare the patterns of shape, like slope, intercept, or size, across icons.
+However, additional clues can make comparisons easier, converting the perceptual task from comparing length to the easier position along a common scale \citep{cleveland:1984}.
+This \emph{visual reference grid} \citep{cleveland:1993a} helps to support the comparisons we are most interested in:
 
@@
 
-The structured spatial arrangement of icons in a glyph-map helps to compare the patterns of shape, like slope, intercept, or size, across icons.
-But, additional clues would make it comparisons easier.
-For example glyph-map in Figure~\ref{fig:gistemp-pred}, contains two additional guides, a reference grid, and reference line.
-Both are drawn in white, because it is minimally perceptible (post-attentive), relative to the dark black ink of the data.
+Each glyph is small, so there is not enough space for a full set of axes.
+Instead, we add reference lines and boxes.
+These need to be minimally perceptible, post-attentive \citep{healey} and de-emphasized, so as not to detract from the raw data.
+For example, glyph-map in Figure~\ref{fig:gistemp-pred}, contains two additional guides, a reference grid and a reference line.
+Both are drawn in white, because it is minimally perceptible relative to the black ink of the data.
 
-The reference grid is a box, framing each icon, representing the spatial grid.
-It helps to read differences in slope and intercept.
+The \textbf{reference grid} is a box, framing each icon, which represents the spatial grid.
+The \textbf{reference line} is a horizontal line at mid-range.
+Both helps to read differences in slope and intercept.
 Slope is read by comparing the left and right endpoints of the predictions (line) against the relative location on the frame.
-Intercept is compared by reading the ``average'' position of the line in the box, near the top, or near the bottom.
-The reference line, drawn as a horizontal line at the average predicted temperature value, provides for better reading of the intercept.
+Intercept is compared by reading the ``average'' position of the line in the box: is it near the top, or near the bottom?
+
+% , making it easier to compare the slopprovides for better reading of the intercept.
 The average value for the temperature anomaly data is not the same across the map: at higher latitudes the temperatures are uniformly higher, but at lower latitudes they are uniformly lower.
@@
 
-For star glyphs, the reference frames are also useful, but the reference line would be replaced by a reference circle, drawn at the average data value.
-Figures \ref{fig:templates} and \ref{fig:gistemp-raw} show examples.
+For star glyphs, the reference frames are also useful and the equivalent of the reference line is a circle. Figures~\ref{fig:templates} and \ref{fig:gistemp-raw} show examples.
 

### 2011-09-28 paper.tex
A bit more restructuring

@@
 
+% \section{What to plot: raw data, trend lines, residuals,  seasonal}
+% 
+% A generalized additive model \citep{wood:2006} with a seasonal term and a long-term smooth term can be useful.
+This is revealing for smooth models of surface temperature, shown in Figure~\ref{fig:scaling}.
+Each location is modeled using a generalized additive model \citep{wood:2006} with a seasonal term and a long-term smooth term.
+The three plots in the figure show the long-term smooth term.
+% 
+% %Because some locations are uniformly warmer than others, the impact of
+% %changes across time are diminished. In the center plot, each location is
+% %scaled by dividing out the maximum, and on the right, each location is scaled
+% %to range $[0, 1]$. These emphasize individual changes but lose the ability to
+% %make global comparisons, and need caution when interpreting - a small
+% %absolute change gets blown up into a large relative change.
+% 
+% *** Need three plots here: trend, seasonality and residuals, shall we do this with NASA ozone data or stick to temperature?
+% 
+% Issues of scale relative to modeling.
+Switching from the data to model summaries brings new scaling issues.
+Often the scale of the raw data is ignored when the model is plotted, one usually shows just the line showing the trend, in the range of the trend.
+This has the effect of exaggerating the significance of the model.
+Scaling the trend lines relative to the range of the data may be useful sometimes.
+
 \section{Construction}~\label{sec:construction}
@@
 
-\section{What to plot: raw data, trend lines, residuals,  seasonal}
-
-A generalized additive model \citep{wood:2006} with a seasonal term and a long-term smooth term can be useful.
-This is revealing for smooth models of surface temperature, shown in Figure~\ref{fig:scaling}.
-Each location is modeled using a generalized additive model \citep{wood:2006} with a seasonal term and a long-term smooth term.
-The three plots in the figure show the long-term smooth term.
-%Because some locations are uniformly warmer than others, the impact of changes across time are diminished.
-In the center plot, each location is scaled by dividing out the maximum, and on the right, each location is scaled to range $[0, 1]$.
-These emphasize individual changes but lose the ability to make global comparisons, and need caution when interpreting - a small absolute change gets blown up into a large relative change.
- %The three plots in the figure show the long-term smooth term. Because some locations are uniformly warmer than others, the impact of changes across time are diminished.
-
-% is a extremely useful technique: product plots, ANOVA, ...
-For glyph plots, it's often useful to partition the glyphs into shape and scale components, decomposing the values into a product of maximum and scaled values.
-In this section we explore how decomposing each series into scale and shape components can be useful.
-
-*** Need three plots here: trend, seasonality and residuals, shall we do this with NASA ozone data or stick to temperature?
-
-Issues of scale relative to modeling.
-Switching from the data to model summaries brings new scaling issues.
-Often the scale of the raw data is ignored when the model is plotted, one usually shows just the line showing the trend, in the range of the trend.
-This has the effect of exaggerating the significance of the model.
-Scaling the trend lines relative to the range of the data may be useful sometimes.
-
-\section{Regular vs irregular locations}~\label{sec:irregular}
+\section{Non-gridded data}~\label{sec:irregular}
 

### 2011-09-28 paper.tex
Move acknowledgements to appropriate places in intro

@@
   measurements. We extracted the locations corresponding to the continental
-  USA.
+  USA. Provided by the NOAA/OAR/ESRL PSD, Boulder, Colorado,
+  USA, from their web site at \url{http://www.esrl.noaa.gov/psd/}.
 
-  \item USHCN (Version 2) ground station network.
+  \item USHCN (Version 2) ground station network. USHCN data provided by NOAA NCDC from their website \url{http://cdiac.ornl.gov/ftp/ushcn_v2_monthly/}.
 
@@
 All datasets and R code used to produce the plots are available in the online supplementary materials.
+All plots were created in R \citep{R} using the {\tt ggplot2} \citep{me:ggplot2} and {\tt plyr} \citep{me:plyr} packages.
 
@@
 
-\section*{Acknowledgements}
-
-GISTEMP data provided by the NOAA/OAR/ESRL PSD, Boulder, Colorado,
-USA, from their web site at \url{http://www.esrl.noaa.gov/psd/}.
-
-USHCN data provided by NOAA NCDC from their website \url{http://cdiac.ornl.gov/ftp/ushcn_v2_monthly/}.
-
-The graphics in this paper were created in R \citep{R} using {\tt ggplot2} \citep{me:ggplot2} and {\tt plyr} \citep{me:plyr}.
-
 \bibliography{references}

### 2011-09-28 paper.tex
Restructure perception sections
Rewrite start of perception section

@@
 Climate data is composed of measurements, such as temperature, precipitation, and winds, with a spatial and temporal context.
-The classic display for data of this type of is a tiled plot using color to display the value of the variable of interest as each location.
+The classic display for data of this type of is ``heatmap'', a tiled plot using color to display the value of the variable of interest as each location.
 When measurements are made at multiple time points it is common to display the data with small multiples, such as in Figure~\ref{fig:nasa-facet}.
@@
 An glyph is produced by mapping each variable to some graphical feature, such as the length of a line.
+The glyph-map can also be thought of as a small multiple display \citet{tufte:2001} of time series with a geographic context.
+
 Figure~\ref{fig:nasa-glyph} shows the glyph-map equivalent of Figure~\ref{fig:nasa-facet}: each location is represented by a small time series glyph.
-As suggested by this plot, the glyph-map can also be thought of as a small multiple display \citet{tufte:2001} of time series with a geographic context.
+The primary visible structure is still El Ni\~no, the regions in the Pacific with bump in the middle of the series, but we can also see other features: temperatures are generally more varied over land and a few locations have dramatically increasing temperature in the later years typically in high elevation areas, along the Andes and the Sierra Madre Occidental in Mexico.
 
@@
 
-\section{Reference frames}~\label{sec:reference}
+\section{Perception}~\label{sec:perception}
+
+All plots facilitate some comparisons and impede others.
+To decide if a plot if useful for a particular data analytic task, we need to check that the comparisons that we are interested in are the ones that a supported by the plot.
+Generally, for climate data, we are most interested in seeing changes in slope, or trend, average value, and variance.
+We argue that glyph-maps support these comparisons better because the comparisons are perceptually easier.
+
+Glyph-maps allow time trends to be read directly from the plot.
+Values are mapped to position, one of the easiest properties to perceive \citep{cleveland:1984}.
+Perceiving time trends in faceted heatmaps is much more difficult: not only do you need to read value from color (a relatively difficult task), you also need to spot the difference between the different maps.
+This is challenging from a cognitive perspective \citep{healey:2011,busey} and comparisons can suffer from \emph{change blindness}, often leading to silent failure.
+
+This is particularly obvious when comparing Figure~\ref{fig:nasa-facet} to Figure~\ref{fig:nasa-glyph}.
+
+% Laying time series out on the spatial grid, the main task expected of the viewer is comparing the temporal patterns from one location to the next.
+In contrast, in the small multiples used in the facetted colored map, the viewer is expected to compare the patterns in the maps from one year to the next, for corresponding months.
+(See \citet{carr:1999} for a discussion on perceptual grouping.) That is, compare and contrast maps with facetted colored map, or compare and contrast time series using the glyph-map.
+
+One draw-back of the glyph-maps, particularly if linear trend is displayed as an icon, is they may suffer from the Z\"ollner Illusion \citep{Zollner}, which makes straight lines look crooked, and is visible in Figure \ref{fig:gistemp-pred}.
+
+An third alternative is to calculate the relevant statistic for each location and display these values as single heatmap.
+For example, to study long-term trends, we could calculate the slope of a linear model at each location and color tiles across the map by the value of the slope.
+The disadvantage of this approach is that it is not exploratory, and requires that we have a very precise definition of the measurement we are interested in, and it still suffers from the relative difficultly of colour perception.
+
+Two other factors are critical for accurate perception of change: reference frames and scaling. These are described in the following two sections.
+
+\subsection{Reference frames}~\label{sec:reference}
 
@@
 However, additional clues can make comparisons easier, converting the perceptual task from comparing length to the easier position along a common scale \citep{cleveland:1984}.
-This \emph{visual reference grid} \citep{cleveland:1993a} helps to support the comparisons we are most interested in:
-
-\begin{itemize} \itemsep 0in
-\item Slope, or trend
-\item Intercept, or average value
-\item Variance, or size
-\end{itemize}
+This is also called a \emph{visual reference grid} \citep{cleveland:1993a}.
 
@@
 
-\section{Perception of structure}~\label{sec:perception}
-
-% From a cognitive perspective, this is a very hard task to do.
-It falls into the framework of {\it change blindness} (see e.g.
-\citet{healey:2011} and often leads to failure without readers realizing, in particular for small areas of change or subtle changes.
-
-Glyph-maps using trend lines enable the reading of trends across spatial locations directly from the plot.
-In comparison, to read trend from facetted colored maps one needs to spot the difference between the different maps.
-Compare the facetted map in Figure \ref{fig:nasa-facet} with the glyph-map of de-seasonalized data in Figure \ref{fig:nasa-glyph}.
-The glyph-map provides a texturing over the map.
-The El Ni\~no event is still the primary visible structure, the region that has a bump in the middle of the time series, corresponding to the hot 1997-1998 period.
-We can also see other temporal features: that the temperatures are generally more varied over land, there are a few locations with dramatically increasing temperature in the later years typically in locations that are the high elevation areas, along the Andes, and Sierra Madre Occiddental in Mexico.
-
-By laying the time series out on the spatial grid, the main task expected of the viewer is comparing the temporal patterns from one location to the next.
-In contrast, in the small multiples used in the facetted colored map, the viewer is expected to compare the patterns in the maps from one year to the next, for corresponding months.
-(See \citet{carr:1999} for a discussion on perceptual grouping.) That is, compare and contrast maps with facetted colored map, or compare and contrast time series using the glyph-map.
-One draw-back of the glyph-maps, particularly if linear trend is displayed as an icon, is they may suffer from the Z\"ollner Illusion \citep{Zollner} which is visible in Figure \ref{fig:gistemp-pred}.
-On the other hand, comparing images require that we play a spot the difference game, which is perceptually difficult \citet{busey}.
-
-An alternative, to both of these approaches, to focus attention on specific parts of the data calculate the relevant statistic for each location and display these values as color on a map.
-For example, to study the warming or cooling trends, we might calculate the slope of a linear model at each location and color tiles across the map by the value of the slope.
-From this we might see the increasing temperatures at higher altitudes in the NASA temperature data.
-However, we would fail to see the finer points of the time series, that the increase is not as simple as a linear increase.
-In addition, Cleveland's \citep{cleveland:1993a} perceptual tests suggest that the reader can perceive the positional difference of the lines better than the color differences.
-Accurate perception of color differences is difficult.
-
-\section{Effect of scale}~\label{sec:scale}
+\subsection{Effect of scale}~\label{sec:scale}
 

### 2011-09-28 paper.tex
Restructure sections and update intro

@@
 
-In this paper we describe the algorithm for creating glyphs-maps, and various modifications that can enable different features in the data to be detected.
-Section~\ref{sec:large-data} describes how glyph-maps scale to large climate data sets, and Section~\ref{sec:construction} explains how to construct them.
-Section~\ref{sec:reference} discusses the use of reference lines and boxes, Section~\ref{sec:perception}, their perceptual properties, and Section~\ref{sec:scale} visits scaling issues.
+We begin in Section~\ref{sec:construction} by describing the algorithm used to create glyphs-maps.
+Section~\ref{sec:perception} discusses their perceptual properties, including the importance of a visual reference grid, and of carefully considering scale.
+Section~\ref{sec:large-data} shows how glyph-maps scale to large data, and the interplay of models with glyph-maps.
 We originally developed glyph-maps for data on a rectangular grid, but many spatiotemporal data sets have irregular spatial locations.
@@
 All plots were created in R \citep{R} using the {\tt ggplot2} \citep{me:ggplot2} and {\tt plyr} \citep{me:plyr} packages.
-
-\section{Glyph-maps for large data}
-\label{sec:large-data}
-
-Climate data is usually large: many locations in space, and many points in time.
-This makes climate data challenging to work with computationally and to visualise.
-Plotting the raw data often reveals little more than a well-known observation that ``temperature is quite varied!'' Careful use of models than Careful use of models to decompose data into long-term trend, seasonal effects and error is vital for to make the problem both computationally and visually tractable.
-
-Figure~\ref{fig:gistemp-pred} shows the a glyph-map for the large GISTEMP dataset.
-Glyphs at each location represent temperature from 1880--2011, starting with Jan 1880 at 12 o'clock and proceeding clockwise to 2011 at 11:59.
-The most important feature to notice is the pattern of missing values.
-Most glyphs do not complete the full circle, typically missing points between 12 and 1, suggesting missing values in the earliest part of the data.
-Another important feature is variability, shown by the thickness of the circle.
-Some locations, particularly in the north western mountain region, have large variations in temperature.
-
-*** Need to add reference circles.
-
-\begin{figure}[htbp]
-  \centering
-  %\includegraphics[width=1\linewidth]{gistemp-pred}%
-
-  \includegraphics[width=1\linewidth]{gistemp-polar-raw}
-
-  \caption{Glyph-map of raw temperature anomaly data, from 1880-2011. Gaps in the full circle indicate missing values, and thickness of the circle indicates more variability in the measurements.}
-  \label{fig:gistemp-raw}
-\end{figure}
-
-The raw data shows us missing values and gives some sense of variability, but gives no information about long-term trend.
-We can look trend by fitting a linear model to each location and displaying the predicted values.
-Figure~\ref{fig:gistemp-pred} shows this for 1950--2010 (limited time period used due to missing data).
-Most locations have increasing temperature over this time period.
-Only a few locations show a decline.
-These locations tend to be different from their neighbors, which might suggest isolated data problems.
-The rate of increase differs across regions: steeper inclines in the north and western mountain regions and across the Great Lakes, and flatter inclines in the southeastern USA.
-
-*** Remove the strange data over ocean, and make the reference lines white?
-
-\begin{figure}[htbp]
-  \centering
-  \includegraphics[width=1\linewidth]{gistemp-pred}%
-  %\includegraphics[width=1\linewidth]{gistemp-polar-pred}
-
-  \caption{Glyph-map of predicted temperature anomaly data, from
-    1950-2010. Increasing trends can be seen over most of the USA,
-    with varying degrees of increase. Some isolated locations exhibit
-    steep declines, and inclines, perhaps indicative of some isolated
-    data problems.}
-  \label{fig:gistemp-pred}
-\end{figure}
-
-%The most seasonality occurs in north American.
-Some seasonal effects can also be seen in the Caribbean and in south American locations.
-Around the equator the temperatures are relatively warm for all seasons.
-Instead of radially displaying the temperature a glyph can be a regular time series, time displayed horizontally and temperature vertically (right plot).
-The structure perceived is different with this change in glyph type: locations at the equator have flat series, on land more varied temperatures.
-
-% \section{What to plot: raw data, trend lines, residuals,  seasonal}
-% 
-% A generalized additive model \citep{wood:2006} with a seasonal term and a long-term smooth term can be useful.
-This is revealing for smooth models of surface temperature, shown in Figure~\ref{fig:scaling}.
-Each location is modeled using a generalized additive model \citep{wood:2006} with a seasonal term and a long-term smooth term.
-The three plots in the figure show the long-term smooth term.
-% 
-% %Because some locations are uniformly warmer than others, the impact of
-% %changes across time are diminished. In the center plot, each location is
-% %scaled by dividing out the maximum, and on the right, each location is scaled
-% %to range $[0, 1]$. These emphasize individual changes but lose the ability to
-% %make global comparisons, and need caution when interpreting - a small
-% %absolute change gets blown up into a large relative change.
-% 
-% *** Need three plots here: trend, seasonality and residuals, shall we do this with NASA ozone data or stick to temperature?
-% 
-% Issues of scale relative to modeling.
-Switching from the data to model summaries brings new scaling issues.
-Often the scale of the raw data is ignored when the model is plotted, one usually shows just the line showing the trend, in the range of the trend.
-This has the effect of exaggerating the significance of the model.
-Scaling the trend lines relative to the range of the data may be useful sometimes.
 
@@
 
+\section{Glyph-maps for large data}
+\label{sec:large-data}
+
+Climate data is usually large: many locations in space, and many points in time.
+This makes climate data challenging to work with computationally and to visualise.
+Plotting the raw data often reveals little more than a well-known observation that ``temperature is quite varied!'' Careful use of models than Careful use of models to decompose data into long-term trend, seasonal effects and error is vital for to make the problem both computationally and visually tractable.
+
+Figure~\ref{fig:gistemp-pred} shows the a glyph-map for the large GISTEMP dataset.
+Glyphs at each location represent temperature from 1880--2011, starting with Jan 1880 at 12 o'clock and proceeding clockwise to 2011 at 11:59.
+The most important feature to notice is the pattern of missing values.
+Most glyphs do not complete the full circle, typically missing points between 12 and 1, suggesting missing values in the earliest part of the data.
+Another important feature is variability, shown by the thickness of the circle.
+Some locations, particularly in the north western mountain region, have large variations in temperature.
+
+*** Need to add reference circles.
+
+\begin{figure}[htbp]
+  \centering
+  %\includegraphics[width=1\linewidth]{gistemp-pred}%
+
+  \includegraphics[width=1\linewidth]{gistemp-polar-raw}
+
+  \caption{Glyph-map of raw temperature anomaly data, from 1880-2011. Gaps in the full circle indicate missing values, and thickness of the circle indicates more variability in the measurements.}
+  \label{fig:gistemp-raw}
+\end{figure}
+
+The raw data shows us missing values and gives some sense of variability, but gives no information about long-term trend.
+We can look trend by fitting a linear model to each location and displaying the predicted values.
+Figure~\ref{fig:gistemp-pred} shows this for 1950--2010 (limited time period used due to missing data).
+Most locations have increasing temperature over this time period.
+Only a few locations show a decline.
+These locations tend to be different from their neighbors, which might suggest isolated data problems.
+The rate of increase differs across regions: steeper inclines in the north and western mountain regions and across the Great Lakes, and flatter inclines in the southeastern USA.
+
+*** Remove the strange data over ocean, and make the reference lines white?
+
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=1\linewidth]{gistemp-pred}%
+  %\includegraphics[width=1\linewidth]{gistemp-polar-pred}
+
+  \caption{Glyph-map of predicted temperature anomaly data, from
+    1950-2010. Increasing trends can be seen over most of the USA,
+    with varying degrees of increase. Some isolated locations exhibit
+    steep declines, and inclines, perhaps indicative of some isolated
+    data problems.}
+  \label{fig:gistemp-pred}
+\end{figure}
+
+%The most seasonality occurs in north American.
+Some seasonal effects can also be seen in the Caribbean and in south American locations.
+Around the equator the temperatures are relatively warm for all seasons.
+Instead of radially displaying the temperature a glyph can be a regular time series, time displayed horizontally and temperature vertically (right plot).
+The structure perceived is different with this change in glyph type: locations at the equator have flat series, on land more varied temperatures.
+
+% \section{What to plot: raw data, trend lines, residuals,  seasonal}
+% 
+% A generalized additive model \citep{wood:2006} with a seasonal term and a long-term smooth term can be useful.
+This is revealing for smooth models of surface temperature, shown in Figure~\ref{fig:scaling}.
+Each location is modeled using a generalized additive model \citep{wood:2006} with a seasonal term and a long-term smooth term.
+The three plots in the figure show the long-term smooth term.
+% 
+% %Because some locations are uniformly warmer than others, the impact of
+% %changes across time are diminished. In the center plot, each location is
+% %scaled by dividing out the maximum, and on the right, each location is scaled
+% %to range $[0, 1]$. These emphasize individual changes but lose the ability to
+% %make global comparisons, and need caution when interpreting - a small
+% %absolute change gets blown up into a large relative change.
+% 
+% *** Need three plots here: trend, seasonality and residuals, shall we do this with NASA ozone data or stick to temperature?
+% 
+% Issues of scale relative to modeling.
+Switching from the data to model summaries brings new scaling issues.
+Often the scale of the raw data is ignored when the model is plotted, one usually shows just the line showing the trend, in the range of the trend.
+This has the effect of exaggerating the significance of the model.
+Scaling the trend lines relative to the range of the data may be useful sometimes.
+
 \section{Non-gridded data}~\label{sec:irregular}

### 2011-09-28 paper.tex
Reference grids and scaling

@@
 These need to be minimally perceptible, post-attentive \citep{healey} and de-emphasized, so as not to detract from the raw data.
-For example, glyph-map in Figure~\ref{fig:gistemp-pred}, contains two additional guides, a reference grid and a reference line.
-Both are drawn in white, because it is minimally perceptible relative to the black ink of the data.
-
 The \textbf{reference grid} is a box, framing each icon, which represents the spatial grid.
 The \textbf{reference line} is a horizontal line at mid-range.
-Both helps to read differences in slope and intercept.
-Slope is read by comparing the left and right endpoints of the predictions (line) against the relative location on the frame.
-Intercept is compared by reading the ``average'' position of the line in the box: is it near the top, or near the bottom?
+Both help to read differences in slope and intercept.
+Slope is read by comparing the position of left and right endpoints on the frame, or by reading the angle between the data line and the reference line.
+Intercept is read from ``average'' position of the line in the box: is it near the top, or near the bottom?
+Figure~\ref{fig:ref-basic} demonstrates both guides.
+We draw them in white, so they are minimally perceptible relative to the black data.
 
@@
 
-%Figure~\ref{fig:ref-basic} shows two reference grid variants: a line drawn at the overall mid-range, and a box drawn around each grid cell.
-We prefer using white for these references because it is minimally perceptible: you can see the grid easily when looking for it, but it does not otherwise distract from the perception of the graphic \citep{carr:1994}.
-When we add a reference to each location, as in Figure~\ref{fig:ref-basic}, we see not only a difference in shape between Northern and Southern hemispheres, but also average value.
-Without reference lines, it is difficult to see that average seasonal coefficients are much higher in the North than in the South.
-
-% Think we might need to show the original figures of data with reference lines
-
-*** Re-do these figures with zooms in on the GISTEMP data
+% *** Re-do these figures with zooms in on the GISTEMP data
 
@@
 
-  \caption{Glyph-maps of seasonal temperature patterns.
-Adding reference lines, (left) mid-range lines and (right) grid-cell boxes, makes it easier to see differences in glyph position, not just shape.}
+  \caption{Glyph-maps of seasonal temperature patterns (averages for each month), using the data-expo data.
+Adding mid-range reference lines (left) and grid-cell reference boxes (right), makes it easier to see differences in glyph position, not just shape.}
   \label{fig:ref-basic}
@@
 
-For star glyphs, the reference frames are also useful and the equivalent of the reference line is a circle. Figures~\ref{fig:templates} and \ref{fig:gistemp-raw} show examples.
+For star glyphs the reference frames are also useful and the equivalent of the reference line is a circle. Figures~\ref{fig:templates} and \ref{fig:gistemp-raw} show examples.
 
@@
 
-%*** Adding a circular reference line helps to pick out these deviations.
+\subsection{Scaling}~\label{sec:scale}
 
-\subsection{Effect of scale}~\label{sec:scale}
-
-%*** Global/local min/max, remove intercept/average, data vs model scale
 %*** This section needs to have plots re-done, once we have decided on the key pieces
 
-Scaling of the data, into the values used for each glyph, emphasizes different aspects of the data.
-For spatiotemporal data the most natural scale is to use the global minimum and maximum, because the temporal variables are measuring the same information, and we want to make comparisons across the spatial region.
-The icons will have size relative to the magnitude of the full range of the data.
-Locations with overall large temperature will catch the focus.
-Figure \ref{fig:nasa-glyph} shows data scaled to 0 to 1 based on the global minimum and maximum, using:
+We can also emphasise different features of the data by varying the scale used with each cell.
+By default, we use a global scale, so the same position within each cell corresponds to the same value in all locations.
+This facilitates the comparison of absolute values, and draws attention to plots with large overall variation.
 
-\begin{equation}
-y_{st} = \frac{y_{st}-y_{min}}{y_{max}-y_{min}} %\times 2 - 1
-\label{scale1}
-\end{equation}
-
-\noindent where $y$ is the variable used on the minor axis in equations \ref{coords.eqn}, \ref{coords.polar.eqn}, $y_{min} = \mbox{minimum}\{y_{st}; s=1, \dots S, t=1, ..., T \}, y_{max} = \mbox{maximum}\{y_{st}; s=1, \dots S, t=1, ..., T\}$, $S=$ is the number of spatial locations and $T=$ is the number of time points.
-In addition, $y_{st}$ may be further scaled by doubling and subtracting 1 to give values between -0.5 and 0.5 for using in equations \ref{coords.eqn}, \ref{coords.polar.eqn}.
-
-Alternatively, scaling locally at each spatial location enables exploring shape differences, particularly seasonal patterns, from one location to another:
-
-\begin{equation}
-y_{st} = \frac{y_{st}-y_{s,min}}{y_{s,max}-y_{s,min}} %\times 2 - 1
-\label{scale2}
-\end{equation}
-
-\noindent where $y_{s, min} = \mbox{minimum}\{y_{st}; t=1, ..., T \}, y_{s,max} = \mbox{maximum}\{y_{st}; t=1, ..., T\}$.
-With the local scaling, the shape of the icon, for example, differences in the seasonal patterns over spatial locations, are emphasized.
-Fig \ref{fig:scaling} shows glyph-maps generated using both global scaling (left) and local scaling (right) for the NASA temperature data.
-On the global scale, the El Ni\~no blip in temperature is primarily visible, along with the generally flat series over most water locations, and somewhat increasing or variable temperatures over land.
-The local scale emphasizes the individual shapes: The El Ni\~no pattern is visible, but extend to a larger spatial region including the Caribbean, dramatic linear increases across the Andes, decreases in the Gulf of Mexico.
-(The local scale is equivalent to using individual scales in the facetted graphics of \texttt{ggplot2}).
-One caution, with local scale, the viewer needs to realize that the scale is not relative, big patterns in some locations might be just tiny effects.
-To indicate this on the locally scaled data plots, using a very pale coloring to code the original scale in the tiles, might help puts patterns in perspective.
-
-Other types of shifting and scaling can be useful.
-The mean and standard deviation might be used to standardize values, or the median and interquartile range, might be useful with some types of data.
-Shifting the values at each location to have mean zero, places more focus on the trend.
-This is particularly important for the star glyphs, where differences in the average value lead to unintuitive patterns in the resulting glyphs (eg Figure \ref{fig:cycle} star glyphs obscure the seasonal pattern).
-In this situation we would use $y_{st}-\bar{y}_s$ instead of $y_{st}$ in equations \ref{scale1}, \ref{scale2}, with minima and maxima calculated on the mean-shifted data.
-
-%Another way to think about these scale options is partitioning the data structure into shape, magnitude and shift components.
-The different ways to scale emphasize these different components.
-The shift component can be extracted from the data in different ways, using the intercept from a linear model, for example, instead of the average.
-
-%**** This is different - I think it just confuses the topic of scaling.
-Because the GAM is not used to rescale the data, you still have to do that using the local or global scling equations above.
-Moving the discussion to ``What to plot''.
+Alternatively, we can re-scale values within each location prior to plotting.
+In local scaling, the values within each location are scaled to range $[0, 1]$ before plot construction.
+This makes it easier to compare shape (ignoring amplitude), and draws attention to locations with large relative variation.
+Figure~\ref{fig:scaling} compares global scaling and local scaling for the smoothed temperature data from the data-expo data.
+On the global scale, the El Ni\~no blip in temperature is just visible, along with variation in the average temperature at each location.
+Local scaling emphasizes the individual shapes: we see the impact of El Ni\~no over a wider region, dramatic linear increases across the Andes, and decreases in the Gulf of Mexico.
 
@@
   \centering
-  %\includegraphics[width=0.33\linewidth]{month-rescale-none}%
-  \includegraphics[width=0.5\linewidth]{month-rescale-max}%
+  \includegraphics[width=0.5\linewidth]{month-rescale-none}%
+  % \includegraphics[width=0.5\linewidth]{month-rescale-max}%
   \includegraphics[width=0.5\linewidth]{month-rescale01}
-  \caption{Smoothed de-seasonalized daily temperature glyph-map. (Left) %unscaled, (center) 
+  \caption{Smoothed de-seasonalized daily temperature glyph-map. (Left)
 globally scaled, (right) locally scaled.}
@@
 
-%A possible compromise is to use line colour to display the standardising variable, as shown in Figure~\ref{fig:scaling-col}.
-This makes it possible to see the individual patterns while still retaining some ability to see which shapes represent small absolute changes.
+% 
+% \begin{equation}
+% y_{st} = \frac{y_{st}-y_{min}}{y_{max}-y_{min}} %\times 2 - 1
+% \label{scale1}
+% \end{equation}
+% 
+% \noindent where $y$ is the variable used on the minor axis in equations \ref{coords.eqn}, \ref{coords.polar.eqn}, $y_{min} = \mbox{minimum}\{y_{st}; s=1, \dots S, t=1, ..., T \}, y_{max} = \mbox{maximum}\{y_{st}; s=1, \dots S, t=1, ..., T\}$, $S=$ is the number of spatial locations and $T=$ is the number of time points.
+In addition, $y_{st}$ may be further scaled by doubling and subtracting 1 to give values between -0.5 and 0.5 for using in equations \ref{coords.eqn}, \ref{coords.polar.eqn}.
+% 
+% \begin{equation}
+% y_{st} = \frac{y_{st}-y_{s,min}}{y_{s,max}-y_{s,min}} %\times 2 - 1
+% \label{scale2}
+% \end{equation}
+% 
+% \noindent where $y_{s, min} = \mbox{minimum}\{y_{st}; t=1, ..., T \}, y_{s,max} = \mbox{maximum}\{y_{st}; t=1, ..., T\}$.
 
-%\begin{figure}[htbp]
-%  \centering
-%  \includegraphics[width=0.33\linewidth]{month-rescale01-col}
-%  \caption{Adding colour to scaled plots. Here each location has been scaled to range $[0, 1]$ and colour mapped to the range of the original predictions.}
-%  \label{fig:scaling-col}
-%\end{figure}
+Other types of shifting and scaling can be useful:
 
-%Other types of scaling may be interesting to explore, such as a classical scaling using the mean and standard deviation.
+\begin{itemize}
 
-%Equivalent to using individual scales in facetted graphics.
+  \item The mean and standard deviation (or robust equivalents) could be used
+  to standardize values to common moments.
 
-One more important point: if a model summary is displayed in the glyph-map, such as the predicted values, it may be important to force the scale to that of the original data.
-Displaying the predicted values in the absence of the raw data scale can skew the interpretation of the magnitude of the effect.
+   \item Scaling to maximum 1 (but leaving the minimum unchanged) can help
+  emphasis the relative shape, but not distort the range as much as rescaling
+  to range $[0, 1]$.
 
-\section{Glyph-maps for large data}
+  \item Shifting the values at each location to have mean zero places more
+  focus on the trend. This is particularly important for the star glyphs,
+  where differences in the average value lead to unintuitive patterns in the
+  resulting glyphs.
+
+\end{itemize}
+
+Note that the use of local scales needs to be clearly marked.
+The viewer must realize that the scale is not relative and that big patterns in some locations might be just tiny effects.
+To indicate this on the locally scaled data plots, one can use an additional aesthetic, like colour, to encode the original scale.
+This is illustrated in Figure~\ref{fig:scaling-col}.
+
+\begin{figure}[htbp]
+ \centering
+ \includegraphics[width=0.33\linewidth]{month-rescale01-col}
+ \caption{Adding colour to scaled plots.
+Here each location has been scaled to range $[0, 1]$ and colour mapped to the range of the original predictions.
+Pale locations have smaller range than darker locations.
+Shapes at all locations are visible, but attention is drawn to locations with larger ranges.}
+ \label{fig:scaling-col}
+\end{figure}
+
+One more important point: if a model summary is displayed in the glyph-map, as discussed in the following section, it may be important to force the scale to that of the original data.
+The range of predicted values is typically smaller than the raw data, so angles of slopes is increase.
+
+% Displaying the predicted values in the absence of the raw data scale typically increases magnitude because the variation is much lower.
+
+\section{Large data and models}
 \label{sec:large-data}

### 2011-09-28 paper.tex
fixed a couple of typos, added more to data explanation

@@
 When measurements are made at multiple time points it is common to display the data with small multiples, such as in Figure~\ref{fig:nasa-facet}.
-In this plot, a separate map is drawn for each month (columns) and year (rows), over a 6 year period of remotely sensed temperature data above Central America \citet{murrell:2010}.
+In this plot, a separate map is drawn for each month (columns) and year (rows), over a 6 year period of remotely sensed temperature data above Central America \citep{murrell:2010}.
 Color is used to display de-seasonalized temperature, with red mapped to high values and blue to low.
@@
   \label{fig:nasa-facet}
+\end{figure*}
+
+\begin{figure*}[htbp]
+  \centering
+  \includegraphics[width=3.5in]{nasa-deseas-glyph}
+  \caption{Glyph-map of de-seasonalized temperature, to compare with Figure \ref{fig:nasa-facet}. }
+  \label{fig:nasa-glyph}
 \end{figure*}
@@
 The structure perceived is different with this change in glyph type: locations at the equator have flat series, on land more varied temperatures.
-
-\begin{figure*}[htbp]
-  \centering
-  \includegraphics[width=3.5in]{nasa-deseas-glyph}
-  \caption{Glyph-map of de-seasonalized temperature, to compare with Figure \ref{fig:nasa-facet}. }
-  \label{fig:nasa-glyph}
-\end{figure*}
 
@@
 
-  \item GISTEMP surface temperature data provided by the Earth System Research
-  Laboratory, Physical Sciences Division, National Oceanic and Atmospheric
-  Administration. This data is provided on $2^{\circ}$ x $2^{\circ}$ grid over
-  the entire globe, continental United States (XXX locations), measured
-  monthly. Ground station data was de-seasonalized, differenced from from the
-  1951-1980 temperature averages, and spatially averaged to obtain gridded
-  measurements. We extracted the locations corresponding to the continental
-  USA. Provided by the NOAA/OAR/ESRL PSD, Boulder, Colorado,
-  USA, from their web site at \url{http://www.esrl.noaa.gov/psd/}.
+\item GISTEMP surface temperature data provided on $2^{\circ}$ x
+  $2^{\circ}$ grid over the entire globe, measured monthly. Ground
+  station data was de-seasonalized, differenced from from the
+  1951-1980 temperature averages, and spatially averaged to obtain
+  gridded measurements. For the purposes of this paper, we extracted
+  the locations corresponding to the continental USA. The data is made
+  available by the Earth System Research Laboratory, Physical Sciences
+  Division, National Oceanic and Atmospheric Administration, from
+  their web site at \url{http://www.esrl.noaa.gov/psd/}.
 

### 2011-09-28 paper.tex
spell check, and fig 2 caption

@@
 
-Climate data is challenging to visualise because it is multivariate and has both spatial (2d) and temporal (1d) components.
+Climate data is challenging to visualize because it is multivariate and has both spatial (2d) and temporal (1d) components.
 In this paper we propose an new display for climate data: glyph-maps.
-Glyph-maps are a specialisation of multivariate glyph plots; each spatial location is display with one glyph that represents the measurements recorded over time at that location.
-Glyph-maps allow the discovery of both local and global structure, making discovery much easier than the classical display using colours.
+Glyph-maps are a specialization of multivariate glyph plots; each spatial location is display with one glyph that represents the measurements recorded over time at that location.
+Glyph-maps allow the discovery of both local and global structure, making discovery much easier than the classical display using colors.
 
 We include a description of how to create glyph-maps using existing statistical software, our initial explorations of their perceptual features, and a discussion of reference frames.
-Our techniques are inspired by rectangular gridded data, but we provide generalisation to deal with non-rectangular grids and irregular locations.
+Our techniques are inspired by rectangular gridded data, but we provide generalization to deal with non-rectangular grids and irregular locations.
 
@@
   \includegraphics[width=3.5in]{nasa-deseas-glyph}
-  \caption{Glyph-map of de-seasonalized temperature, to compare with Figure \ref{fig:nasa-facet}. }
+  \caption{Glyph-map of de-seasonalized temperature, to compare with Figure \ref{fig:nasa-facet}.
+Time series of the six years of monthly temperature are plotted at each spatial grid location.
+The El Ni\~no appears as a bump in the middle of the time series, in the equatorial Pacific region.
+Large variations in temperature can be seen in areas over land, while being fairly constant over water.
+}
   \label{fig:nasa-glyph}
@@
 
-We can also emphasise different features of the data by varying the scale used with each cell.
+We can also emphasize different features of the data by varying the scale used with each cell.
 By default, we use a global scale, so the same position within each cell corresponds to the same value in all locations.
@@
 Climate data is usually large: many locations in space, and many points in time.
-This makes climate data challenging to work with computationally and to visualise.
+This makes climate data challenging to work with computationally and to visualize.
 Plotting the raw data often reveals little more than a well-known observation that ``temperature is quite varied!'' Careful use of models than Careful use of models to decompose data into long-term trend, seasonal effects and error is vital for to make the problem both computationally and visually tractable.

### 2011-09-28 paper.tex
typos

@@
 \date{}
+
+% Remove words like "obvious", "challenge", "scratched the surface"
+% Make data URLs references
 
@@
 This makes climate data challenging to work with computationally and to visualize.
-Plotting the raw data often reveals little more than a well-known observation that ``temperature is quite varied!'' Careful use of models than Careful use of models to decompose data into long-term trend, seasonal effects and error is vital for to make the problem both computationally and visually tractable.
+Plotting the raw data often reveals little more than a well-known observation that ``temperature is quite varied!'' Careful use of models to decompose data into long-term trend, seasonal effects and error is vital for to make the problem both computationally and visually tractable.
 

### 2011-09-28 code/modeling.r
Fill colour reference for fig 6

@@ -174,6 +174,14 @@ ggplot(day_preds2, aes(gx, gy, group = gid)) +
   scale_colour_gradient(high = "black", low = "grey60")
 ggsave("../images/month-rescale01-col.png", width = 4, height = 4)
 
+grid <- unique(day_preds2[c("lat", "long", "range")])
+
+ggplot(day_preds2) + 
+  map + 
+  geom_tile(aes(long, lat, fill = range), data = grid, alpha = 0.5) +
+  geom_path(aes(gx, gy, group = gid)) +
+  theme_fullframe()
+ggsave("../images/month-rescale01-fill.png", width = 4, height = 4)
 
 # Residuals ------------------------------------------------------------------
 

### 2011-09-28 paper.tex
Fill colour reference for fig 6
equations fixed in write-up

@@
 \title{Glyph-maps for Visualizing Climate Data and Models}
-\author{Dianne Cook$^1$, Heike Hofmann$^1$, Charlotte Wickham$^2$, Hadley Wickham$^3$, \\
-$^1$Department of Statistics, Iowa State University\\
-$^2$ Department of Statistics, Oregon State University\\
-$^3$ Department of Statistics, Rice University}
+\author{Hadley Wickham$^1$, Heike Hofmann$^2$, Charlotte Wickham$^2$, Dianne Cook$^2$\\
+$^1$ Department of Statistics, Rice University}
+$^2$Department of Statistics, Iowa State University\\
+$^3$ Department of Statistics, Oregon State University\\
 \date{}
@@
 % Make data URLs references
+% Do we have different aspect ratios between the different US maps?
+% Check where data is described, just use key names given in the description at beginning of paper
 
@@
 Glyph-maps can be generated with existing graphics software after performing a simple pre-processing step, making them easily accessible to a wide audience.
-Creating a glyph-map requires a recognition they combine two structural components of the data: spatial location and data values.
-The spatial location is the major positioning component, and the data values are minor adjustments to those positions.
+Creating a glyph-map requires a recognition of the two structural components of the data: spatial location and data values.
+The spatial location is the major positioning component, while the data values are minor adjustments to those positions.
 For spatiotemporal data, the major axes are latitude ($y_{\amaj}$) and longitude ($x_{\amaj}$), and the minor axes are time ($x_{\amin}$) and some measurement ($y_{\amin}$), for example, temperature, or predicted temperature.
-Assuming the minor axes are rescaled to $[-1, 1]$, the final positions are a simple linear combination:
+Assuming the minor axes are rescaled to $[-1, 1]$, the final coordinates $(x,y)$ on the chart are a simple linear combination:
 
@@
   \begin{array}{lll}
-  x_\text{ coordinate}&=& x_{\amaj} + w \cdot x_{\amin}\\
-  y_\text{ coordinate}&=& y_{\amaj} + h \cdot y_{\amin}, 
+  x &=& x_{\amaj} + \frac{w}{2} \cdot x_{\amin}\\
+  y &=& y_{\amaj} + \frac{h}{2} \cdot y_{\amin}, 
   \end{array}
@@
 
-\noindent where $w$ and $h$ are width and height respectively.
-For gridded data, the width and height be the resolution of the major axes, that is, the minimum difference between consecutive locations.
+\noindent where $w$ and $h$ are maximal width and height of the glyph, respectively.
+For gridded data, it is convenient to make use of the resolution of the major axes, that is, the minimum difference between neighboring locations.
+
 Since the coordinates are linear combinations of two variables, it is possible to build these these line-glyphs interactively, illustrated by software packages that have incorporated tours \citep{cook:2006}, such as DataViewer \citep{buja:1986}, XGobi \citep{swayne:1991} or GGobi \citep{swayne:2003}.
@@
   \begin{array}{lll}
-  x_\text{ coordinate}&=& x_{\amaj} + y_{\amin} \cdot \frac{w}{2} \sin(2 \pi x_{\amin}) \\
-  y_\text{ coordinate}&=& y_{\amaj} + y_{\amin} \cdot \frac{h}{2} \cos(2 \pi x_{\amin}).
+  x &=& x_{\amaj} + y_{\amin} \cdot \frac{w}{2} \sin(2 \pi x_{\amin}) \\
+  y &=& y_{\amaj} + y_{\amin} \cdot \frac{h}{2} \cos(2 \pi x_{\amin}).
   \end{array}
@@
 The viewer must realize that the scale is not relative and that big patterns in some locations might be just tiny effects.
-To indicate this on the locally scaled data plots, one can use an additional aesthetic, like colour, to encode the original scale.
+To indicate this on the locally scaled data plots, one can use an additional aesthetic, like colour, to encode the range of original scale.
 This is illustrated in Figure~\ref{fig:scaling-col}.
@@
  \centering
- \includegraphics[width=0.33\linewidth]{month-rescale01-col}
+ \includegraphics[width=0.5\linewidth]{month-rescale01-col}%
+ \includegraphics[width=0.5\linewidth]{month-rescale01-fill}
  \caption{Adding colour to scaled plots.
 Here each location has been scaled to range $[0, 1]$ and colour mapped to the range of the original predictions.
-Pale locations have smaller range than darker locations.
+(Left) Range mapped to colour of line: pale locations have smaller range than darker locations.
+(Right) Range mapped to fill colour of grid box: blue boxes have larger ranges.
 Shapes at all locations are visible, but attention is drawn to locations with larger ranges.}
@@
 
-*** Need to add reference circles.
-
 \begin{figure}[htbp]
@@
 The rate of increase differs across regions: steeper inclines in the north and western mountain regions and across the Great Lakes, and flatter inclines in the southeastern USA.
-
-*** Remove the strange data over ocean, and make the reference lines white?
 
@@
 When applying a projection to the spatial locations and then constructing a glyph-map, the grid becomes irregular.
-
-*** Example?
 

### 2011-09-28 paper.tex
re-done fig 7, order of names

--- 2011-09-28 paper.tex
@@
  \centering
- \includegraphics[width=0.33\linewidth]{month-rescale01-col}
+ \includegraphics[width=0.5\linewidth]{month-rescale01-col}%
+ \includegraphics[width=0.5\linewidth]{month-rescale01-fill}
  \caption{Adding colour to scaled plots.

### 2011-09-28 code/maps.r

@@ -31,6 +31,9 @@ states$group <- max(world$group) + states$group
 both <- rbind(world, states)
 both <- getbox(both, xlim = c(-126, -65), ylim = c(24, 50))
 
+both.gistemp <- rbind(world, states)
+both.gistemp <- getbox(both, xlim = c(-140, -55), ylim = c(20, 60))
+
 # Enter the coords of the data for the limits
 world <- getbox(world, xlim = c(-114.93, -55.07), ylim = c(-21.53, 36.64))
 
@@ -56,5 +59,12 @@ map_ushcn <- list(
   xlab(NULL),
   ylab(NULL))
 
+map_gistemp <- list(
+  geom_polygon(aes(long, lat, group = group), inherit.aes = FALSE, 
+    data = subset(both.gistemp, region != "Great Lakes"), legend = FALSE, fill = "grey80", colour = "grey90"),
+  scale_x_continuous(breaks = NA, expand = c(0.02, 0)),
+  scale_y_continuous(breaks = NA, expand = c(0.02, 0)), 
+  xlab(NULL),
+  ylab(NULL))
 
 

### 2011-09-28 gridded.temp.r

@@ -63,6 +63,8 @@ colnames(temp.usa.melt)<-c("gridx", "gridy", "time", "temp")
 temp.usa.melt$lon<-lon[temp.usa.melt$gridx+113]
 temp.usa.melt$lat<-lat[temp.usa.melt$gridy+19]
 
+temp.usa.melt<-subset(temp.usa.melt, !(lon>-70 & lat<37))
+
 temp.usa.melt.gly<-glyphs(temp.usa.melt, "lon", "time", "lat", "temp")
 qplot(gx, gy, data=temp.usa.melt.gly, group=gid, geom="line") + map + coord_map() # Testing
 p <-ggplot(temp.usa.melt.gly, aes(gx, gy, group = gid)) + 
@@ -75,9 +77,9 @@ ggsave("../images/gistemp-raw.png", width = 8, height = 4)
 
 temp.usa.melt.gly<-glyphs(temp.usa.melt, "lon", "time", "lat", "temp", polar=T)
 p <- ggplot(temp.usa.melt.gly, aes(gx, gy, group = gid)) + 
-  map  
+  map_gistemp
+p <- p + add_ref_lines(temp.usa.melt.gly) + add_ref_boxes(temp.usa.melt.gly)
 p <- p + geom_path() 
-p <- p +  ref_boxes 
 #  theme_fullframe() + coord_map()
 p
 ggsave("../images/gistemp-polar-raw.png", width = 8, height = 4)

### 2011-09-28 paper.tex
removed *** comments
Tweaks to fig 6

@@
 Here each location has been scaled to range $[0, 1]$ and colour mapped to the range of the original predictions.
-Pale locations have smaller range than darker locations.
+(Left) Range mapped to colour of line: pale locations have smaller range than darker locations.
+(Right) Range mapped to fill colour of grid box: blue boxes have larger ranges.
 Shapes at all locations are visible, but attention is drawn to locations with larger ranges.}

### 2011-09-28 code/modeling.r

@@ -180,7 +180,8 @@ ggplot(day_preds2) +
   map + 
   geom_tile(aes(long, lat, fill = range), data = grid, alpha = 0.5) +
   geom_path(aes(gx, gy, group = gid)) +
-  theme_fullframe()
+  theme_fullframe() + 
+  scale_fill_gradient(low = "white", high = "#3B4FB8")
 ggsave("../images/month-rescale01-fill.png", width = 4, height = 4)
 
 # Residuals ------------------------------------------------------------------

### 2011-09-28 paper.tex
Fix authors bug even though I don't think I should be first author

--- 2011-09-28 paper.tex
@@
 \author{Hadley Wickham$^1$, Heike Hofmann$^2$, Charlotte Wickham$^2$, Dianne Cook$^2$\\
-$^1$ Department of Statistics, Rice University}
+$^1$Department of Statistics, Rice University\\
 $^2$Department of Statistics, Iowa State University\\
-$^3$ Department of Statistics, Oregon State University\\
+$^3$Department of Statistics, Oregon State University}
 \date{}

### 2011-09-28 code/glyph-examples.r
fixed some missing sources

@@ -2,6 +2,7 @@
 # putting all of the glyph-map examples, using gistemp and ushcn data
 # here too
 
+source("glyph.r")
 source("maps.r")
 # world, map are created in maps.r
 
@@ -10,8 +11,8 @@ source("maps.r")
 # to run that code before this
 resids.gly <- glyphs(resids, "long", "day", "lat", "temp_resid", height=2.5) 
 ggplot(resids.gly, aes(gx, gy, group = gid)) + map +
-  ref_lines(resids.gly) +
-  ref_boxes(resids.gly) +
+  add_ref_lines(resids.gly) +
+  add_ref_boxes(resids.gly) +
   geom_path() + theme_fullframe() + coord_map()
 #qplot(gx, gy, data = resids, geom = "line", group = gid) + map
 ggsave("../images/nasa-deseas-glyph.png")

### 2011-09-28 code/glyph-utils.r
missing sources included

@@ -2,6 +2,7 @@ library("maps")
 library("ggplot2")
 
 source("maps.r")
+source("glyph.r")
 
 theme_fullframe <- function (base_size = 12){
   structure(list(

### 2011-09-28 code/maps.r
separated world from map of central america

@@ -31,10 +31,11 @@ states$group <- max(world$group) + states$group
 both <- rbind(world, states)
 both <- getbox(both, xlim = c(-126, -65), ylim = c(24, 52))
 
+central <- getbox(world, xlim=c(-115.05217, -54.94783), ylim = c(-21.81739, 36.81739))
 ##### Maps
 map <- list(
   geom_polygon(aes(long, lat, group = group), inherit.aes = FALSE, 
-    data = world, legend = FALSE, fill = "grey80", colour = "grey90"),
+    data = central, legend = FALSE, fill = "grey80", colour = "grey90"),
   scale_x_continuous(breaks = NA, expand = c(0.02, 0)),
   scale_y_continuous(breaks = NA, expand = c(0.02, 0)), 
   xlab(NULL),

### 2011-09-28 paper.tex

@@
 \title{Glyph-maps for Visualizing Climate Data and Models}
-\author{Hadley Wickham$^1$, Heike Hofmann$^2$, Charlotte Wickham$^2$, Dianne Cook$^2$\\
+\author{Hadley Wickham$^1$, Heike Hofmann$^2$, Charlotte Wickham$^3$, Dianne Cook$^2$\\
 $^1$Department of Statistics, Rice University\\

### 2011-09-28 code/glyph-examples.r
removed coord_map

@@ -13,7 +13,7 @@ resids.gly <- glyphs(resids, "long", "day", "lat", "temp_resid", height=2.5)
 ggplot(resids.gly, aes(gx, gy, group = gid)) + map +
   add_ref_lines(resids.gly) +
   add_ref_boxes(resids.gly) +
-  geom_path() + theme_fullframe() + coord_map()
+  geom_path() + theme_fullframe() 
 #qplot(gx, gy, data = resids, geom = "line", group = gid) + map
 ggsave("../images/nasa-deseas-glyph.png")
 
@@ -25,12 +25,12 @@ nasa.scale<-ddply(nasa, c("long","lat"), summarise,
 #nasa.scat.gly <- glyphs(nasa.scale, "long", "cloudlow.s", "lat", "cloudhigh.s") 
 #ggplot(nasa.scat.gly, aes(gx, gy, group = gid)) + map +
 #  add_ref_boxes(nasa.scat.gly) +
-#  geom_point(size=I(0.7), alpha=I(0.5)) + theme_fullframe() + coord_map()
+#  geom_point(size=I(0.7), alpha=I(0.5)) + theme_fullframe() 
 #ggsave("../images/nasa-scat-glyph.png")
 nasa.scat.gly <- glyphs(nasa.scale, "long", "temp.s", "lat", "cloudhigh.s") 
 ggplot(nasa.scat.gly, aes(gx, gy, group = gid)) + map +
   add_ref_boxes(nasa.scat.gly) +
-  geom_point(size=I(0.7), alpha=I(0.5)) + theme_fullframe() + coord_map()
+  geom_point(size=I(0.7), alpha=I(0.5)) + theme_fullframe() 
 ggsave("../images/nasa-scat-glyph.png")
 nasa.loess<-ddply(nasa.scale, c("long","lat"), summarise,
                   temp.s=temp.s,
@@ -38,5 +38,5 @@ nasa.loess<-ddply(nasa.scale, c("long","lat"), summarise,
 nasa.loess.gly <- glyphs(nasa.loess, "long", "temp.s", "lat", "pcloud") 
 ggplot(nasa.loess.gly, aes(gx, gy, group = gid)) + map +
   add_ref_boxes(nasa.loess.gly) +
-  geom_line() + theme_fullframe() + coord_map()
+  geom_line() + theme_fullframe() 
 ggsave("../images/nasa-loess-glyph.png")

### 2011-09-28 code/data-nasa.r
Fixes for fig 5

@@ -1,8 +1,3 @@
-source("glyph-utils.r")
-source("maps.r")
-
-library("ggplot2")
-
 if (!file.exists("../data/nasadata.rds")) {
   library("lubridate")
 
@@ -16,35 +11,3 @@ if (!file.exists("../data/nasadata.rds")) {
 } else {
   nasa <- readRDS("../data/nasadata.rds")
 }
-
-  
-theme_fullframe <- function (base_size = 12){
-  structure(list(
-    axis.line = theme_blank(), 
-    axis.text.x = theme_blank(), 
-    axis.text.y = theme_blank(),
-    axis.ticks = theme_blank(), 
-    axis.title.x = theme_blank(), 
-    axis.title.y = theme_blank(), 
-    axis.ticks.length = unit(0, "lines"), 
-    axis.ticks.margin = unit(0, "lines"), 
-    legend.position = "none", 
-    panel.border = theme_blank(), 
-    panel.grid.major = theme_blank(), 
-    panel.grid.minor = theme_blank(), 
-    panel.margin = unit(0, "lines"), 
-    plot.background = theme_blank(), 
-    plot.margin = unit(rep(2, 4), "mm")
-  ), class = "options")
-}
-
-ref_boxes <- list(
-  geom_tile(aes(lon, lat), colour = "white", fill = NA, inherit.aes = FALSE),
-  scale_x_continuous(breaks = NA, expand = c(0, 0)),
-  scale_y_continuous(breaks = NA, expand = c(0, 0))
-)
-
-
-map_mini <- c(map, list(
-  coord_cartesian(xlim = c(-114.93, -85), ylim = c(7.55, 36.64), wise = T)))
-

### 2011-09-28 code/maps.r

@@ -29,26 +29,22 @@ world <- map_data("world")
 states <- map_data("state")
 states$group <- max(world$group) + states$group
 both <- rbind(world, states)
-both <- getbox(both, xlim = c(-126, -65), ylim = c(24, 52))
 
-central <- getbox(world, xlim=c(-115.05217, -54.94783), ylim = c(-21.81739, 36.81739))
 ##### Maps
-map <- list(
+nasa <- getbox(both, xlim = c(-115, -55), ylim = c(-21.1, 36.6))
+map_nasa <- list(
   geom_polygon(aes(long, lat, group = group), inherit.aes = FALSE, 
-    data = central, legend = FALSE, fill = "grey80", colour = "grey90"),
+    data = nasa, legend = FALSE, fill = "grey80", colour = "grey90"),
   scale_x_continuous(breaks = NA, expand = c(0.02, 0)),
   scale_y_continuous(breaks = NA, expand = c(0.02, 0)), 
   xlab(NULL),
   ylab(NULL))
 
-
-map_mini <- c(map, list(
-  coord_cartesian(xlim = c(-114.93, -85), ylim = c(7.55, 36.64), wise = T)))
-
-
+ushcn <- subset(getbox(both, xlim = c(-126, -65), ylim = c(24, 52)),
+  region != "Great Lakes")
 map_ushcn <- list(
   geom_polygon(aes(long, lat, group = group), inherit.aes = FALSE, 
-    data = subset(both, region != "Great Lakes"), legend = FALSE, fill = "grey80", colour = "grey90"),
+    data = ushcn, legend = FALSE, fill = "grey80", colour = "grey90"),
   scale_x_continuous(breaks = NA, expand = c(0.02, 0)),
   scale_y_continuous(breaks = NA, expand = c(0.02, 0)), 
   xlab(NULL),

### 2011-09-28 code/modeling.r

@@ -2,37 +2,11 @@ library("ggplot2")
 library("plyr")
 library("mgcv")
 
-source("glyph.r")
 source("data-nasa.r")
+source("glyph.r")
+source("glyph-utils.r")
 source("cache.r")
-
-# Scatterplots ----------------------------------------------------------------
-
-cloud <- glyphs(nasa, "long", "cloudlow", "lat", "cloudhigh") 
-ggplot(cloud, aes(gx, gy, group = gid)) + 
-  map +
-  geom_point(shape = ".") +
-  geom_tile(aes(long, lat), colour = "white", fill = NA)
-ggsave("../images/clouds.png", width = 6, height = 6)
-
-grid_along <- function(x, n = 20) {
-  rng <- range(x, na.rm = TRUE)
-  seq(rng[1], rng[2], length = n)
-}
-
-cloud_mod <- ddply(nasa, c("lat", "long"), function(df) {
-  grid_low <- data.frame(cloudlow = grid_along(df$cloudlow))
-  mod <- loess(cloudhigh ~ cloudlow, data = df, na = na.exclude)
-  mutate(grid_low, cloudhigh = predict(mod, newdata = grid_low))
-})
-cloud_mod <- glyphs(cloud_mod, "long", "cloudlow", "lat", "cloudhigh") 
-
-ggplot(cloud_mod, aes(gx, gy, group = gid)) + 
-  map +
-  geom_line() +
-  geom_tile(aes(long, lat), colour = "white", fill = NA) 
-ggsave("../images/clouds-smooth.png", width = 6, height = 6)
-
+source("maps.r")
 
 # Model fitting --------------------------------------------------------------
 
@@ -62,6 +36,46 @@ day_preds <- ldply(temp_smooth, function(mod) {
   day_grid
 })
 
+# Rescale predictions to individual scales -----------------------------------
+
+day_preds <- glyphs(day_preds, "long", "day", "lat", "pred") 
+ggplot(day_preds, aes(gx, gy, group = gid)) + 
+  map_nasa + 
+  add_ref_boxes(day_preds) +
+  geom_path() +
+  theme_fullframe()
+ggsave("../images/month-rescale-none.png", width = 4, height = 4)
+
+last_plot() %+% glyphs(day_preds, "long", "day", "lat", "pred", 
+  y_scale = range01) 
+ggsave("../images/month-rescale01.png", width = 4, height = 4)
+
[…]

### 2011-09-28 code/modeling.r
Fixes for fig 6

@@ -59,9 +59,9 @@ day_preds2 <- ddply(day_preds, c("lat", "long"), mutate,
 
 day_preds2 <- glyphs(day_preds2, "long", "day", "lat", "pred_s") 
 ggplot(day_preds2, aes(gx, gy, group = gid)) + 
-  map + 
+  map_nasa + 
+  add_ref_boxes(day_preds) +
   geom_path(aes(colour = range)) +
-  ref_boxes +
   theme_fullframe() + 
   scale_colour_gradient(high = "black", low = "grey60")
 ggsave("../images/month-rescale01-col.png", width = 4, height = 4)
@@ -69,11 +69,11 @@ ggsave("../images/month-rescale01-col.png", width = 4, height = 4)
 grid <- unique(day_preds2[c("lat", "long", "range")])
 
 ggplot(day_preds2) + 
-  map + 
+  map_nasa + 
   geom_tile(aes(long, lat, fill = range), data = grid, alpha = 0.5) +
   geom_path(aes(gx, gy, group = gid)) +
   theme_fullframe() + 
-  scale_fill_gradient(low = "white", high = "#3B4FB8")
+  scale_fill_gradient(high = "white", low = "#3B4FB8")
 ggsave("../images/month-rescale01-fill.png", width = 4, height = 4)
 
 

### 2011-09-28 paper.tex
Fixes for fig 6
Add note about Zollner optical illusion

@@
 (Left) Range mapped to colour of line: pale locations have smaller range than darker locations.
-(Right) Range mapped to fill colour of grid box: blue boxes have larger ranges.
+(Right) Range mapped to fill colour of grid box: blue boxes have smaller ranges.
 Shapes at all locations are visible, but attention is drawn to locations with larger ranges.}
@@
     steep declines, and inclines, perhaps indicative of some isolated
-    data problems.}
+    data problems. References line are straight, but look curves because of
+    the Z\"ollner optical illusion}
   \label{fig:gistemp-pred}

### 2011-09-29 paper.tex
abstract and conclusions

@@
 
-Climate data is challenging to visualize because it is multivariate and has both spatial (2d) and temporal (1d) components.
+Climate data poses visual challenges because it is has both spatial and temporal components, and is often multivariate.
 In this paper we propose an new display for climate data: glyph-maps.
-Glyph-maps are a specialization of multivariate glyph plots; each spatial location is display with one glyph that represents the measurements recorded over time at that location.
-Glyph-maps allow the discovery of both local and global structure, making discovery much easier than the classical display using colors.
+Glyph-maps are a specialization of multivariate glyph plots; each spatial location is displayed with one glyph that represents the measurements recorded over time at that location.
+Glyph-maps allow the discovery of both local and global structure, with a particular focus on temporal relationships.
 
-We include a description of how to create glyph-maps using existing statistical software, our initial explorations of their perceptual features, and a discussion of reference frames.
-Our techniques are inspired by rectangular gridded data, but we provide generalization to deal with non-rectangular grids and irregular locations.
+The instructions for computing and displaying glyph-maps are described, along with the design of reference frames, and the effects of different scaling of the data on perception of structure.
+The methods are developed for rectangularly gridded data, but with some clever processing, which is explained, it is possible get good glyph-maps of irregularly gridded data.
 
@@
 
-\section{Conclusions}
+\section{Generalizations}
 
-We have only scratched the surface of the potential of this type of display for climate data.
-Glyph-maps are not limited to time series plots.
-They are readily generalized to use any other type of display as an icon.
+Glyphs can be more general than time series or stars.
+For multivariate data we may want to represent several variables in the glyphs.
 Figure~\ref{fig:cloud} shows scatterplots as icons: temperature values are displayed horizontally and high-cloud values are displayed vertically, both locally scaled.
@@
 
-There is also much work to be done on the perceptual underpinnings of these displays.
-We have noticed that periodic trends are easier to perceive in star-glyphs, and long-term trends in line-glyphs.
+\section{Conclusions}
+
+In summary, glyph-maps enable the study of temporal patterns in multivariate spatiotemporal data.
+Climate change is focused on changes over time, so glyph-maps provide a way to study these changes directly.
+Glyph-maps enable different resolutions of the data to be examined: the raw data to discover data quality issues, early, global trend, seasonality, residuals, and multivariate dependencies.
+When the data is spatially gridded the glyphs are organized in a manner that makes reasonable comparisons.
+For irregularly gridded data the icon size needs to be chosen to minimize overlap but use as much display space as possible, or combined to produce close to gridded icons.
+
+The next direction would be to examine the perceptual elements of these displays, in more detail.
+For example, it appears that periodic trends are easier to perceive in star-glyphs, and long-term trends in line-glyphs.
 Is this always the case?

### 2011-09-29 paper.tex
data refs, nz to us spelling

@@
 When measurements are made at multiple time points it is common to display the data with small multiples, such as in Figure~\ref{fig:nasa-facet}.
-In this plot, a separate map is drawn for each month (columns) and year (rows), over a 6 year period of remotely sensed temperature data above Central America \citep{murrell:2010}.
+In this plot, a separate map is drawn for each month (columns) and year (rows), over a six year period of remotely sensed temperature data above Central America \citep{murrell:2010}.
 Color is used to display de-seasonalized temperature, with red mapped to high values and blue to low.
@@
 
-We begin in Section~\ref{sec:construction} by describing the algorithm used to create glyphs-maps.
-Section~\ref{sec:perception} discusses their perceptual properties, including the importance of a visual reference grid, and of carefully considering scale.
-Section~\ref{sec:large-data} shows how glyph-maps scale to large data, and the interplay of models with glyph-maps.
-We originally developed glyph-maps for data on a rectangular grid, but many spatiotemporal data sets have irregular spatial locations.
-Section~\ref{sec:irregular} discusses how glyph-maps can be adjusted for this type of data.
+The paper is structured as follows.
+Section~\ref{sec:construction} describes the algorithm used to create glyphs-maps.
+Section~\ref{sec:perception} discusses their perceptual properties, including the importance of a visual reference grid, and of carefully consideration of scale.
+Large data and the interplay of models and data are discussed in Section~\ref{sec:large-data}.
+Many spatiotemporal data sets have irregular spatial locations, and Section~\ref{sec:irregular} discusses how glyph-maps can be adjusted for this type of data.
+Three datasets are used for examples:
 
-To demonstrate the glyph-map we will use three datasets:
-
+\begin{quote}
 \begin{itemize}
 
-  \item The ASA 2009 data expo data which consists of monthly observations of
-  several atmospheric variables from the International Satellite Cloud
-  Climatology Project. The dataset includes observations over 72 months
-  (1995--2000) on a 24 x 24 grid (576 locations) stretching from
-  $113.75^{\circ}$W to $56.25^{\circ}$W longitude and $21.25^{\circ}$S to
-  $36.25{^\circ}$N latitude.
+\item[data-expo] The ASA 2009 data expo data \citep{murrell:2010} consists
+  of monthly observations of several atmospheric variables from the
+  International Satellite Cloud Climatology Project. The dataset
+  includes observations over 72 months (1995--2000) on a 24 x 24 grid
+  (576 locations) stretching from $113.75^{\circ}$W to
+  $56.25^{\circ}$W longitude and $21.25^{\circ}$S to $36.25{^\circ}$N
+  latitude.
 
-\item GISTEMP surface temperature data provided on $2^{\circ}$ x
-  $2^{\circ}$ grid over the entire globe, measured monthly. Ground
+\item[GISTEMP] surface temperature data provided on $2^{\circ}$ x
+  $2^{\circ}$ grid over the entire globe, measured monthly \citep{GISTEMP}. Ground
   station data was de-seasonalized, differenced from from the
@@
   gridded measurements. For the purposes of this paper, we extracted
-  the locations corresponding to the continental USA. The data is made
-  available by the Earth System Research Laboratory, Physical Sciences
-  Division, National Oceanic and Atmospheric Administration, from
-  their web site at \url{http://www.esrl.noaa.gov/psd/}.
+  the locations corresponding to the continental USA.
 
-  \item USHCN (Version 2) ground station network. USHCN data provided by NOAA NCDC from their website \url{http://cdiac.ornl.gov/ftp/ushcn_v2_monthly/}.
+\item[USHCN] (Version 2) ground station network of historical
+  temperatures \citep{USHCN}. Temperatures from 1219 stations on the
+  contiguous United States, from 1871 to present.
 
 \end{itemize}
+\end{quote}
 
@@
 
-Since the coordinates are linear combinations of two variables, it is possible to build these these line-glyphs interactively, illustrated by software packages that have incorporated tours \citep{cook:2006}, such as DataViewer \citep{buja:1986}, XGobi \citep{swayne:1991} or GGobi \citep{swayne:2003}.
+Since the coordinates are linear combinations of two variables, it is possible to build these these line-glyphs interactively, illustrated by software packages that have incorporated tours (as described in \citet{cook:2006}), such as DataViewer \citep{buja:1986}, XGobi \citep{swayne:1991} or GGobi \citep{swayne:2003}.
 This technique is used to good effect in \citet{buja:1996a}, and was used to examine the climate data in \citet{hobbs:2010}.
@@
 For example, to study long-term trends, we could calculate the slope of a linear model at each location and color tiles across the map by the value of the slope.
-The disadvantage of this approach is that it is not exploratory, and requires that we have a very precise definition of the measurement we are interested in, and it still suffers from the relative difficultly of colour perception.
+The disadvantage of this approach is that it is not exploratory, and requires that we have a very precise definition of the measurement we are interested in, and it still suffers from the relative difficultly of color perception.
 
@@
 Instead, we add reference lines and boxes.
-These need to be minimally perceptible, post-attentive \citep{healey} and de-emphasized, so as not to detract from the raw data.
+These need to be minimally perceptible, post-attentive \citep{healey} and de-emphasized, so as not to detract from the data.
 The \textbf{reference grid} is a box, framing each icon, which represents the spatial grid.
@@
 %It's also often useful to display other types of reference information.
-You are only limited by your imagination (and the capabilities of your graphics software), but two simple ideas are to (1) vary the colour or transparency of glyphs or references, or (2) use an additional layer to highlight special points.
+You are only limited by your imagination (and the capabilities of your graphics software), but two simple ideas are to (1) vary the color or transparency of glyphs or references, or (2) use an additional layer to highlight special points.
 Figure~\ref{fig:ref-adv} illustrates these ideas to display the month with the highest temperature at each locations.
-Careful colour choice is necessary so that this information is perceptible, but not distracting.
+Careful color choice is necessary so that this information is perceptible, but not distracting.
 
@@
 %  \includegraphics[width=0.5\linewidth]{ref-max-2}
-%  \caption{Glyph-maps of seasonal temperature patterns also showing the month with the highest temperature using (left) an additional layer of points and (right) changing the background colour of the reference grid.}
+%  \caption{Glyph-maps of seasonal temperature patterns also showing the month with the highest temperature using (left) an additional layer of points and (right) changing the background color of the reference grid.}
 %  \label{fig:ref-adv}
@@
 The viewer must realize that the scale is not relative and that big patterns in some locations might be just tiny effects.
-To indicate this on the locally scaled data plots, one can use an additional aesthetic, like colour, to encode the range of original scale.
+To indicate this on the locally scaled data plots, one can use an additional aesthetic, like color, to encode the range of original scale.
 This is illustrated in Figure~\ref{fig:scaling-col}.
@@
  \includegraphics[width=0.5\linewidth]{month-rescale01-fill}
- \caption{Adding colour to scaled plots.
-Here each location has been scaled to range $[0, 1]$ and colour mapped to the range of the original predictions.
-(Left) Range mapped to colour of line: pale locations have smaller range than darker locations.
-(Right) Range mapped to fill colour of grid box: blue boxes have smaller ranges.
+ \caption{Adding intensity and color to scaled plots.
+Here each location has been scaled to range $[0, 1]$ and color mapped to the range of the original predictions.
+(Left) Range mapped to intensity of line: pale locations have smaller range than darker locations.
+(Right) Range mapped to fill color of grid box: blue boxes have smaller ranges.
 Shapes at all locations are visible, but attention is drawn to locations with larger ranges.}
@@
 This makes climate data challenging to work with computationally and to visualize.
-Plotting the raw data often reveals little more than a well-known observation that ``temperature is quite varied!'' Careful use of models to decompose data into long-term trend, seasonal effects and error is vital for to make the problem both computationally and visually tractable.
+Careful use of models to decompose data into long-term trend, seasonal effects and error is vital for to make the problem both computationally and visually tractable.
 
@@
 
-  \caption{Glyph-map of raw temperature anomaly data, from 1880-2011. Gaps in the full circle indicate missing values, and thickness of the circle indicates more variability in the measurements.}
+  \caption{Glyph-map of raw temperature anomaly data (GISTEMP), from 1880-2011.
+Gaps in the full circle indicate missing values, and thickness of the circle indicates more variability in the measurements.}
   \label{fig:gistemp-raw}
@@
 The raw data shows us missing values and gives some sense of variability, but gives no information about long-term trend.
-We can look trend by fitting a linear model to each location and displaying the predicted values.
+We can look at trend by fitting a linear model to each location and displaying the predicted values.
 Figure~\ref{fig:gistemp-pred} shows this for 1950--2010 (limited time period used due to missing data).
@@
 
-  \caption{Glyph-map of predicted temperature anomaly data, from
+  \caption{Glyph-map of predicted temperature anomaly data (GISTEMP), from
     1950-2010. Increasing trends can be seen over most of the USA,
@@
 \subsection{Irregular locations}
-Figure \ref{fig:irregular} shows linear trends for monthly average surface temperature at stations in the USHCN (Version 2) ground station network.
+Figure \ref{fig:irregular} shows linear trends for monthly average surface temperature at stations in the USHCN data.
 It illustrates three challenges in working with irregular locations: there is no natural choice for icon size, icons can potentially overlap, and comparisons rely more heavily on the reference guides.
@@
   \includegraphics[width=0.35\linewidth]{ghcn-mountains}%
-  \caption{Glyph-maps of the linear trend in ground station temperature values, 1950--2010, for all continental US USHCN stations (left) and stations in the mountain states (right).} 
+  \caption{Glyph-maps of the linear trend in ground station temperature values, 1950--2010, for USHCN stations (left) and stations in the mountain states (right).} 
   \label{fig:irregular}
@@
 For multivariate data we may want to represent several variables in the glyphs.
-Figure~\ref{fig:cloud} shows scatterplots as icons: temperature values are displayed horizontally and high-cloud values are displayed vertically, both locally scaled.
+Figure~\ref{fig:cloud} shows scatterplots of the data-expo data as icons: temperature values are displayed horizontally and high-cloud values are displayed vertically, both locally scaled.
 The bivariate relationship between the two variables is explored spatially, and it is clear that it is different across the region.

### 2011-09-29 code/irregular-examples.r
Fix aspect ratio of USHCN plots to match GISTEMP plots

@@ -16,6 +16,8 @@ temp.post1950 <- mutate(temp.post1950,
     season = season_index(time)
     )    
 
+width.mod <- (diff(range(ushcn$lon))/diff(range(gistemp$lon)))/(diff(range(ushcn$lat))/diff(range(gistemp$lat)))
+
 #==== Modelling
 
 # seasonal.fit.post1950 <- dlply(temp.post1950, "stn", function(df){
@@ -55,7 +57,7 @@ ggplot(lin.pred.gly, aes(gx, gy, group = gid)) +
   add_ref_lines(lin.pred.gly) +
   geom_path() +
   theme_fullframe() 
-ggsave("../images/usa-lin-overlap.png", width = 5.5, height = 4)
+ggsave("../images/usa-lin-overlap.png", width = 8, height = 4.5)
 
 # Mountain states: Colorado, Wyoming, Idaho, Montana, Nevada, Utah
 mountain.states <- c("CO", "WY", "ID", "MT", "NV", "UT")
@@ -73,7 +75,7 @@ ggplot(mountains.gly, aes(gx, gy, group = gid)) +
   geom_path() +
   theme_fullframe() +
   coord_cartesian(xlim = c(-121, -101), ylim = c(34, 50)) 
-ggsave("../images/ghcn-mountains.png", width = 3.5, height = 4)
+ggsave("../images/ghcn-mountains.png", width = 4, height = 4.5)
 
 #== one solution combine overlapping glyphs
 stn.dists <- get.stn.dists(ushcn.stations, "lon", "lat")
@@ -116,7 +118,7 @@ ggplot(lin.pred.sum, aes(gx, gy, group = gid)) +
   geom_path(aes(colour = n > 1)) +
   theme_fullframe() +
   scale_colour_manual(values = c("TRUE" = "black", "FALSE" = "grey40"))
-ggsave("../images/usa-lin-collapse.png", width = 6, height = 4)
+ggsave("../images/usa-lin-collapse.png", width = 8, height = 4.5)
 
 
 
@@ -136,7 +138,7 @@ ggplot(lin.pred.grid.gly,aes(gx, gy, group = gid)) +
   geom_path(aes(colour = n > 1)) +
   theme_fullframe() +
   scale_colour_manual(values = c("TRUE" = "black", "FALSE" = "grey40"))
-ggsave("../images/usa-lin-grid.png", width = 6, height = 4)
[…]

### 2011-09-29 paper.tex
Fix aspect ratio of USHCN plots to match GISTEMP plots
typos

@@
 
-Climate data poses visual challenges because it is has both spatial and temporal components, and is often multivariate.
-In this paper we propose an new display for climate data: glyph-maps.
+Understanding the multiple aspects of climate data is a challenge.
+There are both spatial and temporal components to examine, and often multivariate dependencies.
+In this paper we propose a new display for climate data: glyph-maps.
 Glyph-maps are a specialization of multivariate glyph plots; each spatial location is displayed with one glyph that represents the measurements recorded over time at that location.
@@
 
-  \caption{Glyph-map of predicted temperature anomaly data (GISTEMP), from
-    1950-2010. Increasing trends can be seen over most of the USA,
-    with varying degrees of increase. Some isolated locations exhibit
-    steep declines, and inclines, perhaps indicative of some isolated
-    data problems. References line are straight, but look curves because of
-    the Z\"ollner optical illusion}
+  \caption{Glyph-map of predicted temperature anomaly data (GISTEMP),
+    from 1950-2010. Increasing trends can be seen over most of the
+    USA, with varying degrees of increase. Some isolated locations
+    exhibit steep declines, and inclines, perhaps indicative of some
+    isolated data problems. Note that, the reference lines are
+    straight, but they look curved because of the Z\"ollner optical
+    illusion.}
   \label{fig:gistemp-pred}

### 2011-09-30 cover-letter.tex
final edits

\documentclass{letter}
\usepackage{graphics}
\usepackage{fullpage}
\topmargin=-0.2in
%\headheight=0in
\setlength\oddsidemargin{-0.2in}
\textheight=8.5in
\textwidth=7in
\def\apl#1#2{
\begin{letter}{#1}
\signature{Professor Dianne Cook\\Department of Statistics\\Iowa State University 
\\Ames, Iowa 50011-1210\\ Phone (515) 294 8865\\{\tt dicook@iastate.edu}}
\opening{{Dear #2}}

Please find attached our submission for the special issue of Environmetrics on Climate. The title is Glyph-maps for Visualizing Climate Data and Models, with authors Hadley Wickham, Heike Hofmann, Charlotte Wickham and myself. We hope that you find this paper interesting reading, and that it will make a valuable contribution  to the special issue.

\closing{Sincerely yours,
{\pdfimage width 1.5in {signature2.pdf}}}
\end{letter}}

\begin{document}

{{\pdfimage width 4in {letterhead.jpg}}}
\apl{Editors \\Environmetrics Special Issue on Climate}{Profs Guttorp, Sain and Wikle}

\end{document}

### 2011-09-30 paper.tex
final edits

@@
 There are both spatial and temporal components to examine, and often multivariate dependencies.
-In this paper we propose a new display for climate data: glyph-maps.
+This paper proposes a new display for climate data: glyph-maps.
 Glyph-maps are a specialization of multivariate glyph plots; each spatial location is displayed with one glyph that represents the measurements recorded over time at that location.
@@
 
-In this paper we develop a new type of display to resolves these problems: the \emph{glyph-map}.
+This paper develops a new type of display to resolve these problems: the \emph{glyph-map}.
 The glyph-map uses a small glyph, or icon, to represent multiple values at each location.
@@
 Figure~\ref{fig:nasa-glyph} shows the glyph-map equivalent of Figure~\ref{fig:nasa-facet}: each location is represented by a small time series glyph.
-The primary visible structure is still El Ni\~no, the regions in the Pacific with bump in the middle of the series, but we can also see other features: temperatures are generally more varied over land and a few locations have dramatically increasing temperature in the later years typically in high elevation areas, along the Andes and the Sierra Madre Occidental in Mexico.
+The primary visible structure is still El Ni\~no, the regions in the Pacific with bump in the middle of the series, but other features are also visible: temperatures are generally more varied over land and a few locations have dramatically increasing temperature in the later years typically in high elevation areas, along the Andes and the Sierra Madre Occidental in Mexico.
 
@@
 This problem has prevented the widespread adoption of glyph displays, despite some proposed remedies \citep{kleiner:1981,hurley:2010}.
-Using glyphs with space-time data makes the problem go away: we place glyphs according to the location of the measurement, and we order variables by the time they were collected.
+Using glyphs with space-time data eliminates these problems: glyphs are placed according to the location of the measurement, and variables are ordered by the time they were collected.
 Additionally, climate data usually has high correlations between nearby locations and times.
@@
 Others \citet{pickett:1988} have used glyphs on maps to display multivariate spatial data.
-This area of work has morphed into the field of metaphorical data displays, which creates abstract landscapes of spatial data, a digression from glyph-maps.
+This area of work has morphed into the field of metaphorical data displays, which create abstract landscapes of spatial data, a digression from glyph-maps.
 \citet{gribov:2006} describes the use of glyph-maps for multivariate data also, with emphasis on the graphics software Gauguin.
 
-We focus our attention on two types of glyph: lines and stars.
+Two types of glyph -- lines and stars -- are especially useful for temporal displays.
 Figure~\ref{fig:templates} displays 12 iconic time series shapes with line- and star-glyphs.
 The data underlying each glyph is measured at 36 time points.
-The line-glyphs are a simple time series plot.
+The line-glyphs are   time series plots.
 The star-glyphs are formed by considering the 36 axes radiating from a common midpoint, and the data values for the row are plotted on each axis relative to the locations of the minimum and maximum of the variable.
@@
 \begin{quote}
-\begin{itemize}
+\begin{itemize} \itemsep 0in
 
@@
 \end{quote}
-
-All datasets and R code used to produce the plots are available in the online supplementary materials.
-All plots were created in R \citep{R} using the {\tt ggplot2} \citep{me:ggplot2} and {\tt plyr} \citep{me:plyr} packages.
+Supplementary material contains the datasets and code to reproduce the plots in this paper.
+R 2.13.1 \citep{R} with the packages {\tt ggplot2} \citep{me:ggplot2} and {\tt plyr} \citep{me:plyr} were used.
 
@@
 
-Glyph-maps can be generated with existing graphics software after performing a simple pre-processing step, making them easily accessible to a wide audience.
-Creating a glyph-map requires a recognition of the two structural components of the data: spatial location and data values.
+%Glyph-maps can be generated with existing graphics software after performing a simple pre-processing step, making them easily accessible to a wide audience. 
+Creating a glyph-map is elementary.
+It involves recognizing that they are a linear mix of two structural components of the data: spatial location and data values.
 The spatial location is the major positioning component, while the data values are minor adjustments to those positions.
 For spatiotemporal data, the major axes are latitude ($y_{\amaj}$) and longitude ($x_{\amaj}$), and the minor axes are time ($x_{\amin}$) and some measurement ($y_{\amin}$), for example, temperature, or predicted temperature.
-Assuming the minor axes are rescaled to $[-1, 1]$, the final coordinates $(x,y)$ on the chart are a simple linear combination:
+Assuming the minor axes are rescaled to $[-1, 1]$, the final coordinates $(x,y)$ on the chart are the linear combination, given by:
 
@@
 
-Computing the coordinates for star glyphs is achieved with a simple polar transformation, with minor axes scaled to $[0, 1]$:
+Coordinates for star glyphs are computed with a polar transformation, with minor axes scaled to $[0, 1]$:
 
@@
 All plots facilitate some comparisons and impede others.
-To decide if a plot if useful for a particular data analytic task, we need to check that the comparisons that we are interested in are the ones that a supported by the plot.
-Generally, for climate data, we are most interested in seeing changes in slope, or trend, average value, and variance.
-We argue that glyph-maps support these comparisons better because the comparisons are perceptually easier.
+For a plot to be useful for a particular data analytic task, primary comparisons need to be made the simplest by the display.
+For climate data, changes in slope, or trend, average value, and variance over time, are the primary tasks.
+Glyph-maps support these tasks because the pieces of the graphical elements that are needed to make the temporal comparisons are organized, and grouped together.
 
@@
 This is challenging from a cognitive perspective \citep{healey:2011,busey} and comparisons can suffer from \emph{change blindness}, often leading to silent failure.
-
-This is particularly obvious when comparing Figure~\ref{fig:nasa-facet} to Figure~\ref{fig:nasa-glyph}.
+To see this, examine Figures~\ref{fig:nasa-facet} and \ref{fig:nasa-glyph}.
+The heatmap requires comparing color values across maps to assess temporal change but the glyph-map allows direct reading of the temporal trend at each location, and comparison across spatial neighbors.
 
@@
 
-One draw-back of the glyph-maps, particularly if linear trend is displayed as an icon, is they may suffer from the Z\"ollner Illusion \citep{Zollner}, which makes straight lines look crooked, and is visible in Figure \ref{fig:gistemp-pred}.
+One draw-back of the glyph-maps, particularly if linear trend is displayed as an icon, is they may suffer from the Z\"ollner Illusion \citep{Zollner}, visible in Figure \ref{fig:gistemp-pred}, which makes straight lines look crooked.
 
@@
 For example, to study long-term trends, we could calculate the slope of a linear model at each location and color tiles across the map by the value of the slope.
-The disadvantage of this approach is that it is not exploratory, and requires that we have a very precise definition of the measurement we are interested in, and it still suffers from the relative difficultly of color perception.
+The disadvantage of this approach is that it is not exploratory, and requires that we have a very precise definition of the measurement we are interested in, and it still suffers from the relatively difficult color perception.
 
@@
 Each glyph is small, so there is not enough space for a full set of axes.
-Instead, we add reference lines and boxes.
+Instead, minimal reference lines and boxes are incorporated.
 These need to be minimally perceptible, post-attentive \citep{healey} and de-emphasized, so as not to detract from the data.
-The \textbf{reference grid} is a box, framing each icon, which represents the spatial grid.
-The \textbf{reference line} is a horizontal line at mid-range.
+The reference grid is a box, framing each icon, which represents the spatial grid.
+The reference line is a horizontal line at mid-range.
 Both help to read differences in slope and intercept.
@@
 Figure~\ref{fig:ref-basic} demonstrates both guides.
-We draw them in white, so they are minimally perceptible relative to the black data.
+Reference elements are drawn in white, so they are minimally perceptible relative to the black data.
 
@@
 
-For star glyphs the reference frames are also useful and the equivalent of the reference line is a circle. Figures~\ref{fig:templates} and \ref{fig:gistemp-raw} show examples.
+For star glyphs the reference frames are also useful and the equivalent of the reference line is a circle. Figures~\ref{fig:templates} and \ref{fig:gistemp-raw} show more examples.
 
@@
 
-We can also emphasize different features of the data by varying the scale used with each cell.
-By default, we use a global scale, so the same position within each cell corresponds to the same value in all locations.
+Different features of the data can be emphasized by varying the scale used with each cell.
+By default, a global scale is used, so the same position within each cell corresponds to the same value in all locations.
 This facilitates the comparison of absolute values, and draws attention to plots with large overall variation.
 
-Alternatively, we can re-scale values within each location prior to plotting.
+Alternatively, local scale can be used.
 In local scaling, the values within each location are scaled to range $[0, 1]$ before plot construction.
@@
 On the global scale, the El Ni\~no blip in temperature is just visible, along with variation in the average temperature at each location.
-Local scaling emphasizes the individual shapes: we see the impact of El Ni\~no over a wider region, dramatic linear increases across the Andes, and decreases in the Gulf of Mexico.
+Local scaling emphasizes the individual shapes: the impact of El Ni\~no can be seen to be over a wider region, dramatic linear increases across the Andes, and decreases in the Gulf of Mexico.
 
@@
 
-\begin{itemize}
+\begin{itemize} \itemsep 0in
 
@@
 Climate data is usually large: many locations in space, and many points in time.
-This makes climate data challenging to work with computationally and to visualize.
+This makes climate data arduous to work with computationally, and to visualize.
 Careful use of models to decompose data into long-term trend, seasonal effects and error is vital for to make the problem both computationally and visually tractable.
@@
 The raw data shows us missing values and gives some sense of variability, but gives no information about long-term trend.
-We can look at trend by fitting a linear model to each location and displaying the predicted values.
+Trend can be examined by fitting a linear model to each location and displaying the predicted values.
 Figure~\ref{fig:gistemp-pred} shows this for 1950--2010 (limited time period used due to missing data).
@@
   \centering
-  \includegraphics[width=0.55\linewidth]{usa-lin-overlap}%
-  \includegraphics[width=0.35\linewidth]{ghcn-mountains}%
-  \caption{Glyph-maps of the linear trend in ground station temperature values, 1950--2010, for USHCN stations (left) and stations in the mountain states (right).} 
+  \includegraphics[width=0.95\linewidth]{usa-lin-overlap}%
+  %\includegraphics[width=0.25\linewidth]{ghcn-mountains}%
+  \caption{Glyph-maps of the linear trend in ground station temperature values, 1950--2010, for USHCN stations, illustrating three issues for irregularly gridded data: overlapping icons, no natural choice for icon size, and the disordered appearance making comparisons more difficult.
+}% (left) and stations in the mountain states (right).}
   \label{fig:irregular}
 \end{figure}
-The two panels illustrate one approach to the overlapping problem: zooming.
+%The two panels illustrate one approach to the overlapping problem: zooming.  
+One solution to the overlapping problem is to zoom in on smaller regions.
 There is an interplay between icon size (in geographical units), plotting area (in geographical units) and plot size (in display units).
 Icons need to be big enough to be readable on the display, but small enough to avoid overlapping.
-Readability can be maintained and overlapping reduced by decreasing the icon size and simultaneously increasing the display size or decreasing the plotting area (as done in the right panel in Figure \ref{fig:irregular}).
+Readability can be maintained and overlapping reduced by decreasing the icon size and simultaneously increasing the display size or decreasing the plotting area.
+% (as done in the right panel in Figure \ref{fig:irregular}).  
 However, there are generally limitations in the maximum display size (or resolution) and the utility of plotting very small areas.
@@
 In Figure \ref{fig:irregular-collapsed} nearby locations are collapsed to a single icon.
-In the left panel each station still has an individual glyph but locations close to each other share a plotting area.
+In the top panel each station still has an individual glyph but locations close to each other share a plotting area.
 Not all glyphs are suited to this kind of space sharing, another option is to perform a summary of the locations that are collapsed and display that instead of the raw data.
-The right panel shows an example of this where the average trend for the collapsed locations is displayed.
+The bottom panel shows an example of this where the average trend for the collapsed locations is displayed.
 It is helpful to differentiate glyphs that involve summaries of more than one location, here a heavier line indicates at least two locations have been averaged.
@@
   \centering
-  \includegraphics[width=0.5\linewidth]{usa-season-collapsed}%
-  \includegraphics[width=0.5\linewidth]{usa-lin-collapse}%
-  \caption{Glyph-maps of seasonal patterns (left) and linear trends (right) of USHCN stations, 1950--2010.
+  \includegraphics[width=0.95\linewidth]{usa-season-collapsed}%
+
+  \includegraphics[width=0.95\linewidth]{usa-lin-collapse}%
+  \caption{Glyph-maps of seasonal patterns (top) and linear trends (bottom) of USHCN stations, 1950--2010.
 Nearby stations whose icons would overlap have been collapsed.
@@
 
-\section{Generalizations}
-
-Glyphs can be more general than time series or stars.
-For multivariate data we may want to represent several variables in the glyphs.
-Figure~\ref{fig:cloud} shows scatterplots of the data-expo data as icons: temperature values are displayed horizontally and high-cloud values are displayed vertically, both locally scaled.
-The bivariate relationship between the two variables is explored spatially, and it is clear that it is different across the region.
-The equatorial Pacific and Caribbean have positive association between temperature and high cloud, while the south American continent has negative association.
-The second plot displays a loess curve fit to the data instead of the individual points, sharpening the view of association.
-Another option is available in Gauguin \citep{gribov:2006}, a icon based on a histogram, giving small univariate distribution displays for each location.
-
 \begin{figure}[htbp]
@@
 \end{figure}
+
+\section{Generalizations}
+
+Glyphs can be more general than time series or stars.
+For multivariate data we may want to represent several variables in the glyphs.
+Figure~\ref{fig:cloud} shows scatterplots of the data-expo data as icons: temperature values are displayed horizontally and high-cloud values are displayed vertically, both locally scaled.
+The bivariate relationship between the two variables is explored spatially, and it is clear that it is different across the region.
+The equatorial Pacific and Caribbean have positive association between temperature and high cloud, while the south American continent has negative association.
+The second plot displays a loess curve fit to the data instead of the individual points, sharpening the view of association.
+Another option is available in Gauguin \citep{gribov:2006}, a icon based on a histogram, giving small univariate distribution displays for each location.
 

### 2011-09-30 gridded.temp.r
forgor this one

@@ -64,7 +64,10 @@ temp.usa.melt$lon<-lon[temp.usa.melt$gridx+113]
 temp.usa.melt$lat<-lat[temp.usa.melt$gridy+19]
 
 temp.usa.melt<-subset(temp.usa.melt, !(lon>-70 & lat<37))
-
+x<-ddply(temp.usa.melt, c("lon","lat"), function(df) {
+ if (nrow(subset(df,is.na(temp))) == nrow(df)) NULL else df 
+})
+      
 temp.usa.melt.gly<-glyphs(temp.usa.melt, "lon", "time", "lat", "temp")
 qplot(gx, gy, data=temp.usa.melt.gly, group=gid, geom="line") + map + coord_map() # Testing
 p <-ggplot(temp.usa.melt.gly, aes(gx, gy, group = gid)) + 
@@ -110,10 +113,8 @@ year_preds <- ldply(temp_models, function(mod) {
 })
 temp.usa.melt.sub.gly<-glyphs(year_preds, "lon", "year", "lat", "pred")
 ggplot(temp.usa.melt.sub.gly, aes(gx, gy, group = gid)) + 
-  map + geom_line(aes(y = lat), colour = "white", size = 1.5) +
-  geom_path() +
-  ref_boxes +
-  theme_fullframe() #+ coord_map() makes a strange tranformation of boxes
+  map_gistemp + geom_line(aes(y = lat), colour = "white", size = 1.5) +
+  add_ref_lines(temp.usa.melt.sub.gly) + add_ref_boxes(temp.usa.melt.sub.gly) + geom_path()
 ggsave("../images/gistemp-pred.png", width = 8, height = 4)
 # Coord map seems to make the plotting REALLY SLOW!
 

### 2012-02-25 revision/paper.tex
new directory holding revision to paper

--- 2011-09-30 paper.tex
@@
 \usepackage{amstext}
+\usepackage{setspace}
 \bibliographystyle{plainnat}
@@
 \begin{document}
-
-\title{Glyph-maps for Visualizing Climate Data and Models}
+\doublespacing
+\title{Glyph-maps for Visually Exploring Temporal Patterns in Climate Data and Models}
 \author{Hadley Wickham$^1$, Heike Hofmann$^2$, Charlotte Wickham$^3$, Dianne Cook$^2$\\
@@
 Climate data is composed of measurements, such as temperature, precipitation, and winds, with a spatial and temporal context.
-The classic display for data of this type of is ``heatmap'', a tiled plot using color to display the value of the variable of interest as each location.
+The classic display for data of this type of is ``heatmap'', a tiled plot using color to display the value of the variable of interest at each location.
 When measurements are made at multiple time points it is common to display the data with small multiples, such as in Figure~\ref{fig:nasa-facet}.
@@
 Using a movie to render these small multiples can help, but small trends still fail to draw readers' attention and escape unnoticed \citep{simons:gradual}.
+
+%% Reproduce all of the icon displays using the fixed glyph code
+
+This paper develops a new type of display to resolve these problems: the \emph{glyph-map}.
+The glyph-map uses a small glyph, or icon, to represent multiple values at each location.
+Glyph maps are an adaptation of glyphs, tools developed for display multivariate data.
+Multivariate glyphs include the star plots of \citet{mayr:1877}, the semi-graphic displays of \citet{anderson:1960}, and the infamous Chernoff faces \citep{chernoff:1973}.
+An glyph is produced by mapping each variable to some graphical feature, such as the length of a line.
+The glyph-map can also be thought of as a small multiple display \citep{tufte:2001} of time series with a geographic context.
+
+Figure~\ref{fig:nasa-glyph} shows the glyph-map equivalent of Figure~\ref{fig:nasa-facet}: each location is represented by a small time series glyph.
+The primary visible structure is still El Ni\~no, the regions in the Pacific with bump in the middle of the series, but other features are also visible: temperatures are generally more varied over land and a few locations have dramatically increasing temperature in the later years typically in high elevation areas, along the Andes and the Sierra Madre Occidental in Mexico.
+
+% The most seasonality occurs in north American.
+Some seasonal effects can also be seen in the Caribbean and in south American locations.
+Around the equator the temperatures are relatively warm for all seasons.
+Instead of radially displaying the temperature a glyph can be a regular time series, time displayed horizontally and temperature vertically (right plot).
+The structure perceived is different with this change in glyph type: locations at the equator have flat series, on land more varied temperatures.
 
@@
 
-%% Reproduce all of the icon displays using the fixed glyph code
-
-This paper develops a new type of display to resolve these problems: the \emph{glyph-map}.
-The glyph-map uses a small glyph, or icon, to represent multiple values at each location.
-Glyph maps are an adaptation of glyphs, tools developed for display multivariate data.
-Multivariate glyphs include the star plots of \citet{mayr:1877}, the semi-graphic displays of \citet{anderson:1960}, and the infamous Chernoff faces \citep{chernoff:1973}.
-An glyph is produced by mapping each variable to some graphical feature, such as the length of a line.
-The glyph-map can also be thought of as a small multiple display \citet{tufte:2001} of time series with a geographic context.
-
-Figure~\ref{fig:nasa-glyph} shows the glyph-map equivalent of Figure~\ref{fig:nasa-facet}: each location is represented by a small time series glyph.
-The primary visible structure is still El Ni\~no, the regions in the Pacific with bump in the middle of the series, but other features are also visible: temperatures are generally more varied over land and a few locations have dramatically increasing temperature in the later years typically in high elevation areas, along the Andes and the Sierra Madre Occidental in Mexico.
-
-% The most seasonality occurs in north American.
-Some seasonal effects can also be seen in the Caribbean and in south American locations.
-Around the equator the temperatures are relatively warm for all seasons.
-Instead of radially displaying the temperature a glyph can be a regular time series, time displayed horizontally and temperature vertically (right plot).
-The structure perceived is different with this change in glyph type: locations at the equator have flat series, on land more varied temperatures.
-
 One problem with multivariate glyph is that lack of natural ordering of each glyph, or of the mapping between variables and glyph properties.
@@
 This imposes a degree of smoothness that gives the plot an appearance of a textured  landscape, making it easier to digest patterns.
-Others \citet{pickett:1988} have used glyphs on maps to display multivariate spatial data.
+Others \citep{pickett:1988} have used glyphs on maps to display multivariate spatial data.
 This area of work has morphed into the field of metaphorical data displays, which create abstract landscapes of spatial data, a digression from glyph-maps.
@@
 Section~\ref{sec:construction} describes the algorithm used to create glyphs-maps.
-Section~\ref{sec:perception} discusses their perceptual properties, including the importance of a visual reference grid, and of carefully consideration of scale.
+Section~\ref{sec:perception} discusses their perceptual properties, including the importance of a visual reference grid, and of careful consideration of scale.
 Large data and the interplay of models and data are discussed in Section~\ref{sec:large-data}.
@@
 \item[GISTEMP] surface temperature data provided on $2^{\circ}$ x
-  $2^{\circ}$ grid over the entire globe, measured monthly \citep{GISTEMP}. Ground
-  station data was de-seasonalized, differenced from from the
-  1951-1980 temperature averages, and spatially averaged to obtain
-  gridded measurements. For the purposes of this paper, we extracted
-  the locations corresponding to the continental USA.
+  $2^{\circ}$ grid over the entire globe, measured monthly
+  \citep{GISTEMP}. Ground station data was de-seasonalized,
+  differenced from from the 1951-1980 temperature averages, and
+  spatially averaged to obtain gridded temperature anomalies. For the
+  purposes of this paper, we extracted the locations corresponding to
+  the continental USA.
 
@@
 Supplementary material contains the datasets and code to reproduce the plots in this paper.
-R 2.13.1 \citep{R} with the packages {\tt ggplot2} \citep{me:ggplot2} and {\tt plyr} \citep{me:plyr} were used.
+R 2.13.1 \citep{R} with the packages {\tt ggplot2} \citep{me:ggplot2} and {\tt plyr} \citep{me:plyr} was used.
 
@@
 Creating a glyph-map is elementary.
-It involves recognizing that they are a linear mix of two structural components of the data: spatial location and data values.
+It involves recognizing that the glyph-map are a linear mix of two structural components of the data: spatial location and data values.
 The spatial location is the major positioning component, while the data values are minor adjustments to those positions.
@@
 Perceiving time trends in faceted heatmaps is much more difficult: not only do you need to read value from color (a relatively difficult task), you also need to spot the difference between the different maps.
-This is challenging from a cognitive perspective \citep{healey:2011,busey} and comparisons can suffer from \emph{change blindness}, often leading to silent failure.
+This is challenging from a cognitive perspective \citep{healey:2011,busey} and comparisons can suffer from \emph{change blindness}, often leading to failure to notice the change and not knowing that something changed.
 To see this, examine Figures~\ref{fig:nasa-facet} and \ref{fig:nasa-glyph}.
@@
 
-An third alternative is to calculate the relevant statistic for each location and display these values as single heatmap.
+A third alternative is to calculate the relevant statistic for each location and display these values as single heatmap.
 For example, to study long-term trends, we could calculate the slope of a linear model at each location and color tiles across the map by the value of the slope.
@@
 One more important point: if a model summary is displayed in the glyph-map, as discussed in the following section, it may be important to force the scale to that of the original data.
-The range of predicted values is typically smaller than the raw data, so angles of slopes is increase.
+The range of predicted values is typically smaller than the raw data, so angles of slopes is increased.
 
@@
 The raw data shows us missing values and gives some sense of variability, but gives no information about long-term trend.
-Trend can be examined by fitting a linear model to each location and displaying the predicted values.
+Trend can be examined by fitting a linear model to each location and displaying predicted values.
 Figure~\ref{fig:gistemp-pred} shows this for 1950--2010 (limited time period used due to missing data).

### 2012-02-25 revision/code/data-nasa.r

@@ -1,0 +1,13 @@
+if (!file.exists("../data/nasadata.rds")) {
+  library("lubridate")
+
+  nasa <- read.csv("../data/nasadata.csv.bz2", stringsAsFactors = FALSE)
+  nasa$date <- ymd(nasa$date)
+  nasa$day <- (nasa$date - min(nasa$date)) %/% ddays(1)
+  nasa$month <- month(nasa$date)
+  nasa$year <- year(nasa$date)
+  
+  saveRDS(nasa, "../data/nasadata.rds")
+} else {
+  nasa <- readRDS("../data/nasadata.rds")
+}

### 2012-02-25 revision/code/data-ushcn.r

@@ -1,0 +1,42 @@
+library(maps)
+library(ggplot2)
+library(plyr)
+
+source("glyph-utils.r")
+
+source("maps.r")
+
+ushcn.loc <- "../data/ushcndata.rds"
+ushcn.stations.loc <- "../data/ushcnstations.rds"
+
+if (!file.exists(ushcn.loc)) {
+  temp.all<-read.table("../data/9641C_201012_raw.avg")
+  colnames(temp.all)<-c("year", "jan", "feb", "mar", "apr", "may", "jun", "jul",
+    "aug", "sep", "oct", "nov", "dec", "ann")
+  temp.all$stn<-temp.all$year%/%100000
+  temp.all$year<-temp.all$year%%10000
+
+  stn.all<-read.table("../data/ushcn-stations.txt", fill=T)
+  stn.all<-stn.all[,c(1:3,5:6)]
+  colnames(stn.all)<-c("stn", "lat", "lon", "state", "town")
+  stn.all$name<-paste(stn.all$town, stn.all$state)
+  ushcn.stations <- stn.all
+  
+  ushcn.all <- join(temp.all, stn.all)
+  ushcn <- melt(ushcn.all[ , -c(14, 19)], 
+    id=c("stn", "year", "lat", "lon", "name", "state"))
+  ushcn <- rename(ushcn, 
+      c("variable" = "month", "value" = "temp"))
+
+  ushcn <- mutate(ushcn, 
+      temp = temp/10,
+      date = as.Date(paste(year, "-", as.numeric(month), "-01", sep = "")),
+      )    
+  ushcn$temp[ushcn$temp == -999.9] = NA
+  
+  saveRDS(ushcn, ushcn.loc)
+  saveRDS(ushcn.stations, ushcn.stations.loc)
+} else {
+  ushcn <- readRDS(ushcn.loc)
+  ushcn.stations <- readRDS(ushcn.stations.loc)
+}

### 2012-02-25 revision/code/glyph-examples.r

@@ -1,0 +1,42 @@
+# This file contains the glyph-map for nasa data Might think about
+# putting all of the glyph-map examples, using gistemp and ushcn data
+# here too
+
+source("glyph.r")
+source("maps.r")
+# world, map are created in maps.r
+
+
+# De-seasonalized nasa data: matches the facetted, coloured map, need
+# to run that code before this
+resids.gly <- glyphs(resids, "long", "day", "lat", "temp_resid", height=2.5) 
+ggplot(resids.gly, aes(gx, gy, group = gid)) + map +
+  add_ref_lines(resids.gly) +
+  add_ref_boxes(resids.gly) +
+  geom_path() + theme_fullframe() 
+#qplot(gx, gy, data = resids, geom = "line", group = gid) + map
+ggsave("../images/nasa-deseas-glyph.png")
+
+# Scatterplot
+nasa.scale<-ddply(nasa, c("long","lat"), summarise,
+                  cloudlow.s=scale(cloudlow), 
+                  cloudhigh.s=scale(cloudhigh),
+                  temp.s=scale(surftemp))
+#nasa.scat.gly <- glyphs(nasa.scale, "long", "cloudlow.s", "lat", "cloudhigh.s") 
+#ggplot(nasa.scat.gly, aes(gx, gy, group = gid)) + map +
+#  add_ref_boxes(nasa.scat.gly) +
+#  geom_point(size=I(0.7), alpha=I(0.5)) + theme_fullframe() 
+#ggsave("../images/nasa-scat-glyph.png")
+nasa.scat.gly <- glyphs(nasa.scale, "long", "temp.s", "lat", "cloudhigh.s") 
+ggplot(nasa.scat.gly, aes(gx, gy, group = gid)) + map +
+  add_ref_boxes(nasa.scat.gly) +
+  geom_point(size=I(0.7), alpha=I(0.5)) + theme_fullframe() 
+ggsave("../images/nasa-scat-glyph.png")
+nasa.loess<-ddply(nasa.scale, c("long","lat"), summarise,
+                  temp.s=temp.s,
+                  pcloud=predict(loess(cloudhigh.s~temp.s)))
+nasa.loess.gly <- glyphs(nasa.loess, "long", "temp.s", "lat", "pcloud") 
+ggplot(nasa.loess.gly, aes(gx, gy, group = gid)) + map +
+  add_ref_boxes(nasa.loess.gly) +
+  geom_line() + theme_fullframe() 
+ggsave("../images/nasa-loess-glyph.png")

### 2012-02-25 revision/code/glyph-utils.r

@@ -1,0 +1,47 @@
+library("maps")
+library("ggplot2")
+
+source("maps.r")
+source("glyph.r")
+
+theme_fullframe <- function (base_size = 12){
+  structure(list(
+    axis.line = theme_blank(), 
+    axis.text.x = theme_blank(), 
+    axis.text.y = theme_blank(),
+    axis.ticks = theme_blank(), 
+    axis.title.x = theme_blank(), 
+    axis.title.y = theme_blank(), 
+    axis.ticks.length = unit(0, "lines"), 
+    axis.ticks.margin = unit(0, "lines"), 
+    legend.position = "none", 
+    panel.border = theme_blank(), 
+    panel.grid.major = theme_blank(), 
+    panel.grid.minor = theme_blank(), 
+    panel.margin = unit(0, "lines"), 
+    plot.background = theme_blank(), 
+    plot.margin = unit(rep(2, 4), "mm")
+  ), class = "options")
+}
+
+add_ref_lines <- function(data, colour = "white", size = 1.5, ...){
+  rl <- ref_lines(data)
+  geom_line(data = rl, colour = colour, size = size, ...)
+}
+
+add_ref_boxes <- function(data, var_fill = NULL, colour = "white", size = 0.5, 
+    fill = NA, ...){
+  rb <- ref_boxes(data, var_fill)
+  if (!is.null(var_fill)){
+    geom_rect(aes_all(names(rb)), data = rb,
+     colour = colour, size = size, inherit.aes = FALSE, ...)
+   }
+   else{
+     geom_rect(aes_all(names(rb)), data = rb,
+      colour = colour, size = size, inherit.aes = FALSE, fill = fill, ...)
+   }
+}
+
+# Need to have a function for the circles references for polar coords
+# Need to have a function for the reference lines
+# Need to have reference boxes for the irregular grid, too

### 2012-02-25 revision/code/glyph.r

@@ -1,0 +1,172 @@
+#' Create the data needed to generate a glyph plot.
+#'
+#' @param data A data frame containing variables named in \code{x_major},
+#'   \code{x_minor}, \code{y_major} and \code{y_minor}.
+#' @param x_major,x_minor,y_major,y_minor The name of the variable (as a
+#'   string) for the major and minor x and y axes.  Together, each unique
+#    combination of \code{x_major} and \code{y_major} specifies a grid cell.
+#' @param polar A logical of length 1, specifying whether the glyphs should
+#'   be drawn in polar coordinates.  Defaults to \code{FALSE}.
+#' @param height,width The height and width of each glyph. Defaults to 95\% of
+#'  the \code{\link[ggplot2]{resolution}} of the data. Specify the width
+#'  absolutely by supplying a numeric vector of length 1, or relative to the
+#   resolution of the data by using \code{\link{rel}}.
+#' @param y_scale,x_scale The scaling function to be applied to each set of
+#'  minor values within a grid cell.  Defaults to \code{\link{identity}} so
+#'  that no scaling is performed.
+glyphs <- function(data, x_major, x_minor, y_major, y_minor, polar = FALSE, height = rel(0.95), width = rel(0.95), y_scale = identity, x_scale = identity) {
+  data$gid <- interaction(data[[x_major]], data[[y_major]], drop = TRUE)
+  
+  if (is.rel(width)) {
+    width <- resolution(data[[x_major]], zero = FALSE) * unclass(width)
+    message("Using width ", format(width, digits = 3))
+  }
+    
+  if (is.rel(height)) {
+    height <- resolution(data[[y_major]], zero = FALSE) * unclass(height)
+    message("Using height ", format(height, digits = 3))
+  }
+  
+  if (!identical(x_scale, identity) || !identical(y_scale, identity)) {
+    data <- ddply(data, "gid", function(df) {
+      df[[x_minor]] <- x_scale(df[[x_minor]])
+      df[[y_minor]] <- y_scale(df[[y_minor]])
+      df
+    })
+  }
+  
+  if (polar) {
+    theta <- 2 * pi * rescale01(data[[x_minor]])
+    r <- rescale01(data[[y_minor]])
+    
[…]

### 2012-02-25 revision/code/irregular-examples.r

@@ -1,0 +1,198 @@
+library(mgcv)
+
+source("glyph.r")
+source("glyph-utils.r")
+source("overlapping.r")
+source("data-ushcn.r")
+
+temp.post1950 <- subset(ushcn, year >= 1950)
+
+season_index <- function(time){
+  (time %% 12  + 1 )/ 12
+}
+
+temp.post1950 <- mutate(temp.post1950, 
+    time = (year - 1950) * 12 + as.numeric(month),
+    season = season_index(time)
+    )    
+
+width.mod <- (diff(range(ushcn$lon))/diff(range(gistemp$lon)))/(diff(range(ushcn$lat))/diff(range(gistemp$lat)))
+
+#==== Modelling
+
+# seasonal.fit.post1950 <- dlply(temp.post1950, "stn", function(df){
+#   gam(temp ~ s(season, bs = "cc"), 
+#     knots = list(season = c(0, 12)),
+#     data = df, na.action = "na.exclude")
+#   }, .progress = "text"
+# ) 
+fit.post1950 <- dlply(temp.post1950, "stn", function(df){
+  lm(temp ~ year + month, 
+    data = df, na.action = "na.exclude")
+  }, .progress = "text"
+) 
+
+temp.post1950 <- cbind(temp.post1950, 
+  ldply(fit.post1950, function(fit) {
+    data.frame(res = residuals(fit), pred = predict(fit))},
+  .progress = "text"))
+
+lin.pred.post1950 <- ldply(fit.post1950, function(fit){
+    one.df <- data.frame(year = 1950:2010, 
+      month = "jan")
+    one.df$pred <- predict(fit, one.df)
+    one.df
+  }, .progress = "text")
+
+lin.pred.post1950 <- merge(lin.pred.post1950, ushcn.stations)
+
+w <- 1
+h <- 1
+
+lin.pred.gly <- glyphs(lin.pred.post1950, "lon", "year", "lat", "pred", 
+  width=w, height=h, y_scale = mean0)
+
+ggplot(lin.pred.gly, aes(gx, gy, group = gid)) + 
+  map_ushcn + 
+  add_ref_lines(lin.pred.gly) +
+  geom_path() +
+  theme_fullframe() 
+ggsave("../images/usa-lin-overlap.png", width = 8, height = 4.5)
+
+# Mountain states: Colorado, Wyoming, Idaho, Montana, Nevada, Utah
[…]

### 2012-02-25 revision/code/maps.r

@@ -1,0 +1,64 @@
+library("maps")
+library("ggplot2")
+
+getbox <- function (map, xlim, ylim) {
+	# identify all regions involved
+	small <- subset(map, (long > xlim[1]) & (long < xlim[2]) & (lat > ylim[1]) & (lat < ylim[2]))
+	regions <- unique(small$region)
+	small <- subset(map, region %in% regions)	
+	
+	# now shrink all nodes back to the bounding box
+	small$long <- pmax(small$long, xlim[1])
+	small$long <- pmin(small$long, xlim[2])
+	small$lat <- pmax(small$lat, ylim[1])
+	small$lat <- pmin(small$lat, ylim[2])
+
+	# Remove slivvers
+	small <- ddply(small, "group", function(df) {
+	  if (diff(range(df$long)) < 1e-6) return(NULL)
+	  if (diff(range(df$lat)) < 1e-6) return(NULL)
+	  df
+	})
+
+	small
+}
+
+##### Prepping Data
+world <- map_data("world")
+
+states <- map_data("state")
+states$group <- max(world$group) + states$group
+both <- rbind(world, states)
+
+##### Maps
+nasa <- getbox(both, xlim = c(-115, -55), ylim = c(-21.1, 36.6))
+map_nasa <- list(
+  geom_polygon(aes(long, lat, group = group), inherit.aes = FALSE, 
+    data = nasa, legend = FALSE, fill = "grey80", colour = "grey90"),
+  scale_x_continuous(breaks = NA, expand = c(0.02, 0)),
+  scale_y_continuous(breaks = NA, expand = c(0.02, 0)), 
+  xlab(NULL),
+  ylab(NULL))
+
+ushcn <- subset(getbox(both, xlim = c(-126, -65), ylim = c(24, 52)),
+  region != "Great Lakes")
+map_ushcn <- list(
+  geom_polygon(aes(long, lat, group = group), inherit.aes = FALSE, 
+    data = ushcn, legend = FALSE, fill = "grey80", colour = "grey90"),
+  scale_x_continuous(breaks = NA, expand = c(0.02, 0)),
+  scale_y_continuous(breaks = NA, expand = c(0.02, 0)), 
+  xlab(NULL),
+  ylab(NULL))
+
+gistemp <- subset(getbox(both, xlim = c(-135, -57), ylim = c(22, 53)),
+  region != "Great Lakes")
+map_gistemp <- list(
[…]

### 2012-02-25 revision/code/modeling.r

@@ -1,0 +1,151 @@
+library("ggplot2")
+library("plyr")
+library("mgcv")
+
+source("data-nasa.r")
+source("glyph.r")
+source("glyph-utils.r")
+source("cache.r")
+source("maps.r")
+
+# Model fitting --------------------------------------------------------------
+
+temp_models %<-cache% dlply(nasa, c("lat", "long"), function(df) {
+  lm(surftemp ~ year + factor(month), data = df)
+})
+
+temp_smooth %<-cache% dlply(nasa, c("lat", "long"), function(df) {
+  gam(surftemp ~ s(day) + factor(month), data = df)
+})
+
+month_grid <- expand.grid(year = 2000, month = 1:12)
+month_preds <- ldply(temp_models, function(mod) {
+  month_grid$pred <- predict(mod, newdata = month_grid)
+  month_grid
+})
+
+year_grid <- expand.grid(year = unique(nasa$year), month = 1)
+year_preds <- ldply(temp_models, function(mod) {
+  year_grid$pred <- predict(mod, newdata = year_grid)
+  year_grid
+})
+
+day_grid <- expand.grid(day = seq(0, 2161, length = 50), month = 1)
+day_preds <- ldply(temp_smooth, function(mod) {
+  day_grid$pred <- predict(mod, newdata = day_grid)
+  day_grid
+})
+
+# Rescale predictions to individual scales -----------------------------------
+
+day_preds <- glyphs(day_preds, "long", "day", "lat", "pred") 
+ggplot(day_preds, aes(gx, gy, group = gid)) + 
+  map_nasa + 
+  add_ref_boxes(day_preds) +
+  geom_path() +
+  theme_fullframe()
+ggsave("../images/month-rescale-none.png", width = 4, height = 4)
+
+last_plot() %+% glyphs(day_preds, "long", "day", "lat", "pred", 
+  y_scale = range01) 
+ggsave("../images/month-rescale01.png", width = 4, height = 4)
+
+day_preds2 <- ddply(day_preds, c("lat", "long"), mutate, 
+  pred_s = rescale01(pred),
+  pred_m = pred / max(pred),
+  max = max(pred),
+  range = diff(range(pred))
+)
+
+day_preds2 <- glyphs(day_preds2, "long", "day", "lat", "pred_s") 
+ggplot(day_preds2, aes(gx, gy, group = gid)) + 
+  map_nasa + 
+  add_ref_boxes(day_preds) +
[…]

### 2012-02-25 revision/code/templates.r

@@ -1,0 +1,58 @@
+source("glyph.r")
+source("glyph-utils.r")
+# Template time series
+amp <- 30
+mu <- 300
+
+t <- seq(0, pi, length=36)
+x1 <- mu + seq(0, 30, length=36) # linear increase
+x2 <- mu + seq(30, 0, length=36) # linear decrease
+x3 <- mu + 30 + seq(0, 30, length=36) # linear increase + shift
+x4 <- mu + 30 + seq(30, 0, length=36) # linear decrease + shift
+x5 <- mu + amp*sin(t) # inverted U
+x6 <- mu + 30-amp*sin(t) # shifted, inverted U
+x7 <- mu + seq(0, 30, length=36) + amp*sin(t) # inverted U + linear
+x8 <- mu + seq(0, 30, length=36) + 30-amp*sin(t) # shifted, inverted U + linear
+x9 <- mu + amp*sin(t*6) # seasonal
+x10 <- mu + amp*sin(t*6)/2 # seasonal + smaller amplitude
+x11 <- mu + seq(0, 30, length=36) + amp*sin(t*6) # seasonal + linear
+x12 <- mu + seq(0, 30, length=36) + amp*sin(t*6)/2 # seasonal + smaller amplitude + linear
+
+locs <- data.frame(cbind(rep(t,12), c(x1,x2,x3,x4,x5,x6,x7,x8,x9,x10,x11,x12), c(rep(1:6, each=36), rep(1:6, each=36)), rep(2:1,each=36*6)))
+names(locs) <- c("time","temp","x_max", "y_max")
+
+locs2 <- data.frame(cbind(rep(t+pi,12), rep(c(mean(x1),mean(x2),mean(x3),mean(x4),mean(x5),mean(x6),mean(x7),mean(x8),mean(x9),mean(x10),mean(x11),mean(x12)), each=36), c(rep(1:6, each=36), rep(1:6, each=36)), rep(2:1,each=36*6)))
+names(locs2) <- c("time","temp","x_max", "y_max")
+
+library(ggplot2)
+qplot(time,temp, data=subset(locs, x_max==1), geom="line")
+
+# multiple lines
+locs <- glyphs(locs, "x_max", "time", "y_max", "temp")
+ggplot(locs, aes(gx, gy, group = gid)) + 
+  geom_path(colour="grey80", data = ref_lines(locs)) +
+  geom_path(aes(gx, gy, group = gid), colour="black") +
[…]

### 2012-02-28 revision/paper.tex
Re-do first two plots

@@
   \includegraphics[width=3.5in]{nasa-deseas-glyph}
-  \caption{Glyph-map of de-seasonalized temperature, to compare with Figure \ref{fig:nasa-facet}.
+  \includegraphics[width=2in]{nasa-deseas-glyph-leg}
+  \caption{(Left) Glyph-map of de-seasonalized temperature, to compare with Figure \ref{fig:nasa-facet}.
 Time series of the six years of monthly temperature are plotted at each spatial grid location.
@@
 Large variations in temperature can be seen in areas over land, while being fairly constant over water.
-}
+(Right) Glyph-map legend, showing temporal pattern without spatial pattern: this allows you to see the scale of the individual axes.}
   \label{fig:nasa-glyph}

### 2012-02-28 revision/code/1-examples.r

@@ -3,9 +3,9 @@
 # here too
 
 source("glyph.r")
+source("glyph-utils.r")
 source("maps.r")
 source("data-nasa.r")
-source("glyph-utils.r")
 
 temp_seas_models <- dlply(nasa, c("lat", "long"), function(df) {
   lm(surftemp ~ factor(month), data = df)
@@ -30,13 +30,29 @@ preds <- mdply(cbind(d = locs, m = temp_seas_models), function(d, m) {
 
 # First figure: facetted colored map -----------------------------------------
 
+outline_nasa <- list(
+  geom_polygon(aes(long, lat, group = group), inherit.aes = FALSE, 
+    data = .nasa, legend = FALSE, fill = NA, colour = "grey60"),
+  scale_x_continuous(expand = c(0.02, 0)),
+  scale_y_continuous(expand = c(0.02, 0)), 
+  xlab(NULL),
+  ylab(NULL))
+
+# This plot is INCREDIBLY slow in 0.8.9, so I've switched to 0.9.0 to make
+# it a bit faster - it's still not great, but it's at least tolerable.
+
 ggplot(resids, aes(long, lat)) +
-  geom_tile(aes(fill = temp_resid)) +
+  geom_raster(aes(fill = temp_resid)) +
   outline_nasa + 
   facet_grid(year ~ month) + 
   scale_fill_gradient2(mid = "white", high = "red", low = "blue") + 
-  theme_fullframe() +
-  coord_map()
+  theme_fullframe() + 
+  opts(
+    axis.text.x = theme_blank(), 
+    axis.text.y = theme_blank(),
+    axis.ticks = theme_blank(), 
+    axis.ticks.length = unit(0, "lines"), 
+    axis.ticks.margin = unit(0, "lines"))
 ggsave("../images/nasa-colored-map.png", height = 10, width = 18)
 
 # Second figure: same data but with glyphs -----------------------------------
@@ -50,3 +66,10 @@ ggplot(resids.gly, aes(gx, gy, group = gid)) +
   theme_fullframe() 
 ggsave("../images/nasa-deseas-glyph.png", width = 6, height = 6)
 
+# Legend
+ggplot(resids.gly, aes(time, temp_resid)) + 
+  geom_line(aes(group = gid), alpha = 1/0) + 
+  opts(aspect.ratio = 1) + 
+  xlab("Month") + 
+  ylab("Temperature deviation (C)")
+ggsave("../images/nasa-deseas-glyph-leg.png", width = 4, height = 4)

### 2012-02-28 revision/code/glyph-utils.r

@@ -1,19 +1,16 @@
-library("maps")
 library("ggplot2")
-
-source("maps.r")
-source("glyph.r")
+library("grid")
 
 theme_fullframe <- function (base_size = 12){
-  structure(list(
+  opts(
     axis.line = theme_blank(), 
-    axis.text.x = theme_blank(), 
-    axis.text.y = theme_blank(),
-    axis.ticks = theme_blank(), 
+    # axis.text.x = theme_blank(), 
+    # axis.text.y = theme_blank(),
+    # axis.ticks = theme_blank(), 
     axis.title.x = theme_blank(), 
     axis.title.y = theme_blank(), 
-    axis.ticks.length = unit(0, "lines"), 
-    axis.ticks.margin = unit(0, "lines"), 
+    # axis.ticks.length = unit(0, "lines"), 
+    # axis.ticks.margin = unit(0, "lines"), 
     legend.position = "none", 
     panel.border = theme_blank(), 
     panel.grid.major = theme_blank(), 
@@ -21,7 +18,7 @@ theme_fullframe <- function (base_size = 12){
     panel.margin = unit(0, "lines"), 
     plot.background = theme_blank(), 
     plot.margin = unit(rep(2, 4), "mm")
-  ), class = "options")
+  )
 }
 
 add_ref_lines <- function(data, colour = "white", size = 1.5, ...){

### 2012-02-28 revision/code/glyph.r

@@ -1,3 +1,5 @@
+library(plyr)
+
 #' Create the data needed to generate a glyph plot.
 #'
 #' @param data A data frame containing variables named in \code{x_major},

### 2012-02-28 revision/code/maps.r

@@ -31,23 +31,14 @@ states$group <- max(world$group) + states$group
 both <- rbind(world, states)
 
 ##### Maps
-nasa <- getbox(both, xlim = c(-115, -55), ylim = c(-21.1, 36.6))
+.nasa <- getbox(both, xlim = c(-115, -55), ylim = c(-21.1, 36.6))
 map_nasa <- list(
   geom_polygon(aes(long, lat, group = group), inherit.aes = FALSE, 
-    data = nasa, legend = FALSE, fill = "grey80", colour = "grey90"),
+    data = .nasa, legend = FALSE, fill = "grey80", colour = "grey90"),
   scale_x_continuous(expand = c(0.02, 0)),
   scale_y_continuous(expand = c(0.02, 0)), 
   xlab(NULL),
   ylab(NULL))
-
-outline_nasa <- list(
-  geom_polygon(aes(long, lat, group = group), inherit.aes = FALSE, 
-    data = nasa, legend = FALSE, fill = NA, colour = "grey20"),
-  scale_x_continuous(expand = c(0.02, 0)),
-  scale_y_continuous(expand = c(0.02, 0)), 
-  xlab(NULL),
-  ylab(NULL))
-
 
 ushcn <- subset(getbox(both, xlim = c(-126, -65), ylim = c(24, 52)),
   region != "Great Lakes")

### 2012-03-02 revision/paper.tex
added discussion of MAUP

@@
 It is helpful to differentiate glyphs that involve summaries of more than one location, here a heavier line indicates at least two locations have been averaged.
+The aggregation of spatial locations to a single summary can be misleading.
+The sensitivity of analyses to the scale and configuration of aggregation is described by the  modifiable areal unit problem (\cite{Openshaw:1984kx, Fotheringham:1991uq}).
+In this type of glyph map the scale of aggregation is determined, rather arbitrarily, by the icon size.
+This dependence of visual inferences on the icon size is undesirable.
+The ability to interactively modify icon size, and zoom quickly to disaggregate locations would go some way to alleviating the problem.
 

### 2012-03-03 revision/paper.tex
Add colour package to stop complaints
Update plots and add legend to ref frames section

@@
 \graphicspath{{images/}}
+\usepackage[dvipsnames,usenames]{color}
 \usepackage{hyperref}
@@
 
+  \includegraphics[width=0.33\linewidth]{ref-legend}
+
   \caption{Glyph-maps of seasonal temperature patterns (averages for each month), using the data-expo data.
-Adding mid-range reference lines (left) and grid-cell reference boxes (right), makes it easier to see differences in glyph position, not just shape.}
+Adding mid-range reference lines (top-left) and grid-cell reference boxes (top-right), makes it easier to see differences in glyph position, not just shape.
+(Bottom) Legend shows overall scales to aid interpretation.}
   \label{fig:ref-basic}

### 2012-03-03 revision/code/3-scaling.r

@@ -2,39 +2,9 @@ library("ggplot2")
 library("plyr")
 library("mgcv")
 
-source("data-nasa.r")
+source("models-nasa")
 source("glyph.r")
-source("glyph-utils.r")
-source("cache.r")
 source("maps.r")
-
-# Model fitting --------------------------------------------------------------
-
-temp_models %<-cache% dlply(nasa, c("lat", "long"), function(df) {
-  lm(surftemp ~ year + factor(month), data = df)
-})
-
-temp_smooth %<-cache% dlply(nasa, c("lat", "long"), function(df) {
-  gam(surftemp ~ s(day) + factor(month), data = df)
-})
-
-month_grid <- expand.grid(year = 2000, month = 1:12)
-month_preds <- ldply(temp_models, function(mod) {
-  month_grid$pred <- predict(mod, newdata = month_grid)
-  month_grid
-})
-
-year_grid <- expand.grid(year = unique(nasa$year), month = 1)
-year_preds <- ldply(temp_models, function(mod) {
-  year_grid$pred <- predict(mod, newdata = year_grid)
-  year_grid
-})
-
-day_grid <- expand.grid(day = seq(0, 2161, length = 50), month = 1)
-day_preds <- ldply(temp_smooth, function(mod) {
-  day_grid$pred <- predict(mod, newdata = day_grid)
-  day_grid
-})
 
 # Rescale predictions to individual scales -----------------------------------
 

### 2012-03-03 revision/code/3-scaling.r
Legends for scaling

@@ -16,6 +16,13 @@ ggplot(day_preds, aes(gx, gy, group = gid)) +
   theme_fullframe()
 ggsave("../images/month-rescale-none.png", width = 4, height = 4)
 
+ggplot(day_preds, aes(day, pred)) + 
+  geom_line(aes(group = gid), alpha = 1/10) + 
+  opts(aspect.ratio = 1) + 
+  xlab("Days from start") + 
+  ylab("Average temperature (K)")
+ggsave("../images/month-rescale-legend.pdf", width = 4, height = 4)
+
 last_plot() %+% glyphs(day_preds, "long", "day", "lat", "pred", 
   y_scale = range01) 
 ggsave("../images/month-rescale01.png", width = 4, height = 4)
@@ -33,8 +40,13 @@ ggplot(day_preds2, aes(gx, gy, group = gid)) +
   add_ref_boxes(day_preds) +
   geom_path(aes(colour = range)) +
   theme_fullframe() + 
-  scale_colour_gradient(high = "black", low = "grey60")
-ggsave("../images/month-rescale01-col.png", width = 4, height = 4)
+  scale_colour_gradient("Temperature\nrange (K)",
+    high = "black", low = "grey60", limits = c(0, 8.5),
+    breaks = seq(0, 8, by = 2), guide = guide_colourbar(
+      direction = "horizontal", title.vjust = 0.7, 
+      title.theme = theme_text(face = "bold"))) +
+  opts(legend.position = "bottom", aspect.ratio = 1) 
+ggsave("../images/month-rescale01-col.png", width = 4, height = 4.5)
 
 grid <- unique(day_preds2[c("lat", "long", "range")])
 
@@ -43,8 +55,13 @@ ggplot(day_preds2) +
   geom_tile(aes(long, lat, fill = range), data = grid, alpha = 0.5) +
   geom_path(aes(gx, gy, group = gid)) +
   theme_fullframe() + 
-  scale_fill_gradient(high = "white", low = "#3B4FB8")
-ggsave("../images/month-rescale01-fill.png", width = 4, height = 4)
+  scale_fill_gradient("Temperature\nrange (K)",
+    high = "white", low = "#3B4FB8", limits = c(0, 8.5),
+    breaks = seq(0, 8, by = 2), guide = guide_colourbar(
+      direction = "horizontal", title.vjust = 0.7, 
+      title.theme = theme_text(face = "bold"))) +
[…]

### 2012-03-03 revision/paper.tex
Legends for scaling
Legends for generalisations

@@
   \includegraphics[width=0.5\linewidth]{month-rescale01}
-  \caption{Smoothed de-seasonalized daily temperature glyph-map. (Left)
-globally scaled, (right) locally scaled.}
+
+  \includegraphics[width=0.33\linewidth]{month-rescale-legend}
+
+  \caption{Smoothed de-seasonalized daily temperature glyph-map. (Top-left)
+globally scaled, (Top-right) locally scaled. (Bottom) Legend showing ranges of individual glyphs.}
   \label{fig:scaling}
@@
 
-  \caption{Glyph-map showing (left) scatterplot of temperature vs high-cloud and (right) smoothed (loess) curve fit to each location.
+  \includegraphics[width=0.33\linewidth]{nasa-scat-legend}%
+  \hspace{0.17\linewidth}%
+  \includegraphics[width=0.33\linewidth]{nasa-loess-legend}
+
+  \caption{Glyph-map showing (top-left) scatterplot of temperature vs high-cloud and (top-right) smoothed (loess) curve fit to each location.
 The relationship between these two variables varies considerably over the spatial domain.
-}
+(Bottom) Legends show overall patterns and scales.}
   \label{fig:cloud}

### 2012-03-04 revision/paper.tex
Legends for large data section

@@
 
-  \caption{Glyph-map of raw temperature anomaly data (GISTEMP), from 1880-2011.
-Gaps in the full circle indicate missing values, and thickness of the circle indicates more variability in the measurements.}
+  \includegraphics[width=0.33\linewidth]{gistemp-polar-legend}
+
+  \caption{(Top) Glyph-map of raw temperature anomaly data (GISTEMP), from 1880-2011.
+Gaps in the full circle indicate missing values, and thickness of the circle indicates more variability in the measurements.
+(Bottom) Legend showing ranges of individual glyphs}
   \label{fig:gistemp-raw}
@@
   \centering
-  \includegraphics[width=1\linewidth]{gistemp-pred}%
+  \includegraphics[width=1\linewidth]{gistemp-pred}
+
+  \includegraphics[width=0.33\linewidth]{gistemp-pred-legend}
   %\includegraphics[width=1\linewidth]{gistemp-polar-pred}
 
-  \caption{Glyph-map of predicted temperature anomaly data (GISTEMP),
+  \caption{(Top) Glyph-map of predicted temperature anomaly data (GISTEMP),
     from 1950-2010. Increasing trends can be seen over most of the
@@
     straight, but they look curved because of the Z\"ollner optical
-    illusion.}
+    illusion. (Bottom) Legend showing ranges of individual glyphs}
   \label{fig:gistemp-pred}

### 2012-03-04 revision/code/4-large.r

@@ -1,0 +1,140 @@
+library(ncdf)
+library(reshape2)
+library(ggplot2)
+library(lubridate)
+
+# Predictions
+source("maps.r")
+source("glyph.r")
+
+# Load data ------------------------------------------------------------------
+
+if (file.exists("temp_usa.rds")) {
+  temp_usa <- readRDS("temp_usa.rds")
+} else {
+  if (!file.exists("air.2x2.250.mon.anom.land.nc")) {
+  download.file("ftp://ftp.cdc.noaa.gov/Datasets/gistemp/landonly/250km/air.2x2.250.mon.anom.land.nc", "air.2x2.250.mon.anom.land.nc")
+  }
+
+  fo<-open.ncdf("air.2x2.250.mon.anom.land.nc")
+  fo$nvars
+  lon <- get.var.ncdf(fo, "lon")
+  lat <- get.var.ncdf(fo, "lat")
+  time <- get.var.ncdf(fo, "time")
+  temp <- get.var.ncdf(fo, "air")
+  # Convert lon from 1 to 360 to -180 to 180
+  lon <- (lon + 180) %% 360 - 180
+
+  length(lon)
+  dim(temp)
+
+  # Ok, this now works for transforming the lon, and selecting the USA
+  temp.usa <- temp[114:150, 20:33,] # lon -130, -65; lat 25, 50
+  temp_usa <- melt(temp.usa)
+  colnames(temp_usa) <- c("gridx", "gridy", "time", "temp")
+  temp_usa$lon <- lon[temp_usa$gridx + 113]
+  temp_usa$lat <- lat[temp_usa$gridy + 19]
+
+  temp_usa <- subset(temp_usa, !(lon > -70 & lat < 37))
+
+  library(lubridate)
+  temp_usa$date <- ymd(18800101) + months(temp_usa$time - 1)
+  temp_usa$year <- year(temp_usa$date)
+  temp_usa$month <- month(temp_usa$date)  
+  
+  saveRDS(temp_usa, "temp_usa.rds")
+}
+
+# x<-ddply(temp_usa, c("lon","lat"), function(df) {
+#  if (nrow(subset(df,is.na(temp))) == nrow(df)) NULL else df 
+# })
+
+# Plots of raw data -------------------------------------------------------
+
+# temp_glyphs <- glyphs(temp_usa, "lon", "time", "lat", "temp")
+# 
+# ggplot(temp_glyphs, aes(gx, gy, group = gid)) + 
+#   map_gistemp +
[…]

### 2012-03-04 revision/code/maps.r

@@ -1,5 +1,6 @@
 library("maps")
 library("ggplot2")
+library("plyr")
 
 getbox <- function (map, xlim, ylim) {
 	# identify all regions involved

### 2012-03-04 revision/paper.tex
abstract and intro

@@
 
+% resolution issues with large data - reference - aloso long time series.
+
+% emphasize way of thinking, which leads into interactive setting
+
 \begin{abstract}
 
-Understanding the multiple aspects of climate data is a challenge.
+The multivariate spatio-temporal nature of climate data makes it difficult to draw all of the aspects simultaneously.
+This paper describes the conceptualization and construction of a type of display, a glyph-map, that has been used infrequently for climate data.
+Glyph-maps are a specialization of multivariate glyph plots.
+Each spatial location is displayed with one glyph that represents the multiple measurements, recorded over time, at that location.
+Glyph-maps allow the discovery of both local and global structure, with a particular focus on temporal relationships.
+They provide alternatives to colored, facetted maps, or statistical summaries such as principal components.
+With the new conceptualization, developing interactive versions of glyph-maps will be easier, and it will be simpler to explore the perceptual effects of different scales.
+The methods are developed for rectangularly gridded data, but with some clever processing, which is explained, it is possible get good glyph-maps of irregularly gridded data.
+
+%Understanding the multiple aspects of climate data is a challenge.
 There are both spatial and temporal components to examine, and often multivariate dependencies.
@@
 
-The instructions for computing and displaying glyph-maps are described, along with the design of reference frames, and the effects of different scaling of the data on perception of structure.
+%The instructions for computing and displaying glyph-maps are described, along with the design of reference frames, and the effects of different scaling of the data on perception of structure.
 The methods are developed for rectangularly gridded data, but with some clever processing, which is explained, it is possible get good glyph-maps of irregularly gridded data.
@@
 
-Climate data is composed of measurements, such as temperature, precipitation, and winds, with a spatial and temporal context.
+Climate data is composed of multiple variables, such as temperature, ozone, precipitation, and winds, with a spatial and temporal context.
 The classic display for data of this type of is ``heatmap'', a tiled plot using color to display the value of the variable of interest at each location.
-When measurements are made at multiple time points it is common to display the data with small multiples, such as in Figure~\ref{fig:nasa-facet}.
+When measurements are made at multiple time points it is reasonably common to display the data with small multiples~\citep{tufte:2001,IPCC}, such as in Figure~\ref{fig:nasa-facet}.
 In this plot, a separate map is drawn for each month (columns) and year (rows), over a six year period of remotely sensed temperature data above Central America \citep{murrell:2010}.
@@
 The most noticeable feature is the strong red patch in the equatorial Pacific beginning mid-1997 and tapering out during 1998.
-This is the El Ni\~no event, a major temperature anomaly.
-More work is required to see more localized patterns, like the cooler land temperatures seen in early years.
+This is the El Ni\~no event, a major temperature anomaly (see for example, \citet{trenberth} or \citet{ElNino2}.) More work is required to see more localized patterns, like the cooler land temperatures seen in early years.
 
-Reading this type of plot is cognitively challenging: the reader must play ``spot the difference'' from one small image to another.
-Large structures such as El Ni\~no are clear but it is very difficult to mentally difference the images and read off the long-term trend or to notice local deviations.
+To read long-term or localized temporal patterns, this type of plot is cognitively challenging: the reader must play ``spot the difference'' from one small image to another.
+Large spatial structures such as El Ni\~no are clear but it is very difficult to extract the temporal pattern because it requires mentally differencing the images.
 Using a movie to render these small multiples can help, but small trends still fail to draw readers' attention and escape unnoticed \citep{simons:gradual}.
+The \emph{glyph-map} display described in this paper enables easier perception of temporal patterns.
 
@@
 
-This paper develops a new type of display to resolve these problems: the \emph{glyph-map}.
-The glyph-map uses a small glyph, or icon, to represent multiple values at each location.
-Glyph maps are an adaptation of glyphs, tools developed for display multivariate data.
-Multivariate glyphs include the star plots of \citet{mayr:1877}, the semi-graphic displays of \citet{anderson:1960}, and the infamous Chernoff faces \citep{chernoff:1973}.
-An glyph is produced by mapping each variable to some graphical feature, such as the length of a line.
-The glyph-map can also be thought of as a small multiple display \citep{tufte:2001} of time series with a geographic context.
+The glyph-map uses a small glyph, or icon, to represent multiple variables, or time values, at each location.
+Glyph maps derive from glyph plots, which were initially developed to display multivariate data.
+One glyph is drawn for each observation in the data set.
+Types of glyph plots include the star plots \citep{mayr:1877, kleiner:1981}, the semi-graphic displays \citep{anderson:1960}, and the infamous Chernoff faces \citep{chernoff:1973}.
+An glyph is produced by mapping each variable to some graphical feature, such as the length of a line, curvature of the mouth, or size of the face.
 
-Figure~\ref{fig:nasa-glyph} shows the glyph-map equivalent of Figure~\ref{fig:nasa-facet}: each location is represented by a small time series glyph.
-The primary visible structure is still El Ni\~no, the regions in the Pacific with bump in the middle of the series, but other features are also visible: temperatures are generally more varied over land and a few locations have dramatically increasing temperature in the later years typically in high elevation areas, along the Andes and the Sierra Madre Occidental in Mexico.
+Figure~\ref{fig:nasa-glyph} shows the glyph-map equivalent of Figure~\ref{fig:nasa-facet}: each location is represented by a small time series glyph of de-seasonalized temperature.
+The darker grey in the background represents the land mass of the study region.
+Because the data is de-seasonalized, the primary visible structure is still El Ni\~no -- it is both a temporal anomaly and a spatial phenomenon.
+It can be recognized by the bump in the middle of the series of the equatorial Pacific region.
+Other temporal features are visible from the many time series.
+Temperatures are generally more varied over land, than sea.
+A few locations have dramatic jumps in temperature in the later years, all in high elevation areas, along the Andes and the Sierra Madre Occidental in Mexico.
+
+The glyph-map can also be thought of as a small multiple display \citep{tufte:2001}.
+The primary comparison is different.
+Cognitively, elements placed close to each other are more easily and accurately compared \citep{cleveland:1984}.
+The facetted colored map, places greater emphasis on comparison between the spatial locations, because these are positioned together, and secondary emphasis on time because these are separated by facets.
+In the glyph-map primary emphasis is placed on the temporal comparison, and secondarily on spatial comparison, because facetting is on spatial locations.
 

### 2012-03-05 revision/paper.tex
abstract and intro

@@
 The methods are developed for rectangularly gridded data, but with some clever processing, which is explained, it is possible get good glyph-maps of irregularly gridded data.
+Guides, and reference marks, for different types of glyphs used in the glyph-maps are also discussed.
 
@@
 The most noticeable feature is the strong red patch in the equatorial Pacific beginning mid-1997 and tapering out during 1998.
-This is the El Ni\~no event, a major temperature anomaly (see for example, \citet{trenberth} or \citet{ElNino2}.) More work is required to see more localized patterns, like the cooler land temperatures seen in early years.
+This is the El Ni\~no event, a major temperature anomaly (see for example, \citet{trenberth} or \url{http://www.esrl.noaa.gov/psd/enso/enso.description.html}.) More work is required to see more localized patterns, like the cooler land temperatures seen in early years.
 
@@
 Temperatures are generally more varied over land, than sea.
-A few locations have dramatic jumps in temperature in the later years, all in high elevation areas, along the Andes and the Sierra Madre Occidental in Mexico.
+A few locations have dramatic jumps in temperature in the later years.%, all in high elevation areas, along the Andes and the Sierra Madre Occidental in Mexico.
 
-The glyph-map can also be thought of as a small multiple display \citep{tufte:2001}.
-The primary comparison is different.
+The glyph-map can also be thought of as a small multiple display.
+The primary comparison is different from the facetted map.
 Cognitively, elements placed close to each other are more easily and accurately compared \citep{cleveland:1984}.
 The facetted colored map, places greater emphasis on comparison between the spatial locations, because these are positioned together, and secondary emphasis on time because these are separated by facets.
-In the glyph-map primary emphasis is placed on the temporal comparison, and secondarily on spatial comparison, because facetting is on spatial locations.
+In the glyph-map, primary emphasis is placed on the temporal comparison, and secondarily on spatial comparison, because facetting is on spatial locations.
 
@@
 
-One problem with multivariate glyph is that lack of natural ordering of each glyph, or of the mapping between variables and glyph properties.
-This leads to a combinatorial explosion of possibilities, each of which may have dramatically different perceptual properties.
-This problem has prevented the widespread adoption of glyph displays, despite some proposed remedies \citep{kleiner:1981,hurley:2010}.
-Using glyphs with space-time data eliminates these problems: glyphs are placed according to the location of the measurement, and variables are ordered by the time they were collected.
+In general, multivariate glyph displays lack a natural ordering of the glyphs because observations have no ordering.
+Similarly, there are multiple ways that variables can be mapped to glyph property.
+It can result in a combinatorial explosion of possible mappings, each producing a display that may emphasize some entirely different features of the data than another mapping.
+This may have prevented the widespread adoption of glyph displays, despite some proposed ordering solutions \citep{kleiner:1981,hurley:2010}.
+Using glyphs with space-time data eliminates this issue: glyphs are placed according to the location of the measurement, and variables are ordered by the time they were collected.
 Additionally, climate data usually has high correlations between nearby locations and times.
-This imposes a degree of smoothness that gives the plot an appearance of a textured  landscape, making it easier to digest patterns.
+This imposes a degree of smoothness that gives the plot an appearance of a textured  landscape, making it easier to digest patterns from one glyph to another.
 Others \citep{pickett:1988} have used glyphs on maps to display multivariate spatial data.
@@
 \citet{gribov:2006} describes the use of glyph-maps for multivariate data also, with emphasis on the graphics software Gauguin.
+Glyph-maps for climate data have been used in several publications \citep{carr:1992,eden:2010,hobbs:2010}.
+Glyph-maps have commonly been used for displaying wind direction and strength, as can be seen in the Figure 8.1 of \citet{IPCC} and a rudimentary  time series glyph display can be seen in Figure 9.12.
 

### 2012-03-05 revision/paper.tex
finished intro

@@
 \citet{gribov:2006} describes the use of glyph-maps for multivariate data also, with emphasis on the graphics software Gauguin.
-Glyph-maps for climate data have been used in several publications \citep{carr:1992,eden:2010,hobbs:2010}.
-Glyph-maps have commonly been used for displaying wind direction and strength, as can be seen in the Figure 8.1 of \citet{IPCC} and a rudimentary  time series glyph display can be seen in Figure 9.12.
+Glyph-maps for spatiotemporal data have been used in several publications \citep{carr:1992,eden:2010,hobbs:2010}.
+Glyph-maps using arrow glyphs have commonly been used for displaying ocean currents, wind direction and strength, as can be seen in the Figure 8.1 of \citet{IPCC} and a rudimentary  time series glyph display can be seen in Figure 9.12.
 
@@
 
-The paper is structured as follows.
-Section~\ref{sec:construction} describes the algorithm used to create glyphs-maps.
-Section~\ref{sec:perception} discusses their perceptual properties, including the importance of a visual reference grid, and of careful consideration of scale.
-Large data and the interplay of models and data are discussed in Section~\ref{sec:large-data}.
-Many spatiotemporal data sets have irregular spatial locations, and Section~\ref{sec:irregular} discusses how glyph-maps can be adjusted for this type of data.
+Section~\ref{sec:construction} describes the algorithm used to create glyphs-maps of these two types of glyphs.
+Section~\ref{sec:perception} discusses their perceptual properties, including the importance of a visual reference grids, and of careful consideration of scale.
+Large data and the interplay with models and data are discussed in Section~\ref{sec:large-data}.
+Many spatiotemporal data sets have irregular spatial locations, and Section~\ref{sec:irregular} discusses how glyph-maps can be constructed for this type of data.
 Three datasets are used for examples:
@@
 %Glyph-maps can be generated with existing graphics software after performing a simple pre-processing step, making them easily accessible to a wide audience. 
-Creating a glyph-map is elementary.
-It involves recognizing that the glyph-map are a linear mix of two structural components of the data: spatial location and data values.
+To create a glyph-map it is useful to recognize that the glyph-map are a linear mix of two structural components of the data: spatial location and data values.
 The spatial location is the major positioning component, while the data values are minor adjustments to those positions.

### 2012-03-05 revision/paper.tex
finished perception

@@
 Climate data is composed of multiple variables, such as temperature, ozone, precipitation, and winds, with a spatial and temporal context.
-The classic display for data of this type of is ``heatmap'', a tiled plot using color to display the value of the variable of interest at each location.
+The classic display for data of this type is a ``heatmap'', a tiled plot using color to display the value of the variable of interest at each location.
 When measurements are made at multiple time points it is reasonably common to display the data with small multiples~\citep{tufte:2001,IPCC}, such as in Figure~\ref{fig:nasa-facet}.
@@
 The glyph-map can also be thought of as a small multiple display.
-The primary comparison is different from the facetted map.
+The primary comparison is different from the facetted heatmap.
 Cognitively, elements placed close to each other are more easily and accurately compared \citep{cleveland:1984}.
-The facetted colored map, places greater emphasis on comparison between the spatial locations, because these are positioned together, and secondary emphasis on time because these are separated by facets.
+The facetted heatmap, places greater emphasis on comparison between the spatial locations, because these are positioned together, and secondary emphasis on time because these are separated by facets.
 In the glyph-map, primary emphasis is placed on the temporal comparison, and secondarily on spatial comparison, because facetting is on spatial locations.
@@
   \includegraphics[width=5.5in]{nasa-colored-map.png}
-  \caption{Facetted map of de-seasonalized temperature.
+  \caption{Facetted heatmap of de-seasonalized temperature.
 The dominant feature is the El Ni\~no warming in the southern equatorial region in the last half of 1997 and first half of 1998.
@@
 Glyph-maps using arrow glyphs have commonly been used for displaying ocean currents, wind direction and strength, as can be seen in the Figure 8.1 of \citet{IPCC} and a rudimentary time series glyph display can be seen in Figure 9.12.
+
+Another common approach for displaying spatiotemporal data is to calculate a statistical summary for each location and display these values as single heatmap.
+For example, to study long-term trends, the slope of a linear model can be calculated at each location and color tiles used to display the slope value over space.
+This approach is less than exploratory, in the sense that it requires a pre-conceived definition of the relationship of interest, and clean data so that the summary statistic adequately summarizes the pattern.
 
@@
 %Glyph-maps can be generated with existing graphics software after performing a simple pre-processing step, making them easily accessible to a wide audience. 
-To create a glyph-map it is useful to recognize that the glyph-map are a linear mix of two structural components of the data: spatial location and data values.
+To create a glyph-map it is useful to recognize that the glyph-map is a linear mix of two structural components of the data: spatial location and data values.
 The spatial location is the major positioning component, while the data values are minor adjustments to those positions.
-For spatiotemporal data, the major axes are latitude ($y_{\amaj}$) and longitude ($x_{\amaj}$), and the minor axes are time ($x_{\amin}$) and some measurement ($y_{\amin}$), for example, temperature, or predicted temperature.
-Assuming the minor axes are rescaled to $[-1, 1]$, the final coordinates $(x,y)$ on the chart are the linear combination, given by:
+For spatiotemporal data, the major components are latitude ($y_{\amaj}$) and longitude ($x_{\amaj}$), and the minor components are time ($x_{\amin}$) and some measurement ($y_{\amin}$), for example, temperature, or predicted temperature.
+Assuming the minor components are rescaled to $[-1, 1]$, the final coordinates $(x,y)$ on the chart are the linear combination, given by:
 
@@
 
-\noindent where $w$ and $h$ are maximal width and height of the glyph, respectively.
-For gridded data, it is convenient to make use of the resolution of the major axes, that is, the minimum difference between neighboring locations.
+\noindent where $w$ and $h$ are scale parameters, setting the width and height of the glyph, respectively.
+For gridded data, $w$ and $h$ will be the minimum difference between neighboring spatial locations to make use of the resolution of the major components.
 
-Since the coordinates are linear combinations of two variables, it is possible to build these these line-glyphs interactively, illustrated by software packages that have incorporated tours (as described in \citet{cook:2006}), such as DataViewer \citep{buja:1986}, XGobi \citep{swayne:1991} or GGobi \citep{swayne:2003}.
+This way of thinking of the glyph-map is convenenient for several reasons.
+The first is that it moves the plots beyond the small multiples, many little plots, approach.
+That provides flexibility to cover large spatial domains using potentially tiny glyphs.
+A second, and major, reason is that with this approach they can then be readily made to be interactive.
+Different variables can be swept into, and out of, the minor axes using mouse motion by varying $w$ and $h$ proportional to mouse movement.
+Interactive glyph-maps were used to explore the data-expo data using  GGobi \citep{swayne:2003}, and then later reproduced in publication quality resolution using {\tt ggplot2} \citep{me:ggplot2} in \citet{hobbs:2010}.
+In GGobi, a manual tour is used to make the linear combination of spatial variables with other variables \citep{CB95}, and equation \ref{coords.eqn} makes this approach more explicit.
+The third reason is that it enables adjusting scales of glyphs in multiple ways to vary the emphasis, which is discussed in Section \ref{sec:perception}.
+
+%Since the coordinates are linear combinations of two variables, it is possible to build these these line-glyphs interactively, illustrated by software packages that have incorporated tours (as described in \citet{cook:2006}), such as DataViewer \citep{buja:1986}, XGobi \citep{swayne:1991} or.
 This technique is used to good effect in \citet{buja:1996a}, and was used to examine the climate data in \citet{hobbs:2010}.
@@
 
-Figure~\ref{fig:templates} gives some small examples of time series with prototypical trends, shown both in cartesian coordinates and polar coordinates.
+Figure~\ref{fig:templates} gives some small examples of time series with prototypical trends, shown both in Cartesian coordinates and polar coordinates.
 Differences between linear and nonlinear trend are more apparent in the line-glyphs, and are effectively lost in the star-glyphs, which are most effective in exposing cyclical patterns due to seasonality.
@@
 Glyph-maps allow time trends to be read directly from the plot.
-Values are mapped to position, one of the easiest properties to perceive \citep{cleveland:1984}.
-Perceiving time trends in faceted heatmaps is much more difficult: not only do you need to read value from color (a relatively difficult task), you also need to spot the difference between the different maps.
-This is challenging from a cognitive perspective \citep{healey:2011,busey} and comparisons can suffer from \emph{change blindness}, often leading to failure to notice the change and not knowing that something changed.
+Values are mapped to position, one of the easiest properties to perceive, rather than color, which is among the hardest graphical element to parse \citep{cleveland:1984}.
+Perceiving time trends in faceted heatmaps is much more difficult: not only do you need to read value from color (difficult), you also need to spot the difference (major challenge) between the different maps.
+From a cognitive perspective these sorts of comparisons can suffer from \emph{change blindness}, often leading to failure to notice the change and not knowing that something changed \citep{healey:2011,busey}.
 To see this, examine Figures~\ref{fig:nasa-facet} and \ref{fig:nasa-glyph}.
 The heatmap requires comparing color values across maps to assess temporal change but the glyph-map allows direct reading of the temporal trend at each location, and comparison across spatial neighbors.
+Conversely, if the purpose is to read spatial trend for one time point the heatmap may be a better choice, because this is a difficult task with the glyph-map.
 
@@
 
-One draw-back of the glyph-maps, particularly if linear trend is displayed as an icon, is they may suffer from the Z\"ollner Illusion \citep{Zollner}, visible in Figure \ref{fig:gistemp-pred}, which makes straight lines look crooked.
-
-A third alternative is to calculate the relevant statistic for each location and display these values as single heatmap.
-For example, to study long-term trends, we could calculate the slope of a linear model at each location and color tiles across the map by the value of the slope.
-The disadvantage of this approach is that it is not exploratory, and requires that we have a very precise definition of the measurement we are interested in, and it still suffers from the relatively difficult color perception.
+A possible draw-back of the glyph-map, particularly if linear trend is displayed as an icon, is it may suffer from the Z\"ollner Illusion \citep{Zollner}, visible in Figure \ref{fig:gistemp-pred}, which makes straight lines look crooked.
 
@@
 This makes climate data arduous to work with computationally, and to visualize.
-Careful use of models to decompose data into long-term trend, seasonal effects and error is vital for to make the problem both computationally and visually tractable.
+There is a limit to the resolution of human visual system \citep{Kr12} which will prevent perfect perception of large spatial domains or seasonality of many years of data.
+Exploring large data requires reducing it to a resolution that can be reasonably absorbed, and statistical models can be useful for this.
+Models can be used to decompose data into long-term trend, seasonal effects and error, which is vital for making the problem both computationally and visually tractable.
 

### 2012-03-05 revision/paper.tex
getting close to addressing all comments

@@
 
-% resolution issues with large data - reference - aloso long time series.
+% resolution issues with large data - reference - also long time series - large spatial domain.
+
+% order of the algorithm calculations
 
 % emphasize way of thinking, which leads into interactive setting
+
+% refer to the wrapping of times series in eden
 
@@
 One glyph is drawn for each observation in the data set.
-Types of glyph plots include the star plots \citep{mayr:1877, kleiner:1981}, the semi-graphic displays \citep{anderson:1960}, and the infamous Chernoff faces \citep{chernoff:1973}.
+Types of glyph plots include the star plots \citep{mayr:1877, FD01, kleiner:1981}, the semi-graphic displays \citep{anderson:1960}, and the infamous Chernoff faces \citep{chernoff:1973}.
 An glyph is produced by mapping each variable to some graphical feature, such as the length of a line, curvature of the mouth, or size of the face.
@@
   $2^{\circ}$ grid over the entire globe, measured monthly
-  \citep{GISTEMP}. Ground station data was de-seasonalized,
+  \citep{GISTEMP} from 1880-2011. Ground station data was de-seasonalized,
   differenced from from the 1951-1980 temperature averages, and
@@
 
-A possible draw-back of the glyph-map, particularly if linear trend is displayed as an icon, is it may suffer from the Z\"ollner Illusion \citep{Zollner}, visible in Figure \ref{fig:gistemp-pred}, which makes straight lines look crooked.
+A possible draw-back of the glyph-map, particularly if linear trend is displayed as an icon, is it may suffer from the Z\"ollner Illusion \citep{Zollner}, which makes straight lines look crooked (Section \ref{sec:large-data} has an example).
 
@@
 Reference elements are drawn in white, so they are minimally perceptible relative to the black data.
+Average temperatures were calculated for each month, and these were passed to the glyph calculations to produce the glyph-maps.
 
@@
 
-For star glyphs the reference frames are also useful and the equivalent of the reference line is a circle. Figures~\ref{fig:templates} and \ref{fig:gistemp-raw} show more examples.
+For star glyphs the reference frames are also useful and the equivalent of the reference line is a circle (see Figure~\ref{fig:templates}).
 
@@
 Figure~\ref{fig:scaling} compares global scaling and local scaling for the smoothed temperature data from the data-expo data.
-On the global scale, the El Ni\~no blip in temperature is just visible, along with variation in the average temperature at each location.
-Local scaling emphasizes the individual shapes: the impact of El Ni\~no can be seen to be over a wider region, dramatic linear increases across the Andes, and decreases in the Gulf of Mexico.
+(Generalized additive model using smooth terms as described in \citet{wood:2006} were used.)  On the global scale, the El Ni\~no blip in temperature is just visible, along with variation in the average temperature at each location.
+Local scaling emphasizes the individual shapes: the impact of El Ni\~no can be seen to be over a wider region, linear increases across the Andes, and decreases in the Gulf of Mexico.
 

### 2012-03-05 revision/paper.tex
Legends for irregular section

@@
 Nearby stations whose icons would overlap have been collapsed.
-Each location has a glyph in the seasonal plot and color of the tiles corresponds to the average temperature of the locations (red=warm, blue=cool).
-Heavy lines in the linear trend plot correspond to average trends over more than one location.}
+Each location has a glyph in the seasonal plot and color of the tiles corresponds to the average temperature of the locations.
+Heavy lines in the linear trend plot correspond to average trends over more than one location.
+Legend in Figure~\ref{fig:irregular-legend}}
   \label{fig:irregular-collapsed}
@@
   \caption{Glyph-maps of seasonal patterns (left) and linear trends (right) of USHCN stations, 1950--2010.
-Station locations have been rounded to the nearest degree Each location has a glyph in the seasonal plot and color of the tiles corresponds to the average temperature of the locations (red=warm, blue=cool).
-Heavy lines in the linear trend plot correspond to average trends over more than one location.}
+Station locations have been rounded to the nearest degree Each location has a glyph in the seasonal plot and color of the tiles corresponds to the average temperature of the locations.
+Heavy lines in the linear trend plot correspond to average trends over more than one location.
+Legend in Figure~\ref{fig:irregular-legend}}
   \label{fig:irregular-grid}
+\end{figure}
+
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=0.33\linewidth]{usa-season-legend}%
+  \hspace{0.18 \linewidth}%
+  \includegraphics[width=0.33\linewidth]{usa-lin-legend}
+
+  \caption{Legend for Figures~\ref{fig:irregular-collapsed} aad \ref{fig:irregular-grid}.}
+  \label{fig:irregular-legend}
 \end{figure}

### 2012-03-05 revision/code/5-irregular.r

@@ -5,44 +5,28 @@ source("glyph-utils.r")
 source("5-overlapping.r")
 source("data-ushcn.r")
 
-temp.post1950 <- subset(ushcn, year >= 1950)
 
 season_index <- function(time){
-  (time %% 12  + 1 )/ 12
+  (time %% 12  + 1) / 12
 }
 
+temp.post1950 <- subset(ushcn, year >= 1950)
+
 temp.post1950 <- mutate(temp.post1950, 
-    time = (year - 1950) * 12 + as.numeric(month),
-    season = season_index(time)
-    )    
+  time = (year - 1950) * 12 + as.numeric(month),
+  season = season_index(time))    
 
-width.mod <- (diff(range(ushcn$lon))/diff(range(gistemp$lon)))/(diff(range(ushcn$lat))/diff(range(gistemp$lat)))
+# Modelling ------------------------------------------------------------------
 
-#==== Modelling
+fit.post1950 <- dlply(temp.post1950, "stn", function(df) {
+  lm(temp ~ year + month, data = df, na.action = "na.exclude")
+}, .progress = "text")
 
-# seasonal.fit.post1950 <- dlply(temp.post1950, "stn", function(df){
-#   gam(temp ~ s(season, bs = "cc"), 
-#     knots = list(season = c(0, 12)),
-#     data = df, na.action = "na.exclude")
-#   }, .progress = "text"
-# ) 
-fit.post1950 <- dlply(temp.post1950, "stn", function(df){
-  lm(temp ~ year + month, 
-    data = df, na.action = "na.exclude")
-  }, .progress = "text"
-) 
-
-temp.post1950 <- cbind(temp.post1950, 
-  ldply(fit.post1950, function(fit) {
-    data.frame(res = residuals(fit), pred = predict(fit))},
-  .progress = "text"))
-
-lin.pred.post1950 <- ldply(fit.post1950, function(fit){
-    one.df <- data.frame(year = 1950:2010, 
-      month = "jan")
-    one.df$pred <- predict(fit, one.df)
-    one.df
-  }, .progress = "text")
+year_grid <- expand.grid(year = 1950:2010, month = "jan")
+lin.pred.post1950 <- ldply(fit.post1950, function(mod) {
+  year_grid$pred <- predict(mod, newdata = year_grid)
+  year_grid
+})
 
[…]

### 2012-03-05 revision/code/data-ushcn.r

@@ -1,5 +1,6 @@
 library(maps)
 library(ggplot2)
+library(reshape2)
 library(plyr)
 
 source("maps.r")
@@ -8,7 +9,7 @@ ushcn.loc <- "../../data/ushcndata.rds"
 ushcn.stations.loc <- "../../data/ushcnstations.rds"
 
 if (!file.exists(ushcn.loc)) {
-  temp.all<-read.table("../data/9641C_201012_raw.avg")
+  temp.all<-read.table("../../data/9641C_201012_raw.avg")
   colnames(temp.all)<-c("year", "jan", "feb", "mar", "apr", "may", "jun", "jul",
     "aug", "sep", "oct", "nov", "dec", "ann")
   temp.all$stn<-temp.all$year%/%100000

### 2012-03-05 revision/paper.tex
fixed urls and finished discussion

--- 2012-03-05 revision/paper.tex
@@
 The classic display for data of this type is a ``heatmap'', a tiled plot using color to display the value of the variable of interest at each location.
-When measurements are made at multiple time points it is reasonably common to display the data with small multiples~\citep{tufte:2001,IPCC}, such as in Figure~\ref{fig:nasa-facet}.
+When measurements are made at multiple time points it is reasonably common to display the data with small multiples~\citep{tufte:2001}, such as in Figure~\ref{fig:nasa-facet}.
 In this plot, a separate map is drawn for each month (columns) and year (rows), over a six year period of remotely sensed temperature data above Central America \citep{murrell:2010}.
@@
 
-\item[data-expo] The ASA 2009 data expo data \citep{murrell:2010} consists
+\item[EXPO] The ASA 2009 data expo data \citep{murrell:2010} consists
   of monthly observations of several atmospheric variables from the
@@
 Different variables can be swept into, and out of, the minor axes using mouse motion by varying $w$ and $h$ proportional to mouse movement.
-Interactive glyph-maps were used to explore the data-expo data using  GGobi \citep{swayne:2003}, and then later reproduced in publication quality resolution using {\tt ggplot2} \citep{me:ggplot2} in \citet{hobbs:2010}.
+Interactive glyph-maps were used to explore the EXPO data using  GGobi \citep{swayne:2003}, and then later reproduced in publication quality resolution using {\tt ggplot2} \citep{me:ggplot2} in \citet{hobbs:2010}.
 In GGobi, a manual tour is used to make the linear combination of spatial variables with other variables \citep{CB95}, and equation \ref{coords.eqn} makes this approach more explicit.
@@
 \noindent This is a non-standard conversion to polar coordinates, but it creates a timeline that starts at 12 o'clock and proceeds clockwise.
+\citet{eden:2010} wraps the time series, overplotting the multiple years, and this could be done also, by adjusting the scale of $x_{\amin}$.
 
@@
 
-  \caption{Glyph-maps of seasonal temperature patterns (averages for each month), using the data-expo data.
+  \caption{Glyph-maps of seasonal temperature patterns (averages for each month), using the EXPO data.
 Adding mid-range reference lines (top-left) and grid-cell reference boxes (top-right), makes it easier to see differences in glyph position, not just shape.
@@
 This makes it easier to compare shape (ignoring amplitude), and draws attention to locations with large relative variation.
-Figure~\ref{fig:scaling} compares global scaling and local scaling for the smoothed temperature data from the data-expo data.
-(Generalized additive model using smooth terms as described in \citet{wood:2006} were used.)  On the global scale, the El Ni\~no blip in temperature is just visible, along with variation in the average temperature at each location.
+Figure~\ref{fig:scaling} compares global scaling and local scaling for the smoothed temperature data from the EXPO data.
+(Generalized additive models using smooth terms as described in \citet{wood:2006} were used.)  On the global scale, the El Ni\~no blip in temperature is just visible, along with variation in the average temperature at each location.
 Local scaling emphasizes the individual shapes: the impact of El Ni\~no can be seen to be over a wider region, linear increases across the Andes, and decreases in the Gulf of Mexico.
@@
 
-Climate data is usually large: many locations in space, and many points in time.
-This makes climate data arduous to work with computationally, and to visualize.
+Climate data is usually large: many locations in space, and many points in time, which can make it arduous to work with computationally.
+The glyph-map calculations are as efficient as possible computationally, being linear in the number of spatial locations.
+It is possible that the calculations could also be paralellized for massive amounts of climate model data.
+Visualizing large data also hits a perceptual boundary.
 There is a limit to the resolution of human visual system \citep{Kr12} which will prevent perfect perception of large spatial domains or seasonality of many years of data.
-Exploring large data requires reducing it to a resolution that can be reasonably absorbed, and statistical models can be useful for this.
-Models can be used to decompose data into long-term trend, seasonal effects and error, which is vital for making the problem both computationally and visually tractable.
+Exploring large data requires reducing it to a resolution that can be reasonably absorbed visually.
 
-Figure~\ref{fig:gistemp-pred} shows the a glyph-map for the large GISTEMP dataset.
+Figure~\ref{fig:gistemp-pred} shows the glyph-map for the GISTEMP dataset, which is more realistically sized than the EXPO data.
 Glyphs at each location represent temperature from 1880--2011, starting with Jan 1880 at 12 o'clock and proceeding clockwise to 2011 at 11:59.
@@
 The raw data shows us missing values and gives some sense of variability, but gives no information about long-term trend.
-Trend can be examined by fitting a linear model to each location and displaying predicted values.
-Figure~\ref{fig:gistemp-pred} shows this for 1950--2010 (limited time period used due to missing data).
+Statistical models are useful for pre-processing climate data, to decompose data into long-term trend, seasonal effects and error, which is vital for making the problem both computationally and visually tractable.
+Trend is examined by fitting a linear model to each location and displaying predicted values in Figure~\ref{fig:gistemp-pred}.
+This plot shows long-term (de-seasonalized) temperature trend for 1950--2010 (the limited time period used due to missing data).
 Most locations have increasing temperature over this time period.
-Only a few locations show a decline.
-These locations tend to be different from their neighbors, which might suggest isolated data problems.
+Only a few locations show a decline, and these few locations tend to be different from their neighbors, and on the edge of the data collection region, indicating that there might be isolated data problems.
 The rate of increase differs across regions: steeper inclines in the north and western mountain regions and across the Great Lakes, and flatter inclines in the southeastern USA.
@@
 For multivariate data we may want to represent several variables in the glyphs.
-Figure~\ref{fig:cloud} shows scatterplots of the data-expo data as icons: temperature values are displayed horizontally and high-cloud values are displayed vertically, both locally scaled.
+Figure~\ref{fig:cloud} shows scatterplots of the EXPO data as icons: temperature values are displayed horizontally and high-cloud values are displayed vertically, both locally scaled.
 The bivariate relationship between the two variables is explored spatially, and it is clear that it is different across the region.
@@
 
-\section{Conclusions}
+\section{Discussion}
+
+There are two natural next directions for this work: perceptual experiments, and development of interactive graphics.
+Several careful studies comparing the perceptual elements of these displays, using different scale and glyph options, and in comparison with facetted heatmaps, would be recommended.
+For example, it appears that periodic trends are easier to perceive in star-glyphs, and long-term trends in line-glyphs.
+Is this always the case?
+Are their certain types of periodicity that are particularly easy to spot?
+These questions need to be answered with rigorous perceptual studies to help guide the use of these plots.
+
+The conceptualization and equations for computing the glyph-maps lend themselves to producing these displays interactively.
+Different variables can be swept into the display using mouse motion, resolution could be increased upon zooming in to a spatial neighborhood.
+This experimentation is ideally done using R, so that computation can be linked with displays.
+The hurdle is the lagging development of interactive graphics in R.
+Several new packages that support interactive graphics in R have become available in the past year \citep{qtbase, qtpaint, plumbr} which will enable these experiments.
 
@@
 
-The next direction would be to examine the perceptual elements of these displays, in more detail.
-For example, it appears that periodic trends are easier to perceive in star-glyphs, and long-term trends in line-glyphs.
-Is this always the case?
-Are their certain types of periodicity that are particularly easy to spot?
-These questions need to be answered with rigorous perceptual studies to help guide the use of these plots.
-
 \bibliography{references}

### 2012-03-05 revision/paper.tex
clarified different de-seasonalization, i hope

@@
 In this plot, a separate map is drawn for each month (columns) and year (rows), over a six year period of remotely sensed temperature data above Central America \citep{murrell:2010}.
-Color is used to display de-seasonalized temperature, with red mapped to high values and blue to low.
+Color is used to display de-seasonalized temperature (residuals from a linear model containing year and month terms), where red is mapped to high values and blue to low.
 The most noticeable feature is the strong red patch in the equatorial Pacific beginning mid-1997 and tapering out during 1998.
@@
 The dominant feature is the El Ni\~no warming in the southern equatorial region in the last half of 1997 and first half of 1998.
-Smaller features are only noticeable on closer inspection, or if pointed out: such as the relative warming on the mountain regions in south and north America in the later years.}
+Smaller features are only noticeable on closer inspection, or if pointed out: such as the relative warming on the mountain regions in south and north America which are red in all months in the later years.}
   \label{fig:nasa-facet}
@@
 Large variations in temperature can be seen in areas over land, while being fairly constant over water.
-(Right) Glyph-map legend, showing temporal pattern without spatial pattern: this allows you to see the scale of the individual axes.}
+(Right) Glyph-map legend, showing temporal pattern without spatial pattern, giving the scale of the glyphs.}
   \label{fig:nasa-glyph}
@@
 
-  \caption{Icon plots for 12 iconic time series shapes (linear increasing, decreasing, shifted, single peak, single dip, combined linear and nonlinear, seasonal trends with different scales, and a combined linear and seasonal trend) in Euclidean coordinates, time series icons (left) and polar coordinates, star plots (right).}
+  \caption{Icon plots for 12 iconic time series shapes (linear increasing, decreasing, shifted, single peak, single dip, combined linear and nonlinear, seasonal trends with different scales, and a combined linear and seasonal trend) in Euclidean coordinates, time series icons (left) and polar coordinates, star plots (right).
+White reference grids, lines and circles are added to help compare the shapes.}
   \label{fig:templates}
@@
 
-  \caption{Glyph-maps of seasonal temperature patterns (averages for each month), using the EXPO data.
+  \caption{Glyph-maps of seasonal temperature patterns (averages for each month over all years), using the EXPO data.
 Adding mid-range reference lines (top-left) and grid-cell reference boxes (top-right), makes it easier to see differences in glyph position, not just shape.
-(Bottom) Legend shows overall scales to aid interpretation.}
+(Bottom) Legend shows temperature values across all locations to aid interpretation.}
   \label{fig:ref-basic}
@@
 Figure~\ref{fig:scaling} compares global scaling and local scaling for the smoothed temperature data from the EXPO data.
-(Generalized additive models using smooth terms as described in \citet{wood:2006} were used.)  On the global scale, the El Ni\~no blip in temperature is just visible, along with variation in the average temperature at each location.
+(The model used to get the temperature values is a more complicated than that used previously to de-seasonalize.
+A generalized additive model using a month, and a smoothed day, term \citep{wood:2006} was fit to the observed data.
+The smoothed day was computed from the monthly resolution.
+Predicted values were computed for each day of the six year period.
+Season is removed but not overall average and long-term and trend.) On the global scale, the El Ni\~no blip in temperature is just visible, along with variation in the average temperature at each location.
 Local scaling emphasizes the individual shapes: the impact of El Ni\~no can be seen to be over a wider region, linear increases across the Andes, and decreases in the Gulf of Mexico.

### 2012-03-05 revision/paper.tex
changed figure 5 caption
Model desc tweaks

@@
 Figure~\ref{fig:scaling} compares global scaling and local scaling for the smoothed temperature data from the EXPO data.
-(The model used to get the temperature values is a more complicated than that used previously to de-seasonalize.
-A generalized additive model using a month, and a smoothed day, term \citep{wood:2006} was fit to the observed data.
-The smoothed day was computed from the monthly resolution.
-Predicted values were computed for each day of the six year period.
-Season is removed but not overall average and long-term and trend.) On the global scale, the El Ni\~no blip in temperature is just visible, along with variation in the average temperature at each location.
+(The model is more complicated than the model used previously to de-seasonalize the data.
+A generalized additive model \citep{wood:2006} with qualitative month and smoothed day terms was fit to the observed data and predicted values were computed for each day of the six year period.
+Seasonal effects are removed but not overall average or long-term trend.) On the global scale, the El Ni\~no blip in temperature is just visible, along with variation in the average temperature at each location.
 Local scaling emphasizes the individual shapes: the impact of El Ni\~no can be seen to be over a wider region, linear increases across the Andes, and decreases in the Gulf of Mexico.
@@
 
-  \caption{Smoothed de-seasonalized daily temperature glyph-map. (Top-left)
+  \caption{Glyph-map using predicted daily temperature values from a generalized additive model fit, where seasonality has been extracted. (Top-left)
 globally scaled, (Top-right) locally scaled. (Bottom) Legend showing ranges of individual glyphs.}

### 2012-03-05 revision/paper.tex
figure 6 caption

--- 2012-03-05 revision/paper.tex
@@
 This is illustrated in Figure~\ref{fig:scaling-col}.
+In the encoding, emphasis is placed on the locations with larger range, stronger effects: the line intensity is stronger (left plot), and the background is lightened to provide greater contrast (right plot) drawing the eye to these regions.
 
@@
 Here each location has been scaled to range $[0, 1]$ and color mapped to the range of the original predictions.
-(Left) Range mapped to intensity of line: pale locations have smaller range than darker locations.
-(Right) Range mapped to fill color of grid box: blue boxes have smaller ranges.
-Shapes at all locations are visible, but attention is drawn to locations with larger ranges.}
+(Left) Range mapped to intensity (grey scale) of the line: pale locations have smaller range, smaller effect, than darker locations, stronger effect.
+(Right) Range mapped to fill color of grid box: blue boxes have smaller ranges, and light boxes have larger effects.
+Shapes at all locations are visible, but attention is drawn to locations with lighter backgrounds, larger ranges, because of the higher contrast.}
  \label{fig:scaling-col}

### 2012-03-05 revision/paper.tex
added . fig 7

@@
 One more important point: if a model summary is displayed in the glyph-map, as discussed in the following section, it may be important to force the scale to that of the original data.
-The range of predicted values is typically smaller than the raw data, so angles of slopes is increased.
+The range of predicted values is typically smaller than the raw data, so angles of slopes is increased, giving an inaccurate sense of a stronger effect.
 
@@
 Gaps in the full circle indicate missing values, and thickness of the circle indicates more variability in the measurements.
-(Bottom) Legend showing ranges of individual glyphs}
+(Bottom) Legend showing temperature and temporal ranges of individual glyphs.}
   \label{fig:gistemp-raw}

### 2012-03-05 revision/paper.tex
fixed a ref, and .

@@
 
-  \caption{(Top) Glyph-map of predicted temperature anomaly data (GISTEMP),
-    from 1950-2010. Increasing trends can be seen over most of the
-    USA, with varying degrees of increase. Some isolated locations
-    exhibit steep declines, and inclines, perhaps indicative of some
-    isolated data problems. Note that, the reference lines are
-    straight, but they look curved because of the Z\"ollner optical
-    illusion. (Bottom) Legend showing ranges of individual glyphs}
+  \caption{(Top) Glyph-map of predicted temperature anomaly data
+    (GISTEMP), on reduced range temperature 1950-2010. Increasing
+    trends can be seen over most of the USA, with varying degrees of
+    increase. Some isolated locations exhibit steep declines, and
+    inclines, perhaps indicative of some isolated data problems. Note
+    that, the reference lines are straight, but they look curved
+    because of the Z\"ollner optical illusion. (Bottom) Legend showing
+    temperature and temporal ranges of individual glyphs.}
   \label{fig:gistemp-pred}
@@
 Glyph-maps work best when displayed in the coordinate system in which the grid is rectangular and regular.
-For example, the GISTEMP grid is 2$^o$ square and is displayed treating longitude and latitude as cartesian coordinates in Figure \ref{fig:gistemp-pred}.
+For example, the GISTEMP grid is 2$^o$ square and is displayed treating longitude and latitude as Cartesian coordinates in Figure \ref{fig:gistemp-pred}.
 For regularly gridded data, this results in icons that are rectangles of equal size.

### 2012-03-05 revision/paper.tex
captions

@@
 
-% resolution issues with large data - reference - also long time series - large spatial domain.
+% re-work equations, to incorporate scale, and interaction, use spatial location letter and time/var letters
 
-% order of the algorithm calculations
-
-% emphasize way of thinking, which leads into interactive setting
-
-% refer to the wrapping of times series in eden
+% overlay legends on the plots?
 
@@
   \caption{Glyph-maps of seasonal temperature patterns (averages for each month over all years), using the EXPO data.
-Adding mid-range reference lines (top-left) and grid-cell reference boxes (top-right), makes it easier to see differences in glyph position, not just shape.
+Adding mid-range reference lines (top left) and grid-cell reference boxes (top right), makes it easier to see differences in glyph position, not just shape.
 (Bottom) Legend shows temperature values across all locations to aid interpretation.}
@@
 
-  \caption{Glyph-map using predicted daily temperature values from a generalized additive model fit, where seasonality has been extracted. (Top-left)
-globally scaled, (Top-right) locally scaled. (Bottom) Legend showing ranges of individual glyphs.}
+  \caption{Glyph-map using predicted daily temperature values from a generalized additive model fit, where seasonality has been extracted: globally scaled (top left), locally scaled (top right).
+(Bottom) Legend showing ranges of individual glyphs.}
   \label{fig:scaling}
@@
   \centering
-  \includegraphics[width=0.95\linewidth]{usa-lin-overlap}%
+  \includegraphics[width=1\linewidth]{usa-lin-overlap}%
+
+\includegraphics[width=0.33\linewidth]{usa-lin-legend}
   %\includegraphics[width=0.25\linewidth]{ghcn-mountains}%
-  \caption{Glyph-maps of the linear trend in ground station temperature values, 1950--2010, for USHCN stations, illustrating three issues for irregularly gridded data: overlapping icons, no natural choice for icon size, and the disordered appearance making comparisons more difficult.
-}% (left) and stations in the mountain states (right).}
+  \caption{(Top) Glyph-maps of the linear trend in ground station temperature values, 1950--2010, for USHCN stations, illustrating three issues for irregularly gridded data: overlapping icons, no natural choice for icon size, and the disordered appearance making comparisons more difficult.
+(Bottom) Legend showing
+    temperature and temporal ranges of individual glyphs. }% (left) and stations in the mountain states (right).} 
   \label{fig:irregular}
@@
   \centering
-  \includegraphics[width=0.95\linewidth]{usa-season-collapsed}%
+  \includegraphics[width=0.5\linewidth]{usa-season-collapsed}%
+  \includegraphics[width=0.5\linewidth]{usa-lin-collapse}%
 
-  \includegraphics[width=0.95\linewidth]{usa-lin-collapse}%
-  \caption{Glyph-maps of seasonal patterns (top) and linear trends (bottom) of USHCN stations, 1950--2010.
+\includegraphics[width=0.20\linewidth]{usa-season-legend}\hspace{2in}%
+\includegraphics[width=0.20\linewidth]{usa-lin-legend}
+  \caption{Glyph-maps of seasonal patterns (top left) and linear trends (top right) of USHCN stations, 1950--2010.
 Nearby stations whose icons would overlap have been collapsed.
@@
 Heavy lines in the linear trend plot correspond to average trends over more than one location.
-Legend in Figure~\ref{fig:irregular-legend}}
+(Bottom) Legends showing scale of glyphs.} % Legend in Figure~\ref{fig:irregular-legend}}
   \label{fig:irregular-collapsed}
@@
   \includegraphics[width=0.5\linewidth]{usa-lin-grid}%
-  \caption{Glyph-maps of seasonal patterns (left) and linear trends (right) of USHCN stations, 1950--2010.
+
+  \includegraphics[width=0.20\linewidth]{usa-season-legend}\hspace{2in}%
+  \includegraphics[width=0.20\linewidth]{usa-lin-legend}
+  \caption{Glyph-maps of seasonal patterns (top left) and linear trends (top right) of USHCN stations, 1950--2010.
 Station locations have been rounded to the nearest degree Each location has a glyph in the seasonal plot and color of the tiles corresponds to the average temperature of the locations.
 Heavy lines in the linear trend plot correspond to average trends over more than one location.
-Legend in Figure~\ref{fig:irregular-legend}}
+(Bottom) Legends showing scale of glyphs.} %Legend in Figure~\ref{fig:irregular-legend}}
   \label{fig:irregular-grid}
@@
 
-\begin{figure}[htbp]
-  \centering
-  \includegraphics[width=0.33\linewidth]{usa-season-legend}%
-  \hspace{0.18 \linewidth}%
-  \includegraphics[width=0.33\linewidth]{usa-lin-legend}
+%\begin{figure}[htbp]
+%  \centering
+%  \includegraphics[width=0.33\linewidth]{usa-season-legend}%
+%  \hspace{0.18 \linewidth}%
+%  \includegraphics[width=0.33\linewidth]{usa-lin-legend}
 
-  \caption{Legend for Figures~\ref{fig:irregular-collapsed} aad \ref{fig:irregular-grid}.}
-  \label{fig:irregular-legend}
-\end{figure}
+%  \caption{Legend for Figures~\ref{fig:irregular-collapsed} aad \ref{fig:irregular-grid}.}
+%  \label{fig:irregular-legend}
+%\end{figure}
 
@@
 
-  \includegraphics[width=0.33\linewidth]{nasa-scat-legend}%
+  \includegraphics[width=0.20\linewidth]{nasa-scat-legend}%
   \hspace{0.17\linewidth}%
-  \includegraphics[width=0.33\linewidth]{nasa-loess-legend}
+  \includegraphics[width=0.20\linewidth]{nasa-loess-legend}
 
-  \caption{Glyph-map showing (top-left) scatterplot of temperature vs high-cloud and (top-right) smoothed (loess) curve fit to each location.
+  \caption{Glyph-map showing (top left) scatterplot of temperature vs high-cloud and (top right) smoothed (loess) curve fit to each location.
 The relationship between these two variables varies considerably over the spatial domain.

### 2012-03-05 revision/paper.tex
minor changes from proofreading

@@
 Types of glyph plots include the star plots \citep{mayr:1877, FD01, kleiner:1981}, the semi-graphic displays \citep{anderson:1960}, and the infamous Chernoff faces \citep{chernoff:1973}.
-An glyph is produced by mapping each variable to some graphical feature, such as the length of a line, curvature of the mouth, or size of the face.
+A glyph is produced by mapping each variable to some graphical feature, such as the length of a line, curvature of the mouth, or size of the face.
 
@@
 The dominant feature is the El Ni\~no warming in the southern equatorial region in the last half of 1997 and first half of 1998.
-Smaller features are only noticeable on closer inspection, or if pointed out: such as the relative warming on the mountain regions in south and north America which are red in all months in the later years.}
+Smaller features are only noticeable on closer inspection, or if pointed out: such as the relative warming on the mountain regions in South and North America which are red in all months in the later years.}
   \label{fig:nasa-facet}
@@
 
-In general, multivariate glyph displays lack a natural ordering of the glyphs because observations have no ordering.
-Similarly, there are multiple ways that variables can be mapped to glyph property.
-It can result in a combinatorial explosion of possible mappings, each producing a display that may emphasize some entirely different features of the data than another mapping.
+In general, multivariate glyph displays lack a natural ordering of the glyphs because the observations have no ordering.
+Similarly, there are multiple ways that variables can be mapped to glyph properties.
+It can result in a combinatorial explosion of possible mappings, each producing a display that may emphasize some entirely different features of the data compared to another mapping.
 This may have prevented the widespread adoption of glyph displays, despite some proposed ordering solutions \citep{kleiner:1981,hurley:2010}.
@@
 
-  \caption{Icon plots for 12 iconic time series shapes (linear increasing, decreasing, shifted, single peak, single dip, combined linear and nonlinear, seasonal trends with different scales, and a combined linear and seasonal trend) in Euclidean coordinates, time series icons (left) and polar coordinates, star plots (right).
+  \caption{Icon plots for 12 iconic time series shapes (linear increasing, decreasing, shifted, single peak, single dip, combined linear and nonlinear, seasonal trends with different scales, and a combined linear and seasonal trend) in Cartesian coordinates, time series icons (left) and polar coordinates, star plots (right).
 White reference grids, lines and circles are added to help compare the shapes.}
@@
 
-This way of thinking of the glyph-map is convenenient for several reasons.
+This way of thinking of the glyph-map is convenient for several reasons.
 The first is that it moves the plots beyond the small multiples, many little plots, approach.
 That provides flexibility to cover large spatial domains using potentially tiny glyphs.
-A second, and major, reason is that with this approach they can then be readily made to be interactive.
-Different variables can be swept into, and out of, the minor axes using mouse motion by varying $w$ and $h$ proportional to mouse movement.
+A second, and major, reason is that with this approach they can then readily be made interactive.
+Different variables can be swept into, and out of, the minor axes using mouse motion, the scale can be adjusted by varying $w$ and $h$ proportional to mouse movement.
 Interactive glyph-maps were used to explore the EXPO data using  GGobi \citep{swayne:2003}, and then later reproduced in publication quality resolution using {\tt ggplot2} \citep{me:ggplot2} in \citet{hobbs:2010}.
@@
 All plots facilitate some comparisons and impede others.
-For a plot to be useful for a particular data analytic task, primary comparisons need to be made the simplest by the display.
+For a plot to be useful for a particular data analytic task, the display needs to make the primary comparisons the simplest to make.
 For climate data, changes in slope, or trend, average value, and variance over time, are the primary tasks.
@@
 
-Different features of the data can be emphasized by varying the scale used with each cell.
+Different features of the data can be emphasized by varying the scale used within each cell.
 By default, a global scale is used, so the same position within each cell corresponds to the same value in all locations.
-This facilitates the comparison of absolute values, and draws attention to plots with large overall variation.
+This facilitates the comparison of absolute values, and draws attention to locations with large overall variation.
 
@@
 Here each location has been scaled to range $[0, 1]$ and color mapped to the range of the original predictions.
-(Left) Range mapped to intensity (grey scale) of the line: pale locations have smaller range, smaller effect, than darker locations, stronger effect.
-(Right) Range mapped to fill color of grid box: blue boxes have smaller ranges, and light boxes have larger effects.
+(Left) Range mapped to intensity (grey scale) of the line: pale locations have smaller effect, smaller range, than darker locations, stronger effect.
+(Right) Range mapped to fill color of grid box: blue boxes have smaller effects, and light boxes have larger effects.
 Shapes at all locations are visible, but attention is drawn to locations with lighter backgrounds, larger ranges, because of the higher contrast.}
@@
 One more important point: if a model summary is displayed in the glyph-map, as discussed in the following section, it may be important to force the scale to that of the original data.
-The range of predicted values is typically smaller than the raw data, so angles of slopes is increased, giving an inaccurate sense of a stronger effect.
+The range of predicted values is typically smaller than the raw data, so angles of slopes are increased, giving an inaccurate sense of a stronger effect.
 
@@
 Trend is examined by fitting a linear model to each location and displaying predicted values in Figure~\ref{fig:gistemp-pred}.
-This plot shows long-term (de-seasonalized) temperature trend for 1950--2010 (the limited time period used due to missing data).
+This plot shows long-term (de-seasonalized) temperature trends for 1950--2010 (the limited time period used due to missing data).
 Most locations have increasing temperature over this time period.
@@
   \caption{(Top) Glyph-map of predicted temperature anomaly data
-    (GISTEMP), on reduced range temperature 1950-2010. Increasing
+    (GISTEMP), for the reduced period 1950-2010. Increasing
     trends can be seen over most of the USA, with varying degrees of
@@
 When spatial locations are not on a regular grid, some difficulties arise.
-In this situation it is not clear what the scale of icons should be, there is no regular spacing.
+In this situation it is not clear what the scale of icons should be, as there is no regular spacing.
 Some locations may be close to each other which would generate icons that overlap.
@@
 One solution to the overlapping problem is to zoom in on smaller regions.
-There is an interplay between icon size (in geographical units), plotting area (in geographical units) and plot size (in display units).
+There is an interplay between icon size (in geographical units), area plotted (in geographical units) and plot size (in display units).
 Icons need to be big enough to be readable on the display, but small enough to avoid overlapping.
-Readability can be maintained and overlapping reduced by decreasing the icon size and simultaneously increasing the display size or decreasing the plotting area.
+Readability can be maintained and overlapping reduced by decreasing the icon size and simultaneously increasing the display size or decreasing the area plotted.
 % (as done in the right panel in Figure \ref{fig:irregular}).  
@@
 In Figure \ref{fig:irregular-collapsed} nearby locations are collapsed to a single icon.
-In the top panel each station still has an individual glyph but locations close to each other share a plotting area.
+In the left panel each station still has an individual glyph but locations close to each other share a plotting area.
 Not all glyphs are suited to this kind of space sharing, another option is to perform a summary of the locations that are collapsed and display that instead of the raw data.
-The bottom panel shows an example of this where the average trend for the collapsed locations is displayed.
+The right panel shows an example of this where the average trend for the collapsed locations is displayed.
 It is helpful to differentiate glyphs that involve summaries of more than one location, here a heavier line indicates at least two locations have been averaged.
@@
 Climate change is focused on changes over time, so glyph-maps provide a way to study these changes directly.
-Glyph-maps enable different resolutions of the data to be examined: the raw data to discover data quality issues, early, global trend, seasonality, residuals, and multivariate dependencies.
+Glyph-maps enable different resolutions of the data to be examined: the raw data to discover data quality issues, global trend, seasonality, residuals, and multivariate dependencies.
 When the data is spatially gridded the glyphs are organized in a manner that makes reasonable comparisons.

### 2012-03-05 revision/paper.tex
remove reference to mountain states plot

@@
 It illustrates three challenges in working with irregular locations: there is no natural choice for icon size, icons can potentially overlap, and comparisons rely more heavily on the reference guides.
-The icon sizes were chosen manually for these two plots, although the icons are displayed at about the same size, they correspond to 1$^o$ and 0.5$^o$ square respectively in the geographical coordinates.
+The icon size was chosen manually for this plot, corresponding to 1$^o$ square in the geographical coordinates.
 \begin{figure}[htbp]

### 2012-03-06 revision/code/5-irregular.r
more files for legends in irregular section

@@ -46,7 +46,7 @@ ggsave("../images/usa-lin-overlap.png", width = 8, height = 4.5)
 ggplot(lin.pred.gly, aes(year, pred)) + 
   geom_line(aes(group = gid), alpha = 1/20) + 
   xlab("Year") + 
-  ylab("Predicted January temperature (F)")
+  ylab("Linear Temp Anomaly (F)")
 ggsave("../images/usa-lin-legend.pdf", width = 4, height = 4)
 
 # # Mountain states: Colorado, Wyoming, Idaho, Montana, Nevada, Utah
@@ -112,6 +112,16 @@ ggplot(lin.pred.sum, aes(gx, gy, group = gid)) +
   opts(legend.position = "bottom")
 ggsave("../images/usa-lin-collapse.png", width = 8, height = 5)
 
+ggplot(lin.pred.sum, aes(year, pred)) + 
+  geom_line(aes(group = gid, colour = n > 1), alpha = 1/5) + 
+  xlab("Year") + 
+  scale_colour_manual("", values = c("TRUE" = "black", "FALSE" = "grey40"), 
+  breaks = c(F, T), labels = c("Single", "Multiple")) + 
+  ylab("Linear Temp Anomaly (F)") +
+  opts(legend.position = "none")
+ggsave("../images/usa-lin-collapse-legend.pdf", width = 4, height = 4)
+
+
 # Gridding -------------------------------------------------------------------
 lin.pred.grid <- grid.all(lin.pred.post1950, "lon", "lat", w, h)
 
@@ -131,6 +141,15 @@ ggplot(lin.pred.grid.gly,aes(gx, gy, group = gid)) +
     breaks = c(F, T), labels = c("Single", "Multiple")) +
   opts(legend.position = "bottom")
 ggsave("../images/usa-lin-grid.png", width = 8, height = 5)
+
+ggplot(lin.pred.grid.gly, aes(year, pred)) + 
+  geom_line(aes(group = gid, colour = n > 1), alpha = 1/5) + 
+  xlab("Year") + 
+  scale_colour_manual("", values = c("TRUE" = "black", "FALSE" = "grey40"), 
+  breaks = c(F, T), labels = c("Single", "Multiple")) + 
+  ylab("Linear Temp Anomaly (F)") +
+  opts(legend.position = "none")
+ggsave("../images/usa-lin-grid-legend.pdf", width = 4, height = 4)
 
[…]

### 2012-03-06 revision/paper.tex
more files for legends in irregular section
explaining legends

@@
 Average temperatures were calculated for each month, and these were passed to the glyph calculations to produce the glyph-maps.
+For star glyphs the reference frames are also useful and the equivalent of the reference line is a circle (see Figure~\ref{fig:templates}).
 
@@
 
-For star glyphs the reference frames are also useful and the equivalent of the reference line is a circle (see Figure~\ref{fig:templates}).
+To observe the scale of the measurements represented by the glyphs we recommend a full legend, alongside the glyph-map, that contains the glyphs overplotted ignoring the spatial location, with complete axes.
+See, for example, Figure \ref{fig:nasa-glyph} where from the legend it can be seen that months range from 0 to more than 60 (actually 72) and de-seasonalized temperature ranges from -8 to 8$^0$K.
+In Figure \ref{fig:ref-basic}, months range from 1 to 12, and temperature from 270 to 315$^o$K.
 

### 2012-03-06 revision/paper.tex
tweaked abstract

@@
 The multivariate spatio-temporal nature of climate data makes it difficult to draw all of the aspects simultaneously.
-This paper describes the conceptualization and construction of a type of display, a glyph-map, that has been used infrequently for climate data.
+This paper describes the conceptualization and construction of a type of display, a glyph-map, that can show these multiple aspects.
 Glyph-maps are a specialization of multivariate glyph plots.
-Each spatial location is displayed with one glyph that represents the multiple measurements, recorded over time, at that location.
-Glyph-maps allow the discovery of both local and global structure, with a particular focus on temporal relationships.
+Each spatial location is displayed with one glyph that represents the multiple measurements, often recorded over time, at that location.
+Glyph-maps allow the discovery of both local and global structure, with a particular focus on temporal relationships, important for studying climate change.
 They provide alternatives to colored, facetted maps, or statistical summaries such as principal components.
-With the new conceptualization, developing interactive versions of glyph-maps will be easier, and it will be simpler to explore the perceptual effects of different scales.
-The methods are developed for rectangularly gridded data, but with some clever processing, which is explained, it is possible get good glyph-maps of irregularly gridded data.
-Guides, and reference marks, for different types of glyphs used in the glyph-maps are also discussed.
+Glyph-maps have been used sporadically for spatio-temporal data, and with the ideas and software described in this paper it will be easier to produce them.
+The conceptualization described here also will enable developing interactive versions of glyph-maps, and make it simpler to explore the perceptual effects of different scales.
+The methods are developed for rectangular gridded data, but with some clever processing, which is explained, it is possible get good glyph-maps of irregularly gridded data.
+Guides and reference marks, for different types of glyphs used in the glyph-maps, are also discussed.
 
@@
   \caption{(Top) Glyph-maps of the linear trend in ground station temperature values, 1950--2010, for USHCN stations, illustrating three issues for irregularly gridded data: overlapping icons, no natural choice for icon size, and the disordered appearance making comparisons more difficult.
+Local centering, subtracting the location mean from the linear fit, is used as the variation in average temperature is of much greater magnitude than the variation in linear trend.
 (Bottom) Legend showing
-    temperature and temporal ranges of individual glyphs. }% (left) and stations in the mountain states (right).} 
+    temperature and temporal ranges of individual glyphs.
+The local scaling is apparent here with all lines going through zero at the mean year.
+}% (left) and stations in the mountain states (right).}
   \label{fig:irregular}
@@
 \includegraphics[width=0.20\linewidth]{usa-season-legend}\hspace{2in}%
-\includegraphics[width=0.20\linewidth]{usa-lin-legend}
+\includegraphics[width=0.20\linewidth]{usa-lin-collapse-legend}
   \caption{Glyph-maps of seasonal patterns (top left) and linear trends (top right) of USHCN stations, 1950--2010.
@@
   \includegraphics[width=0.20\linewidth]{usa-season-legend}\hspace{2in}%
-  \includegraphics[width=0.20\linewidth]{usa-lin-legend}
+  \includegraphics[width=0.20\linewidth]{usa-lin-grid-legend}
   \caption{Glyph-maps of seasonal patterns (top left) and linear trends (top right) of USHCN stations, 1950--2010.

### 2012-03-06 revision/paper.tex
tweaked intro

@@
 Climate data is composed of multiple variables, such as temperature, ozone, precipitation, and winds, with a spatial and temporal context.
-The classic display for data of this type is a ``heatmap'', a tiled plot using color to display the value of the variable of interest at each location.
+The classic display for data of this type is a ``heatmap'', a tiled plot using color to display the value of the variable of interest at each location (see many examples in \citet{IPCC}, like Figure 2.11).
 When measurements are made at multiple time points it is reasonably common to display the data with small multiples~\citep{tufte:2001}, such as in Figure~\ref{fig:nasa-facet}.
@@
 The most noticeable feature is the strong red patch in the equatorial Pacific beginning mid-1997 and tapering out during 1998.
-This is the El Ni\~no event, a major temperature anomaly (see for example, \citet{trenberth} or \url{http://www.esrl.noaa.gov/psd/enso/enso.description.html}.) More work is required to see more localized patterns, like the cooler land temperatures seen in early years.
+This is the El Ni\~no event, a major temperature anomaly (see for example, \citet{trenberth} or \url{http://www.esrl.noaa.gov/psd/enso/enso.description.html}.) More work is required to see more localized patterns, like cooler land temperatures seen in early years.
 
@@
 %% Reproduce all of the icon displays using the fixed glyph code
-
 The glyph-map uses a small glyph, or icon, to represent multiple variables, or time values, at each location.
@@
 A glyph is produced by mapping each variable to some graphical feature, such as the length of a line, curvature of the mouth, or size of the face.
-
-Figure~\ref{fig:nasa-glyph} shows the glyph-map equivalent of Figure~\ref{fig:nasa-facet}: each location is represented by a small time series glyph of de-seasonalized temperature.
-The darker grey in the background represents the land mass of the study region.
-Because the data is de-seasonalized, the primary visible structure is still El Ni\~no -- it is both a temporal anomaly and a spatial phenomenon.
-It can be recognized by the bump in the middle of the series of the equatorial Pacific region.
-Other temporal features are visible from the many time series.
-Temperatures are generally more varied over land, than sea.
-A few locations have dramatic jumps in temperature in the later years.%, all in high elevation areas, along the Andes and the Sierra Madre Occidental in Mexico.
-
-The glyph-map can also be thought of as a small multiple display.
-The primary comparison is different from the facetted heatmap.
-Cognitively, elements placed close to each other are more easily and accurately compared \citep{cleveland:1984}.
-The facetted heatmap, places greater emphasis on comparison between the spatial locations, because these are positioned together, and secondary emphasis on time because these are separated by facets.
-In the glyph-map, primary emphasis is placed on the temporal comparison, and secondarily on spatial comparison, because facetting is on spatial locations.
-
-% The most seasonality occurs in north American.
-Some seasonal effects can also be seen in the Caribbean and in south American locations.
-Around the equator the temperatures are relatively warm for all seasons.
-Instead of radially displaying the temperature a glyph can be a regular time series, time displayed horizontally and temperature vertically (right plot).
-The structure perceived is different with this change in glyph type: locations at the equator have flat series, on land more varied temperatures.
 
@@
 
-In general, multivariate glyph displays lack a natural ordering of the glyphs because the observations have no ordering.
-Similarly, there are multiple ways that variables can be mapped to glyph properties.
-It can result in a combinatorial explosion of possible mappings, each producing a display that may emphasize some entirely different features of the data compared to another mapping.
+Figure~\ref{fig:nasa-glyph} shows the glyph-map equivalent of Figure~\ref{fig:nasa-facet}: each location is represented by a small time series glyph of de-seasonalized temperature.
+The darker grey in the background represents the land mass of the study region.
+Because the data is de-seasonalized, the primary visible structure is still El Ni\~no, recognized by the bump in the middle of the series of the equatorial Pacific region.
+It is both a temporal anomaly and a spatial phenomenon, so this is not surprising.
+Other temporal features are visible from the many time series.
+Temperatures are generally more varied over land, than sea.
+A few locations, near the coast of South America and the southern part of North America, have dramatic jumps in temperature in the later years.%, all in high elevation areas, along the Andes and the Sierra Madre Occidental in Mexico.
+
+The glyph-map can also be thought of as a small multiple display~\citep{tufte:2001}.
+The primary comparison is different from the facetted heatmap.
+Cognitively, elements placed close to each other are more easily and accurately compared \citep{cleveland:1984}.
+The facetted heatmap, places greater emphasis on comparison between the spatial locations, because these are positioned together, and secondary emphasis on time because these are separated by facets.
+In the glyph-map, primary emphasis is placed on the temporal comparison, and secondarily on spatial comparison, because facetting is on spatial locations.
+
+% The most seasonality occurs in north American.
+Some seasonal effects can also be seen in the Caribbean and in south American locations.
+Around the equator the temperatures are relatively warm for all seasons.
+Instead of radially displaying the temperature a glyph can be a regular time series, time displayed horizontally and temperature vertically (right plot).
+The structure perceived is different with this change in glyph type: locations at the equator have flat series, on land more varied temperatures.
+
+In general, multivariate glyph displays lack a natural ordering because the observations have no ordering.
+Similarly, there are multiple ways that variables can be mapped to glyph properties, which can result in a combinatorial explosion of possible mappings, each producing a display that may emphasize some entirely different features of the data compared to another mapping.
 This may have prevented the widespread adoption of glyph displays, despite some proposed ordering solutions \citep{kleiner:1981,hurley:2010}.
-Using glyphs with space-time data eliminates this issue: glyphs are placed according to the location of the measurement, and variables are ordered by the time they were collected.
-Additionally, climate data usually has high correlations between nearby locations and times.
-This imposes a degree of smoothness that gives the plot an appearance of a textured  landscape, making it easier to digest patterns from one glyph to another.
+Using glyphs with space-time data eliminates the ordering issue: glyphs are placed according to the location of the measurement, and variables are ordered by the time they were collected.
+Additionally, climate data usually has high correlations between nearby locations and times, imposing a degree of smoothness that gives the plot an appearance of a textured  landscape, which makes it easier to digest patterns from one glyph to another.
 Others \citep{pickett:1988} have used glyphs on maps to display multivariate spatial data.
@@
 Glyph-maps for spatiotemporal data have been used in several publications \citep{carr:1992,eden:2010,hobbs:2010}.
-Glyph-maps using arrow glyphs have commonly been used for displaying ocean currents, wind direction and strength, as can be seen in the Figure 8.1 of \citet{IPCC} and a rudimentary time series glyph display can be seen in Figure 9.12.
+Glyph-maps using arrow glyphs have commonly been used for displaying ocean currents, wind direction and strength, as can be seen in the Figure 8.1  and a rudimentary time series glyph display can be seen in Figure 9.12 of \citet{IPCC}.
 
@@
 Two types of glyph -- lines and stars -- are especially useful for temporal displays.
-Figure~\ref{fig:templates} displays 12 iconic time series shapes with line- and star-glyphs.
+Figure~\ref{fig:templates} plots 12 iconic time series shapes (linear increasing, decreasing, shifted, single peak, single dip, combined linear and nonlinear, seasonal trends with different scales, and a combined linear and seasonal trend) as line- and star-glyphs.
 The data underlying each glyph is measured at 36 time points.
@@
 Section~\ref{sec:construction} describes the algorithm used to create glyphs-maps of these two types of glyphs.
-Section~\ref{sec:perception} discusses their perceptual properties, including the importance of a visual reference grids, and of careful consideration of scale.
+Section~\ref{sec:perception} discusses their perceptual properties, including the importance of visual reference grids, and of careful consideration of scale.
 Large data and the interplay with models and data are discussed in Section~\ref{sec:large-data}.
@@
 
-\item[GISTEMP] surface temperature data provided on $2^{\circ}$ x
+\item[GISTEMP] Surface temperature data provided on $2^{\circ}$ x
   $2^{\circ}$ grid over the entire globe, measured monthly
-  \citep{GISTEMP} from 1880-2011. Ground station data was de-seasonalized,
-  differenced from from the 1951-1980 temperature averages, and
-  spatially averaged to obtain gridded temperature anomalies. For the
-  purposes of this paper, we extracted the locations corresponding to
-  the continental USA.
+  \citep{GISTEMP} from 1880-2011. To produce this data irregularly
+  gridded ground station data was de-seasonalized, differenced from
+  the 1951-1980 temperature averages, and spatially averaged to obtain
+  gridded temperature anomalies. For the purposes of this paper, we
+  extracted the locations corresponding only to the continental USA.
 
-\item[USHCN] (Version 2) ground station network of historical
+\item[USHCN] (Version 2) Ground station network of historical
   temperatures \citep{USHCN}. Temperatures from 1219 stations on the
@@
 \end{quote}
-Supplementary material contains the datasets and code to reproduce the plots in this paper.
-R 2.13.1 \citep{R} with the packages {\tt ggplot2} \citep{me:ggplot2} and {\tt plyr} \citep{me:plyr} was used.
+Supplementary material contains the datasets and code to reproduce the plots in this paper.  R \citep{R} with the packages {\tt ggplot2} \citep{me:ggplot2} and {\tt plyr} \citep{me:plyr} was used.
 

### 2012-03-06 revision/paper.tex
fixed some glyph(-)maps

@@
 The glyph-map uses a small glyph, or icon, to represent multiple variables, or time values, at each location.
-Glyph maps derive from glyph plots, which were initially developed to display multivariate data.
+Glyph-maps derive from glyph plots, which were initially developed to display multivariate data.
 One glyph is drawn for each observation in the data set.
@@
 
-Section~\ref{sec:construction} describes the algorithm used to create glyphs-maps of these two types of glyphs.
+Section~\ref{sec:construction} describes the algorithm used to create glyph-maps of these two types of glyphs.
 Section~\ref{sec:perception} discusses their perceptual properties, including the importance of visual reference grids, and of careful consideration of scale.
@@
 The sensitivity of analyses to the scale and configuration of aggregation is described by the  modifiable areal unit problem (\cite{Openshaw:1984kx, Fotheringham:1991uq}).
-In this type of glyph map the scale of aggregation is determined, rather arbitrarily, by the icon size.
+In this type of glyph-map the scale of aggregation is determined, rather arbitrarily, by the icon size.
 This dependence of visual inferences on the icon size is undesirable.

### 2012-03-06 revision/paper.tex
done with construction

@@
 
-\section{Construction}~\label{sec:construction}
+\section{Structural components}~\label{sec:construction}
 
 %Glyph-maps can be generated with existing graphics software after performing a simple pre-processing step, making them easily accessible to a wide audience. 
-To create a glyph-map it is useful to recognize that the glyph-map is a linear mix of two structural components of the data: spatial location and data values.
+It is useful to recognize that the glyph-map is a linear mix of two structural components of the data: spatial location and data values.
 The spatial location is the major positioning component, while the data values are minor adjustments to those positions.
-For spatiotemporal data, the major components are latitude ($y_{\amaj}$) and longitude ($x_{\amaj}$), and the minor components are time ($x_{\amin}$) and some measurement ($y_{\amin}$), for example, temperature, or predicted temperature.
-Assuming the minor components are rescaled to $[-1, 1]$, the final coordinates $(x,y)$ on the chart are the linear combination, given by:
+For spatiotemporal data, the major components are longitude ($s_x$) and latitude ($s_y$), and the minor components are time ($t$) and some variable ($z$), for example, temperature, or predicted temperature.
+Assuming the minor components ($t, z$) are scaled into the range $[-1, 1]$, the final coordinates $(x,y)$ of the values in the plot are the linear combination, given by:
 
@@
   \begin{array}{lll}
-  x &=& x_{\amaj} + \frac{w}{2} \cdot x_{\amin}\\
-  y &=& y_{\amaj} + \frac{h}{2} \cdot y_{\amin}, 
+  x &=& s_x + \alpha_x \cdot w \cdot t\\
+  y &=& s_y + \alpha_y \cdot h \cdot z, 
   \end{array}
@@
 
-\noindent where $w$ and $h$ are scale parameters, setting the width and height of the glyph, respectively.
+\noindent where $\alpha_x, \alpha_y$ are scale parameters setting the fraction the width $w$ and height $h$ of the glyph, respectively.
 For gridded data, $w$ and $h$ will be the minimum difference between neighboring spatial locations to make use of the resolution of the major components.
+For irregularly gridded data, $w$ and $h$ would be set to minimize overlap between glyphs.
 
 This way of thinking of the glyph-map is convenient for several reasons.
-The first is that it moves the plots beyond the small multiples, many little plots, approach.
+The first is, that it moves the plots beyond the small multiples, many little plots, approach.
 That provides flexibility to cover large spatial domains using potentially tiny glyphs.
 A second, and major, reason is that with this approach they can then readily be made interactive.
-Different variables can be swept into, and out of, the minor axes using mouse motion, the scale can be adjusted by varying $w$ and $h$ proportional to mouse movement.
+Different variables can be swept into, and out of, the minor axes using mouse motion that maps to $\alpha_x, \alpha_y$.
 Interactive glyph-maps were used to explore the EXPO data using  GGobi \citep{swayne:2003}, and then later reproduced in publication quality resolution using {\tt ggplot2} \citep{me:ggplot2} in \citet{hobbs:2010}.
-In GGobi, a manual tour is used to make the linear combination of spatial variables with other variables \citep{CB95}, and equation \ref{coords.eqn} makes this approach more explicit.
+In GGobi, a manual tour is used to make the linear combination of spatial variables with other variables \citep{CB95}, and Equation \ref{coords.eqn} makes this approach more explicit, and generalizable beyond tours.
 The third reason is that it enables adjusting scales of glyphs in multiple ways to vary the emphasis, which is discussed in Section \ref{sec:perception}.
@@
 
-Coordinates for star glyphs are computed with a polar transformation, with minor axes scaled to $[0, 1]$:
+Coordinates for star glyphs are computed with a polar transformation of the minor components $(t, z)$:
 
@@
   \begin{array}{lll}
-  x &=& x_{\amaj} + y_{\amin} \cdot \frac{w}{2} \sin(2 \pi x_{\amin}) \\
-  y &=& y_{\amaj} + y_{\amin} \cdot \frac{h}{2} \cos(2 \pi x_{\amin}).
+  x &=& s_x + \alpha_x \cdot z \cdot w \cdot \sin(2 \pi t) \\
+  y &=& s_y + \alpha_y \cdot z \cdot h \cdot \cos(2 \pi t).
   \end{array}
@@
 
-\noindent This is a non-standard conversion to polar coordinates, but it creates a timeline that starts at 12 o'clock and proceeds clockwise.
-\citet{eden:2010} wraps the time series, overplotting the multiple years, and this could be done also, by adjusting the scale of $x_{\amin}$.
+\noindent where $(t, z)$ have been scaled into the range $[0, 1]$.
+This is a non-standard conversion to polar coordinates, but it creates a timeline that starts at 12 o'clock and proceeds clockwise.
+\citet{eden:2010} wraps the time series, overplotting the multiple years, and this could be done also, by adjusting the scale of $t$.
 
@@
 
-Figure~\ref{fig:templates} gives some small examples of time series with prototypical trends, shown both in Cartesian coordinates and polar coordinates.
+Figure~\ref{fig:templates} gives some iconic examples of time series, shown both in Cartesian coordinates and polar coordinates.
 Differences between linear and nonlinear trend are more apparent in the line-glyphs, and are effectively lost in the star-glyphs, which are most effective in exposing cyclical patterns due to seasonality.
 Star-glyphs show seasonality as floral cartoons, with peaks forming petals.
-It is surprising that time series with opposing trends in line-glyphs don't show this symmetry when displayed with star-glyphs.
-The area of star-glyphs coordinates mainly reflect the average value of a time series, while its shape shows deviations from the average.
+Opposing linear trends (up vs down) in line-glyphs are not readily seen in the star-glyphs, appearing only as differences in the gaps at the top of each glyph.
+The internal area of the star-glyphs reflects to the average value of a time series, while its shape shows deviations from the average.
 

### 2012-03-06 revision/paper.tex
done with perception

@@
 The structured spatial arrangement of icons in a glyph-map helps to compare the patterns of shape, like slope, intercept, or size, across icons.
-However, additional clues can make comparisons easier, converting the perceptual task from comparing length to the easier position along a common scale \citep{cleveland:1984}.
+However, additional clues can make comparisons easier, converting the perceptual task from comparing length to the easier task of position along a common scale \citep{cleveland:1984}.
 This is also called a \emph{visual reference grid} \citep{cleveland:1993a}.
 
-Each glyph is small, so there is not enough space for a full set of axes.
+Each glyph is small, so there is not enough space for a full set of axes on each, and it is not desirable beause it would be repetitive, redundant information.
 Instead, minimal reference lines and boxes are incorporated.
@@
 
-To observe the scale of the measurements represented by the glyphs we recommend a full legend, alongside the glyph-map, that contains the glyphs overplotted ignoring the spatial location, with complete axes.
-See, for example, Figure \ref{fig:nasa-glyph} where from the legend it can be seen that months range from 0 to more than 60 (actually 72) and de-seasonalized temperature ranges from -8 to 8$^0$K.
+To observe the scale of the measurements represented by the glyphs secondary legend plot is generated and displayed alongside the glyph-map, of all the glyphs overplotted, with complete axes.
+See, for example, Figure \ref{fig:nasa-glyph} where from the legend it can be seen that months range from 0 to 72 and de-seasonalized temperature ranges from -8 to 8$^0$K.
 In Figure \ref{fig:ref-basic}, months range from 1 to 12, and temperature from 270 to 315$^o$K.
@@
 
-Different features of the data can be emphasized by varying the scale used within each cell.
-By default, a global scale is used, so the same position within each cell corresponds to the same value in all locations.
+Different features of the data can be emphasized by varying the scale of $t, z$ (Equation \ref{coords.eqn}, \ref{coords.polar.eqn}) used within each cell.
+By default, a global scale, using the overall minimum and maximum of all $t, z$ values, is used to scale the measurements into the range $[-1, 1]$ (or $[0, 1]$ for polar coordinates).
+Then the range of values in one cell is comparable to those in other cells -- the same position within each cell corresponds to the same value in all locations.
 This facilitates the comparison of absolute values, and draws attention to locations with large overall variation.
+Alternatively, local scale can be used.
+With local scaling, the minimum and maximum values of $t, z$ are calculated for each location, and these are used to scale $t, z$ into the range $[-1, 1]$ (or $[0, 1]$).
+This makes it easier to compare shape (ignoring amplitude), and draws attention to locations with large relative variation.
 
-Alternatively, local scale can be used.
-In local scaling, the values within each location are scaled to range $[0, 1]$ before plot construction.
-This makes it easier to compare shape (ignoring amplitude), and draws attention to locations with large relative variation.
 Figure~\ref{fig:scaling} compares global scaling and local scaling for the smoothed temperature data from the EXPO data.
@@
 Seasonal effects are removed but not overall average or long-term trend.) On the global scale, the El Ni\~no blip in temperature is just visible, along with variation in the average temperature at each location.
-Local scaling emphasizes the individual shapes: the impact of El Ni\~no can be seen to be over a wider region, linear increases across the Andes, and decreases in the Gulf of Mexico.
+Local scaling emphasizes the individual shapes: the impact of El Ni\~no can be seen to be over a wider region, along with linear increases in parts of South America, and decreases in the Gulf of Mexico.
 
@@
 
-  \item The mean and standard deviation (or robust equivalents) could be used
-  to standardize values to common moments.
+\item The mean and standard deviation (or robust equivalents) could be
+  used to standardize values to common moments. This may lead to some
+  glyphs overlapping neighbors.
 
    \item Scaling to maximum 1 (but leaving the minimum unchanged) can help
-  emphasis the relative shape, but not distort the range as much as rescaling
-  to range $[0, 1]$.
+  emphasis the relative shape.
 
@@
 Note that the use of local scales needs to be clearly marked.
-The viewer must realize that the scale is not relative and that big patterns in some locations might be just tiny effects.
+The viewer must realize that the scale is not relative and that big patterns in some locations might be just tiny, nugatory effects.
 To indicate this on the locally scaled data plots, one can use an additional aesthetic, like color, to encode the range of original scale.
 This is illustrated in Figure~\ref{fig:scaling-col}.
-In the encoding, emphasis is placed on the locations with larger range, stronger effects: the line intensity is stronger (left plot), and the background is lightened to provide greater contrast (right plot) drawing the eye to these regions.
+In the encoding, emphasis is placed on the locations with larger range, stronger effects: the line intensity is darker (left plot), and the background is lightened to provide greater contrast (right plot) drawing the eye to these regions.
 
@@
 One more important point: if a model summary is displayed in the glyph-map, as discussed in the following section, it may be important to force the scale to that of the original data.
-The range of predicted values is typically smaller than the raw data, so angles of slopes are increased, giving an inaccurate sense of a stronger effect.
+The range of predicted values is typically much smaller than the raw data, so angles of slopes are increased, giving an inaccurate sense of a stronger effect.
 

### 2012-03-06 revision/paper.tex
done with large data

@@
 The glyph-map calculations are as efficient as possible computationally, being linear in the number of spatial locations.
-It is possible that the calculations could also be paralellized for massive amounts of climate model data.
+It is possible that the calculations could also be parallelized for massive amounts of climate model data.
 Visualizing large data also hits a perceptual boundary.
@@
 Another important feature is variability, shown by the thickness of the circle.
-Some locations, particularly in the north western mountain region, have large variations in temperature.
+Some locations, particularly in the north and western mountain region, have large variations in temperature.
 
@@
 Trend is examined by fitting a linear model to each location and displaying predicted values in Figure~\ref{fig:gistemp-pred}.
-This plot shows long-term (de-seasonalized) temperature trends for 1950--2010 (the limited time period used due to missing data).
-Most locations have increasing temperature over this time period.
+This plot shows long-term (de-seasonalized) temperature trends for 1950--2010.
+(The reduced time period is used to eliminate the influence of the missing data values in early years.) Most locations have increasing temperature over this time period.
 Only a few locations show a decline, and these few locations tend to be different from their neighbors, and on the edge of the data collection region, indicating that there might be isolated data problems.

### 2012-03-06 revision/paper.tex
done with irregular

@@
 % (as done in the right panel in Figure \ref{fig:irregular}).  
-However, there are generally limitations in the maximum display size (or resolution) and the utility of plotting very small areas.
+%However, there are generally limitations in the maximum display size (or resolution) and the utility of plotting very small areas.
 
@@
 In Figure \ref{fig:irregular-collapsed} nearby locations are collapsed to a single icon.
-In the left panel each station still has an individual glyph but locations close to each other share a plotting area.
-Not all glyphs are suited to this kind of space sharing, another option is to perform a summary of the locations that are collapsed and display that instead of the raw data.
-The right panel shows an example of this where the average trend for the collapsed locations is displayed.
-It is helpful to differentiate glyphs that involve summaries of more than one location, here a heavier line indicates at least two locations have been averaged.
+In the left panel each station still has an individual glyph but locations close to each other share a plotting area (note the multiple lines visible in some reference boxes).
+The right panel shows an example where glyphs are generated on aggregated data: the average trend for the collapsed locations is displayed.
+To differentiate glyphs that involve summaries of more than one location, heavier lines (barely visible) indicate at least two locations that have been averaged.
 The aggregation of spatial locations to a single summary can be misleading.
@@
 
-In both examples of collapsing locations, the icon locations still represent real geographical locations of data, but contain data from up to an icon width away.
-An alternative way to combine stations is to round locations to the icon size, Figure \ref{fig:irregular-grid}.
-This produces a regular grid with the advantage of easier perception of structure.
-However, it hides the irregular nature of the locations and icons may end up centered over nonsense places (i.e.
-land measurements in the ocean).
+In both of the examples where locations have been collapsed, the icon locations still represent an actual geographic location of collected data, albeit including data from up to an icon-width away.
+An alternative way to combine stations is to round locations to the icon size, as done in Figure \ref{fig:irregular-grid}.
+This produces a regular grid with the advantage of easier perception of structure \citet{Kr12b}.
+(Figure 9.12 of \citet{IPCC} uses a regular gridded time series over a spatial domain to illustrate temporal patterns in climate change data.) It hides the irregular nature of the locations, though, and a complication is that icons may end up centered over nonsense places, such as land measurements in the ocean.
 
@@
 
-Reference guides become especially important in the irregular location case.
-Comparisons are harder in this case, as the ability to easily compare icons along a common baseline, be it along a row or column of glyphs, is absent.
-Without reference it is hard to tell if short series are missing data at the start or end.
+Reference guides become especially important in the irregularly gridded data.
+Comparisons are harder because comparing icons along a common baseline, be it along a row or column of glyphs, is absent.
+Without reference lines or boxes it is also hard to compare line lengths, to tell if short series are missing data at the start or end.
+In each of the plots in Figures \ref{fig:irregular}, \ref{fig:irregular-collapsed} and \ref{fig:irregular-grid}, reference lines or boxes have been included.
 

### 2012-03-06 revision/paper.tex
done completely!

@@
 An alternative way to combine stations is to round locations to the icon size, as done in Figure \ref{fig:irregular-grid}.
-This produces a regular grid with the advantage of easier perception of structure \citet{Kr12b}.
+This produces a regular grid with the advantage of easier perception of structure \citep{Kr12b}.
 (Figure 9.12 of \citet{IPCC} uses a regular gridded time series over a spatial domain to illustrate temporal patterns in climate change data.) It hides the irregular nature of the locations, though, and a complication is that icons may end up centered over nonsense places, such as land measurements in the ocean.
@@
 
-Glyphs can be more general than time series or stars.
-For multivariate data we may want to represent several variables in the glyphs.
-Figure~\ref{fig:cloud} shows scatterplots of the EXPO data as icons: temperature values are displayed horizontally and high-cloud values are displayed vertically, both locally scaled.
-The bivariate relationship between the two variables is explored spatially, and it is clear that it is different across the region.
-The equatorial Pacific and Caribbean have positive association between temperature and high cloud, while the south American continent has negative association.
-The second plot displays a loess curve fit to the data instead of the individual points, sharpening the view of association.
-Another option is available in Gauguin \citep{gribov:2006}, a icon based on a histogram, giving small univariate distribution displays for each location.
+Glyphs, of course, can be more general than lines or stars.
+For multivariate data we may want to represent several variables ($z_1, \dots, z_p; p=$ number of variables) in the glyphs.
+Figure~\ref{fig:cloud} shows scatterplots of the EXPO data as icons: temperature values ($z_1$ instead of $t$) are displayed horizontally and high-cloud values ($z_2$) are displayed vertically, both locally scaled.
+The bivariate relationship between the two variables is explored spatially: the equatorial Pacific and Caribbean have positive association between temperature and high cloud, while the south American continent has negative association.
+The right side plot displays a loess curve fit to the data instead of the individual points, sharpening the association signal realtive to the variation.
+Gauguin \citep{gribov:2006}, has an icon based on a histogram, giving small univariate distribution displays for each location.
+\citet{carr:1992} use ray glyphs to show bivariate relationships.
 
@@
 Are their certain types of periodicity that are particularly easy to spot?
-These questions need to be answered with rigorous perceptual studies to help guide the use of these plots.
+Rigorous perceptual studies will help guide the use of glyph-maps.
 
 The conceptualization and equations for computing the glyph-maps lend themselves to producing these displays interactively.
-Different variables can be swept into the display using mouse motion, resolution could be increased upon zooming in to a spatial neighborhood.
+Different variables can be swept into the display using mouse motion mapped to $\alpha_x, \alpha_y$ (Equation \ref{coords.eqn}), resolution could be increased, and aggregated data icons disengaged, upon zooming into a spatial neighborhood.
 This experimentation is ideally done using R, so that computation can be linked with displays.
-The hurdle is the lagging development of interactive graphics in R.
-Several new packages that support interactive graphics in R have become available in the past year \citep{qtbase, qtpaint, plumbr} which will enable these experiments.
+The hurdle has been the lagging development of interactive graphics in R.
+Several new packages  in R have become available in the past year \citep{qtbase, qtpaint, plumbr} that support interactive graphics which will enable these developments.
 
 In summary, glyph-maps enable the study of temporal patterns in multivariate spatiotemporal data.
-Climate change is focused on changes over time, so glyph-maps provide a way to study these changes directly.
+Climate change is focused on changes over time, so glyph-maps provide a way to explore these changes directly.
 Glyph-maps enable different resolutions of the data to be examined: the raw data to discover data quality issues, global trend, seasonality, residuals, and multivariate dependencies.
@@
 
+\section*{Acknowledgements}
+
+This work was partially supported National Science Research grant DMS0706949.
+
 \bibliography{references}

### 2012-03-06 revision/code/1-examples.r

@@ -45,7 +45,7 @@ ggplot(resids, aes(long, lat)) +
   geom_raster(aes(fill = temp_resid)) +
   outline_nasa + 
   facet_grid(year ~ month) + 
-  scale_fill_gradient2(mid = "white", high = "red", low = "blue") + 
+  scale_fill_gradient2("Residuals", mid = "white", high = "red", low = "blue") + 
   theme_fullframe() + 
   opts(
     axis.text.x = theme_blank(), 

### 2012-03-07 revision/paper.tex
made data plural everywhere

@@
 
-Climate data is composed of multiple variables, such as temperature, ozone, precipitation, and winds, with a spatial and temporal context.
+Climate data are composed of multiple variables, such as temperature, ozone, precipitation, and winds, with a spatial and temporal context.
 The classic display for data of this type is a ``heatmap'', a tiled plot using color to display the value of the variable of interest at each location (see many examples in \citet{IPCC}, like Figure 2.11).
@@
 The darker grey in the background represents the land mass of the study region.
-Because the data is de-seasonalized, the primary visible structure is still El Ni\~no, recognized by the bump in the middle of the series of the equatorial Pacific region.
+Because the data are de-seasonalized, the primary visible structure is still El Ni\~no, recognized by the bump in the middle of the series of the equatorial Pacific region.
 It is both a temporal anomaly and a spatial phenomenon, so this is not surprising.
@@
 
-\item[EXPO] The ASA 2009 data expo data \citep{murrell:2010} consists
+\item[EXPO] The ASA 2009 expo data \citep{murrell:2010} consists
   of monthly observations of several atmospheric variables from the
-  International Satellite Cloud Climatology Project. The dataset
+  International Satellite Cloud Climatology Project. The data set
   includes observations over 72 months (1995--2000) on a 24 x 24 grid
@@
 \end{quote}
-Supplementary material contains the datasets and code to reproduce the plots in this paper.  R \citep{R} with the packages {\tt ggplot2} \citep{me:ggplot2} and {\tt plyr} \citep{me:plyr} was used.
+Supplementary material contains the data sets and code to reproduce the plots in this paper.  R \citep{R} with the packages {\tt ggplot2} \citep{me:ggplot2} and {\tt plyr} \citep{me:plyr} was used.
 
@@
 Figure~\ref{fig:ref-basic} demonstrates both guides.
-Reference elements are drawn in white, so they are minimally perceptible relative to the black data.
+Reference elements are drawn in white, so they are minimally perceptible relative to the black data elements.
 Average temperatures were calculated for each month, and these were passed to the glyph calculations to produce the glyph-maps.
@@
 
-Climate data is usually large: many locations in space, and many points in time, which can make it arduous to work with computationally.
+Climate data are usually large: many locations in space, and many points in time, which can make it arduous to work with computationally.
 The glyph-map calculations are as efficient as possible computationally, being linear in the number of spatial locations.
@@
 
-Figure~\ref{fig:gistemp-pred} shows the glyph-map for the GISTEMP dataset, which is more realistically sized than the EXPO data.
+Figure~\ref{fig:gistemp-pred} shows the glyph-map for the GISTEMP data set, which is more realistically sized than the EXPO data.
 Glyphs at each location represent temperature from 1880--2011, starting with Jan 1880 at 12 o'clock and proceeding clockwise to 2011 at 11:59.

### 2012-03-07 revision/cover-letter.tex
all done, trying to resubmit now

\documentclass{letter}
\usepackage{graphics}
\usepackage{fullpage}
\topmargin=-0.2in
%\headheight=0in
\setlength\oddsidemargin{-0.2in}
\textheight=7in
\textwidth=6in
\def\apl#1#2{
\begin{letter}{#1}
\signature{Professor Dianne Cook\\Department of Statistics\\Iowa State University 
\\Ames, Iowa 50011-1210\\ Phone (515) 294 8865\\{\tt dicook@iastate.edu}}
\opening{Dear #2,}

Please find attached our revision of  "Glyph-maps for Visually Exploring Temporal Patterns in Climate Data and Models" for the special issue of Environmetrics on Climate, with authors Hadley Wickham, Heike Hofmann, Charlotte Wickham and myself.

We appreciate the hard work and comments of the reviewers. We have made many changes to the paper reflecting the reviews, as follows:

\begin{enumerate}
\item All reviewers pointed to additional literature to cite. We have revised the literature accordingly and added more references that we found to be relevant, including the most recent IPCC report.  We have to admit that we were embarassed to have missed the Eden reference! This was a major oversight - yes, we had a paper in the same issue, and we had a poster in the same competition as these authors, but we reviewed a different paper of the issue and we had not realized how similar their work for the poster was to ours! 
\item We believe that our paper adds substantially to the literature. The major contributions are coordinating the ideas used in previous publications on glyph plots in a way that will facilitate future work on interactive versions and perception experiments, and the adaption of the methods for irregularly gridded data.  This has been made more explicit in the paper.
\item The section on large data has been expanded to better explain the approach and constraints. The actual bounds on data size are two-fold: one is computational, but the algorithm we have is efficient in this regard, and will depend somewhat on the memory of the computer used. The visual resolution is the second limit, and this requires some resolution reduction methods.
\item The section on Construction has been re-posed as Structural components to better explain the key ideas behind the glyph-map development.
\item Models used have been explained in more detail. Code and data are provided as supplementary material so that all work can be reproduced.
\item Developing rigorous perceptual studies requires some careful work. Given the time frame to revise the paper, this would be impossible, but time is not the only reason for delaying this work. It is better tackled as a separate body of work that has the appropriate human subjects approval, and experimental design.
\item Wording that was not clear or accurate and references to particular geographic locations have been fixed.
\item The reviewers asked for axes and legends for the glyph maps, which is a good idea. We adopted a style that creates an additional plot that is displayed alongside the glyph-map, that ignores spatial location, to use as a legend. This has the advantage that the relationship between, trend and variance of, the glyphs can be directly compared with the overplotting, and the axes on this plot provide the scales of the glyphs in the glyph-map. It is also a style that can be generalized to many different types of glyphs that might be used in glyph-maps. One reviewer suggested annotating the plots with details of the scale, or geographic region. We feel that with today's technology, this should be avoided. In past years we have carefully re-processed each plot in a drawing program. But there is a distinct advantage to being able to re-generate plots exactly as they appear in a paper, for reproducibility, and for speed of re-working ideas.
\item References and explanation of spatial aggregation issues (MAUP) have been added.
\item Forward references have been removed.
\item The typos and inconsistencies corrected, and a careful proof-read of the final version.
\end{enumerate}

\closing{Sincerely yours,
{\pdfimage width 1.5in {signature2.pdf}}}
\end{letter}}

\begin{document}

{{\pdfimage width 4in {letterhead.jpg}}}
\apl{Editors \\Environmetrics Special Issue on Climate}{Professors Guttorp, Sain and Wikle}

\end{document}

### 2012-05-14 revision-final/code/1-examples.r
final revision

@@ -1,0 +1,77 @@
+# This file contains the glyph-map for nasa data Might think about
+# putting all of the glyph-map examples, using gistemp and ushcn data
+# here too
+
+source("glyph.r")
+source("glyph-utils.r")
+source("maps.r")
+source("data-nasa.r")
+
+temp_seas_models <- dlply(nasa, c("lat", "long"), function(df) {
+  lm(surftemp ~ factor(month), data = df)
+})
+
+# Including trend
+temp_trend_models <- dlply(nasa, c("lat", "long"), function(df) {
+  lm(surftemp ~ year + factor(month), data = df)
+})
+
+locs <- dlply(nasa, c("lat", "long"))
+
+resids <- mdply(cbind(d = locs, m = temp_seas_models), function(d, m) {
+  d$temp_resid <- d$surftemp - predict(m, newdata = d)
+  d
+})
+
+preds <- mdply(cbind(d = locs, m = temp_seas_models), function(d, m) {
+  d$temp_pred <- predict(m, newdata = d)
+  d
+})
+
+# First figure: facetted colored map -----------------------------------------
+
+outline_nasa <- list(
+  geom_polygon(aes(long, lat, group = group), inherit.aes = FALSE, 
+    data = .nasa, show_guide = FALSE, fill = NA, colour = "grey60"),
+  scale_x_continuous(expand = c(0.02, 0)),
+  scale_y_continuous(expand = c(0.02, 0)), 
+  xlab(NULL),
+  ylab(NULL))
+
+# This plot is INCREDIBLY slow in 0.8.9, so I've switched to 0.9.0 to make
+# it a bit faster - it's still not great, but it's at least tolerable.
+
+ggplot(resids, aes(long, lat)) +
+  geom_raster(aes(fill = temp_resid)) +
+  outline_nasa + 
+  facet_grid(year ~ month) + 
+  scale_fill_gradient2("Residuals", mid = "white", high = "red", 
+    low = "blue") + 
+  theme_fullframe() + 
+  opts(
+    axis.text.x = theme_blank(), 
+    axis.text.y = theme_blank(),
+    axis.ticks = theme_blank(), 
+    axis.ticks.length = unit(0, "lines"), 
+    axis.ticks.margin = unit(0, "lines"),
+    legend.position = "right")
+ggsave("../images/nasa-colored-map.png", height = 5, width = 9, dpi = 72)
+
[…]

### 2012-05-14 revision-final/code/3-reference-frames.r

@@ -1,0 +1,27 @@
+source("models-nasa.r")
+source("maps.r")
+source("glyph.r")
+
+month_preds <- glyphs(month_preds, "long", "month", "lat", "pred")
+
+ggplot(month_preds, aes(gx, gy, group = gid)) + 
+  map_nasa +
+  add_ref_lines(month_preds) +
+  geom_path() + 
+  theme_fullframe() 
+ggsave("../images/ref-line.pdf", width = 6, height = 6)
+
+ggplot(month_preds, aes(gx, gy, group = gid)) + 
+  map_nasa +
+  geom_path() + 
+  add_ref_boxes(month_preds) +
+  theme_fullframe() 
+ggsave("../images/ref-box.pdf", width = 6, height = 6)
+
+ggplot(month_preds, aes(month, pred)) + 
+  geom_line(aes(group = gid), alpha = 1/10) + 
+  opts(aspect.ratio = 1) + 
+  xlab("Month") + 
+  ylab("Average temperature (K)")
+ggsave("../images/ref-legend.pdf", width = 4, height = 4)
+

### 2012-05-14 revision-final/code/3-scaling.r

@@ -1,0 +1,110 @@
+library("ggplot2")
+library("plyr")
+library("mgcv")
+
+source("models-nasa.r")
+source("glyph.r")
+source("maps.r")
+
+# Rescale predictions to individual scales -----------------------------------
+
+day_preds <- glyphs(day_preds, "long", "day", "lat", "pred") 
+ggplot(day_preds, aes(gx, gy, group = gid)) + 
+  map_nasa + 
+  add_ref_boxes(day_preds) +
+  geom_path() +
+  theme_fullframe()
+ggsave("../images/month-rescale-none.png", width = 4, height = 4)
+
+last_plot() %+% glyphs(day_preds, "long", "day", "lat", "pred", 
+  y_scale = range01) 
+ggsave("../images/month-rescale01.png", width = 4, height = 4)
+
+ggplot(day_preds, aes(day, pred)) + 
+  geom_line(aes(group = gid), alpha = 1/10) + 
+  opts(aspect.ratio = 1) + 
+  xlab("Days from start") + 
+  ylab("Average temperature (K)")
+ggsave("../images/month-rescale-legend.pdf", width = 4, height = 4)
+
+day_preds2 <- ddply(day_preds, c("lat", "long"), mutate, 
+  pred_s = rescale01(pred),
+  pred_m = pred / max(pred),
+  max = max(pred),
+  range = diff(range(pred))
+)
+
+day_preds2 <- glyphs(day_preds2, "long", "day", "lat", "pred_s") 
+ggplot(day_preds2, aes(gx, gy, group = gid)) + 
+  map_nasa + 
+  add_ref_boxes(day_preds) +
+  geom_path(aes(colour = range)) +
+  theme_fullframe() + 
+  scale_colour_gradient("Temperature\nrange (K)",
+    high = "black", low = "grey60", limits = c(0, 8.5),
+    breaks = seq(0, 8, by = 2), guide = guide_colourbar(
+      direction = "horizontal", title.vjust = 0.7, 
+      title.theme = theme_text(face = "bold"))) +
+  opts(legend.position = "bottom", aspect.ratio = 1) 
+ggsave("../images/month-rescale01-col.png", width = 4, height = 4.5)
+
+grid <- unique(day_preds2[c("lat", "long", "range")])
+
+ggplot(day_preds2) + 
+  map_nasa + 
+  geom_tile(aes(long, lat, fill = range), data = grid, alpha = 0.5) +
+  geom_path(aes(gx, gy, group = gid)) +
[…]

### 2012-05-14 revision-final/code/4-large.r

@@ -1,0 +1,140 @@
+library(ncdf)
+library(reshape2)
+library(ggplot2)
+library(lubridate)
+
+# Predictions
+source("maps.r")
+source("glyph.r")
+
+# Load data ------------------------------------------------------------------
+
+if (file.exists("temp_usa.rds")) {
+  temp_usa <- readRDS("temp_usa.rds")
+} else {
+  if (!file.exists("air.2x2.250.mon.anom.land.nc")) {
+  download.file("ftp://ftp.cdc.noaa.gov/Datasets/gistemp/landonly/250km/air.2x2.250.mon.anom.land.nc", "air.2x2.250.mon.anom.land.nc")
+  }
+
+  fo<-open.ncdf("air.2x2.250.mon.anom.land.nc")
+  fo$nvars
+  lon <- get.var.ncdf(fo, "lon")
+  lat <- get.var.ncdf(fo, "lat")
+  time <- get.var.ncdf(fo, "time")
+  temp <- get.var.ncdf(fo, "air")
+  # Convert lon from 1 to 360 to -180 to 180
+  lon <- (lon + 180) %% 360 - 180
+
+  length(lon)
+  dim(temp)
+
+  # Ok, this now works for transforming the lon, and selecting the USA
+  temp.usa <- temp[114:150, 20:33,] # lon -130, -65; lat 25, 50
+  temp_usa <- melt(temp.usa)
+  colnames(temp_usa) <- c("gridx", "gridy", "time", "temp")
+  temp_usa$lon <- lon[temp_usa$gridx + 113]
+  temp_usa$lat <- lat[temp_usa$gridy + 19]
+
+  temp_usa <- subset(temp_usa, !(lon > -70 & lat < 37))
+
+  library(lubridate)
+  temp_usa$date <- ymd(18800101) + months(temp_usa$time - 1)
+  temp_usa$year <- year(temp_usa$date)
+  temp_usa$month <- month(temp_usa$date)  
+  
+  saveRDS(temp_usa, "temp_usa.rds")
+}
+
+temp_usa <- ddply(temp_usa, c("lon","lat"), function(df) {
+ if (all(is.na(df$temp))) NULL else df 
+})
+
+# Plots of raw data -------------------------------------------------------
+
+# temp_glyphs <- glyphs(temp_usa, "lon", "time", "lat", "temp")
+# 
+# ggplot(temp_glyphs, aes(gx, gy, group = gid)) + 
+#   map_gistemp +
[…]

### 2012-05-14 revision-final/code/5-irregular.r

@@ -1,0 +1,232 @@
+library(mgcv)
+
+source("glyph.r")
+source("glyph-utils.r")
+source("5-overlapping.r")
+source("data-ushcn.r")
+
+
+season_index <- function(time){
+  (time %% 12  + 1) / 12
+}
+
+temp.post1950 <- subset(ushcn, year >= 1950)
+
+temp.post1950 <- mutate(temp.post1950, 
+  time = (year - 1950) * 12 + as.numeric(month),
+  season = season_index(time))    
+
+# Modelling ------------------------------------------------------------------
+
+fit.post1950 <- dlply(temp.post1950, "stn", function(df) {
+  lm(temp ~ year + month, data = df, na.action = "na.exclude")
+}, .progress = "text")
+
+year_grid <- expand.grid(year = 1950:2010, month = "jan")
+lin.pred.post1950 <- ldply(fit.post1950, function(mod) {
+  year_grid$pred <- predict(mod, newdata = year_grid)
+  year_grid
+})
+
+lin.pred.post1950 <- merge(lin.pred.post1950, ushcn.stations)
+
+w <- 1
+h <- 1
+
+lin.pred.gly <- glyphs(lin.pred.post1950, "lon", "year", "lat", "pred", 
+  width = w, height = h, y_scale = mean0)
+
+ggplot(lin.pred.gly, aes(gx, gy, group = gid)) + 
+  map_ushcn + 
+  add_ref_lines(lin.pred.gly) +
+  geom_path() +
+  theme_fullframe() 
+ggsave("../images/usa-lin-overlap.png", width = 8, height = 4.5)
+
+ggplot(lin.pred.gly, aes(year, pred)) + 
+  geom_line(aes(group = gid), alpha = 1/20) + 
+  xlab("Year") + 
+  ylab("Linear Temp Anomaly (F)")
+ggsave("../images/usa-lin-legend.pdf", width = 4, height = 4)
+
+# # Mountain states: Colorado, Wyoming, Idaho, Montana, Nevada, Utah
+# mountain.states <- c("CO", "WY", "ID", "MT", "NV", "UT")
+# mountains.gly<-glyphs(subset(lin.pred.post1950, state %in% mountain.states),
+#   "lon", "year", "lat", "pred", width=.5, height=.5, y_scale = mean0)
+# 
+# ggplot(mountains.gly, aes(gx, gy, group = gid)) + 
[…]

### 2012-05-14 revision-final/code/5-overlapping.r

@@ -1,0 +1,67 @@
+require(plyr)
+get.stn.dists <- function(stn.df, x, y){
+  get.one <- function(stn1){
+    stn.id <- stn.df[stn1, "stn"]
+    stn.pairs <- stn.df[stn.df$stn > stn.id, ]
+    data.frame(stn1 = stn.id, stn2 = stn.pairs[ , "stn"],
+      stn.df[rep(stn1, nrow(stn.pairs)), c(x, y)] - 
+      stn.pairs[ , c(x, y)])
+  }
+  stn.dists <- ldply(1:(nrow(stn.df) - 1), get.one, .progress = "text")
+  names(stn.dists)[3:4] <- paste(names(stn.dists)[3:4], ".dist", sep = "")
+  stn.dists
+}
+
+overlap <- function(x.dist, y.dist, width, height){
+  abs(x.dist) < width & abs(y.dist) < height
+}
+
+# dist.df contains results of get.stn.dists (i.e. stn1, stn2 columns)
+combine.overlapping <- function(dist.df, x.dists, y.dists, width, height){
+  ids <- unique(unlist(dist.df[,c("stn1","stn2")]))
+  overlapping <- overlap(dist.df[[x.dists]], dist.df[[y.dists]],
+    width, height)
+  overlap.df <- subset(dist.df, overlapping)
+  overlap.df <- overlap.df[order(overlap.df$stn1), ]
+  overlap.ids <- unique(unlist(overlap.df[, c("stn1", "stn2")]))
+
+  groups <- data.frame(stn = overlap.ids, group = NA)
+  for (row in 1:nrow(overlap.df)){
+    stns <- overlap.df[row, c("stn1", "stn2"), drop = TRUE]
+    current.groups <- c(groups[groups$stn == stns[1], "group"], 
+      groups[groups$stn == stns[2], "group"])
+  
+    if (all(is.na(current.groups))){ # if no groups assigned assign to stn1
+      groups[groups$stn %in% stns, "group"] <- stns[1]
+      # then assign all other stns that overlap with stn1 to stn1
+      overlap.with.1 <- unique(unlist(overlap.df[(overlap.df$stn1 == stns[1] |
+        overlap.df$stn2 == stns[1]), c("stn1", "stn2")]))
+      groups[groups$stn %in% overlap.with.1, "group"] <- stns[1]
+    }
+  }
+  for (row in 1:nrow(overlap.df)){
+    stns <- overlap.df[row, c("stn1", "stn2"), drop = TRUE]
[…]

### 2012-05-14 revision-final/code/6-generalizations.r

@@ -1,0 +1,42 @@
+# Generalisations: scatterplots of arbitrary variables.
+source("glyph.r")
+source("glyph-utils.r")
+source("maps.r")
+source("data-nasa.r")
+
+nasa.scale <- ddply(nasa, c("long","lat"), summarise,
+  cloudlow.s = scale(cloudlow), 
+  cloudhigh.s = scale(cloudhigh),
+  temp.s = scale(surftemp))
+
+nasa.scat.gly <- glyphs(nasa.scale, "long", "temp.s", "lat", "cloudhigh.s") 
+ggplot(nasa.scat.gly, aes(gx, gy, group = gid)) + 
+  map_nasa +
+  add_ref_boxes(nasa.scat.gly) +
+  geom_point(size = 0.7, alpha = 0.5) + 
+  theme_fullframe() 
+ggsave("../images/nasa-scat-glyph.png", width = 6, height = 6)
+
+ggplot(nasa, aes(surftemp, cloudhigh)) + 
+  geom_point(size = 0.7, alpha = 1 / 10) + 
+  ylab("High cloud") + 
+  xlab("Surface temperature (K)")
+ggsave("../images/nasa-scat-legend.png", width = 4, height = 4)
+
+nasa.loess <- ddply(nasa.scale, c("long","lat"), summarise,
+  temp.s = temp.s,
+  pcloud = predict(loess(cloudhigh.s ~ temp.s)))
+nasa.loess.gly <- glyphs(nasa.loess, "long", "temp.s", "lat", "pcloud") 
+ggplot(nasa.loess.gly, aes(gx, gy, group = gid)) + 
+  map_nasa +
+  add_ref_boxes(nasa.loess.gly) +
+  geom_line() + theme_fullframe() 
+ggsave("../images/nasa-loess-glyph.png", width = 6, height = 6)
+
+nasa <- ddply(nasa, c("long","lat"), transform,
+  pcloud = predict(loess(cloudhigh ~ surftemp)))
+ggplot(nasa, aes(surftemp, pcloud)) + 
+  geom_line(aes(group = interaction(long, lat)), alpha = 1/10) + 
+  ylab("Smoothed high cloud") + 
+  xlab("Surface temperature (K)")
+ggsave("../images/nasa-loess-legend.png", width = 4, height = 4)

### 2012-05-14 revision-final/code/data-nasa.r

@@ -1,0 +1,13 @@
+if (!file.exists("../../data/nasadata.rds")) {
+  library("lubridate")
+
+  nasa <- read.csv("../../data/nasadata.csv.bz2", stringsAsFactors = FALSE)
+  nasa$date <- ymd(nasa$date)
+  nasa$day <- (nasa$date - min(nasa$date)) %/% ddays(1)
+  nasa$month <- month(nasa$date)
+  nasa$year <- year(nasa$date)
+  
+  saveRDS(nasa, "../../data/nasadata.rds")
+} else {
+  nasa <- readRDS("../../data/nasadata.rds")
+}

### 2012-05-14 revision-final/code/data-ushcn.r

@@ -1,0 +1,41 @@
+library(maps)
+library(ggplot2)
+library(reshape2)
+library(plyr)
+
+source("maps.r")
+
+ushcn.loc <- "../../data/ushcndata.rds"
+ushcn.stations.loc <- "../../data/ushcnstations.rds"
+
+if (!file.exists(ushcn.loc)) {
+  temp.all<-read.table("../../data/9641C_201012_raw.avg")
+  colnames(temp.all)<-c("year", "jan", "feb", "mar", "apr", "may", "jun", "jul",
+    "aug", "sep", "oct", "nov", "dec", "ann")
+  temp.all$stn<-temp.all$year%/%100000
+  temp.all$year<-temp.all$year%%10000
+
+  stn.all<-read.table("../../data/ushcn-stations.txt", fill=T)
+  stn.all<-stn.all[,c(1:3,5:6)]
+  colnames(stn.all)<-c("stn", "lat", "lon", "state", "town")
+  stn.all$name<-paste(stn.all$town, stn.all$state)
+  ushcn.stations <- stn.all
+  
+  ushcn.all <- join(temp.all, stn.all)
+  ushcn <- melt(ushcn.all[ , -c(14, 19)], 
+    id=c("stn", "year", "lat", "lon", "name", "state"))
+  ushcn <- rename(ushcn, 
+      c("variable" = "month", "value" = "temp"))
+
+  ushcn <- mutate(ushcn, 
+      temp = temp/10,
+      date = as.Date(paste(year, "-", as.numeric(month), "-01", sep = "")),
+      )    
+  ushcn$temp[ushcn$temp == -999.9] = NA
+  
+  saveRDS(ushcn, ushcn.loc)
+  saveRDS(ushcn.stations, ushcn.stations.loc)
+} else {
+  ushcn <- readRDS(ushcn.loc)
+  ushcn.stations <- readRDS(ushcn.stations.loc)
+}

### 2012-05-14 revision-final/code/glyph-utils.r

@@ -1,0 +1,44 @@
+library("ggplot2")
+library("grid")
+
+theme_fullframe <- function (base_size = 12){
+  opts(
+    axis.line = theme_blank(), 
+    # axis.text.x = theme_blank(), 
+    # axis.text.y = theme_blank(),
+    # axis.ticks = theme_blank(), 
+    axis.title.x = theme_blank(), 
+    axis.title.y = theme_blank(), 
+    # axis.ticks.length = unit(0, "lines"), 
+    # axis.ticks.margin = unit(0, "lines"), 
+    legend.position = "none", 
+    panel.border = theme_blank(), 
+    panel.grid.major = theme_blank(), 
+    panel.grid.minor = theme_blank(), 
+    panel.margin = unit(0, "lines"), 
+    plot.background = theme_blank(), 
+    plot.margin = unit(rep(2, 4), "mm")
+  )
+}
+
+add_ref_lines <- function(data, colour = "white", size = 1.5, ...){
+  rl <- ref_lines(data)
+  geom_line(data = rl, colour = colour, size = size, ...)
+}
+
+add_ref_boxes <- function(data, var_fill = NULL, colour = "white", size = 0.5, 
+    fill = NA, ...){
+  rb <- ref_boxes(data, var_fill)
+  if (!is.null(var_fill)){
+    geom_rect(aes_all(names(rb)), data = rb,
+     colour = colour, size = size, inherit.aes = FALSE, ...)
+   }
+   else{
+     geom_rect(aes_all(names(rb)), data = rb,
+      colour = colour, size = size, inherit.aes = FALSE, fill = fill, ...)
+   }
+}
+
+# Need to have a function for the circles references for polar coords
+# Need to have a function for the reference lines
+# Need to have reference boxes for the irregular grid, too

### 2012-05-14 revision-final/code/glyph.r

@@ -1,0 +1,175 @@
+library(plyr)
+source("glyph-utils.r")
+
+#' Create the data needed to generate a glyph plot.
+#'
+#' @param data A data frame containing variables named in \code{x_major},
+#'   \code{x_minor}, \code{y_major} and \code{y_minor}.
+#' @param x_major,x_minor,y_major,y_minor The name of the variable (as a
+#'   string) for the major and minor x and y axes.  Together, each unique
+#    combination of \code{x_major} and \code{y_major} specifies a grid cell.
+#' @param polar A logical of length 1, specifying whether the glyphs should
+#'   be drawn in polar coordinates.  Defaults to \code{FALSE}.
+#' @param height,width The height and width of each glyph. Defaults to 95\% of
+#'  the \code{\link[ggplot2]{resolution}} of the data. Specify the width
+#'  absolutely by supplying a numeric vector of length 1, or relative to the
+#   resolution of the data by using \code{\link{rel}}.
+#' @param y_scale,x_scale The scaling function to be applied to each set of
+#'  minor values within a grid cell.  Defaults to \code{\link{identity}} so
+#'  that no scaling is performed.
+glyphs <- function(data, x_major, x_minor, y_major, y_minor, polar = FALSE, height = rel(0.95), width = rel(0.95), y_scale = identity, x_scale = identity) {
+  data$gid <- interaction(data[[x_major]], data[[y_major]], drop = TRUE)
+  
+  if (is.rel(width)) {
+    width <- resolution(data[[x_major]], zero = FALSE) * unclass(width)
+    message("Using width ", format(width, digits = 3))
+  }
+    
+  if (is.rel(height)) {
+    height <- resolution(data[[y_major]], zero = FALSE) * unclass(height)
+    message("Using height ", format(height, digits = 3))
+  }
+  
+  if (!identical(x_scale, identity) || !identical(y_scale, identity)) {
+    data <- ddply(data, "gid", function(df) {
+      df[[x_minor]] <- x_scale(df[[x_minor]])
+      df[[y_minor]] <- y_scale(df[[y_minor]])
+      df
+    })
+  }
+  
+  if (polar) {
+    theta <- 2 * pi * rescale01(data[[x_minor]])
[…]

### 2012-05-14 revision-final/code/maps.r

@@ -1,0 +1,65 @@
+library("maps")
+library("ggplot2")
+library("plyr")
+
+getbox <- function (map, xlim, ylim) {
+	# identify all regions involved
+	small <- subset(map, (long > xlim[1]) & (long < xlim[2]) & (lat > ylim[1]) & (lat < ylim[2]))
+	regions <- unique(small$region)
+	small <- subset(map, region %in% regions)	
+	
+	# now shrink all nodes back to the bounding box
+	small$long <- pmax(small$long, xlim[1])
+	small$long <- pmin(small$long, xlim[2])
+	small$lat <- pmax(small$lat, ylim[1])
+	small$lat <- pmin(small$lat, ylim[2])
+
+	# Remove slivvers
+	small <- ddply(small, "group", function(df) {
+	  if (diff(range(df$long)) < 1e-6) return(NULL)
+	  if (diff(range(df$lat)) < 1e-6) return(NULL)
+	  df
+	})
+
+	small
+}
+
+##### Prepping Data
+world <- map_data("world")
+
+states <- map_data("state")
+states$group <- max(world$group) + states$group
+both <- rbind(world, states)
+
+##### Maps
+.nasa <- getbox(both, xlim = c(-115, -55), ylim = c(-21.1, 36.6))
+map_nasa <- list(
+  geom_polygon(aes(long, lat, group = group), inherit.aes = FALSE, 
+    data = .nasa, show_guide = FALSE, fill = "grey80", colour = "grey90"),
+  scale_x_continuous(expand = c(0.02, 0)),
+  scale_y_continuous(expand = c(0.02, 0)), 
+  xlab(NULL),
+  ylab(NULL))
+
+ushcn <- subset(getbox(both, xlim = c(-126, -65), ylim = c(24, 52)),
+  region != "Great Lakes")
+map_ushcn <- list(
+  geom_polygon(aes(long, lat, group = group), inherit.aes = FALSE, 
+    data = ushcn, show_guide = FALSE, fill = "grey80", colour = "grey90"),
+  scale_x_continuous(expand = c(0.02, 0)),
+  scale_y_continuous(expand = c(0.02, 0)), 
+  xlab(NULL),
+  ylab(NULL))
+
+gistemp <- subset(getbox(both, xlim = c(-135, -57), ylim = c(22, 53)),
+  region != "Great Lakes")
+map_gistemp <- list(
[…]

### 2012-05-14 revision-final/code/modeling-adv.r

@@ -1,0 +1,39 @@
+source("data-nasa.r")
+source("glyph.r")
+
+library(ggplot2)
+library(plyr)
+library(mgcv)
+
+space <- dlply(nasa, "time", function(df) { 
+   gam(surftemp ~ s(lat, long), data =df)
+})
+
+loc_grid <- unique(nasa[c("lat", "long")])
+
+space_preds <- ldply(space, function(mod) {
+  loc_grid$pred <- predict(mod, newdata = loc_grid)
+  loc_grid
+})
+
+space_preds <- glyphs(space_preds, "long", "time", "lat", "pred")
+qplot(gx, gy, data = space_preds, geom = "line", group = gid) + map
+
+qplot(long, lat, data = space_preds, fill = pred, geom = "tile") + facet_wrap(~ time)
+
+
+big_mod <- gam(surftemp ~ s(long, lat, time) + factor(month), data = nasa)
+loc_grid <- expand.grid(
+  lat = unique(nasa$lat), 
+  long = unique(nasa$long), 
+  month = 1, 
+  time = seq(0, 2161, length = 50))
+loc_grid$pred <- predict(big_mod, newdata = loc_grid)
+
+loc_grid <- glyphs(loc_grid, "long", "time", "lat", "pred")
+qplot(gx, gy, data = loc_grid, geom = "line", group = gid) + map
+
+
+nasa$resid <- resid(big_mod)
+nasa <- glyphs(nasa, "long", "time", "lat", "resid")
+qplot(gx, gy, data = nasa, geom = "line", group = gid) + map

### 2012-05-14 revision-final/code/models-nasa.r

@@ -1,0 +1,32 @@
+library("plyr")
+library("mgcv")
+source("cache.r")
+source("data-nasa.r")
+
+# Model fitting --------------------------------------------------------------
+
+temp_models %<-cache% dlply(nasa, c("lat", "long"), function(df) {
+  lm(surftemp ~ year + factor(month), data = df)
+})
+
+temp_smooth %<-cache% dlply(nasa, c("lat", "long"), function(df) {
+  gam(surftemp ~ s(day) + factor(month), data = df)
+})
+
+month_grid <- expand.grid(year = 2000, month = 1:12)
+month_preds <- ldply(temp_models, function(mod) {
+  month_grid$pred <- predict(mod, newdata = month_grid)
+  month_grid
+})
+
+year_grid <- expand.grid(year = unique(nasa$year), month = 1)
+year_preds <- ldply(temp_models, function(mod) {
+  year_grid$pred <- predict(mod, newdata = year_grid)
+  year_grid
+})
+
+day_grid <- expand.grid(day = seq(0, 2161, length = 50), month = 1)
+day_preds <- ldply(temp_smooth, function(mod) {
+  day_grid$pred <- predict(mod, newdata = day_grid)
+  day_grid
+})

### 2012-05-14 revision-final/paper.tex
final revision
fixed figure 7 ref, and typo in section 5

--- 2012-03-07 revision/paper.tex
@@
 \newcommand\amaj{\text{major}}
+
+\setlength{\topmargin}{0.3in}
+\setlength{\textheight}{7.5in}
 
@@
 
-Figure~\ref{fig:gistemp-pred} shows the glyph-map for the GISTEMP data set, which is more realistically sized than the EXPO data.
+Figure~\ref{fig:gistemp-raw} shows the glyph-map for the GISTEMP data set, which is more realistically sized than the EXPO data.
 Glyphs at each location represent temperature from 1880--2011, starting with Jan 1880 at 12 o'clock and proceeding clockwise to 2011 at 11:59.
@@
 Glyph-maps work best when displayed in the coordinate system in which the grid is rectangular and regular.
-For example, the GISTEMP grid is 2$^o$ square and is displayed treating longitude and latitude as Cartesian coordinates in Figure \ref{fig:gistemp-pred}.
+For example, the GISTEMP grid is 2$^o$ square and is displayed treating longitude and latitude as Cartesian coordinates in Figures~\ref{fig:gistemp-raw}, \ref{fig:gistemp-pred}.
 For regularly gridded data, this results in icons that are rectangles of equal size.
 It is often preferable to display maps as a projection of the geographical coordinates which preserves a property of interest, area for example.
-When a projection is desired a choice needs to be made between applying the projection to the completed glyph-map, or to just the spatial locations.
+When a projection is desired, a choice needs to be made between applying the projection to the completed glyph-map, or to just the spatial locations.
 When the completed glyph-map is projected the icons are no longer rectangles of the same size but they do retain their space filling property.

### 2012-05-14 revision-final/paper.tex
fixed typo in abstract and removed comments in latex

@@
 
-% Remove words like "obvious", "challenge", "scratched the surface"
-% Make data URLs references
-% Do we have different aspect ratios between the different US maps?
-% Check where data is described, just use key names given in the description at beginning of paper
-
-% Found data: http://www.esrl.noaa.gov/psd/data/gridded/data.HADCRUT2.html
-%     air.mon.anom.nc, var adj, this is 5x5 degree grid, a bit sparse
-% Or: http://www.esrl.noaa.gov/psd/data/gridded/data.gisstemp.html
-% air.2x2.mon.anom.land.nc 250k smooth data, is on a 2x2 grid
-
-% Drawing trend lines on map
-%   - How to
-%   - Comparison to coloring slope
-%   - Reference to Grinstein icon plots
-% Pickett RM and Grinstein GG (1988). Iconographics 
-% displays for visualizing multidimensional data. 
-% In Proc. IEEE Conference on Systems, Man, and Cybernetics, pages 514-19.
-% Note the later work on metaphoric data display, garbage work!
-%   - Alexander Gribov's work http://rosuda.org/software/Gauguin/gauguin.html
-
-%   - Spatial star plots by Andrews http://www.udallas.edu:8080/~andrews/software/software.html
-%   - Look at what Bertin does
-%
-% Interactive graphics
-% Data processing
-%
-% Perhaps the GHCN data? Not gridded
-% NOAA data, almost grid, but over ocean, so maps more tricky
-% NARCAP from NCAR?
-% Fill in with NASA data, to start, try to put new data for final version
 \maketitle
-
-% re-work equations, to incorporate scale, and interaction, use spatial location letter and time/var letters
-
-% overlay legends on the plots?
 
@@
 The conceptualization described here also will enable developing interactive versions of glyph-maps, and make it simpler to explore the perceptual effects of different scales.
-The methods are developed for rectangular gridded data, but with some clever processing, which is explained, it is possible get good glyph-maps of irregularly gridded data.
+The methods are developed for rectangular gridded data, but with some clever processing, which is explained, it is possible to get good glyph-maps of irregularly gridded data.
 Guides and reference marks, for different types of glyphs used in the glyph-maps, are also discussed.
-
-%Understanding the multiple aspects of climate data is a challenge.
-There are both spatial and temporal components to examine, and often multivariate dependencies.
-This paper proposes a new display for climate data: glyph-maps.
-Glyph-maps are a specialization of multivariate glyph plots; each spatial location is displayed with one glyph that represents the measurements recorded over time at that location.
-Glyph-maps allow the discovery of both local and global structure, with a particular focus on temporal relationships.
-
-%The instructions for computing and displaying glyph-maps are described, along with the design of reference frames, and the effects of different scaling of the data on perception of structure.
-The methods are developed for rectangularly gridded data, but with some clever processing, which is explained, it is possible get good glyph-maps of irregularly gridded data.
 
@@
 
-%% Reproduce all of the icon displays using the fixed glyph code
 The glyph-map uses a small glyph, or icon, to represent multiple variables, or time values, at each location.
@@
 Temperatures are generally more varied over land, than sea.
-A few locations, near the coast of South America and the southern part of North America, have dramatic jumps in temperature in the later years.%, all in high elevation areas, along the Andes and the Sierra Madre Occidental in Mexico.
+A few locations, near the coast of South America and the southern part of North America, have dramatic jumps in temperature in the later years.
 
@@
 In the glyph-map, primary emphasis is placed on the temporal comparison, and secondarily on spatial comparison, because facetting is on spatial locations.
-
-% The most seasonality occurs in north American.
-Some seasonal effects can also be seen in the Caribbean and in south American locations.
-Around the equator the temperatures are relatively warm for all seasons.
-Instead of radially displaying the temperature a glyph can be a regular time series, time displayed horizontally and temperature vertically (right plot).
-The structure perceived is different with this change in glyph type: locations at the equator have flat series, on land more varied temperatures.
 
@@
 \end{figure}
-
-%Glyph-maps have attributes of both glyphs (stars, faces, profiles, ...), inviting exploration at the global level, and small-multiple time series, allowing local exploration of individual series.
-Allow to see both geographic context and individual locations simultaneously.
-
-%These displays tend to be particularly effective when printed in large format, as the much higher resolution of print (600 dpi) compared to computer  screen (72-120 dpi) allows for much finer reading.
-These are high-density data displays in the best tradition of Tufte.
-Wall size maps are not usually practical during analysis, but are very engaging.
 
@@
 
-%Glyph-maps can be generated with existing graphics software after performing a simple pre-processing step, making them easily accessible to a wide audience. 
 It is useful to recognize that the glyph-map is a linear mix of two structural components of the data: spatial location and data values.
@@
 
-%Since the coordinates are linear combinations of two variables, it is possible to build these these line-glyphs interactively, illustrated by software packages that have incorporated tours (as described in \citet{cook:2006}), such as DataViewer \citep{buja:1986}, XGobi \citep{swayne:1991} or.
-This technique is used to good effect in \citet{buja:1996a}, and was used to examine the climate data in \citet{hobbs:2010}.
-
 Coordinates for star glyphs are computed with a polar transformation of the minor components $(t, z)$:
@@
 \citet{eden:2010} wraps the time series, overplotting the multiple years, and this could be done also, by adjusting the scale of $t$.
-
-% HW: I removed this because I don't think it adds anything at this point
-%
-% Figure~\ref{fig:cycle} shows average monthly temperature as time (Cartesian coordinates) and star (polar coordinates) glyphs, for a subset of the spatial region.
-In this case, the star glyphs are not as effective as the linear glyphs: perhaps because of the absence of cyclical pattern.
-% 
-% *** Perhaps replace these figures with something from the GISTEMP data.
-% 
-% \begin{figure}[htbp]
-%   \centering
-%   \includegraphics[width=3in]{month-cartesian}
-%   \includegraphics[width=3in]{month-polar}
-%     
-%   \caption{Monthly averages in (left) Cartesian coordinates and (right) polar coordinates.}
-%   
-%   \label{fig:cycle}
-% \end{figure}
 
@@
 Conversely, if the purpose is to read spatial trend for one time point the heatmap may be a better choice, because this is a difficult task with the glyph-map.
-
-% Laying time series out on the spatial grid, the main task expected of the viewer is comparing the temporal patterns from one location to the next.
-In contrast, in the small multiples used in the facetted colored map, the viewer is expected to compare the patterns in the maps from one year to the next, for corresponding months.
-(See \citet{carr:1999} for a discussion on perceptual grouping.) That is, compare and contrast maps with facetted colored map, or compare and contrast time series using the glyph-map.
 
@@
 
-% , making it easier to compare the slopprovides for better reading of the intercept.
-The average value for the temperature anomaly data is not the same across the map: at higher latitudes the temperatures are uniformly higher, but at lower latitudes they are uniformly lower.
-(**** Can this be right?
-Would it mean that at these locations the temperatures 1950-2010 are much higher than the reference year temperatures?)
-
-% *** Re-do these figures with zooms in on the GISTEMP data
-
 \begin{figure}[htbp]
@@
 
-%*** Re-visit this next paragraph - its a little too airy-fairy, and the plot with color needs to have the color dampened more.
-Its way too dominant.
-This paragraph is too speculative for the paper, so removing it from first draft.
-
-%It's also often useful to display other types of reference information.
-You are only limited by your imagination (and the capabilities of your graphics software), but two simple ideas are to (1) vary the color or transparency of glyphs or references, or (2) use an additional layer to highlight special points.
-Figure~\ref{fig:ref-adv} illustrates these ideas to display the month with the highest temperature at each locations.
-Careful color choice is necessary so that this information is perceptible, but not distracting.
-
-%\begin{figure}[htbp]
-%  \centering
-%  \includegraphics[width=0.5\linewidth]{ref-max-1}%
-%  \includegraphics[width=0.5\linewidth]{ref-max-2}
-%  \caption{Glyph-maps of seasonal temperature patterns also showing the month with the highest temperature using (left) an additional layer of points and (right) changing the background color of the reference grid.}
-%  \label{fig:ref-adv}
-%\end{figure}
-
-%A particularly useful summary when to add when displaying model predictions, is some measure of model fit. This makes it easier to avoid over-interpreting predictions from poorly-fitting models.
-
 \subsection{Scaling}~\label{sec:scale}
-
-%*** This section needs to have plots re-done, once we have decided on the key pieces
 
@@
   \includegraphics[width=0.5\linewidth]{month-rescale-none}%
-  % \includegraphics[width=0.5\linewidth]{month-rescale-max}%
   \includegraphics[width=0.5\linewidth]{month-rescale01}
@@
 \end{figure}
-
-% 
-% \begin{equation}
-% y_{st} = \frac{y_{st}-y_{min}}{y_{max}-y_{min}} %\times 2 - 1
-% \label{scale1}
-% \end{equation}
-% 
-% \noindent where $y$ is the variable used on the minor axis in equations \ref{coords.eqn}, \ref{coords.polar.eqn}, $y_{min} = \mbox{minimum}\{y_{st}; s=1, \dots S, t=1, ..., T \}, y_{max} = \mbox{maximum}\{y_{st}; s=1, \dots S, t=1, ..., T\}$, $S=$ is the number of spatial locations and $T=$ is the number of time points.
-In addition, $y_{st}$ may be further scaled by doubling and subtracting 1 to give values between -0.5 and 0.5 for using in equations \ref{coords.eqn}, \ref{coords.polar.eqn}.
-% 
-% \begin{equation}
-% y_{st} = \frac{y_{st}-y_{s,min}}{y_{s,max}-y_{s,min}} %\times 2 - 1
-% \label{scale2}
-% \end{equation}
-% 
-% \noindent where $y_{s, min} = \mbox{minimum}\{y_{st}; t=1, ..., T \}, y_{s,max} = \mbox{maximum}\{y_{st}; t=1, ..., T\}$.
 

### 2012-05-14 revision-final/paper.tex
all comments removed from latex

@@
 
-% Displaying the predicted values in the absence of the raw data scale typically increases magnitude because the variation is much lower.
-
 \section{Large data and models}
@@
   \centering
-  %\includegraphics[width=1\linewidth]{gistemp-pred}%
 
@@
   \includegraphics[width=0.33\linewidth]{gistemp-pred-legend}
-  %\includegraphics[width=1\linewidth]{gistemp-polar-pred}
 
@@
 \end{figure}
-
-%The most seasonality occurs in north American.
-Some seasonal effects can also be seen in the Caribbean and in south American locations.
-Around the equator the temperatures are relatively warm for all seasons.
-Instead of radially displaying the temperature a glyph can be a regular time series, time displayed horizontally and temperature vertically (right plot).
-The structure perceived is different with this change in glyph type: locations at the equator have flat series, on land more varied temperatures.
-
-% \section{What to plot: raw data, trend lines, residuals,  seasonal}
-% 
-% A generalized additive model \citep{wood:2006} with a seasonal term and a long-term smooth term can be useful.
-This is revealing for smooth models of surface temperature, shown in Figure~\ref{fig:scaling}.
-Each location is modeled using a generalized additive model \citep{wood:2006} with a seasonal term and a long-term smooth term.
-The three plots in the figure show the long-term smooth term.
-% 
-% %Because some locations are uniformly warmer than others, the impact of
-% %changes across time are diminished. In the center plot, each location is
-% %scaled by dividing out the maximum, and on the right, each location is scaled
-% %to range $[0, 1]$. These emphasize individual changes but lose the ability to
-% %make global comparisons, and need caution when interpreting - a small
-% %absolute change gets blown up into a large relative change.
-% 
-% *** Need three plots here: trend, seasonality and residuals, shall we do this with NASA ozone data or stick to temperature?
-% 
-% Issues of scale relative to modeling.
-Switching from the data to model summaries brings new scaling issues.
-Often the scale of the raw data is ignored when the model is plotted, one usually shows just the line showing the trend, in the range of the trend.
-This has the effect of exaggerating the significance of the model.
-Scaling the trend lines relative to the range of the data may be useful sometimes.
 
@@
 \includegraphics[width=0.33\linewidth]{usa-lin-legend}
-  %\includegraphics[width=0.25\linewidth]{ghcn-mountains}%
   \caption{(Top) Glyph-maps of the linear trend in ground station temperature values, 1950--2010, for USHCN stations, illustrating three issues for irregularly gridded data: overlapping icons, no natural choice for icon size, and the disordered appearance making comparisons more difficult.
@@
 (Bottom) Legend showing
-    temperature and temporal ranges of individual glyphs.
-The local scaling is apparent here with all lines going through zero at the mean year.
-}% (left) and stations in the mountain states (right).}
-  \label{fig:irregular}
+    temperature and temporal ranges of individual glyphs. The local scaling is apparent here with all lines going through zero at the mean year. }  \label{fig:irregular}
 \end{figure}
-%The two panels illustrate one approach to the overlapping problem: zooming.  
 One solution to the overlapping problem is to zoom in on smaller regions.
@@
 Readability can be maintained and overlapping reduced by decreasing the icon size and simultaneously increasing the display size or decreasing the area plotted.
-% (as done in the right panel in Figure \ref{fig:irregular}).  
-%However, there are generally limitations in the maximum display size (or resolution) and the utility of plotting very small areas.
 
@@
 Heavy lines in the linear trend plot correspond to average trends over more than one location.
-(Bottom) Legends showing scale of glyphs.} % Legend in Figure~\ref{fig:irregular-legend}}
+(Bottom) Legends showing scale of glyphs.}
   \label{fig:irregular-collapsed}
@@
 Heavy lines in the linear trend plot correspond to average trends over more than one location.
-(Bottom) Legends showing scale of glyphs.} %Legend in Figure~\ref{fig:irregular-legend}}
+(Bottom) Legends showing scale of glyphs.}
   \label{fig:irregular-grid}
 \end{figure}
-
-%\begin{figure}[htbp]
-%  \centering
-%  \includegraphics[width=0.33\linewidth]{usa-season-legend}%
-%  \hspace{0.18 \linewidth}%
-%  \includegraphics[width=0.33\linewidth]{usa-lin-legend}
-
-%  \caption{Legend for Figures~\ref{fig:irregular-collapsed} aad \ref{fig:irregular-grid}.}
-%  \label{fig:irregular-legend}
-%\end{figure}
 

### 2012-05-14 revision-final/paper.tex
fixed typo in discussion
fixed typo in generalization, and another near Gauguin ref

@@
 This area of work has morphed into the field of metaphorical data displays, which create abstract landscapes of spatial data, a digression from glyph-maps.
-\citet{gribov:2006} describes the use of glyph-maps for multivariate data also, with emphasis on the graphics software Gauguin.
+\citet{gribov:2006} describes the use of glyph-maps for multivariate data also, with emphasis on the graphics software, Gauguin.
 Glyph-maps for spatiotemporal data have been used in several publications \citep{carr:1992,eden:2010,hobbs:2010}.
@@
 The bivariate relationship between the two variables is explored spatially: the equatorial Pacific and Caribbean have positive association between temperature and high cloud, while the south American continent has negative association.
-The right side plot displays a loess curve fit to the data instead of the individual points, sharpening the association signal realtive to the variation.
+The right side plot displays a loess curve fit to the data instead of the individual points, sharpening the association signal relative to the variation.
 Gauguin \citep{gribov:2006}, has an icon based on a histogram, giving small univariate distribution displays for each location.
@@
 Is this always the case?
-Are their certain types of periodicity that are particularly easy to spot?
+Are there certain types of periodicity that are particularly easy to spot?
 Rigorous perceptual studies will help guide the use of glyph-maps.

### 2012-05-14 revision-final/paper.tex
fixed wraps

@@
 
-\noindent where $\alpha_x, \alpha_y$ are scale parameters setting the fraction the width $w$ and height $h$ of the glyph, respectively.
+\noindent where $\alpha_x, \alpha_y$ are scale parameters setting the fraction of width, $w$, and height, $h$, of the glyph, respectively.
 For gridded data, $w$ and $h$ will be the minimum difference between neighboring spatial locations to make use of the resolution of the major components.
@@
 This is a non-standard conversion to polar coordinates, but it creates a timeline that starts at 12 o'clock and proceeds clockwise.
-\citet{eden:2010} wraps the time series, overplotting the multiple years, and this could be done also, by adjusting the scale of $t$.
+\citet{eden:2010} wrap the time series, overplotting the multiple years, and this could be done also, by adjusting the scale of $t$.
 

### 2012-05-14 revision-final/paper.tex
fixed typos around pages 8,9

@@
 
-Another common approach for displaying spatiotemporal data is to calculate a statistical summary for each location and display these values as single heatmap.
+Another common approach for displaying spatiotemporal data is to calculate a statistical summary for each location and display these values as asingle heatmap.
 For example, to study long-term trends, the slope of a linear model can be calculated at each location and color tiles used to display the slope value over space.
@@
 
-A possible draw-back of the glyph-map, particularly if linear trend is displayed as an icon, is it may suffer from the Z\"ollner Illusion \citep{Zollner}, which makes straight lines look crooked (Section \ref{sec:large-data} has an example).
+A possible drawback of the glyph-map, particularly if linear trend is displayed as an icon, is it that may suffer from the Z\"ollner Illusion \citep{Zollner}, which makes straight lines look crooked (Section \ref{sec:large-data} has an example).
 
@@
 
-Each glyph is small, so there is not enough space for a full set of axes on each, and it is not desirable beause it would be repetitive, redundant information.
+Each glyph is small, so there is not enough space for a full set of axes on each, and it is not desirable because it would be repetitive, redundant information.
 Instead, minimal reference lines and boxes are incorporated.

### 2012-05-15 revision-final/paper.tex
fixed Kelvin refs

@@
 
-To observe the scale of the measurements represented by the glyphs secondary legend plot is generated and displayed alongside the glyph-map, of all the glyphs overplotted, with complete axes.
-See, for example, Figure \ref{fig:nasa-glyph} where from the legend it can be seen that months range from 0 to 72 and de-seasonalized temperature ranges from -8 to 8$^0$K.
-In Figure \ref{fig:ref-basic}, months range from 1 to 12, and temperature from 270 to 315$^o$K.
+To observe the scale of the measurements represented by the glyphs, a secondary legend plot of all the glyphs overplotted, with complete axes is generated, and displayed alongside the glyph-map.
+See, for example, Figure \ref{fig:nasa-glyph} where from the legend it can be seen that months range from 0 to 72 and de-seasonalized temperature ranges from -8 to 8 K.
+In Figure \ref{fig:ref-basic}, months range from 1 to 12, and temperature from 270 to 315 K.
 

### 2012-05-15 revision-final/paper.tex
emphasis to emphasize
relative to comparable

@@
 
-   \item Scaling to maximum 1 (but leaving the minimum unchanged) can help
-  emphasis the relative shape.
+   \item Scaling to maximum of 1 (but leaving the minimum unchanged) can help
+  emphasize the relative shape.
 
@@
 Note that the use of local scales needs to be clearly marked.
-The viewer must realize that the scale is not relative and that big patterns in some locations might be just tiny, nugatory effects.
+The viewer must realize that the size is not comparable and that big patterns in some locations might be just tiny, nugatory effects.
 To indicate this on the locally scaled data plots, one can use an additional aesthetic, like color, to encode the range of original scale.

### 2012-05-15 revision-final/paper.tex
fixed figure captions

@@
  \includegraphics[width=0.5\linewidth]{month-rescale01-fill}
- \caption{Adding intensity and color to scaled plots.
-Here each location has been scaled to range $[0, 1]$ and color mapped to the range of the original predictions.
+ \caption{Adding intensity and color to indicate global scale to locally scaled data glyph-maps.
 (Left) Range mapped to intensity (grey scale) of the line: pale locations have smaller effect, smaller range, than darker locations, stronger effect.
@@
 
-  \caption{(Top) Glyph-map of raw temperature anomaly data (GISTEMP), from 1880-2011.
+  \caption{(Top) Glyph-map of raw temperature anomaly data (GISTEMP), from 1880--2011.
 Gaps in the full circle indicate missing values, and thickness of the circle indicates more variability in the measurements.
@@
   \caption{(Top) Glyph-map of predicted temperature anomaly data
-    (GISTEMP), for the reduced period 1950-2010. Increasing
+    (GISTEMP), for the reduced period 1950--2010. Increasing
     trends can be seen over most of the USA, with varying degrees of
@@
 \subsection{Irregular locations}
-Figure \ref{fig:irregular} shows linear trends for monthly average surface temperature at stations in the USHCN data.
-It illustrates three challenges in working with irregular locations: there is no natural choice for icon size, icons can potentially overlap, and comparisons rely more heavily on the reference guides.
-The icon size was chosen manually for this plot, corresponding to 1$^o$ square in the geographical coordinates.
+
 \begin{figure}[htbp]
@@
 \includegraphics[width=0.33\linewidth]{usa-lin-legend}
-  \caption{(Top) Glyph-maps of the linear trend in ground station temperature values, 1950--2010, for USHCN stations, illustrating three issues for irregularly gridded data: overlapping icons, no natural choice for icon size, and the disordered appearance making comparisons more difficult.
-Local centering, subtracting the location mean from the linear fit, is used as the variation in average temperature is of much greater magnitude than the variation in linear trend.
-(Bottom) Legend showing
-    temperature and temporal ranges of individual glyphs. The local scaling is apparent here with all lines going through zero at the mean year. }  \label{fig:irregular}
+  \caption{(Top) Glyph-maps of the linear trend in ground station temperature values, 1950--2010, for USHCN stations, illustrating three issues for irregularly gridded data: overlapping icons, no natural choice for icon size, and the disordered appearance making comparisons difficult.
+(Bottom) Legend showing  temperature and temporal ranges of individual glyphs.
+All lines intersecting at zero is indicative of the local centering of the data.
+}
+\label{fig:irregular}
 \end{figure}
+
+Figure \ref{fig:irregular} shows linear trends for monthly average surface temperature at stations in the USHCN data.
+It illustrates three challenges in working with irregular locations: there is no natural choice for icon size, icons can potentially overlap, and comparisons between icons are difficult.
+Local centering was done on the data, where the mean temperature for each location was subtracted from the linear fit, to focus on trend,  because the variation in average temperature is much greater than the variation in linear trend.
+The icon size was chosen manually for this plot, corresponding to 1$^o$ square in the geographical coordinates.
 One solution to the overlapping problem is to zoom in on smaller regions.
@@
 \includegraphics[width=0.20\linewidth]{usa-lin-collapse-legend}
-  \caption{Glyph-maps of seasonal patterns (top left) and linear trends (top right) of USHCN stations, 1950--2010.
-Nearby stations whose icons would overlap have been collapsed.
-Each location has a glyph in the seasonal plot and color of the tiles corresponds to the average temperature of the locations.
-Heavy lines in the linear trend plot correspond to average trends over more than one location.
-(Bottom) Legends showing scale of glyphs.}
+  \caption{Glyph-maps of seasonal patterns (left) and linear trends (right) of USHCN stations, 1950--2010, where data from close neighboring stations have been collapsed into a single location.
+In the seasonal glyph-map, the color of the tile corresponds to the average temperature of stations represented at that location, and in the trend glyph-map, line width indicates the number of locations represented.
+(Bottom) Legends showing the range of values represented by the glyphs.}
   \label{fig:irregular-collapsed}
@@
   \includegraphics[width=0.20\linewidth]{usa-lin-grid-legend}
-  \caption{Glyph-maps of seasonal patterns (top left) and linear trends (top right) of USHCN stations, 1950--2010.
-Station locations have been rounded to the nearest degree Each location has a glyph in the seasonal plot and color of the tiles corresponds to the average temperature of the locations.
-Heavy lines in the linear trend plot correspond to average trends over more than one location.
-(Bottom) Legends showing scale of glyphs.}
+  \caption{Glyph-maps of seasonal patterns (left) and linear trends (right) of USHCN stations, 1950--2010,  where locations have been rounded to the nearest degree.
+Similar to the glyph-maps in Figure \ref{fig:irregular-collapsed},  the color of the tile in the seasonal glyph-map corresponds to the average temperature of stations represented at that location, and  line width indicates the number of locations represented in the trend glyph-map.
+(Bottom) Legends showing the range of values represented by the glyphs.}
   \label{fig:irregular-grid}
@@
 
-  \caption{Glyph-map showing (top left) scatterplot of temperature vs high-cloud and (top right) smoothed (loess) curve fit to each location.
+  \caption{Glyph-map showing (left) scatterplot of temperature vs high-cloud and (right) smoothed (loess) curve fit to each location.
 The relationship between these two variables varies considerably over the spatial domain.

### 2012-06-21 cover-letter.tex
made pdf image for publishers

@@
 Please find attached our submission for the special issue of Environmetrics on Climate.
-The title is Glyph-maps for Visualizing Climate Data and Models, with authors Hadley Wickham, Heike Hofmann, Charlotte Wickham and myself.
+The title is Glyph-maps for Visually Exploring Temporal Patterns in Climate Data and Models, with authors Hadley Wickham, Heike Hofmann, Charlotte Wickham and myself.
 We hope that you find this paper interesting reading, and that it will make a valuable contribution  to the special issue.

### 2012-06-21 nasa.r

@@ -26,7 +26,7 @@ temp_trend_models <- dlply(nasa, c("lat", "long"), function(df) {
 
 locs <- dlply(nasa, c("lat", "long"))
 
-resids <- mdply(cbind(d = locs, m = temp_models), function(d, m) {
+resids <- mdply(cbind(d = locs, m = temp_seas_models), function(d, m) {
   d$temp_resid <- d$surftemp - predict(m, newdata = d)
   d
 })
@@ -55,7 +55,7 @@ theme_nothing <- function (base_size = 12) {
 }
 
 qplot(long, lat, data=resids, geom="tile", fill=temp_resid, facets=year~month) + scale_fill_gradient2(midpoint=0, mid="white", high="red", low="blue") + map + theme_nothing() + coord_map()
-ggsave("nasa-colored-map.png", height=10, width=18)
+ggsave("../images/nasa-colored-map.png", height=10, width=18)
        
 resids <- glyphs(resids, "long", "day", "lat", "temp_resid") 
 qplot(gx, gy, data = resids, geom = "line", group = gid) + map

### 2012-06-21 revision-final/code/1-examples.r

@@ -55,7 +55,7 @@ ggplot(resids, aes(long, lat)) +
     axis.ticks.length = unit(0, "lines"), 
     axis.ticks.margin = unit(0, "lines"),
     legend.position = "right")
-ggsave("../images/nasa-colored-map.png", height = 5, width = 9, dpi = 72)
+ggsave("../images/nasa-colored-map.png", height = 5, width = 9, dpi = 600)
 
 # Second figure: same data but with glyphs -----------------------------------
 
